### 2020-03-28 research/sigbovik2020/paper.tex
research/sigbovik2020: write paper and README about symsh

\documentclass[sigplan]{acmart}

\usepackage{hyperref}
\usepackage{listings}
\usepackage{booktabs} % For formal tables

% Copyright
%\setcopyright{none}
%\setcopyright{acmcopyright}
%\setcopyright{acmlicensed}
\setcopyright{rightsretained}
%\setcopyright{usgov}
%\setcopyright{usgovmixed}
%\setcopyright{cagov}
%\setcopyright{cagovmixed}

% DOI
\acmDOI{10.475/123_4}

% ISBN
\acmISBN{123-4567-24-567/08/06}

%Conference
\acmConference[SIGBOVIK 2020]{SIGBOVIK}{April 2020}{Pittsburgh, PA USA}
\acmYear{2020}
\copyrightyear{2020}

%\acmBadgeL[http://ctuning.org/ae/ppopp2016.html]{ae-logo}
%\acmBadgeR[http://ctuning.org/ae/ppopp2016.html]{ae-logo}

% Document starts
\begin{document}

% Title portion. Note the short title for running heads
\title[Type-directed decompilation of shell scripts]{Type-directed decompilation of shell scripts}

\author{Spencer Baugh}
\affiliation{%
  \institution{University of Carcosa}
  \streetaddress{104 Lost Beach}
  \city{Carcaso}
  \country{Hyades}}
\email{first.last@gmail.com}
\author{Dougal Pugson}
\affiliation{%
  \institution{Pugson's C++ Crypt LLC}
  \city{New York}
  \country{USA}
}
\email{dougalpugson@gmail.com}

\begin{abstract}
Maintenance shell programmers are often faced with inscrutable shell scripts without human-readable source code.
We apply techniques pioneered by the type-directed partial evaluation community
to create a decompiler which can take an executable shell script and recover its original source code. 
This technique has surprising generality,
and our decompiler can also be used as a pretty-printer, or in general,
as a compiler from any language into shell.
\end{abstract}

%
% The code below should be generated by the tool at
% http://dl.acm.org/ccs.cfm
% Please copy and paste the code instead of the example below.
%
\begin{CCSXML}
<ccs2012>
   <concept>
       <concept_id>10011007.10011006.10011050.10010517</concept_id>
       <concept_desc>Software and its engineering~Scripting languages</concept_desc>
       <concept_significance>500</concept_significance>
       </concept>
   <concept>
       <concept_id>10011007.10011006.10011041</concept_id>
       <concept_desc>Software and its engineering~Compilers</concept_desc>
       <concept_significance>500</concept_significance>
       </concept>
   <concept>
       <concept_id>10003752.10003790.10011740</concept_id>
       <concept_desc>Theory of computation~Type theory</concept_desc>
       <concept_significance>500</concept_significance>
       </concept>
   <concept>
       <concept_id>10011007.10011006.10011073</concept_id>
       <concept_desc>Software and its engineering~Software maintenance tools</concept_desc>
       <concept_significance>500</concept_significance>
       </concept>
 </ccs2012>
\end{CCSXML}

\ccsdesc[500]{Software and its engineering~Scripting languages}
\ccsdesc[500]{Software and its engineering~Compilers}
\ccsdesc[500]{Theory of computation~Type theory}
\ccsdesc[500]{Software and its engineering~Software maintenance tools}

\keywords{Partial evaluation, type systems, shell scripting}

\maketitle

\section{Introduction}\label{introduction}
Among its many beneficial properties, shell has a unique feature:
The language automatically compiles itself as it is written.
Immediately after a shell script is typed into a computer and saved into a file,
it is transformed into a compiled form which is unreadable by humans.
\footnote{Use of the authors' previous work on delimited continuations in bash, \cite{bashcc}, seems to accelerate this process.}
Such a script can be immediately used for all your most important financial data, medical procedures, etc.,
while you are safe in the knowledge that no-one can read your important proprietary shell scripts.

Unfortunately, sometimes such scripts might exhibit minor bugs.
Since these executable files that are leftover from the process of shell scripting are unreadable garbage,
maintenance programmers are often forced to rewrite from scratch.
Sometimes, very brave programmers will try open one of these old, buggy "shell script files",
but generally they are instantly turned to stone upon seeing the inscrutable contents.
And since shell is so convenient, typically the original programmer is long gone - probably retired early.

Our contribution in this paper is a decompiler for these files,
which is able to recover the original source code.
This source code can then be viewed by the maintenance programmer,
to help them reimplement the script in their non-shell language of choice.

As outlined in our previous work, \cite{bashcc},
the Unix shell is closely historically related to functional programming.
Thus it should be no surprise that we are able to transplant techniques
from the functional programming community, to the shell,
in the tradition of \cite{frankenstein}.

In this case, we use partial evaluation techniques to achieve our goal.
As described in \cite{oleg}, as citing \cite{danvy}, as blowing the minds of undergrads everywhere,
sufficiently general partial evaluation techniques
can be applied to "reify" a compiled program and recover the program source code,
under the constraints that the program 1. has a type and 2. terminates.
Pretty easy constraints, I think we can make that happen for bash!

As a more approachable introduction (note that "more approachable" doesn't mean "approachable"),
these techniques can be compared to tagless-final-style techniques.
Decompilation of a typed, compiled program is essentially identical to pretty-printing of a tagless-final-style term.
The magic is that any typed, closed program can be treated as a TFS term.

Our paper is organized in some number of sections.
Section \ref{introduction} contains an introduction and a description of the organization of the paper.
Section \ref{theory} contains a exploration of the requirements for applying these techniques,
and establishes their firm grounding in theory.
Section \ref{applications} demonstrates several applications.
Section \ref{implementation} gives an overview of the implementation.
Section \ref{conclusion} concludes the paper, discusses future work, and affirms that this was the right thing to do.
\section{Theory}\label{theory}

As mentioned above and in \cite{danvy},
we have two requirements to apply type-directed partial evaluation:
The program must be typed and must terminate.
\subsection{Typed}
What is the type of a shell script?
Well, a shell script takes the PATH environment variable,
and runs commands (identified by strings) out of PATH,
each of which has some side effects.
A shell script doesn't return anything, it just has side effects,
so let's say that its output type is unit.

So that means a shell script has type $path -> unit$,
where $path$ is $(str, [str]) -> unit$.

Then in total, a shell script has type $((str, [str]) -> unit) -> unit$.

We can confirm this is correct by applying double negation elimination
\footnote{DNE is an axiom because this paper is unintuitive.}
which shows us that shell scripts have type $(str, [str])$.
This is correct because a shell script is indeed a bunch of strings.

Let's be a little more specific with our type, though, and model each executable as a function.
So then the type of PATH is $(str -> ([str] -> unit))$,
and the type of a shell script is then $(str -> ([str] -> unit)) -> unit$

For now, just think of a shell script as taking PATH and running commands out of it.
\subsection{Termination}
Termination, on the other hand, is a much harder problem than assigning a strict static type to shell scripts.
This is because of the presence of the dreaded D-wait in Unix.
A process can get into an uninterruptible state when making a filesystem request,
and just hang forever, ignoring all signals, including SIGKILL.
This is really annoying for people developing filesystems, which we had to do for this paper,
so we want to complain about it here.

Nevertheless, if the program hangs, this issue can be solved by simply mashing Ctrl-C.
Even D-wait can be solved by throwing the computer out the window
(assuming a sufficiently portable computer, and great enough height for it to be destroyed on impact).

So termination is ultimately not a problem either.
\subsection{Background}
In brief, the principle of the technique we will be applying is this:
A closed, abstract function,
which takes in other functions and combines them through application in some way to eventually return a result,
can be passed functions which, instead of performing actual operations and returning real results,
take ASTs and return ASTs.

For example, a parameter with type $(a, a) -> a$,
which might normally be addition of two integers or something,
can be passed at the specific type $(ast, ast) -> ast$,
and be implemented as $\lambda x. \lambda y. Plus (x, y)$
where $Plus$ is some data-type constructor.

A shell script's single argument (in our model) is the PATH environment variable.
\footnote{This has sufficiently abstract types, since everything is a string, so we don't know what anything is.}
We will pass in a PATH which, when an executable name is looked up in it,
returns the executables (functions) of our choice.
These executables will in turn, when executed,
construct an AST instead of actually doing anything.
\section{Applications}\label{applications}
Let's demonstrate our tool before getting to the actually interesting part.

\subsection{Decompiling bash}
Suppose we save the following shell script to a file and mark it executable,
which instantaneously makes it unreadable.
\begin{lstlisting}[language=bash]
ls; which ls
stat /
foo|bar
\end{lstlisting}
Nevertheless our decompiler can run on the script and produces the following output:
\begin{lstlisting}[language=bash]
ls
which ls
stat /
foo | bar
\end{lstlisting}
As you can see, our decompiler even pretty-prints the shell script.
\subsection{Decompiling arbitrary executables}
It also works on C programs, and in general, arbitrary executables.
We can compile the following normal C program, and run our decompiler on it.
\begin{lstlisting}[language=C]
int main() {
  int rc;
  rc = fork();
  if (rc == 0) {
    execlp("foo",
           "foo", "bar", "baz", NULL);
  } else { wait(NULL); }
  rc = fork();
  if (rc == 0) {
    execlp("whatever",
           "whatever", "quux", NULL);
  } else { wait(NULL); }
  return 0;
}
\end{lstlisting}
And we get the following shell script out:
\begin{lstlisting}[language=bash]
foo bar baz
whatever quux
\end{lstlisting}

Useful!

\subsection{Optimizing decompiler}
Our decompiler is so advanced that it in fact transparently applies optimizations in the process of decompilation.
Consider the following C program:
\begin{lstlisting}[language=C]
int main() {
  printf("hello world\n");
}
\end{lstlisting}
This program decompiles to the following shell script:
\begin{lstlisting}[language=bash]
\end{lstlisting}
Our decompiler correctly recognizes that this program,
since it doesn't execute any other programs from the filesystem,
is in fact utterly worthless, and optimizes it away to nothing.

\section{Implementation}\label{implementation}
Much to our surprise, we actually implemented this.

We have implemented a filesystem (using FUSE),
which pretends to have any possible executable you want.
We point the shell script at this filesystem using PATH,
and each time the shell script goes to run a command,
it instead runs a stub under our control.
Using some real technology we developed earlier and just thought it would be funny to use for this,
this stub connects back to the filesystem server,
where our decompiler is able to query its argv, stat its stdin/out/err, and tell it to exit with a specific exit code.

To mount the filesystem without requiring privileges or setuid executables,
we use a user namespace and mount namespace, and run the script inside those namespaces.

After the shell script finishes execution,
we reconstruct its source code from the trace of executed commands
using a highly advanced for loop.

Note that this doesn't use ``LD\_PRELOAD'' or strace,
so it can even be used on statically linked, setuid shell scripts.
There are lots of those!

The code is on Github at
\url{https://github.com/catern/rsyscall/tree/master/research/sigbovik2020}.

\section{Conclusion}\label{conclusion}
\subsection{Future work}
\subsubsection{Support for niche shell features}
There are some niche, minor features of the shell language which are not supported by our decompiler,
such as ``if'' and ``while''.
As any true shell programmer uses ``xargs'' instead,
which our framework would decompile just fine,
this isn't a problem.

Nevertheless it might be nice to figure out some ridiculous hack that would allow such features
(and shell builtins in general)
to be visible to our decompiler.

Maybe we could execute the shell script multiple times,
returning different exit codes from commands different times,
and thereby get a collection of traces through the control flow graph,
which we could then piece back together.
But this is beginning to sound like real work.
\subsection{Conclusion conclusion}
In conclusion,
we hope that this tool proves useful for maintenance shell programmers,
who will finally have a way to read those shell scripts that they always complain are unreadable.
Hopefully this will increase their productivity,
massively increase global GDP,
and cause my 401K to recover all the value it lost due to the coronavirus.

\bibliographystyle{ACM-Reference-Format}
\bibliography{bibliography}

\end{document}

### 2020-03-28 research/sigbovik2020/example.sh

@@ -1,0 +1,4 @@
+#!/bin/sh
+ls; which ls
+stat /
+foo|bar

### 2020-03-28 research/sigbovik2020/prog.c

@@ -1,0 +1,14 @@
+#include <unistd.h>
+#include <stdio.h>
+#include <sys/wait.h>
+
+int main() {
+        int rc;
+        rc = fork();
+        if (rc == 0) { execlp("foo", "foo", "bar", "baz", NULL); }
+        else { wait(NULL); }
+        rc = fork();
+        if (rc == 0) { execlp("whatever", "whatever", "quux", NULL); }
+        else { wait(NULL); }
+        return 0;
+}

### 2020-05-03 research/paper/paper.tex
texxx

% \documentclass[sigplan]{acmart}
\documentclass{acmart}

\usepackage{hyperref}
\usepackage{float}
\usepackage{listings}
\usepackage{booktabs} % For formal tables

% Copyright
%\setcopyright{none}
%\setcopyright{acmcopyright}
%\setcopyright{acmlicensed}
\setcopyright{rightsretained}
%\setcopyright{usgov}
%\setcopyright{usgovmixed}
%\setcopyright{cagov}
%\setcopyright{cagovmixed}

% DOI
%% \acmDOI{10.475/123_4}

% ISBN
%% \acmISBN{123-4567-24-567/08/06}

%Conference
\acmConference[OSDI 2020]{SIGBOVIK}{November 2020}{Banff, Alberta, Canada}
\acmYear{2020}
\copyrightyear{2020}

%\acmBadgeL[http://ctuning.org/ae/ppopp2016.html]{ae-logo}
%\acmBadgeR[http://ctuning.org/ae/ppopp2016.html]{ae-logo}

% Document starts
\begin{document}

% Title portion. Note the short title for running heads
\title[Direct-style process creation on Linux]{Direct-style process creation on Linux}

\author{Spencer Baugh}
\affiliation{%
  \institution{Two Sigma}
  \streetaddress{100 6th Avenue}
  \city{New York City, New York}
  \country{USA}}
\email{sbaugh@twosigma.com}

\begin{abstract}
Traditional process creation interfaces,
such as fork and spawn,
are complex to use, limited in power, and difficult to abstract over.
We develop a new process creation interface for Linux
which allows a program to create a child process in an non-running state
and initialize the new process by operating on it from the outside.
This method of process creation results in more comprehensible programs, 
has better error handling,
is more efficient,
and is more amenable to abstraction.
Our implementation is immediately deployable without kernel modifications on any recent Linux kernel version.
\end{abstract}

%
% The code below should be generated by the tool at
% http://dl.acm.org/ccs.cfm
% Please copy and paste the code instead of the example below.
%
\begin{CCSXML}
<ccs2012>
   <concept>
       <concept_id>10011007.10010940.10010941.10010949.10010957</concept_id>
       <concept_desc>Software and its engineering~Process management</concept_desc>
       <concept_significance>500</concept_significance>
       </concept>
   <concept>
       <concept_id>10011007.10010940.10010941.10010949.10010950.10010951</concept_id>
       <concept_desc>Software and its engineering~Virtual memory</concept_desc>
       <concept_significance>300</concept_significance>
       </concept>
   <concept>
       <concept_id>10010147.10010169</concept_id>
       <concept_desc>Computing methodologies~Parallel computing methodologies</concept_desc>
       <concept_significance>500</concept_significance>
       </concept>
 </ccs2012>
\end{CCSXML}

\ccsdesc[500]{Software and its engineering~Process management}
\ccsdesc[300]{Software and its engineering~Virtual memory}
\ccsdesc[500]{Computing methodologies~Parallel computing methodologies}

\keywords{Process creation, Linux, fork, spawn}

\maketitle

\section{Introduction}\label{introduction}
Existing process creation interfaces can be divided into three categories:
Fork-style, spawn-style, and direct-style.

Fork-style (such as \texttt{fork}) and spawn-style (such as \texttt{posix\_spawn}) each have advantages and disadvantages.
Correct and performant usage of fork-style places significant constraints on the calling process.
Spawn-style lifts many of these constraints,
but is fundamentally limited in the kinds of processes it can create.
With both fork-style and spawn-style,
it is difficult to express
branching behavior based on results or errors encountered while initializing the new process;
fork-style requires IPC if it wishes to communicate errors back to the rest of the program,
while robust indication of errors in spawn-style requires substantial growth in the spawn interface.

Direct-style is rare in general, but relatively common on academic operating systems.
Direct-style involves creating a new child process in a non-running state,
and then calling individual syscalls from the parent to initialize the child.
Direct-style does not constrain the calling process,
can create any kind of process supported by the underlying system,
and easily expresses errors in any individual step of the initialization.

Our contribution is an implementation of direct-style process creation for Linux.
Our implementation has all the advantages of direct-style
over Linux's fork-style \texttt{fork} syscall and spawn-style \texttt{posix\_spawn} libc function.

In our implementation, a parent process calls the \texttt{clone} syscall
to create a child process which shares memory and other resources with the parent,
and which runs no user code.
The parent then uses \texttt{rsyscall}, a cross-process syscall library,
to call arbitrary syscalls in the child process to initialize it;
for example, the parent might call \texttt{chdir} to change the child's current working directory.
After the process is fully initialized,
the parent can call \texttt{execve} in the child,
causing the process to start running an arbitrary executable
and detaching the child from the control of \texttt{rsyscall}.

We have used direct-style process creation on Linux to do TODO and TODO,
and we've benchmarked our implementation as TODO.

Our implementation is immediately deployable on recent Linux kernels,
and is distributed as a feature of the open source \texttt{rsyscall} library,
available from \url{https://github.com/catern/rsyscall}.

This paper is organized in several sections.
In section \ref{background}, we provide background on fork-style, spawn-style, and direct-style process creation,
and consider their advantages and disadvantages in more depth.
In section \ref{examples} we show examples of usage of direct-style process creation on Linux,
including use of process features which are difficult to use on Linux with \texttt{fork} or \texttt{posix\_spawn},
and briefly examine the user API.
In section \ref{implementation}, we explore our implementation in depth,
and discuss several difficult process-creation related details of \texttt{rsyscall}'s implementation.
In section \ref{evaluation}, we evaluate, somehow (TODO).
In section \ref{future_work},
we list several directions for future work related to direct-style process creation and to \texttt{rsyscall}.
In section \ref{conclusion}, we briefly and optimistically conclude.

\section{Background}\label{background}
Processes are a widespread feature of operating systems,
with substantial variation in characteristics between systems.
One of the areas of variation is in process creation mechanisms.

On most systems today,
the interfaces for process creation
can be divided into "fork-style" and "spawn-style".
A few systems use a third style, which we refer to as "direct-style".
\subsection{Fork-style}
Fork-style interfaces are those that follow the model used by Unix's fork \cite{forkhist}.
A process calls fork, and a copy of that process is made,
with the return value of fork being the only differentiation between the process and its copy.
The same program continues executing in both processes.
Typically, the program will case on the return value to determine if it's running in the parent or the child,
and run process setup and initialization code if the program is running in the child,
calling various syscalls to set up the state of the child.

Fork has a number of flaws,
and much ink has been spilled detailing them.
We'll cover the most prominent briefly here,
but see Baumann 2019 \cite{forkroad} for a more detailed evaluation.

Fork, when called successfully, returns twice,
creating two different processes which execute concurrently.
This immediately makes fork difficult to integrate;
for example, errors can't be easily communicated between the two processes.
If there is an error in the initialization of the child process,
the part of the program running in the child process
needs to use some form of inter-process communication (IPC) to communicate it back to the parent process.
The child process can run arbitrary code to make decisions about its initialization,
but that code needs to use IPC if it wishes to communicate with the rest of the program,
which is still in the parent process.

The most common complaint about fork is its poor performance \cite{forkroad}.
It copies many attributes about the parent process when creating the child process,
including setting up copy-on-write memory-mappings in the child process.
This becomes slower as the parent process has more memory-mappings,
eventually taking a significant amount of time.
This copying is required to robustly implement fork's model,
where the same program continues executing without changes in both the parent process and child process;
the copy of the program running in the child process might access any memory at any time.

Multi-threaded programs generally cannot use fork safely.
Typical Unix fork implementations duplicate only the thread calling fork from the parent process to the child process.
In a multi-threaded program, this can cause various issues;
for example, another thread might be holding a memory allocator lock at the time of the fork,
which in the child process will never be unlocked,
causing the child process to deadlock if it tries to allocate memory.
Some thread libraries provide partial mitigations for this issue,
but it's up to user code to make use of those mitigations.
In combination with fork's poor performance in large-memory programs,
a user of fork must think carefully
about the characteristics of the process from which fork is being called.
\subsection{Spawn-style}
Spawn-style interfaces are those that follow the model used by \texttt{posix\_spawn} \cite{posix_spawn}
and Windows' \texttt{CreateProcess} \cite{create_process}.
All the details about the new process are provided up-front as arguments to a syscall,
which creates the new process from a mostly-clean slate, initialized with the provided details.

Spawn-style interfaces typically still transparently copy some details from the parent process;
for example, various security contexts,
and any other process attribute which is not explicitly specified in the arguments to the spawn interface.
The key difference between fork-style and spawn-style is not how much they copy;
it is how the attributes which are not copied are specified:
by mutation from code running in the new process in the case of fork-style,
or by explicit argument passing in the case of spawn-style.

Spawn-style interfaces lift the constraints on the calling process that fork-style interfaces impose.
Since spawn-style interfaces don't run user code in the new process during initialization,
there is no need to copy memory,
and there are no inherent issues when the interface is used from a multi-threaded program
(though operating systems which implement spawn-style interfaces on top of \texttt{fork} can still have bugs).

However, the arguments that can be provided to a spawn-style process creation syscall
do not cover all the possible attributes that one might want to set for the new process.
Most systems have a large number of syscalls which can mutate the state of a process during its lifetime;
for a spawn-style interface to work in all scenarios,
all those possible mutations must be reproduced in the interface.
For example, the \texttt{posix\_spawn} function provided by glibc does not support creating processes in new namespaces,
as is required for container functionality on Linux.

Spawn-style process creation also does not allow for conditional logic during the setup;
if the setup of the new process encounters an error at some point,
the only option is to return from the entire spawn call with an error.
Such errors returned from spawn-style calls
are typically much less informative
than the errors returned by the syscalls which directly mutate the process attributes.
Other forms of conditional logic are also impossible in spawn-style;
one modification to the process cannot depend on the result of some other modification.
\subsection{Direct-style}
A few academic operating systems, such as KeyKOS \cite{keykos}, seL4 \cite{sel4}
and some others \cite{exokernel} \cite{fuschia} \cite{singularity},
use another style of process creation, which we refer to as "direct-style".

In direct-style, a process is created by its parent in a non-running state,
and then supplied with various resources,
and then started running once it is fully set up.
In operating systems with true direct-style process creation,
the syscalls that can mutate a process
take explicit arguments to indicate which process they should operate on.
In this way, the same syscalls that can mutate a process while it is running,
are called by the parent process to mutate the process while it is being set up.

We refer to this as "direct-style" process creation,
because the parent creating the process operates on it directly and imperatively
rather than dispatching a distinct unit of code to perform setup from inside the context of the new process,
as in fork-style,
or building up a declarative specification of what the new process should look like ahead of time,
as in spawn-style.

Since everything happens directly from the parent process,
process initialization is compatible with a variety of techniques
which were otherwise incompatible with process creation.
For example, a program could provide a UI which directly customizes the new process,
or might initialize several processes at once and share data between them.

Most significantly,
since all operations are performed through syscalls called directly from the main program,
errors are indicated in the same way as any other syscall error.
This is unlike fork-style, which needs IPC to indicate errors,
and unlike spawn-style, which typically indicates errors at a very coarse-grained level.

Like fork-style interfaces,
direct-style interfaces can set up arbitrary attributes in the new process.
Any attribute that can be changed by a syscall
can be manipulated by a direct-style interface,
just as with fork-style interfaces.
This is unlike spawn-style interaces,
which only can change attributes that are supported by the interface.

Like spawn-style interfaces,
direct-style interfaces have no constraints on the calling process.
No user code runs in the child process,
so, unlike fork-style interfaces,
direct-style interfaces can be used from multi-threaded or large-memory processes without issues.

Direct-style can be more complex to use;
it is most typically used in capability-oriented operating systems,
where a great deal of resources and information must be provided explicitly to initialize the new process.
In a truly capability-oriented operating system,
nothing is copied implicitly to the new process;
everything must be explicitly specified.
This can appear more complex
when compared to process creation on Unix systems or Windows.

However, we believe that a natural port of direct-style process creation to Linux
provides an interface that is just as simple to use as fork-style or spawn-style on Linux,
without their disadvantages.
Such an interface, like fork-style and spawn-style,
would implicitly copy some attributes to the new process,
rather than being capability-oriented and requiring all attributes to be specified explicitly.
This is the kind of direct-style interface that we contribute in this paper.

\section{Overview and examples}\label{examples}
In most operating system APIs, including the ones available on Linux,
when a program makes a syscall, it implicitly operates on the current process.
To perform direct-style process creation on Linux,
we need a Linux API where we instead explicitly specify in which process we want to make a syscall.

\texttt{rsyscall} provides this cross-process API for Linux.
\texttt{rsyscall} is a library which allows us to create processes in which we can call arbitrary syscalls.
These processes run no other code; they do nothing except for execute syscalls at our request.
We will discuss the implementation of rsyscall in greater detail in section \ref{implementation}.

From the perspective of an program using rsyscall,
there are multiple processes which are under its control.
We use the term ``thread'' to refer to all such controlled processes,
including the main "thread" on which the program is running.
On Linux, the shared-memory "threads" provided by libraries such as \texttt{pthreads}
are implemented as processes,
and their lifetime and execution is completely controlled by a single program.
The same is true of the controlled-process "threads" provided by rsyscall.
The most significant difference is that rsyscall "threads"
do not run their own code concurrently with the main program;
nevertheless rsyscall "threads" do provide the opportunity for parallel execution of system calls,
and so the terminology provides useful intuition.

To perform direct-style process creation,
we use \texttt{rsyscall} to call \texttt{clone} on our main thread to create a new child thread.
We call various syscalls in the child thread to mutate it until it reaches the desired state,
at which point we can call \texttt{execve} in the child thread to start it running on its own.
After that, the process is no longer under our control, and it functions like any other child process.
(Hence we call it ``process'' rather than ``thread'')
The child process can be monitored using normal Linux child monitoring syscalls from our main thread,
such as \texttt{waitid},
which we can also do with \texttt{rsyscall}.

Resource management across multiple processes can be complex,
as we'll see in section \ref{implementation}.
To ease the complexity for user code,
rsyscall provides substantial language-specific libraries,
which allow for resource management in a language-idiomatic fashion.

To reduce distracting non-essential complexity,
we will show all our examples with the Python rsyscall library.
We will not use any substantial abstractions;
besides process creation, which has complexities we'll discuss in section \ref{clone},
each function call corresponds to exactly one syscall.
Python-specific asynchronous execution syntax (\texttt{await}) has been removed for clarity,
but the examples are otherwise real working code.

\subsection{Example: Create and exec}
\begin{lstlisting}[float,language=Python,label={basic},caption={Creating a new process and immediately execing}]
# Use the clone syscall in the local thread to create a new thread;
# we use a wrapper that supplies defaults for all arguments.
child = local.thread.clone()
# Call execve to run a different executable in the child thread;
# We pass the executable path as the first argument in the argument list, as is traditional.
# We use a wrapper that defaults envp to an unchanged environment, so we don't pass envp.
child.execve(hello_path, [hello_path])
\end{lstlisting}
In Listing \ref{basic}, we simply create a new process under our control (a thread)
and immediately exec a binary with it.
As exceptions are used for error-handling in the Python API,
there is no need for error-checking code.
%% **** note to self: concepts introduced in this example
%% local.thread
%% clone
%% execve
\subsection{Example: Passing down FDs}
\begin{lstlisting}[float,language=Python,label={fds},caption={Passing down FDs}]
child = local.thread.clone()
sock = child.socket(AF.INET, SOCK.DGRAM)
# Bind the socket to a sockaddr_in;
# the sockaddr is allocated in memory with child.ptr and is garbage collected.
sock.bind(child.ptr(SockaddrIn(0, 0)))
sock.listen(10)
sock.disable_cloexec()
child.execve(executable_path, [executable_path, "--listening-socket", str(int(sock))])
\end{lstlisting}
In Listing \ref{fds}, we create a new thread,
then create a listening socket bound to a random port in that thread,
then call exec, 
passing down the socket by disabling cloexec and passing its file descriptor number as an argument to the new program.

File descriptors, here, are object oriented and have relevant syscalls as methods.
They make syscalls in the process they are created in by default;
we can create more objects referring to the same file descriptor from different processes
if we want to make the syscalls from another process.
%% **** note to self: concepts introduced in this example
%% disable_cloexec
%% socket creation
%% using int on sock
%% .ptr
\subsection{Example: Pipeline}
\begin{lstlisting}[float,language=Python,label={pipe},caption={Creating a pipeline}]
# create the pipe
pipe = local.thread.pipe()
child1 = local.thread.clone()
# inherit the write-end of the pipe to child1, and replace child1.stdout with it
child1.inherit_fd(pipe.write).dup2(child1.stdout)
child1_proc = child1.execve(yes_path, [yes_path])
child2 = local.thread.clone()
# inherit the read-end of the pipe to child2, and replace child2.stdin with it
child2.inherit_fd(pipe.read).dup2(child2.stdin)
child2_proc = child2.execve(head_path, [head_path, "-n", "15"])
\end{lstlisting}
In Listing \ref{pipe},
we do the same as the Unix shell pipeline "yes | head -n 15".
We create a pipe,
then create two threads,
connect them with a pipe,
and exec a different program in each thread.

After a process is created with clone,
it may have inherited file descriptors;
here we inherit the pipe.
We make this inheritance explicit with \texttt{inherit\_fd},
a helper method on our thread object,
which takes a file descriptor from a different thread
and performs a runtime check that the file descriptor actually was inherited.
If so, it returns a new handle to the file descriptor which performs syscalls from the new thread.

Then we simply =dup2= as normal to replace child1's stdout with the write end of the pipe;
=dup2= disables CLOEXEC by default on the target.
%% **** note to self: concepts introduced in this example
%% dup2
%% pipe
\subsection{Example: Mount namespace}
\begin{lstlisting}[float,language=Python,label={mount},caption={Mount namespace}]
child = local.thread.clone(CLONE.NEWUSER|CLONE.NEWNS)
child.mkdir(rootdir/"proc")
child.mount(Path("/proc"), rootdir/"proc", "", MS.BIND, "")
child.chroot(rootdir)
child.execve(executable_path, [executable_path])
\end{lstlisting}
In Listing \ref{mount},
we make a new mount namespace and rearrange the filesystem tree for the child process.
We bind-mount /proc at /proc inside the chroot directory,
chroot into the directory,
and exec an executable which will run inside the chroot.
%% **** note to self: concepts introduced in this example
%% namespaces
%% path slash syntax sugar
\subsection{Example: Nested clone and network namespace}
\begin{lstlisting}[float,language=Python,label={nested},caption={Nested clone and network namespace}]
ns_thread = local.thread.clone(CLONE.NEWNET|CLONE.NEWUSER)

listening_child = ns_thread.clone()
sock = listening_child.socket(AF.INET, SOCK.DGRAM)
sockaddr = SockaddrIn(22, "127.0.0.1")
sock.bind(listening_child.ptr(sockaddr))
sock.listen(10)
sock.disable_cloexec()
child.execve(server_path, [server_path, "--listening-socket", str(int(sock))])

connecting_child = ns_thread.clone()
child.execve(client_path, [
  client_path, "--connect-address", str(sockaddr.address) + ":" + str(sockaddr.port)])
\end{lstlisting}
In Listing \ref{nested},
we make a process (\texttt{ns\_thread}) in a new network namespace.
Then, we create two more child processes of \texttt{ns\_thread},
which are also in the new network namespace.
This nested creation of child processes is fully supported,
like all other syscalls,
and allows us to set up complex graphs of processes and namespaces.

We bind to a privileged port on localhost inside the namespace,
and create one child to listen on that socket,
and another child to connect to it.
%% **** note to self: concepts introduced in this example
%% network namespace
%% nested clone
% \subsection{Example: Starting a complex system}
% \begin{lstlisting}[float,language=Python,label={miredo},caption={Starting a complex system}]
% ### create socket outside network namespace that Miredo will use for internet access
% inet_sock = local.thread.socket(AF.INET, SOCK.DGRAM)
% inet_sock.bind(local.thread.ptr(SockaddrIn(0, 0)))
% # set some miscellaneous additional sockopts that Miredo wants
% set_miredo_sockopts(local.thread, inet_sock)
% ### create main network namespace thread
% ns_thread = local.thread.clone(CLONE.NEWNET|CLONE.NEWUSER)
% ### create in-network-namespace raw INET6 socket which Miredo will use to relay pings
% icmp6_fd = ns_thread.socket(AF.INET6, SOCK.RAW, IPPROTO.ICMPV6)
% ### create in-network-namespace socket,
% ### which Miredo will use for unassociated Ifreq ioctls
% reqsock = ns_thread.socket(AF.INET, SOCK.STREAM)
% ### create and set up the TUN interface
% tun_fd, tun_index = make_tun(ns_thread, "miredo", reqsock)
% ### create socketpair which Miredo will use to communicate
% ### between privileged process and Teredo client
% privproc_pair = ns_thread.socketpair(AF.UNIX, SOCK.STREAM)
% ### start up privileged process which manipulates the network setup in the namespace
% privproc_thread = ns_thread.clone()
% # We preserve NET_ADMIN capability over exec so that privproc can
% # manipulate the TUN interface. add_to_ambient_caps is a helper function
% # used because manipulating Linux ambient capabilities is fairly verbose
% add_to_ambient_caps(privproc_thread, {CAP.NET_ADMIN})
% # privproc expects to communicate with the main client over stdin and stdout
% privproc_side = privproc_thread.inherit_fd(privproc_pair.first)
% privproc_side.dup2(privproc_thread.stdin)
% privproc_side.dup2(privproc_thread.stdout)
% privproc_child = privproc_thread.execve(miredo_privproc_executable_path, [
%     miredo_privproc_executable_path, str(tun_index)
% ])
% ### start up Miredo client process which communicates
% ### over the internet to implement the tunnel
% # the client process doesn't need to be in the same network namespace,
% # since it is passed all the resources it needs as fds at startup.
% client_thread = ns_thread.clone(CLONE.NEWUSER|CLONE.NEWNET|CLONE.NEWNS|CLONE.NEWPID)
% # lightly sandbox by unmounting everything except for the executable
% # and its deps (known via package manager)
% unmount_everything_except(client_thread, miredo_exec.run_client.executable_path)
% # a helper function for preparing the fds that are passed as command line arguments
% async def pass_fd(fd: FileDescriptor) -> str:
%     client_thread.inherit_fd(fd).disable_cloexec()
%     return str(int(fd))
% client_child = client_thread.execve(miredo_client_executable_path, [
%     miredo_client_executable_path,
%     pass_fd(inet_sock), pass_fd(tun_fd), pass_fd(reqsock),
%     pass_fd(icmp6_fd), pass_fd(privproc_pair.second),
%     "teredo.remlab.net", "teredo.remlab.net"
% ])
% \end{lstlisting}

% In Listing \ref{miredo},
% we show non-trivial code for launching a real application:
% the Miredo IPv6 tunneling software.
% We use a few helper functions in this listing to keep the attention focused on the interesting parts.

% Miredo is separated into two components, a privileged process which sets up network interfaces,
% and an unprivileged process which talks to the network.
% With minimal modifications to Miredo,
% we launch Miredo entirely unprivileged inside a user namespace and network namespace,
% with all resources created outside and explicitly passed in.

%% **** note to self: concepts introduced in this example
%% socketpair
%% several helper functions
%% **** note to self
%% maybe I should remove some of the unnecessary networking stuff?
%% maybe show the corresponding fork-based code?
%% except doing something with fork with the same error handling requires a bunch of crazy IPC
\section{Implementation}\label{implementation}
\subsection{Basics about rsyscall}
Our main need for implementing direct-style process creation
is a robust system for cross-process syscalls.
We provided this in the rsyscall project.
rsyscall is a toolkit for cross-process syscalls on Linux,
with several language-specific library implementations.

In this section, we'll give a brief overview of rsyscall,
and focus on implementation issues specific to process creation.

rsyscall can be conceptually divided in two parts:
the basic cross-process syscall primitive,
and a language-specific library built on top
to handle the complexities of manipulating resources across multiple processes.
The Python language-specific library has already been demonstrated above.
Such libraries only need to be able to call syscalls and explicitly specify a process in some way;
they are, for the most part, agnostic to how the cross-process syscall is implemented.

Using the Python library as an example,
it provides Python wrappers for Linux system calls and structs
which are type-safe using Python 3 type annotations and runtime checks
while still providing low-abstraction access to a large subset of native Linux functionality.
It also provides garbage collection for remote file descriptors, memory and other resources.
Such features are independent of the precise implementation of the cross-process syscall primitive.

On Linux \verb|x86_64|, a syscall is specified by a syscall number plus six register-sized arguments;
a syscall returns one register-sized value.
rsyscall's default implementation of cross-process syscalls sends those seven integers over a pipe,
and waits for a response on another pipe.
Processes are created running an infinite loop which, at each iteration,
reads a syscall request off the pipe,
performs that syscall,
and writes the return value back over the return pipe.
In this way, a cross-process syscall works much like a very primitive remote procedure call.

Many syscalls either take or return pointers to memory,
and require the caller to read or write that memory to provide arguments or receive results.
Therefore, an rsyscall library needs a way to access memory in the target process.
We implement this through another set of pipes,
by explicitly copying memory into and out of those pipes using the \texttt{read} and \texttt{write} system calls.
When we wish to read \texttt{N} bytes of memory at address \texttt{A} in the target process,
we first perform a \verb|write(memory_pipe, A, N)| in the target process,
and then read that data off the other end of the pipe in the parent process.
When we wish to write \texttt{N} bytes of data at address \texttt{A} in the target process,
we first write that data to the pipe in the parent process,
then perform a \verb|read(memory_pipe, A, N)| in the target process to copy that data from the pipe into memory.

ptrace provides an alternative means to perform arbitrary actions on other processes.
However, among other issues, it has the unavoidable substantial disadvantage of not permitting multiple ptracers.
A ptrace-based implementation would prevent using strace or gdb on rsyscall-controlled processes,
which is an unacceptable limitation for a general-purpose utility.

The \verb|process_vm_readv| and \verb|process_vm_writev| system calls
allow the caller to read and write memory from the virtual address space of other processes.
However, they require that the caller have specific credentials relative to the process being accessed,
which may not always be the case.
Additionally, these system calls are disabled if ptrace is disabled system-wide,
which is a niche but possible system configuration.
To ensure that rsyscall can be used for arbitrary purposes and on arbitrary systems, we avoided these calls.
\subsection{\texttt{clone}}\label{clone}
Now that we've established the basic operations which rsyscall provides,
let's consider the specific issues related to process creation and initialization.

There are three Linux system calls which create processes:
\texttt{fork}, \texttt{vfork} and \texttt{clone}.
\texttt{clone} provides a superset of the functionality of the other two,
so we focused our attention on \texttt{clone}.

\texttt{clone} (along with \texttt{fork}) creates a new process
which immediately starts executing at the next instruction after the syscall instruction,
in parallel with the parent process,
with its registers in generally the same state as the parent process.
\footnote{Note that =glibc= defines a wrapper for the clone syscall;
we are talking about the raw kernel syscall.}
In the style of Plan 9's \texttt{fork} syscall\cite{rfork}, which inspired \texttt{clone},
\texttt{clone} takes a mask of flags which determines whether several attributes of the new process
are either shared with, or copied from, the parent process.

\texttt{clone} only lets us change the stack register for the new process.
We would like to be able to set arbitrary registers for the new process,
so that we can control where it begins executing and the stack it executes on.
Fortunately, changing the stack is sufficient.

We ensure that the next instruction executed after any syscall
is (in x86 terms) a \texttt{RET};
this is always the case, so we have no need to special case the execution of \texttt{clone}.
Since we control the stack of the new process,
the \texttt{RET} will jump to a code address that we control.
We can then supply additional arguments to this code
by putting them on the stack.

We typically cause the new process to jump to a trampoline provided by the rsyscall library
which sets all registers to values found on the stack
and then jumps to another address.
\footnote{This is also a generally useful utility for hackers performing return-oriented-programming attacks;
but similar functionality exists in any standards-compliant C library,
so there is no increase in attack surface.}
With this trampoline,
we can provide a helper Python function that,
when given a function pointer following C calling conventions, and some arguments,
will prepare a stack for a call to clone such that the new process will call that function with those arguments.

With our new ability to call arbitrary C-compatible functions,
we can now call \texttt{clone} so that it launches a process running our infinite syscall loop,
which is implented in C and, as described in the previous section,
uses two pipes passed as arguments to receive syscall requests and respond with syscall results.

% TODO is this section necessary?
The addresses of these functions and trampolines are discovered through a linking procedure.
When the process being created is in the same address space as the main process which is running user code,
the location of the rsyscall library in memory, and the addresses of code within it,
are known through normal language-specific linking mechanisms.
However, when a process is created with a different address space,
such as when we establish a connection to a process after it's been started,
we need to perform linking to learn the addresses of functions.
This linking procedure is performed while bootstrapping the connection,
and involves the target process sending a table of important addresses to the connecting process.

After using \texttt{clone} to create a new process running our syscall loop,
most system calls can be called as normal.
The new process can be modified freely through chdir, dup2, and other system calls.
Out of system calls related to process creation,
only \texttt{execve} and \texttt{unshare} need substantial further attention.
\subsection{\texttt{exec}}
Eventually, most programs will want to call \texttt{execve} in the processes they create.
\texttt{execve} is unusual and requires careful design,
because when it is successful, it does not return.
Therefore we need a way to determine if \texttt{execve} is successful;
naively waiting for a response to the syscall request will leave us waiting forever.

One traditional means to detect a successful \texttt{execve} is to create a pipe before forking,
ensure both ends are marked \verb|O_CLOEXEC|,
perform the fork,
call \texttt{execve} in the child,
close the write end of the pipe,
and wait for EOF on the read end.
If the child process has neither successfully called \texttt{execve}, nor exited for some other reason,
then the write end of the pipe will still be open in the child process's fd table,
and the read end of the pipe will not return EOF.
But once the child process calls \texttt{execve} successfully,
\verb|O_CLOEXEC| will cause the write end of the pipe to be closed,
and the read end of the pipe will return EOF.

This trick works well with \texttt{fork};
but it's not general enough to work with \texttt{clone}.
Child processes can be created with the \verb|CLONE_FILES| flag passed to \texttt{clone},
which causes the parent process and child process to share a single fd table.
This means that when the parent process closes the write end of the pipe,
it will also be closed in the child process,
and the read end of the pipe will immediately return EOF,
regardless of whether the child has called \texttt{execve} or exited.

Fortunately, there is an alternative solution, which does work with \verb|CLONE_FILES|.
The \texttt{ctid} argument to \texttt{clone} specifies a memory address which,
when the \verb|CLONE_CHILD_CLEARTID| flag is set,
the kernel sets to zero when the child exits or execs,
and then, crucially, performs a futex wakeup on.
More specifically,
the kernel clears and does a futex wakeup on \texttt{ctid} when the child process leaves its current address space;
this precisely coincides with exiting or execing,
since those are the only way to change address space in Linux as of this writing.

A futex is a Linux-specific feature,
which is generally used for the implementation of userspace shared-memory synchronization constructs,
such as mutexes and condition variables.
The relevant detail for us here is that we can wait on an address
until a futex wakeup is performed on that address;
that means we can wait on \texttt{ctid} until the futex wakeup is performed,
and in this way get notified of the child process calling \texttt{execve}.

Unfortunately, futexes in current Linux integrate poorly:
There is no way for a single process to wait for more than one futex at a time,
and no way to monitor a futex with file-descriptor-monitoring syscalls such as \texttt{poll}.
The best we can do is create a dedicated child process for each futex we want to wait on,
and have this child process exit when the futex has a wakeup.
Monitoring child processes can be straightforwardly integrated into an event loop.

While slightly complex to implement, this solution works well.
We provide \texttt{ctid} whenever we call \texttt{clone},
and set up a process to wait on that futex.
Then, when we call \texttt{execve},
we wait for either the \texttt{execve} to return an error or the futex process to exit,
whichever comes first.
If the futex process exits,
and the child process doesn't itself exit,
we know that the child has successfully called \texttt{execve}.

If the futex process and child process both exit,
it's ambiguous whether the child process successfully called \texttt{execve};
this ambiguity is unfortunate, but it is also present in the pipe-based approach.
This is, we believe, the best solution currently available.

We would prefer for Linux to natively provide functionality to wait for a child's \texttt{execve}.
Some other Unix-like systems provide this;
kqueue, on FreeBSD, allows waiting for exec in arbitrary processes through kqueue's \verb|EVFILT_PROC|.
One approach for Linux would be to add a new \texttt{clone} flag to opt-in to receiving \texttt{WEXECED} events through \texttt{waitid};
note that a \texttt{waitid} flag alone is not sufficient,
since it's necessary to receive \texttt{SIGCHLD} signals for the \texttt{WEXECED} event if waiting for it from an event loop.
Alternatively, some way to wait for futex wakeups through a file descriptor could be added,
so we can use file-descriptor-monitoring syscalls to wait for the \texttt{ctid} futex;
such a feature used to exist in the form of \verb|FUTEX_FD|,
but was removed from Linux long ago due to race conditions in its design.
\subsection{Managing file descriptor tables}\label{fdtables}
As mentioned in the previous section,
the \verb|CLONE_FILES| flag can be passed to \texttt{clone}.
When this flag is passed,
the file descriptor table is shared between the parent process and child process.
The same file descriptors are open in both processes at the same numbers,
and if new file descriptors are opened in either process,
they are also visible in the other process.
This is simple to model,
and convenient for many purposes;
for example, the child process might be in a different network namespace from the parent,
and the shared file descriptor table would allow the child to bind a socket
and the parent to use it.

If \verb|CLONE_FILES| is not passed to \texttt{clone},
then \texttt{clone} has the same behavior as \texttt{fork}:
The new process has a new file descriptor table,
containing copies of all the file descriptors existing in the parent at the time of the system call.
This same behavior can also be triggered after process creation by calling \verb|unshare(CLONE_FILES)| or \texttt{execve};
if \verb|unshare(CLONE_FILES)| or \texttt{execve} (ignoring \texttt{CLOEXEC}, which we'll discuss later),
are called in a process currently sharing its file descriptor table with another process,
then after the call that process will have a new, private file descriptor table,
again with a copy of all the file descriptors existing at the time of the system call.

After a system call has copied all the file descriptors in the old table into the new table,
we need to decide which file descriptors we want to keep open in the new table,
and which file descriptors should be closed.
Keeping some file descriptors from the old table in the new table
is referred to as "file descriptor inheritance".
\subsubsection{Inheriting file descriptors}
The rule about which file descriptors stay open is simple:
We want to keep a file descriptor open in the new table
if there is a process using that file descriptor.
We track which processes are using which file descriptors as part of our file descriptor garbage collection system.
File descriptors are used through garbage-collected handles,
each of which is associated with a process.
If there is an existing handle for a (process, file descriptor number) combination,
this means that the file descriptor with that number in that process's file descriptor table is in use by that process.

This makes inheritance simple in the case of \verb|unshare(CLONE_FILES)|.
Every file descriptor in use by a process
is referenced by a (process, file descriptor number) combination;
these handles are still valid after a call to \verb|unshare(CLONE_FILES)|,
they simply now refer to the new file descriptor in the process's new file descriptor table.
So, these file descriptors will automatically be kept open;
that is, they will be automatically inherited into the new table.
This same mechanism can work across \texttt{execve},
if control over the process is re-established after the \texttt{execve},
and if we carefully manage \texttt{CLOEXEC}.

When a process is newly created,
no file descriptor handles yet exist which involve that process.
Once a handle is created with that process and some file descriptor number,
the referenced file descriptor will be inherited into (that is, kept open in) the new file descriptor table.
These handles can be created through the \verb|inherit_fd| function.

At the time of creating a new file descriptor table,
we perform some bookkeeping:
We make a list of all the file descriptors that existed in the old table
at the time of the creation of the new table.
These file descriptors are the ones which were copied into the new table.

The \verb|inherit_fd| function uses this list.
It takes as arguments a new process and a file descriptor handle from another process,
and checks (with the list) that that file descriptor was copied into the new process's file descriptor table
and hasn't since been closed.
If so, it creates a new handle for that (process, file descriptor number) combination.
This causes the file descriptor to stay open in the new file descriptor table,
and it can be used in the new process through the handle.
\subsubsection{Closing file descriptors after inheritance}
After file descriptor inheritance is complete,
we must promptly close other file descriptors we don't want to inherit.
Leaving these file descriptors open in the new table is a form of resource leakage.
It can also cause erroneous behavior.
For example, it's a common practice to close the write end of a pipe
and expect an EOF on the read end;
if the write end is copied into the new file descriptor table before being closed,
and the write end is never closed in the new table,
the read end will never get an EOF.

However, we can't simply close all other file descriptors.
The possibility of implicit inheritance of file descriptors is a traditional Unix feature,
which is useful in a wide variety of situations,
in much the same way as implicit inheritance of environment variables;
it can allow a resource to be passed down a process hierarchy without intervening programs being aware.

Here is where \texttt{CLOEXEC} becomes relevant:
\texttt{CLOEXEC}, in practice,
is a tag, set by userspace, for file descriptors which should not be implicitly inherited.
If \texttt{CLOEXEC} is set on a file descriptor,
we should close that file descriptor if we don't explicitly want to inherit it;
if \texttt{CLOEXEC} is not set, we should not close the file descriptor,
but instead we should allow it to be implicitly inherited.

This interpretation of \texttt{CLOEXEC} is a consequence of \texttt{CLOEXEC}'s primary purpose:
Managing inheritance of file descriptors over \texttt{execve},
in programs where the caller of \texttt{execve}
doesn't know all the file descriptors that may have been opened by the rest of the program;
this describes, in practice, all programs which call \texttt{execve}.
When \texttt{CLOEXEC} is set on a file descriptor,
it will not be copied into the new file descriptor table created after \texttt{execve}.
Thus \texttt{CLOEXEC} is a way of saying, before an \texttt{execve},
that this file descriptor will not be used by the new program after an \texttt{execve},
and therefore should not be inherited.
Since, in a general purpose program, \texttt{execve} may be called at any time,
libraries must have \texttt{CLOEXEC} set correctly at all times.
So we can check \texttt{CLOEXEC} at any time to see whether a file descriptor should be inherited or not,
even if the file descriptor is used by a library unrelated to rsyscall.

The implementation of closing non-inherited file descriptors is then simple:
We close all file descriptors which have \texttt{CLOEXEC} set
and which aren't referenced by an rsyscall file descriptor handle.
We also clear the list of copied file descriptors which \verb|inherit_fd| uses.
We do this closing operation in userspace;
a syscall to perform this would be a useful addition to Linux.

\texttt{CLOEXEC} is set by default on all file descriptors opened by rsyscall,
though it may be unset by a user program.
Many user programs will unset \texttt{CLOEXEC} on some file descriptors immediately before calling \texttt{execve}
so that the executable they run will inherit those file descriptors.
The new program run by the \texttt{execve} will typically immediately set \texttt{CLOEXEC} again.

An additional argument to the \texttt{execve} syscall
which allows specifying an explicit list of file descriptor numbers to inherit despite the \texttt{CLOEXEC} flag being set
would allow programs to avoid this pointless behavior of un-setting and then immediately re-setting \texttt{CLOEXEC}.
It would also, more significantly, allow programs to inherit file descriptors across \texttt{execve}
while the file descriptor table is shared;
unsetting \texttt{CLOEXEC} while the file descriptor table is shared will cause race conditions,
because the other processes sharing the file descriptor table might call \texttt{execve} at any time,
and so the file descriptor table must first be unshared before unsetting \texttt{CLOEXEC} and calling \texttt{execve}.
\section{Evaluation}\label{evaluation}
\subsection{performance ideas}
Can mmap a bunch of stuff in a process and then create some children,
and show that without having to copy page tables (due to \verb|CLONE_VM|), it's faster.

Can make the classic chart of fork vs glibc's \verb|posix_spawn| vs direct-style.
\subsection{abstraction ideas}
Implementing Python's popen/subprocess with rsyscall?
But that's not fair, theirs is portable, we'll be way simpler for free.

Rewrite some existing system with rsyscall?
Write a shell?
\section{Future work}\label{future_work}
\subsection{Other applications of rsyscall}
Most avenues of future work focus on rsyscall.
rsyscall was not developed solely for the purpose of this paper,
and it has many uses unrelated to direct-style process creation,
such as asynchronous system calls, exceptionless system calls\cite{flexsc}, cross-host operations, among others.
We are actively exploring such applications,
as well as broadening rsyscall's language support.
\subsection{Kernel support}
rsyscall's cross-process syscalls can be performed entirely in userspace,
which has substantial benefits for deployability.
Nevertheless, direct kernel support for creating a stub process and performing syscalls in the context of that process
may provide efficiency benefits, as well as reducing userspace-visible complexity.

Several other aspects of our implementation would be improved by kernel support.
We discussed these in the implementation section;
in brief, we would most benefit from kernel support for
detecting when a child process finishes =execve=,
closing all =CLOEXEC= file descriptors except for an explicitly specified list,
and explicitly specifying a list of =CLOEXEC= file descriptors to inherit over =excve=.
Implementing these features in the kernel in a generally useful way, and upstreaming them,
is an important direction for our future work.
\subsection{Portability to other Unix systems}
Other non-Linux systems
could adopt the techniques of this paper
to provide direct-style process creation.
Currently, our focus is on Linux,
but others may wish to explore porting these techniques to other operating systems.
\subsection{Large scale open source usage}
We have made use of the techniques described in this paper
in proprietary software at Two Sigma.
While this gives us personally greater confidence in these techniques,
it would be better to use them in a publicly available, open source system.
Either porting an existing system to use these techniques,
or using these techniques to create a substantial new system from scratch,
would provide a meaningful demonstration of the viability of these techniques.
% \subsection{file descriptor lifetime management}
% % TODO this is hard to explain, probably best to drop this
% Keeping file descriptors open if and only if there is a specific process using that file descriptor
% is not the only possibility.
% Keep in mind that "a process using the file descriptor" is a slight abuse of terminology;
% processes don't use file descriptors, programs do, and in our system there is only one program.
% To reflect this, we could instead more directly use file descriptors in a process-agnostic way;
% this would support the creation of objects
% which work transparently across multiple processes which share a file descriptor table.
% Such objects would automatically use the relevant process to perform the syscalls for any specific operation.
% A process-centric view of file descriptors instead forces each object to be associated with one process.
% Nevertheless, we found that the process-centric perspective better matches the existing intuitions of users,
% especially those with prior experience in programming with processes.
% We hope that future systems for multi-process programming
% might explore an object-centric approach for managing resources.
\section{Conclusion}\label{conclusion}
Direct-style process creation is much less known and much less used than fork-style and spawn-style.
We have implemented direct-style process creation for Linux.
Our implementation is immediately deployable on today's Linux systems.
We have discussed various applications of processes,
and demonstrated the use of Linux direct-style process creation
to implement them.
We hope that this work will help encourage more use of the process abstraction,
which, though widespread,
is still not used to its full potential.

\bibliographystyle{ACM-Reference-Format}
\bibliography{bibliography}

\end{document}

### 2020-05-04 research/paper/paper.tex
DEV REVISE partial stuff

@@
 child = local.thread.clone()
-# Call execve to run a different executable in the child thread;
+# Call execv to run a different executable in the child thread;
 # We pass the executable path as the first argument in the argument list, as is traditional.
-# We use a wrapper that defaults envp to an unchanged environment, so we don't pass envp.
-child.execve(hello_path, [hello_path])
+child.execv(hello_path, [hello_path])
 \end{lstlisting}
@@
 sock.disable_cloexec()
-child.execve(executable_path, [executable_path, "--listening-socket", str(int(sock))])
+child.execv(executable_path, [executable_path, "--listening-socket", str(int(sock))])
 \end{lstlisting}
@@
 child1.inherit_fd(pipe.write).dup2(child1.stdout)
-child1_proc = child1.execve(yes_path, [yes_path])
+child1_proc = child1.execv(yes_path, [yes_path])
 child2 = local.thread.clone()
@@
 child2.inherit_fd(pipe.read).dup2(child2.stdin)
-child2_proc = child2.execve(head_path, [head_path, "-n", "15"])
+child2_proc = child2.execv(head_path, [head_path, "-n", "15"])
 \end{lstlisting}
@@
 child.chroot(rootdir)
-child.execve(executable_path, [executable_path])
+child.execv(executable_path, [executable_path])
 \end{lstlisting}
@@
 sock.disable_cloexec()
-child.execve(server_path, [server_path, "--listening-socket", str(int(sock))])
+child.execv(server_path, [server_path, "--listening-socket", str(int(sock))])
 
 connecting_child = ns_thread.clone()
-child.execve(client_path, [
+child.execv(client_path, [
   client_path, "--connect-address", str(sockaddr.address) + ":" + str(sockaddr.port)])
@@
 % privproc_side.dup2(privproc_thread.stdout)
-% privproc_child = privproc_thread.execve(miredo_privproc_executable_path, [
+% privproc_child = privproc_thread.execv(miredo_privproc_executable_path, [
 %     miredo_privproc_executable_path, str(tun_index)
@@
 %     return str(int(fd))
-% client_child = client_thread.execve(miredo_client_executable_path, [
+% client_child = client_thread.execv(miredo_client_executable_path, [
 %     miredo_client_executable_path,
@@
 Write a shell?
+
+Right okay so we can do some fork-style code, sure.
+Show the IPC, not too complex, but show it.
+
+We can do it in Python with rsyscall - have a hidden experimental secret fork thing.
+Using write_all_bytes will be annoying but fine.
+Ugh, but we'll have to decode it too.
+Eh, no, we can wrap that.
+Yeah then it's fine.
 \section{Future work}\label{future_work}

### 2020-05-05 python/rsyscall/environ.py
environ: don't construct a useless ExecutablePathCache on inherit

@@ -96,13 +96,17 @@ class ExecutablePathCache:
 
 class Environment:
     "A representation of Unix environment variables."
-    def __init__(self, task: Task, ram: RAM, environment: t.Dict[str, str],
-                 arglist_ptr: WrittenPointer[ArgList]=None,
+    @staticmethod
+    def make_from_environ(task: Task, ram: RAM, environment: t.Dict[str, str]) -> Environment:
+        return Environment(environment, ExecutablePathCache(task, ram, environment.get("PATH", "").split(":")))
+
+    def __init__(self, environment: t.Dict[str, str], path: ExecutablePathCache,
+                 arglist_ptr: t.Optional[WrittenPointer[ArgList]]=None,
     ) -> None:
         self.data = environment
         self.sh = Command(Path("/bin/sh"), ['sh'], {})
         self.tmpdir = Path(self.get("TMPDIR", "/tmp"))
-        self.path = ExecutablePathCache(task, ram, self.get("PATH", "").split(":"))
+        self.path = path
         self.arglist_ptr = arglist_ptr
 
     def __getitem__(self, key: str) -> str:
@@ -139,9 +143,7 @@ class Environment:
         they're shared between all threads.
 
         """
-        env = Environment(task, ram, dict(self.data), self.arglist_ptr)
-        env.path = self.path
-        return env
+        return Environment(dict(self.data), self.path, self.arglist_ptr)
 
     async def as_arglist(self, ram: RAM) -> WrittenPointer[ArgList]:
         if self.arglist_ptr is None:

### 2020-05-05 python/rsyscall/tasks/local.py

@@ -149,7 +149,7 @@ async def _make_local_thread() -> Thread:
         NativeLoader.make_from_symbols(task, lib),
         epoller,
         await ChildProcessMonitor.make(ram, task, epoller),
-        Environment(task, ram, {**os.environ}),
+        Environment.make_from_environ(task, ram, {**os.environ}),
         stdin=task.make_fd_handle(near.FileDescriptor(0)),
         stdout=task.make_fd_handle(near.FileDescriptor(1)),
         stderr=task.make_fd_handle(near.FileDescriptor(2)),

### 2020-05-05 python/rsyscall/tasks/ssh.py

@@ -298,7 +298,7 @@ async def ssh_bootstrap(
         loader=NativeLoader.make_from_symbols(new_base_task, describe_struct.symbols),
         epoller=epoller,
         child_monitor=child_monitor,
-        environ=Environment(new_base_task, new_ram, environ),
+        environ=Environment.make_from_environ(new_base_task, new_ram, environ),
         stdin=new_base_task.make_fd_handle(near.FileDescriptor(0)),
         stdout=new_base_task.make_fd_handle(near.FileDescriptor(1)),
         stderr=new_base_task.make_fd_handle(near.FileDescriptor(2)),

### 2020-05-05 python/rsyscall/tasks/stdin_bootstrap.py

@@ -138,7 +138,7 @@ async def stdin_bootstrap(
         loader=NativeLoader.make_from_symbols(base_task, describe_struct.symbols),
         epoller=epoller,
         child_monitor=child_monitor,
-        environ=Environment(base_task, ram, environ),
+        environ=Environment.make_from_environ(base_task, ram, environ),
         stdin=base_task.make_fd_handle(near.FileDescriptor(0)),
         stdout=base_task.make_fd_handle(near.FileDescriptor(1)),
         stderr=base_task.make_fd_handle(near.FileDescriptor(2)),

### 2020-05-05 python/rsyscall/tasks/stub.py

@@ -179,7 +179,7 @@ async def _setup_stub(
         loader=NativeLoader.make_from_symbols(base_task, describe_struct.symbols),
         epoller=epoller,
         child_monitor=child_monitor,
-        environ=Environment(base_task, ram, environ),
+        environ=Environment.make_from_environ(base_task, ram, environ),
         stdin=base_task.make_fd_handle(near.FileDescriptor(0)),
         stdout=base_task.make_fd_handle(near.FileDescriptor(1)),
         stderr=base_task.make_fd_handle(near.FileDescriptor(2)),

### 2020-05-05 python/rsyscall/loader.py
loader: clean up trampoline to_bytes
This is a more sane way to do this.

@@ -58,23 +58,23 @@ class Trampoline(Borrowable):
 
 class TrampolineSerializer(Serializer[Trampoline]):
     def to_bytes(self, val: Trampoline) -> bytes:
-        args: t.List[int] = []
-        for arg in val.args:
+        # at most six arguments
+        args: t.List[int] = [0]*6
+        for i, arg in enumerate(val.args):
             if isinstance(arg, FileDescriptor):
-                args.append(int(arg))
+                args[i] = int(arg)
             elif isinstance(arg, Pointer):
-                args.append(int(arg.near))
+                args[i] = int(arg.near)
             else:
-                args.append(int(arg))
-        arg1, arg2, arg3, arg4, arg5, arg6 = args + [0]*(6 - len(args))
+                args[i] = int(arg)
         struct = ffi.new('struct rsyscall_trampoline_stack*', {
             'function': ffi.cast('void*', int(val.function.near)),
-            'rdi': int(arg1),
-            'rsi': int(arg2),
-            'rdx': int(arg3),
-            'rcx': int(arg4),
-            'r8':  int(arg5),
-            'r9':  int(arg6),
+            'rdi': args[0],
+            'rsi': args[1],
+            'rdx': args[2],
+            'rcx': args[3],
+            'r8':  args[4],
+            'r9':  args[5],
         })
         return bytes(ffi.buffer(struct))
 

### 2020-05-05 python/rsyscall/tasks/local.py
local: provide optimized syscall method

@@ -61,6 +61,24 @@ class LocalSyscall(SyscallInterface):
     async def close_interface(self) -> None:
         pass
 
+    async def syscall(self, number: SYS, arg1=0, arg2=0, arg3=0, arg4=0, arg5=0, arg6=0) -> int:
+        """Send a syscall and wait for it to complete, throwing on error results.
+
+        See SyscallInterface.syscall.
+
+        SyscallInterface provides a default implementation for this method, but we override it for performance reasons;
+        since we immediately get the result, we don't need to create a trio.CancelScope as SyscallInterface.syscall
+        does, which is sadly quite expensive.
+
+        """
+        log_syscall(self.logger, number, arg1, arg2, arg3, arg4, arg5, arg6)
+        result = await _direct_syscall(
+            number,
+            arg1=int(arg1), arg2=int(arg2), arg3=int(arg3),
+            arg4=int(arg4), arg5=int(arg5), arg6=int(arg6))
+        raise_if_error(result)
+        return result
+
     async def submit_syscall(self, number, arg1=0, arg2=0, arg3=0, arg4=0, arg5=0, arg6=0) -> LocalSyscallResponse:
         """Make a syscall in the local thread; return SyscallResponse already containing the result.
 

### 2020-05-05 python/rsyscall/sched.py
sched: tweak Stack.to_bytes
This change doesn't detectably speed it up, but at least it's done
now, so the next time I look at this function for optimization, it
will be more obvious that the slow part is the byte concatenation.

@@ -53,6 +53,8 @@ class Borrowable:
 
 T_borrowable = t.TypeVar('T_borrowable', bound=Borrowable)
 
+_address = struct.Struct("Q")
+
 @dataclass
 class Stack(Serializable, t.Generic[T_borrowable]):
     """The stack argument to clone
@@ -76,7 +78,7 @@ class Stack(Serializable, t.Generic[T_borrowable]):
         self.data.borrow_with(stack, task)
 
     def to_bytes(self) -> bytes:
-        return struct.Struct("Q").pack(int(self.function.near)) + self.serializer.to_bytes(self.data)
+        return _address.pack(int(self.function.near)) + self.serializer.to_bytes(self.data)
 
     T_stack = t.TypeVar('T_stack', bound='Stack')
     @classmethod

### 2020-05-07 python/rsyscall/tests/test_fork.py
add test_fork

@@ -1,0 +1,96 @@
+from rsyscall.trio_test_case import TrioTestCase
+import rsyscall.tasks.local as local
+import os
+import time
+import subprocess
+from rsyscall.sys.wait import W
+import typing as t
+
+class TestProc(TrioTestCase):
+    async def asyncSetUp(self) -> None:
+        self.local = local.thread
+
+    async def test_fork(self) -> None:
+        pid = os.fork()
+        if pid == 0:
+            try:
+                os.execv(echo_path, [echo_path, "hello"])
+            except OSError as e:
+                await ipc.send(e)
+                os.exit(1)
+        else:
+            result = await ipc.recv()
+            if result.is_eof:
+                print("successfully exec'd")
+            elif result.is_exception:
+                raise result.exception
+
+    # equivalent to
+    async def test_direct(self) -> None:
+        child = self.local.clone()
+        child.execv(echo_path, [echo_path, "Hello world!"])
+        print("successfully exec'd")
+
+    async def test_bench(self) -> None:
+        count = 100
+        cmd = (await self.local.environ.which("echo")).args("hello world")
+        prep = 5
+        counts = [[], [], [], []]
+        async def rsys_run() -> None:
+            counts[0].append(time.time())
+            thread = await self.local.clone()
+            counts[1].append(time.time())
+            child = await thread.execv(cmd.executable_path, [cmd.executable_path, "-n", "hello"])
+            counts[2].append(time.time())
+            await child.waitpid(W.EXITED)
+            counts[3].append(time.time())
+            await thread.close()
+        async def subp_run() -> None:
+            subprocess.run(["echo", "hello world"])
+        run = rsys_run
+        # run = subp_run
+        before_prep = time.time()
+        for _ in range(prep):
+            await run()
+        before = time.time()
+        for _ in range(count):
+            await run()
+        after = time.time()
+        print("prep time", (before - before_prep)/prep)
+        print("real time", (after - before)/count)
+        
+        def process(starts: t.List[float], ends: t.List[float]) -> float:
+            return sum(end - start for start, end in zip(starts, ends))/count
+        print("clone", process(counts[0], counts[1]))
+        print("exec", process(counts[1], counts[2]))
+        print("waitpid", process(counts[2], counts[3]))
+
+import cProfile
+import trio
+async def prep() -> None:
[…]

### 2020-05-07 python/rsyscall/tests/test_benchmark.py
test_benchmark: add
This is a useful starting point for performance tuning.

@@ -1,0 +1,29 @@
+from rsyscall.tests.trio_test_case import TrioTestCase
+from rsyscall import local_thread, Command
+from rsyscall.sys.wait import W
+import typing as t
+import cProfile
+import pstats
+
+async def rsys_run(cmd: Command, count: int) -> None:
+    for _ in range(count):
+        thread = await local_thread.clone()
+        child = await thread.exec(cmd)
+        await child.waitpid(W.EXITED)
+
+class TestBenchmark(TrioTestCase):
+    async def test_bench(self):
+        "Run a simple clone and exec case inside cProfile (without asserting the results)"
+
+        await local_thread.environ.as_arglist(local_thread.ram)
+        cmd = (await local_thread.environ.which("echo")).args("-n")
+        pr = cProfile.Profile()
+        warm_up_runs = 1
+        real_runs = 1
+        await rsys_run(cmd, warm_up_runs)
+        pr.enable()
+        await rsys_run(cmd, real_runs)
+        pr.disable()
+        # pr.print_stats(sort='cumtime')
+        # ps = pstats.Stats(pr).strip_dirs().sort_stats('cumulative')
+        # ps.print_callees()

### 2020-05-07 research/paper/paper.tex
notes and paper

@@
 With both fork-style and spawn-style,
-it is difficult to express
-branching behavior based on results or errors encountered while initializing the new process;
+it is difficult to
+branch based on results or errors encountered while initializing the new process;
 fork-style requires IPC if it wishes to communicate errors back to the rest of the program,
 while robust indication of errors in spawn-style requires substantial growth in the spawn interface.
+Additionally, modern operating systems such as Linux have features
+which are difficult or impossible to fully exploit through fork-style or spawn-style.
 
@@
 Direct-style does not constrain the calling process,
-can create any kind of process supported by the underlying system,
-and easily expresses errors in any individual step of the initialization.
-
-Our contribution is an implementation of direct-style process creation for Linux.
-Our implementation has all the advantages of direct-style
-over Linux's fork-style \texttt{fork} syscall and spawn-style \texttt{posix\_spawn} libc function.
+easily branches at any step of the initialization,
+and can create any kind of process supported by the underlying system.
+
+Our contribution is an implementation of direct-style process creation for Linux,
+making use of our work on the rsyscall project.
+The rsyscall project develops language-specific ``process-independent'' syscall libraries,
+replacing libc for interaction with Linux,
+allowing a program to perform cross-process syscalls.
+rsyscall currently has best support for Python,
+but can be ported to other languages;
+the examples we show in this paper will be in Python,
+but generalize easily to other languages.
 
@@
 and which runs no user code.
-The parent then uses \texttt{rsyscall}, a cross-process syscall library,
+The parent then uses \texttt{rsyscall}
 to call arbitrary syscalls in the child process to initialize it;
@@
 
+A direct-style \texttt{clone} has all the advantages of direct-style process creation
+over Linux's fork-style \texttt{fork} syscall and spawn-style \texttt{posix\_spawn} libc function.
+Direct-style \texttt{clone} can be used correctly and performantly from any process.
+Cross-process syscalls made with rsyscall individually report errors, as normal, and can be easily branched on.
+Arbitrary processes involving complex namespaces and initialization can be created
+using techniques that are impossible with \texttt{fork}.
+
 We have used direct-style process creation on Linux to do TODO and TODO,
 and we've benchmarked our implementation as TODO.
-
-Our implementation is immediately deployable on recent Linux kernels,
-and is distributed as a feature of the open source \texttt{rsyscall} library,
-available from \url{https://github.com/catern/rsyscall}.
+Direct-style \texttt{clone} has minimal performance cost in normal situations,
+and can outperform \texttt{fork} in some cases.
+We have deployed direct-style process creation in production internally at Two Sigma.
+
+Our implementation is distributed as a feature of \texttt{rsyscall},
+and is immediately deployable on recent Linux kernels.
+rsyscall is open source and available from \url{https://github.com/catern/rsyscall}.
 
@@
 and consider their advantages and disadvantages in more depth.
-In section \ref{examples} we show examples of usage of direct-style process creation on Linux,
+In section \ref{examples} we show examples of usage of direct-style process creation in Python on Linux,
 including use of process features which are difficult to use on Linux with \texttt{fork} or \texttt{posix\_spawn},
@@
 
+\section{Background on the use of processes}
+Why is it important to have a high-quality interface for creating processes?
+Processes are already widely used;
+most software is distributed as an executable which runs in a dedicated process.
+This basic usage of processes can be performed with even a complex and inefficient process creation interface.
+But processes have many uses beyond this simple and widespread one;
+here we examine some more sophisticated applications of processes,
+which benefit from a better process creation interface.
+\subsection{Abstraction over resources}
+In Unix, the mechanism of file descriptor inheritance
+allows a process to be provided a resource by its creator,
+while abstracting over the precise identity of that resource.[fn:ucspi]
+For example, a process can be provided a file descriptor,
+which can correspond to any file in a filesystem,
+without the process being aware of what specific file it is accessing.
+This is further enhanced by Unix's "everything is a file" design;
+the passed file descriptor could also be a pipe, a network connection, or some other resource,
+without the process knowing.
+As another example,
+a process can be provided a socket file descriptor on which it can call \texttt{accept} to receive connections,
+without being aware of whether those connections come from the internet or from a local Unix socket.[fn:ucspi]
+This abstraction mechanism is the basic principle of pipelines and redirection in the Unix shell,
+but it is rarely used outside of the shell.
+\subsection{Customization without explicit support}
+In many systems,
+it's possible to modify a process's view of nominally "global" resources.
+In Unix-derived systems, this ability is most influentially provided in Plan 9[fn:plan9],
+which allows each process to customize its view of the filesystem with private mounts and union directories[fn:plan9ns].
+In Linux, these concepts were implemented as per-process namespaces[fn:linuxns].
+Fundamentally,
+this allows customizing a process's environment and therefore a program's behavior,
+without having to write explicit support code for customization.
+For example, Plan 9, unlike most other Unix-derived systems,
+did not have a \texttt{PATH} environment variable which was searched by code in the process to find executables;
+instead, each process was executed with a \texttt{/bin} directory at the root of the filesystem,
+which was a union of many other directories,
+and simply executed \texttt{/bin/foo} to run the program named \texttt{foo}.
+In this way the set of executables provided to a process could be customized,
+without any code to parse and handle \texttt{PATH} or any other executable-lookup-specific customization code.
+\subsection{Sandboxing}
+The basic isolation powers of processes are used to simplify application development:
+it is beneficial to have a private virtual memory space when developing a stand-alone program.
+But most systems have additional mechanisms of isolation between processes,
+such as different privilege levels and access to global resources,
+which can be used to provide a form of sandboxing.
+For example, components which may exposed to hostile network requests
+can be run in a separate process, at a lower privilege level than the main program;
+in this way, even if an attacker gains control over that component,
+the attacker will only have access to the lower level of privileges of that component,
+rather than the full privileges of the main program.
+\subsection{Capability-based security}
+As a further development of process-based sandboxing,
+the privileges of a process can be explicitly enumerated
+in a capability-based security model.[fn:capsicum]
+By using previously-mentioned resource passing mechanisms,
+such as file descriptor inheritance or namespace manipulation,
+and by disabling the process's access to global resources such as the shared filesystem,
+we can enforce that all resources used by the process are passed at creation time.
+\subsection{Non-shared-memory concurrency}
+Processes run concurrently,
+which allows exploiting parallelism in the hardware.
+Since processes don't share memory,
+they can provide a less complex parallel programming environment
+than shared-memory thread-based approaches.
+The most popular parallel programming environment in existence today is the Unix shell,
+which obtains its parallelism by running multiple processes connected via pipes.
+The Unix shell has a relatively constrained form of parallel processing,
+but it's also possible to create more complex webs of parallel processes,
+where, for example, one process might take multiple inputs over multiple pipes,
+or produce multiple outputs.
+\subsection{Conclusion}
+These techniques, and more, are available through the process interface.
+Most software would benefit from abstraction over resources, sandboxing, and parallelism.
+Yet these features of processes are used only rarely.
+There are multiple reasons for this,
+but one of the primary reasons is the complexity of current process creation interfaces.
+
+Many of these techniques are used today by specialized software and services.
+Often, such software only allows use of one of these techniques;
+for example, the Unix shell allows piping together process, but not namespacing them;
+container systems allow sandboxing processes, but not piping them together.
+By delegating these features to specific separate services,
+we lose the ability to use them in combination.
+
+By improving the process creation interface,
+we can make it possible both for programs to directly manipulate processes to use these techniques,
+and to use and share composable libraries which use these techniques.
+We believe this potential justifies the investment of substantial effort
+into improving the process creation interface.
 \section{Overview and examples}\label{examples}
@@
 # Bind the socket to a sockaddr_in;
-# the sockaddr is allocated in memory with child.ptr and is garbage collected.
+# the sockaddr is written to memory accessible by the child with child.ptr
 sock.bind(child.ptr(SockaddrIn(0, 0)))
@@
 # inherit the write-end of the pipe to child1, and replace child1.stdout with it
-child1.inherit_fd(pipe.write).dup2(child1.stdout)
+child1_pipe_write = child1.inherit_fd(pipe.write)
+child1_pipe_write.dup2(child1.stdout)
 child1_proc = child1.execv(yes_path, [yes_path])
@@
 # inherit the read-end of the pipe to child2, and replace child2.stdin with it
-child2.inherit_fd(pipe.read).dup2(child2.stdin)
+child2_pipe_read = child2.inherit_fd(pipe.read)
+child2_pipe_read.dup2(child2.stdin)
 child2_proc = child2.execv(head_path, [head_path, "-n", "15"])
@@
 and another child to connect to it.
-%% **** note to self: concepts introduced in this example
-%% network namespace
-%% nested clone
-% \subsection{Example: Starting a complex system}
-% \begin{lstlisting}[float,language=Python,label={miredo},caption={Starting a complex system}]
-% ### create socket outside network namespace that Miredo will use for internet access
-% inet_sock = local.thread.socket(AF.INET, SOCK.DGRAM)
-% inet_sock.bind(local.thread.ptr(SockaddrIn(0, 0)))
-% # set some miscellaneous additional sockopts that Miredo wants
-% set_miredo_sockopts(local.thread, inet_sock)
-% ### create main network namespace thread
-% ns_thread = local.thread.clone(CLONE.NEWNET|CLONE.NEWUSER)
-% ### create in-network-namespace raw INET6 socket which Miredo will use to relay pings
-% icmp6_fd = ns_thread.socket(AF.INET6, SOCK.RAW, IPPROTO.ICMPV6)
-% ### create in-network-namespace socket,
-% ### which Miredo will use for unassociated Ifreq ioctls
-% reqsock = ns_thread.socket(AF.INET, SOCK.STREAM)
-% ### create and set up the TUN interface
-% tun_fd, tun_index = make_tun(ns_thread, "miredo", reqsock)
-% ### create socketpair which Miredo will use to communicate
-% ### between privileged process and Teredo client
-% privproc_pair = ns_thread.socketpair(AF.UNIX, SOCK.STREAM)
-% ### start up privileged process which manipulates the network setup in the namespace
-% privproc_thread = ns_thread.clone()
-% # We preserve NET_ADMIN capability over exec so that privproc can
-% # manipulate the TUN interface. add_to_ambient_caps is a helper function
-% # used because manipulating Linux ambient capabilities is fairly verbose
-% add_to_ambient_caps(privproc_thread, {CAP.NET_ADMIN})
-% # privproc expects to communicate with the main client over stdin and stdout
-% privproc_side = privproc_thread.inherit_fd(privproc_pair.first)
-% privproc_side.dup2(privproc_thread.stdin)
-% privproc_side.dup2(privproc_thread.stdout)
-% privproc_child = privproc_thread.execv(miredo_privproc_executable_path, [
-%     miredo_privproc_executable_path, str(tun_index)
-% ])
-% ### start up Miredo client process which communicates
-% ### over the internet to implement the tunnel
-% # the client process doesn't need to be in the same network namespace,
-% # since it is passed all the resources it needs as fds at startup.
-% client_thread = ns_thread.clone(CLONE.NEWUSER|CLONE.NEWNET|CLONE.NEWNS|CLONE.NEWPID)
-% # lightly sandbox by unmounting everything except for the executable
-% # and its deps (known via package manager)
-% unmount_everything_except(client_thread, miredo_exec.run_client.executable_path)
-% # a helper function for preparing the fds that are passed as command line arguments
-% async def pass_fd(fd: FileDescriptor) -> str:
-%     client_thread.inherit_fd(fd).disable_cloexec()
-%     return str(int(fd))
-% client_child = client_thread.execv(miredo_client_executable_path, [
-%     miredo_client_executable_path,
-%     pass_fd(inet_sock), pass_fd(tun_fd), pass_fd(reqsock),
-%     pass_fd(icmp6_fd), pass_fd(privproc_pair.second),
-%     "teredo.remlab.net", "teredo.remlab.net"
-% ])
-% \end{lstlisting}
-
-% In Listing \ref{miredo},
-% we show non-trivial code for launching a real application:
-% the Miredo IPv6 tunneling software.
-% We use a few helper functions in this listing to keep the attention focused on the interesting parts.
-
-% Miredo is separated into two components, a privileged process which sets up network interfaces,
-% and an unprivileged process which talks to the network.
-% With minimal modifications to Miredo,
-% we launch Miredo entirely unprivileged inside a user namespace and network namespace,
-% with all resources created outside and explicitly passed in.
-
-%% **** note to self: concepts introduced in this example
-%% socketpair
-%% several helper functions
-%% **** note to self
-%% maybe I should remove some of the unnecessary networking stuff?
-%% maybe show the corresponding fork-based code?
-%% except doing something with fork with the same error handling requires a bunch of crazy IPC
+\subsection{Example: Starting a complex system}
+\begin{lstlisting}[float,language=Python,label={miredo},caption={Starting a complex system}]
+### create socket outside network namespace that Miredo will use for internet access
+inet_sock = local.thread.socket(AF.INET, SOCK.DGRAM)
+inet_sock.bind(local.thread.ptr(SockaddrIn(0, 0)))
+# set some miscellaneous additional sockopts that Miredo wants
+set_miredo_sockopts(local.thread, inet_sock)
+### create main network namespace thread
+ns_thread = local.thread.clone(CLONE.NEWNET|CLONE.NEWUSER)
+### create in-network-namespace raw INET6 socket which Miredo will use to relay pings
+icmp6_fd = ns_thread.socket(AF.INET6, SOCK.RAW, IPPROTO.ICMPV6)
+### create in-network-namespace socket,
+### which Miredo will use for unassociated Ifreq ioctls
+reqsock = ns_thread.socket(AF.INET, SOCK.STREAM)
+### create and set up the TUN interface
+tun_fd, tun_index = make_tun(ns_thread, "miredo", reqsock)
+### create socketpair which Miredo will use to communicate
+### between privileged process and Teredo client
+privproc_pair = ns_thread.socketpair(AF.UNIX, SOCK.STREAM)
+### start up privileged process which manipulates the network setup in the namespace
+privproc_thread = ns_thread.clone()
+# We preserve NET_ADMIN capability over exec so that privproc can
+# manipulate the TUN interface. add_to_ambient_caps is a helper function
+# used because manipulating Linux ambient capabilities is fairly verbose
+add_to_ambient_caps(privproc_thread, {CAP.NET_ADMIN})
+# privproc expects to communicate with the main client over stdin and stdout
+privproc_side = privproc_thread.inherit_fd(privproc_pair.first)
+privproc_side.dup2(privproc_thread.stdin)
+privproc_side.dup2(privproc_thread.stdout)
+privproc_child = privproc_thread.execv(miredo_privproc_executable_path, [
+    miredo_privproc_executable_path, str(tun_index)
+])
+### start up Miredo client process which communicates
+### over the internet to implement the tunnel
+# the client process doesn't need to be in the same network namespace,
+# since it is passed all the resources it needs as fds at startup.
+client_thread = ns_thread.clone(CLONE.NEWUSER|CLONE.NEWNET|CLONE.NEWNS|CLONE.NEWPID)
+# lightly sandbox by unmounting everything except for the executable
+# and its deps (known via package manager)
+unmount_everything_except(client_thread, miredo_exec.run_client.executable_path)
+# a helper function for preparing the fds that are passed as command line arguments
+async def pass_fd(fd: FileDescriptor) -> str:
+    client_thread.inherit_fd(fd).disable_cloexec()
+    return str(int(fd))
+client_child = client_thread.execv(miredo_client_executable_path, [
+    miredo_client_executable_path,
+    pass_fd(inet_sock), pass_fd(tun_fd), pass_fd(reqsock),
+    pass_fd(icmp6_fd), pass_fd(privproc_pair.second),
+    "teredo.remlab.net", "teredo.remlab.net"
+])
+\end{lstlisting}
+
+In Listing \ref{miredo},
+we show non-trivial code for launching a real application:
+the Miredo IPv6 tunneling software.
+We use a few helper functions in this listing to keep the attention focused on the interesting parts.
+
+Miredo is separated into two components, a privileged process which sets up network interfaces,
+and an unprivileged process which talks to the network.
+With minimal modifications to Miredo,
+we launch Miredo entirely unprivileged inside a user namespace and network namespace,
+with all resources created outside and explicitly passed in.
 \section{Implementation}\label{implementation}
@@
 Write a shell?
-
-Right okay so we can do some fork-style code, sure.
-Show the IPC, not too complex, but show it.
-
-We can do it in Python with rsyscall - have a hidden experimental secret fork thing.
-Using write_all_bytes will be annoying but fine.
-Ugh, but we'll have to decode it too.
-Eh, no, we can wrap that.
-Yeah then it's fine.
 \section{Future work}\label{future_work}

### 2020-05-07 python/rsysapps/miredo.py
fix miredo

@@ -176,7 +176,7 @@ class TestMiredo(TrioTestCase):
         print("c", time.time())
         # bash = await rsc.which(self.thread, "bash")
         # await (await thread.exec(bash)).check()
-        await add_to_ambient(thread, {CAP.NET_RAW})
+        await add_to_ambient_caps(thread, {CAP.NET_RAW})
         await (await thread.exec(ping6.args('-c', '1', 'google.com'))).check()
         print("d", time.time())
 

### 2020-05-08 research/paper/paper.tex
notes and paper

@@
 it is difficult to
-branch based on results or errors encountered while initializing the new process;
+branch based on results or errors encountered while initializing the new process:
 fork-style requires IPC if it wishes to communicate errors back to the rest of the program,
-while robust indication of errors in spawn-style requires substantial growth in the spawn interface.
+and robust indication of errors in spawn-style requires substantial growth in the spawn interface.
 Additionally, modern operating systems such as Linux have features
@@
 Direct-style is rare in general, but relatively common on academic operating systems.
+% insert citations to direct-style papers here
 Direct-style involves creating a new child process in a non-running state,
@@
 making use of our work on the rsyscall project.
-The rsyscall project develops language-specific ``process-independent'' syscall libraries,
-replacing libc for interaction with Linux,
-allowing a program to perform cross-process syscalls.
+The rsyscall project develops language-specific ``process-independent'' syscall libraries for Linux,
+which replace libc syscall wrappers and allow a program to perform cross-process syscalls.
 rsyscall currently has best support for Python,
@@
 Arbitrary processes involving complex namespaces and initialization can be created
-using techniques that are impossible with \texttt{fork}.
+using techniques that are impractical with \texttt{fork}.
+The performance cost from high-level code varies, but the underlying primitives are substantially faster.
 
@@
 we list several directions for future work related to direct-style process creation and to \texttt{rsyscall}.
-In section \ref{conclusion}, we briefly and optimistically conclude.
+In section \ref{conclusions}, we briefly and optimistically conclude.
 
@@
 into improving the process creation interface.
-\section{Overview and examples}\label{examples}
+\section{Overview}\label{examples}
 In most operating system APIs, including the ones available on Linux,
@@
 
-\texttt{rsyscall} provides this cross-process API for Linux.
-\texttt{rsyscall} is a library which allows us to create processes in which we can call arbitrary syscalls.
-These processes run no other code; they do nothing except for execute syscalls at our request.
-We will discuss the implementation of rsyscall in greater detail in section \ref{implementation}.
+\texttt{rsyscall} provides this API for Linux.
+
+The \texttt{rsyscall} project develops
+language-specific, object-capability-model, low-abstraction libraries for the Linux syscall interface,
+bypassing libc.
+Unlike POSIX C libraries such as glibc or musl,
+an rsyscall library is organized based on the object-capability model.
+The capabilities to make syscalls in any given process are reified as language objects.
+If the capability to make syscalls in a specific process is not passed (in some way) to a function,
+then that function cannot make syscalls in that process.
+
+Due to this design, an \texttt{rsyscall} program can make use of multiple processes at once,
+by manipulating capabilities for multiple processes.
+The relevant two capabilities for this paper are the initial capability to the ``local'' process,
+and capabilities to child processes.
+The ``local'' process is the one which hosts the runtime for the running program,
+and in which a legacy libc would implicitly make syscalls;
+every rsyscall program starts with the ``local'' process
+and uses it to bootstrap the initial capabilities for other processes.
 
 From the perspective of an program using rsyscall,
-there are multiple processes which are under its control.
+there are multiple processes which are under its control,
+through multiple capabilities.
 We use the term ``thread'' to refer to all such controlled processes,
@@
 and their lifetime and execution is completely controlled by a single program.
-The same is true of the controlled-process "threads" provided by rsyscall.
+The same is mostly true of the controlled-process "threads" provided by rsyscall.
 The most significant difference is that rsyscall "threads"
@@
 To perform direct-style process creation,
-we use \texttt{rsyscall} to call \texttt{clone} on our main thread to create a new child thread.
+we use \texttt{rsyscall} to call \texttt{clone} on the ``local'' thread to create a new child thread.
 We call various syscalls in the child thread to mutate it until it reaches the desired state,
@@
 
-Resource management across multiple processes can be complex,
-as we'll see in section \ref{implementation}.
-To ease the complexity for user code,
-rsyscall provides substantial language-specific libraries,
-which allow for resource management in a language-idiomatic fashion.
-
-To reduce distracting non-essential complexity,
-we will show all our examples with the Python rsyscall library.
-We will not use any substantial abstractions;
-besides process creation, which has complexities we'll discuss in section \ref{clone},
-each function call corresponds to exactly one syscall.
-Python-specific asynchronous execution syntax (\texttt{await}) has been removed for clarity,
-but the examples are otherwise real working code.
-
-\subsection{Example: Create and exec}
-\begin{lstlisting}[float,language=Python,label={basic},caption={Creating a new process and immediately execing}]
-# Use the clone syscall in the local thread to create a new thread;
-# we use a wrapper that supplies defaults for all arguments.
-child = local.thread.clone()
-# Call execv to run a different executable in the child thread;
-# We pass the executable path as the first argument in the argument list, as is traditional.
-child.execv(hello_path, [hello_path])
+\begin{lstlisting}[float,language=Python,label={basic},caption={Creating a new process, changing CWD, and execing}]
+child = local_thread.clone()
+child.chdir("/etc")
+child.execve(ls_path, [ls_path], local_thread.environ)
 \end{lstlisting}
-In Listing \ref{basic}, we simply create a new process under our control (a thread)
-and immediately exec a binary with it.
-As exceptions are used for error-handling in the Python API,
-there is no need for error-checking code.
-%% **** note to self: concepts introduced in this example
-%% local.thread
-%% clone
-%% execve
+
+Listing \ref{basic} shows this in action in Python.
+We create a new child thread using \texttt{clone} on \texttt{local_thread},
+and receive back a capability to control it.
+We pass no flags to use \texttt{clone}'s defaults
+(which essentially match \texttt{fork}) to create the process.
+We call \texttt{chdir} in the thread to change its working directory to \texttt{/etc}.
+Then we call \texttt{execve},
+passing the path of the \texttt{ls} executable,
+the arguments to the new executable,
+and the unchanged set of environment variables.
+The path of the executable is passed as the first argument,
+as is traditional in Unix.
+Calling \texttt{execve} consumes the capability,
+and later use of this child thread capability will fail with an exception.
+
+The API for direct-style process creation on Linux is not new;
+it is the same Linux system call API that systems programmers are already familiar with.
+The novelty is constrained to the newly explicit process capabilities;
+once that is understood, the rest of the API is mundane,
+and the representation of these programs is obvious in any language.
+
+The familiar and mundane nature of the API is important not just for the learning curve,
+but also for expressivity.
+Since the API is simply normal Linux system calls,
+anything that could be expressed with fork and syscalls
+can also be expressed with direct-style clone.
+\section{Applications}
+In this section,
+we'll discuss several more advanced uses of processes,
+using more sophisticated Linux features,
+and accompany them with Python examples using direct-style process creation.
+As always, since these examples are using the Linux system call API directly,
+they can be translated straightforwardly into any language with an \texttt{rsyscall} API.
+
 \subsection{Example: Passing down FDs}
@@
 is a robust system for cross-process syscalls.
-We provided this in the rsyscall project.
-rsyscall is a toolkit for cross-process syscalls on Linux,
+The rsyscall project provides this.
+
+The rsyscall project develops language-specific libraries
+for the Linux syscall interface, with minimal abstraction.
+The rsyscall project develops a collection of language-specific libraries tightly coupled to Linux,
+which provide access to Linux syscalls with minimal abstraction,
+bypassing libc.
+
+Unlike POSIX C libraries, an rsyscall library is organized in a capability-oriented style;
+the capability to make syscalls in a process is explicicitly passed around as a language object.
+This organization allows an rsyscall program to be process-independent,
+and straightforwardly make system calls in any of the processes under its control,
+so-called ``cross-process'' system calls.
+
+for 
+These 
+for the is an alternative, low-abstraction to the Linux
+The rsyscall project is a toolkit for cross-process syscalls on Linux,
 with several language-specific library implementations.
@@
 and the shared file descriptor table would allow the child to bind a socket
-and the parent to use it.
+and the parent to use it;
+the file descriptor table is unshared once the child calls \texttt{execve}.
 
@@
 % might explore an object-centric approach for managing resources.
-\section{Conclusion}\label{conclusion}
+\section{Conclusions}\label{conclusions}
 Direct-style process creation is much less known and much less used than fork-style and spawn-style.

### 2020-05-09 research/paper/paper.tex
more work

@@
 \subsection{Fork-style}
-Fork-style interfaces are those that follow the model used by Unix's fork \cite{forkhist}.
-A process calls fork, and a copy of that process is made,
+By far the most well-known fork-style interface is Unix's fork.\cite{forkhist}
+With a fork-style interface,
+a process calls \texttt{fork}, and a copy of that process is made,
 with the return value of fork being the only differentiation between the process and its copy.
@@
 \subsection{Spawn-style}
-Spawn-style interfaces are those that follow the model used by \texttt{posix\_spawn} \cite{posix_spawn}
-and Windows' \texttt{CreateProcess} \cite{create_process}.
-All the details about the new process are provided up-front as arguments to a syscall,
+In a spawn-style interface,
+all the details about the new process are provided up-front as arguments to some function,
 which creates the new process from a mostly-clean slate, initialized with the provided details.
+The most well known spawn-style interfaces are Unix's \texttt{posix\_spawn} \cite{posix_spawn}
+and Windows \texttt{CreateProcess} \cite{create_process};
+the Unix shell also serves as a spawn-style process creation interface,
+especially when used with chain-loading techniques.[fn:chainloading]
 
@@
 and any other process attribute which is not explicitly specified in the arguments to the spawn interface.
-The key difference between fork-style and spawn-style is not how much they copy;
+The key difference between fork-style and spawn-style is not how much they copy,
 it is how the attributes which are not copied are specified:
@@
 as is required for container functionality on Linux.
+
+Furthermore, spawn-style process creation
+creates a new domain-specific language for specifying the attributes of the process.
+Fork-style allows existing system calls,
+with which programmers are already familiar,
+to be used to set up the new process.
+Spawn-style requires some form of declarative description of the new process,
+which must be learned in additional to the normal system call interface.
 
@@
 
-\section{Background on the use of processes}
-Why is it important to have a high-quality interface for creating processes?
-Processes are already widely used;
-most software is distributed as an executable which runs in a dedicated process.
-This basic usage of processes can be performed with even a complex and inefficient process creation interface.
-But processes have many uses beyond this simple and widespread one;
-here we examine some more sophisticated applications of processes,
-which benefit from a better process creation interface.
-\subsection{Abstraction over resources}
-In Unix, the mechanism of file descriptor inheritance
-allows a process to be provided a resource by its creator,
-while abstracting over the precise identity of that resource.[fn:ucspi]
-For example, a process can be provided a file descriptor,
-which can correspond to any file in a filesystem,
-without the process being aware of what specific file it is accessing.
-This is further enhanced by Unix's "everything is a file" design;
-the passed file descriptor could also be a pipe, a network connection, or some other resource,
-without the process knowing.
-As another example,
-a process can be provided a socket file descriptor on which it can call \texttt{accept} to receive connections,
-without being aware of whether those connections come from the internet or from a local Unix socket.[fn:ucspi]
-This abstraction mechanism is the basic principle of pipelines and redirection in the Unix shell,
-but it is rarely used outside of the shell.
-\subsection{Customization without explicit support}
-In many systems,
-it's possible to modify a process's view of nominally "global" resources.
-In Unix-derived systems, this ability is most influentially provided in Plan 9[fn:plan9],
-which allows each process to customize its view of the filesystem with private mounts and union directories[fn:plan9ns].
-In Linux, these concepts were implemented as per-process namespaces[fn:linuxns].
-Fundamentally,
-this allows customizing a process's environment and therefore a program's behavior,
-without having to write explicit support code for customization.
-For example, Plan 9, unlike most other Unix-derived systems,
-did not have a \texttt{PATH} environment variable which was searched by code in the process to find executables;
-instead, each process was executed with a \texttt{/bin} directory at the root of the filesystem,
-which was a union of many other directories,
-and simply executed \texttt{/bin/foo} to run the program named \texttt{foo}.
-In this way the set of executables provided to a process could be customized,
-without any code to parse and handle \texttt{PATH} or any other executable-lookup-specific customization code.
-\subsection{Sandboxing}
-The basic isolation powers of processes are used to simplify application development:
-it is beneficial to have a private virtual memory space when developing a stand-alone program.
-But most systems have additional mechanisms of isolation between processes,
-such as different privilege levels and access to global resources,
-which can be used to provide a form of sandboxing.
-For example, components which may exposed to hostile network requests
-can be run in a separate process, at a lower privilege level than the main program;
-in this way, even if an attacker gains control over that component,
-the attacker will only have access to the lower level of privileges of that component,
-rather than the full privileges of the main program.
-\subsection{Capability-based security}
-As a further development of process-based sandboxing,
-the privileges of a process can be explicitly enumerated
-in a capability-based security model.[fn:capsicum]
-By using previously-mentioned resource passing mechanisms,
-such as file descriptor inheritance or namespace manipulation,
-and by disabling the process's access to global resources such as the shared filesystem,
-we can enforce that all resources used by the process are passed at creation time.
-\subsection{Non-shared-memory concurrency}
-Processes run concurrently,
-which allows exploiting parallelism in the hardware.
-Since processes don't share memory,
-they can provide a less complex parallel programming environment
-than shared-memory thread-based approaches.
-The most popular parallel programming environment in existence today is the Unix shell,
-which obtains its parallelism by running multiple processes connected via pipes.
-The Unix shell has a relatively constrained form of parallel processing,
-but it's also possible to create more complex webs of parallel processes,
-where, for example, one process might take multiple inputs over multiple pipes,
-or produce multiple outputs.
-\subsection{Conclusion}
-These techniques, and more, are available through the process interface.
-Most software would benefit from abstraction over resources, sandboxing, and parallelism.
-Yet these features of processes are used only rarely.
-There are multiple reasons for this,
-but one of the primary reasons is the complexity of current process creation interfaces.
-
-Many of these techniques are used today by specialized software and services.
-Often, such software only allows use of one of these techniques;
-for example, the Unix shell allows piping together process, but not namespacing them;
-container systems allow sandboxing processes, but not piping them together.
-By delegating these features to specific separate services,
-we lose the ability to use them in combination.
-
-By improving the process creation interface,
-we can make it possible both for programs to directly manipulate processes to use these techniques,
-and to use and share composable libraries which use these techniques.
-We believe this potential justifies the investment of substantial effort
-into improving the process creation interface.
 \section{Overview}\label{examples}
@@
 \texttt{rsyscall} provides this API for Linux.
-
 The \texttt{rsyscall} project develops
@@
 at which point we can call \texttt{execve} in the child thread to start it running on its own.
-After that, the process is no longer under our control, and it functions like any other child process.
-(Hence we call it ``process'' rather than ``thread'')
+After that, the process is no longer under our control, and it functions like any other child process
+(hence we say ``process'' rather than ``thread'').
 The child process can be monitored using normal Linux child monitoring syscalls from our main thread,
@@
 child = local_thread.clone()
-child.chdir("/etc")
-child.execve(ls_path, [ls_path], local_thread.environ)
+child.chdir("/dev")
+child.execve("/bin/cat",
+  ["cat", "./null"], local_thread.environ)
 \end{lstlisting}
-
 Listing \ref{basic} shows this in action in Python.
-We create a new child thread using \texttt{clone} on \texttt{local_thread},
+We create a new child thread using \texttt{clone} on \verb|local_thread|,
 and receive back a capability to control it.
@@
 (which essentially match \texttt{fork}) to create the process.
-We call \texttt{chdir} in the thread to change its working directory to \texttt{/etc}.
+We call \texttt{chdir} in the thread to change its working directory to \texttt{/dev}.
 Then we call \texttt{execve},
-passing the path of the \texttt{ls} executable,
+passing the path of the \texttt{cat} executable,
 the arguments to the new executable,
 and the unchanged set of environment variables.
-The path of the executable is passed as the first argument,
-as is traditional in Unix.
+The name of the executable is passed as the first argument,
+as is typical in Unix.
 Calling \texttt{execve} consumes the capability,
@@
 and accompany them with Python examples using direct-style process creation.
-As always, since these examples are using the Linux system call API directly,
-they can be translated straightforwardly into any language with an \texttt{rsyscall} API.
-
-\subsection{Example: Passing down FDs}
+
+% TODO this intro is all a bit muddled, and also not true.
+These examples are all conventional applications of Linux features;
+they are all possible using Linux's existing \texttt{fork} system call,
+and in some cases, \verb|posix_spawn|,
+with varying degrees of difficulty and workarounds.
+But direct-style makes these easier to write;
+we'll compare our implementations in this section to fork-style implementations in section \ref{evaluation}.
+Furthermore, since the examples are using the Linux system call API directly,
+they can be translated straightforwardly into any language with an \texttt{rsyscall} API;
+Python is chosen simply for convenience.
+
+Some of these examples showcase functionality that can be done today by specialized software and services.
+Such software often allows use of one or two of these techniques, but prevents the use of others.
+
+By improving the process creation interface,
+all these techniques can be used together, flexibly,
+to more closely meet the specific needs of individual real-world applications.
+
+\subsection{Abstraction through file descriptor passing}\label{fd_abstraction}
+In Linux (and Unix in general),
+a program can use file descriptor inheritance
+to provide file descriptors created by the parent process to child processes.
+Put simply,
+a new Linux child process receives copies of all file descriptors in the parent process.
+When calling \texttt{execve}, these file descriptors are normally closed automatically,
+but selected file descriptors can be preserved over \texttt{execve}
+by disabling the \texttt{CLOEXEC} flag on individual file descriptors.
+We will discuss file descriptor inheritance in more detail in section \label{fdtables},
+as part of our discussion of resource management in the implementation of \texttt{rsyscall}.
+
+Most resources in Linux are managed through file descriptors,
+so this allows the parent process to pass a wide variety of resources to the child process.
+An executable program can be provided a file descriptor,
+which can correspond to any file in any filesystem,
+or to a network connection\cite{ucspi},
+or a pipe connected to another process,
+without the process being aware of what specific resource it is accessing.
+In this way, we immediately know "for free"\cite{theoremsforfree}
+that the behavior of the program does not depend on the name of the file,
+or the hostname to it which the socket is connected,
+or the executable running in the process on the other end of the pipe.
 \begin{lstlisting}[float,language=Python,label={fds},caption={Passing down FDs}]
-child = local.thread.clone()
-sock = child.socket(AF.INET, SOCK.DGRAM)
-# Bind the socket to a sockaddr_in;
-# the sockaddr is written to memory accessible by the child with child.ptr
-sock.bind(child.ptr(SockaddrIn(0, 0)))
-sock.listen(10)
-sock.disable_cloexec()
-child.execv(executable_path, [executable_path, "--listening-socket", str(int(sock))])
+db_fd = local_thread.open("/var/db/database.db", O.RDWR)
+child = local_thread.clone()
+child_fd = child.inherit_fd(db_fd)
+child_fd.fcntl(F.SETFD, 0)
+child.execv("/bin/fooserver",
+  ["fooserver", "--database-fd", str(int(child_fd))])
 \end{lstlisting}
-In Listing \ref{fds}, we create a new thread,
-then create a listening socket bound to a random port in that thread,
-then call exec, 
-passing down the socket by disabling cloexec and passing its file descriptor number as an argument to the new program.
-
-File descriptors, here, are object oriented and have relevant syscalls as methods.
-They make syscalls in the process they are created in by default;
-we can create more objects referring to the same file descriptor from different processes
-if we want to make the syscalls from another process.
-%% **** note to self: concepts introduced in this example
-%% disable_cloexec
-%% socket creation
-%% using int on sock
-%% .ptr
-\subsection{Example: Pipeline}
-\begin{lstlisting}[float,language=Python,label={pipe},caption={Creating a pipeline}]
-# create the pipe
-pipe = local.thread.pipe()
-child1 = local.thread.clone()
-# inherit the write-end of the pipe to child1, and replace child1.stdout with it
-child1_pipe_write = child1.inherit_fd(pipe.write)
-child1_pipe_write.dup2(child1.stdout)
-child1_proc = child1.execv(yes_path, [yes_path])
-child2 = local.thread.clone()
-# inherit the read-end of the pipe to child2, and replace child2.stdin with it
-child2_pipe_read = child2.inherit_fd(pipe.read)
-child2_pipe_read.dup2(child2.stdin)
-child2_proc = child2.execv(head_path, [head_path, "-n", "15"])
+In Listing \ref{fds},
+we first open a file with read-write permission.
+Then we create a child thread,
+which inherits all file descriptors from its parent.
+We indicate that we want to operate on the child's copy of the file descriptor with \verb|child.inherit_fd|.
+\verb|inherit_fd| only updates resource-tracking bookkeeping in our program;
+it makes no system calls.
+Then we disable \texttt{CLOEXEC} on the child's file descriptor,
+which on Linux we can do by clearing the FD flags with \verb|fcntl(fd, F_SETFD, 0)|.
+Finally, we execute some executable,
+passing the file descriptor number as an argument;
+when the executable starts up, that file descriptor number will already be pointing to this file descriptor.
+We call the \texttt{execv} wrapper rather than \texttt{execve} as in listing \ref{basic}
+so that we don't need to pass the (unmodified) environment to every exec call in these examples.
+\subsection{Non-shared-memory concurrency}
+Processes run concurrently,
+which allows exploiting the parallelism of the underlying hardware.
+Since processes don't share memory,
+they can provide a less complex parallel programming environment
+than shared-memory thread-based approaches.
+
+A popular parallel programming environment is the Unix shell,
+which obtains its parallelism by running multiple processes connected via pipes.
+The Unix shell has a relatively constrained form of parallel processing,
+but it's also possible to create more complex webs of parallel processes,
+where, for example, one process might take multiple inputs over multiple pipes,
+or produce multiple outputs.
+\begin{lstlisting}[float,language=Python,label={pipe},caption={Creating a parallel processing pipeline}]
+def as_argument(child, fd):
+  child_fd = child.inherit_fd(fd)
+  child_fd.fcntl(F.SETFD, 0) # unset CLOEXEC
+  return str(int(child_fd)) # turn fd number to string
+
+audio_pipe = local_thread.pipe()
+video_pipe = local_thread.pipe()
+source = local.thread.clone()
+source.execve('/bin/source', ['source',
+  '--audio-output', as_argument(source, audio_pipe.write),
+  '--video-output', as_argument(source, video_pipe.write)])
+video_sink = local.thread.clone()
+video_sink.execve('/bin/video_sink', ['video_sink',
+  '--video-input', as_argument(video_sink, video_pipe.read)])
+audio_sink = local.thread.clone()
+audio_sink.execve('/bin/audio_sink', ['audio_sink',
+  '--audio-input', as_argument(audio_sink, audio_pipe.read)])
 \end{lstlisting}
-In Listing \ref{pipe},
-we do the same as the Unix shell pipeline "yes | head -n 15".
-We create a pipe,
-then create two threads,
-connect them with a pipe,
-and exec a different program in each thread.
-
-After a process is created with clone,
-it may have inherited file descriptors;
-here we inherit the pipe.
-We make this inheritance explicit with \texttt{inherit\_fd},
-a helper method on our thread object,
-which takes a file descriptor from a different thread
-and performs a runtime check that the file descriptor actually was inherited.
-If so, it returns a new handle to the file descriptor which performs syscalls from the new thread.
-
-Then we simply =dup2= as normal to replace child1's stdout with the write end of the pipe;
-=dup2= disables CLOEXEC by default on the target.
-%% **** note to self: concepts introduced in this example
-%% dup2
-%% pipe
-\subsection{Example: Mount namespace}
-\begin{lstlisting}[float,language=Python,label={mount},caption={Mount namespace}]
-child = local.thread.clone(CLONE.NEWUSER|CLONE.NEWNS)
-child.mkdir(rootdir/"proc")
-child.mount(Path("/proc"), rootdir/"proc", "", MS.BIND, "")
-child.chroot(rootdir)
-child.execv(executable_path, [executable_path])
+In listing \ref{pipe},
+we execute a few programs in parallel,
+connected by pipes.
+We first create two pipes,
+then inherit them down to several processes using a helper function, \texttt{as_argument},
+which uses \verb|inherit_fd| and \texttt{fcntl} as described in section \ref{fd_abstraction}.
+We exec each program passing some file descriptors as arguments.
+The whole system then runs in parallel,
+as normal child processes.
+\subsection{Customization without explicit support}
+While, ideally, all resources would be passed by file descriptor,[fn:capsicum]
+Unix has a number of resources that are global,
+such as the pid space or the mount table,
+and which cannot be changed on a per-process basis.
+
+In Linux, most global resources have been made process-local
+through the namespaces system.
+Like other process-local resources such as the current working directory,
+namespaces are typically set up at process creation time.
+
+The most popular usage of namespaces is for large-scale container systems such as Docker;
+however, this is far from the only use.
+With an adaquate process creation primitive,
+namespaces can be used easily on a much smaller scale.
+
+Namespaces, like all process-local resources,
+can be used to customize a program's behavior
+in ways that were not anticipated by the author
+by customizing the environment the process runs in.
+
+\begin{lstlisting}[float,language=Python,label={mount},caption={Overriding absolute path using a mount namespace}]
+child = local.thread.clone(CLONE.NEWNS)
+child.mount("/home/foo/custom_foo.conf", "/etc/foo.conf", "", MS.BIND, "")
+child.execv('/bin/fooserver', ['fooserver'])
 \end{lstlisting}
-In Listing \ref{mount},
-we make a new mount namespace and rearrange the filesystem tree for the child process.
-We bind-mount /proc at /proc inside the chroot directory,
-chroot into the directory,
-and exec an executable which will run inside the chroot.
-%% **** note to self: concepts introduced in this example
-%% namespaces
-%% path slash syntax sugar
-\subsection{Example: Nested clone and network namespace}
-\begin{lstlisting}[float,language=Python,label={nested},caption={Nested clone and network namespace}]
-ns_thread = local.thread.clone(CLONE.NEWNET|CLONE.NEWUSER)
-
-listening_child = ns_thread.clone()
-sock = listening_child.socket(AF.INET, SOCK.DGRAM)
-sockaddr = SockaddrIn(22, "127.0.0.1")
-sock.bind(listening_child.ptr(sockaddr))
-sock.listen(10)
-sock.disable_cloexec()
-child.execv(server_path, [server_path, "--listening-socket", str(int(sock))])
-
-connecting_child = ns_thread.clone()
-child.execv(client_path, [
-  client_path, "--connect-address", str(sockaddr.address) + ":" + str(sockaddr.port)])
+
+One classic form of customization enabled by namespaces
+is overriding files at paths that were hardcoded into the program.
+In listing \ref{mount},
+we create a new child thread as normal,
+but for the first time,
+pass an argument to \texttt{clone}.
+\verb|CLONE_NEWNS| causes clone to create the process in a new mount namespace,
+which allows us to use the \texttt{mount} system call
+to manipulate the filesystem tree in a way that only this process will see.
+We call mount, passing \texttt{MS.BIND}, to make the file \verb|/home/foo/custom_foo.conf|
+appear at the path \verb|/etc/foo.conf|;
+this is known as a bind mount.
+Then we execute some executable,
+which will see the \verb|/etc/foo.conf| as we specified.
+\subsection{Sandboxing}
+At process creation time,
+we can not only pass resources and customize the process's environment,
+we can also deny the process access to resources that it otherwise gets by default.
+This is a key part of creating a secure sandbox for potentially malicious code.
+
+Historically, system calls such as \texttt{chroot} were the main mechanism for doing this;
+today, Linux's namespaces system is the most important component.
+
+\begin{lstlisting}[float,language=Python,label={unmount},caption={Unmount all and run executable via fexec}]
+thread = local_thread.clone(CLONE.NEWNS)
+executable_fd = child.open("/bin/static_executable", O.RDONLY)
+db_fd = child.open("/var/db/database.db", O.RDWR)
+child.umount("/", MNT.DETACH)
+db_fd.fcntl(F.SETFD, 0)
+child.fexec(executable_fd, ["static_executable", "--database-fd", str(int(db_fd))])
 \end{lstlisting}
-In Listing \ref{nested},
-we make a process (\texttt{ns\_thread}) in a new network namespace.
-Then, we create two more child processes of \texttt{ns\_thread},
-which are also in the new network namespace.
-This nested creation of child processes is fully supported,
-like all other syscalls,
-and allows us to set up complex graphs of processes and namespaces.
-
-We bind to a privileged port on localhost inside the namespace,
-and create one child to listen on that socket,
-and another child to connect to it.
-\subsection{Example: Starting a complex system}
-\begin{lstlisting}[float,language=Python,label={miredo},caption={Starting a complex system}]
-### create socket outside network namespace that Miredo will use for internet access
-inet_sock = local.thread.socket(AF.INET, SOCK.DGRAM)
-inet_sock.bind(local.thread.ptr(SockaddrIn(0, 0)))
-# set some miscellaneous additional sockopts that Miredo wants
-set_miredo_sockopts(local.thread, inet_sock)
-### create main network namespace thread
-ns_thread = local.thread.clone(CLONE.NEWNET|CLONE.NEWUSER)
-### create in-network-namespace raw INET6 socket which Miredo will use to relay pings
-icmp6_fd = ns_thread.socket(AF.INET6, SOCK.RAW, IPPROTO.ICMPV6)
-### create in-network-namespace socket,
-### which Miredo will use for unassociated Ifreq ioctls
-reqsock = ns_thread.socket(AF.INET, SOCK.STREAM)
-### create and set up the TUN interface
-tun_fd, tun_index = make_tun(ns_thread, "miredo", reqsock)
-### create socketpair which Miredo will use to communicate
-### between privileged process and Teredo client
-privproc_pair = ns_thread.socketpair(AF.UNIX, SOCK.STREAM)
-### start up privileged process which manipulates the network setup in the namespace
-privproc_thread = ns_thread.clone()
-# We preserve NET_ADMIN capability over exec so that privproc can
-# manipulate the TUN interface. add_to_ambient_caps is a helper function
-# used because manipulating Linux ambient capabilities is fairly verbose
-add_to_ambient_caps(privproc_thread, {CAP.NET_ADMIN})
-# privproc expects to communicate with the main client over stdin and stdout
-privproc_side = privproc_thread.inherit_fd(privproc_pair.first)
-privproc_side.dup2(privproc_thread.stdin)
-privproc_side.dup2(privproc_thread.stdout)
-privproc_child = privproc_thread.execv(miredo_privproc_executable_path, [
-    miredo_privproc_executable_path, str(tun_index)
-])
-### start up Miredo client process which communicates
-### over the internet to implement the tunnel
-# the client process doesn't need to be in the same network namespace,
-# since it is passed all the resources it needs as fds at startup.
-client_thread = ns_thread.clone(CLONE.NEWUSER|CLONE.NEWNET|CLONE.NEWNS|CLONE.NEWPID)
-# lightly sandbox by unmounting everything except for the executable
-# and its deps (known via package manager)
-unmount_everything_except(client_thread, miredo_exec.run_client.executable_path)
-# a helper function for preparing the fds that are passed as command line arguments
-async def pass_fd(fd: FileDescriptor) -> str:
-    client_thread.inherit_fd(fd).disable_cloexec()
-    return str(int(fd))
-client_child = client_thread.execv(miredo_client_executable_path, [
-    miredo_client_executable_path,
-    pass_fd(inet_sock), pass_fd(tun_fd), pass_fd(reqsock),
-    pass_fd(icmp6_fd), pass_fd(privproc_pair.second),
-    "teredo.remlab.net", "teredo.remlab.net"
-])
+
+In listing \ref{unmount}
+we create a new child thread,
+again in a new mount namespace using \texttt{CLONE.NEWNS}.
+We open the executable that we'll later exec in this child thread,
+as well as another file as in listing \ref{fds},
+since we won't be able to do exec or open files normally after the next step.
+Then we use \texttt{umount},
+passing \texttt{MNT>DETACH} to perform a recursive unmount of the entire filesystem tree,
+removing it all from the view of this process.
+As in listing \ref{fds}, we unset the \texttt{CLOEXEC} flag from the database file descriptor,
+so that it can be inherited across \texttt{exec};
+we don't need to call \verb|inherit_fd| since, in this case, the database fd was opened directly from the child.
+We then exec the file we opened earlier, using \texttt{fexec},
+which allows execing a file descriptor,
+and also pass the database file descriptor as an argument,
+as in listing \ref{fds}.
+(The underlying system call for \texttt{fexec} is \texttt{execveat};
+this \texttt{fexec} is a helper function.)
+Note that this executable must be statically linked,
+or it wouldn't work in an empty filesystem namespace,
+with no libraries to dynamically link against.
+
+By removing the filesystem tree from the view of this process,
+we can run this executable with greater confidence
+that it won't be able to tamper with the rest of the system.
+A sandbox which is truly robust against malicious or compromised programs requires some additional steps,
+but this is a substantial start.
+Such a technique can also be used when a full sandbox is not relevant,
+to ease reasoning about the behavior of the program being run,
+and protect against bugs.
+
+Even if the process needs additional resources,
+those can be explicitly passed down through file descriptor passing,
+as we do here with the database file descriptor.
+This is the essence of capability-based security,
+which Linux can increasingly closely approximate.
+
+\subsection{Nested clone and pid namespaces}
+Process-local resource can also be used to control the lifetime of resources used by that process.
+Many Linux resources are not automatically cleaned up on process exit;
+a poorly coded program may allocate global resources
+and not ensure that they are cleaned up,
+leaving behind unuseds litter on the system.
+
+One resource which is not automatically cleaned up is processes themselves;
+if we run a program which itself spawns subprocesses,
+those subprocesses may unintentionally leak,
+and be left running on the system even after the program itself has stopped.
+
+We can use a pid namespace, in conjunction with nested cloning, to solve this issue.
+The lifetime of all processes in a pid namespace is tied to the first process created in it,
+the init process.
+When the init process dies,
+all other processes in the pid namespace are destroyed.
+
+\begin{lstlisting}[float,language=Python,label={pidns},caption={Nested clone and pid namespace}]
+init = local_thread.clone(CLONE.NEWPID)
+grandchild = init.clone()
+grandchild.execv('/bin/fooserver', ['fooserver'])
 \end{lstlisting}
 
-In Listing \ref{miredo},
-we show non-trivial code for launching a real application:
-the Miredo IPv6 tunneling software.
-We use a few helper functions in this listing to keep the attention focused on the interesting parts.
-
-Miredo is separated into two components, a privileged process which sets up network interfaces,
-and an unprivileged process which talks to the network.
-With minimal modifications to Miredo,
-we launch Miredo entirely unprivileged inside a user namespace and network namespace,
-with all resources created outside and explicitly passed in.
+In listing \ref{pidns},
+we create a child process which we call \texttt{init},
+passing \verb|CLONE_NEWPID| to create it in a new pid namespace.
+To create another process in the pid namespace,
+we clone again, this time from \texttt{init};
+note that this is the first time we've cloned from one of our child processes.
+We exec in the grandchild,
+and we can monitor the grandchild process from \texttt{init},
+just as we would monitor \texttt{init} from its parent process.
+
+\texttt{init} in a pid namespace has some special powers and responsibilities.
+We can handle these responsibilities ourselves, directly from this program,
+or we can continue on by execing an init daemon in \texttt{init} to handle it for us.
+
+In either case, when \texttt{init} dies,
+the pid namespace will be destroyed,
+and \texttt{grandchild} will be cleaned up.
+
+\subsection{Shared file descriptor tables}
+A library can create a child process to perform a useful service.
+This can sometimes be done more easily when the child process shares file descriptors with the parent.
+\begin{lstlisting}[float,language=Python,label={fuse},caption={Shared file descriptor tables}]
+ns_child = local_thread.clone(CLONE.FILES|CLONE.NEWNS)
+server_child = ns_child.clone()
+await server_child.execve('/bin/foofs', ['foofs', "--mount-at", "/"])
+fd = ns_child.open("/foo/bar", O.RDONLY)
+parent_fd = local_thread.use_fd(fd)
+\end{lstlisting}
+In listing \ref{fuse},
+we first create a child process, \verb|ns_child|, in a new mount namespace with \verb|CLONE_NEWNS|,
+and for the first time, also pass \verb|CLONE_FILES|,
+which causes the file descriptor table to be shared between the parent process and the child process.
+We create a grandchild off of \verb|ns_child|,
+and use it to exec a FUSE filesystem,
+which will appear only in the mount namespace of \verb|ns_child| and its descendents.
+Then we can open FUSE files in \verb|ns_child|
+and use those files in the parent process,
+through the shared file descriptor table.
 \section{Implementation}\label{implementation}

### 2020-05-11 python/rsyscall/environ.py
environ: fix executable lookup to only check for executability
Not readability; a file can be executable and not readable.

@@ -66,7 +66,7 @@ class ExecutablePathCache:
             try:
                 # TODO hmm this returns fine for directories tho. hm. hm.
                 # oh well we'll just fail at exec time, that was always possible
-                await fd.faccessat(name, OK.R|OK.X)
+                await fd.faccessat(name, OK.X)
             except OSError:
                 return False
             else:

### 2020-05-11 research/paper/paper.tex
things

@@
-% \documentclass[sigplan]{acmart}
-\documentclass{acmart}
+\documentclass[sigplan]{acmart}
+% \documentclass{acmart}
 
@@
 Direct-style is rare in general, but relatively common on academic operating systems.
-% insert citations to direct-style papers here
+\cite{keykos}\cite{sel4}\cite{exokernel}\cite{fuschia}\cite{singularity}
 Direct-style involves creating a new child process in a non-running state,
@@
 and consider their advantages and disadvantages in more depth.
-In section \ref{examples} we show examples of usage of direct-style process creation in Python on Linux,
-including use of process features which are difficult to use on Linux with \texttt{fork} or \texttt{posix\_spawn},
-and briefly examine the user API.
+In section \ref{overview} we discuss the design of the API,
+including our usage of \texttt{rsyscall}.
+In section \ref{applications} we show examples of usage of direct-style process creation in Python on Linux,
+including use of process features which are difficult to use on Linux with \texttt{fork} or \texttt{posix\_spawn}.
 In section \ref{implementation}, we explore our implementation in depth,
@@
 
-\section{Overview}\label{examples}
+\section{Overview}\label{overview}
 In most operating system APIs, including the ones available on Linux,
@@
 
-\begin{lstlisting}[float,language=Python,label={basic},caption={Creating a new process, changing CWD, and execing}]
+\begin{lstlisting}[float=*,language=Python,label={basic},caption={Creating a new process, changing CWD, and execing}]
 child = local_thread.clone()
@@
 can also be expressed with direct-style clone.
-\section{Applications}
+\section{Applications}\label{applications}
 In this section,
@@
 to more closely meet the specific needs of individual real-world applications.
-
 \subsection{Abstraction through file descriptor passing}\label{fd_abstraction}
@@
 or the executable running in the process on the other end of the pipe.
-\begin{lstlisting}[float,language=Python,label={fds},caption={Passing down FDs}]
+
+\begin{lstlisting}[float=*,language=Python,label={fds},caption={Passing down FDs}]
 db_fd = local_thread.open("/var/db/database.db", O.RDWR)
@@
 or produce multiple outputs.
-\begin{lstlisting}[float,language=Python,label={pipe},caption={Creating a parallel processing pipeline}]
+
+\begin{lstlisting}[float=*,language=Python,label={pipe},caption={Creating a parallel processing pipeline}]
 def as_argument(child, fd):
@@
 We first create two pipes,
-then inherit them down to several processes using a helper function, \texttt{as_argument},
+then inherit them down to several processes using a helper function, \verb|as_argument|,
 which uses \verb|inherit_fd| and \texttt{fcntl} as described in section \ref{fd_abstraction}.
@@
 
-\begin{lstlisting}[float,language=Python,label={mount},caption={Overriding absolute path using a mount namespace}]
+\begin{lstlisting}[float=*,language=Python,label={mount},caption={Overriding absolute path using a mount namespace}]
 child = local.thread.clone(CLONE.NEWNS)
@@
 \end{lstlisting}
-
 One classic form of customization enabled by namespaces
@@
 
-\begin{lstlisting}[float,language=Python,label={unmount},caption={Unmount all and run executable via fexec}]
+\begin{lstlisting}[float=*,language=Python,label={unmount},caption={Unmount all and run executable via fexec}]
 thread = local_thread.clone(CLONE.NEWNS)
@@
 \end{lstlisting}
-
 In listing \ref{unmount}
@@
 Then we use \texttt{umount},
-passing \texttt{MNT>DETACH} to perform a recursive unmount of the entire filesystem tree,
+passing \texttt{MNT.DETACH} to perform a recursive unmount of the entire filesystem tree,
 removing it all from the view of this process.
@@
 we don't need to call \verb|inherit_fd| since, in this case, the database fd was opened directly from the child.
+
 We then exec the file we opened earlier, using \texttt{fexec},
@@
 which Linux can increasingly closely approximate.
-
 \subsection{Nested clone and pid namespaces}
@@
 
-\begin{lstlisting}[float,language=Python,label={pidns},caption={Nested clone and pid namespace}]
+\begin{lstlisting}[float=*,language=Python,label={pidns},caption={Nested clone and pid namespace}]
 init = local_thread.clone(CLONE.NEWPID)
@@
 \end{lstlisting}
-
 In listing \ref{pidns},
@@
 and \texttt{grandchild} will be cleaned up.
-
 \subsection{Shared file descriptor tables}
@@
 This can sometimes be done more easily when the child process shares file descriptors with the parent.
-\begin{lstlisting}[float,language=Python,label={fuse},caption={Shared file descriptor tables}]
+
+\begin{lstlisting}[float=*,language=Python,label={fuse},caption={Shared file descriptor tables}]
 ns_child = local_thread.clone(CLONE.FILES|CLONE.NEWNS)
@@
 \section{Evaluation}\label{evaluation}
-\subsection{performance ideas}
+    We quantitatively answer the following questions:
+
+    - Does direct-style process creation on Linux allow for simpler programs than the alternatives?
+      We implement and compare similar functionality using both direct-style and fork-style.\ref{ease}
+
+    - Does a clone-based implementation avoid the well-known performance overheads of fork?
+      We evaluate clone against fork on several microbenchmarks.\ref{microbench}
+
+    - Does a high-level Python direct-style process creation interface
+      have acceptable performance relative to the Python standard library's spawn-style interface?
+      We benchmark direct-style process creation against spawn-style.\ref{subprocess_run}
+\subsection{Ease of programming with direct-style}\label{ease}
+Implementing Python's popen/subprocess with rsyscall?
+But that's not fair, theirs is portable, we'll be way simpler for free.
+
+Rewrite some existing system with rsyscall?
+Write a shell?
+\subsection{Clone versus fork}\label{microbench}
 Can mmap a bunch of stuff in a process and then create some children,
@@
 Can make the classic chart of fork vs glibc's \verb|posix_spawn| vs direct-style.
-\subsection{abstraction ideas}
-Implementing Python's popen/subprocess with rsyscall?
-But that's not fair, theirs is portable, we'll be way simpler for free.
-
-Rewrite some existing system with rsyscall?
-Write a shell?
+\subsection{High-level direct-style versus spawn-style}\label{subprocess_run}
+Compare clone/exec with rsyscall against Python's subprocess.run.
+\section{Related work}\label{related_work}
+% TODO all this is just slapped onto the page without much thought, needs revising
+\subsection{Direct-style process creation}
+As we discuss in section \ref{background},
+many academic operating systems have direct-style process creation.
+We are not aware of any previous instances of direct-style process creation in a Unix-like environment.
+
+The closest related work known to us
+is efforts to build Unix compatibility environments, including fork-style process creation,
+on academic operating systems which natively use direct-style process creation.\cite{exokernel}
+Such compatibility environments could theoretically be bypassed to perform direct-style process creation
+using the underlying primitives,
+but regular Unix system calls could not be used to create new Unix processes in a direct-style way
+in such an environment.
+\subsection{Remote system calls}
+There are a large number of Unix-based systems which do something called "remote system calls".
+Most such systems do not allow a single program to manipulate multiple processes.
+Those systems that do allow manipulating multiple processes
+are generally oriented towards debugging and introspection,
+and are unsuitable for a general purpose system.
+
+Many systems use system call forwarding to implement migration in a computing cluster.
+HTCondor\cite{condor} and Popcorn\cite{popcorn} are two examples.
+In both systems, processes can be live-migrated between hosts;
+when this occurs, the system will automatically forward IO-related system calls
+back to the original host.
+
+Some systems intercept system calls and forward them to another process to be implemented
+as part of a virtualization system.
+gvisor\cite{gvisor} and ViewOS\cite{viewos} are two examples.
+In both systems,
+system calls made in one process are intercepted,
+and serviced by another process.
+Also relevant is seccomp system call interception,
+which can be used to intercept system calls and emulate them in another process.
+
+Some systems forward system calls elsewhere for asynchronous execution.
+FlexSC\cite{flexsc} is one example.
+System calls were forwarded to dedicated system call execution threads.
+The recently added Linux feature \verb|io_uring| is, in some sense, another example;
+it supports sending a subset of system calls to the kernel,
+which executes them asynchronously,
+rather than blocking the thread.
+
+Some systems call system calls in other processes for debugging or introspection purposes.
+strace, gdb, and CRIU\cite{criu} are examples of this.
+These systems typically uses ptrace,
+which is capable of operating on multiple processes at once.
+Unfortunately, as discussed in section \ref{implementation},
+ptrace is unsuitable as a mechanism for a general purpose system since it is not re-entrant;
+a process which is already being ptraced cannot be ptraced again,
+so using ptrace as a cross-process system call mechanism would disallow use of strace or debuggers.
+\subsection{Process capabilities}
+We mentioned the concept of process capabilities while discussing the design of rsyscall in section \ref{overview}.
+The Capsicum system provides something called process descriptors\cite{capsicum},
+which is a file descriptor handle for a process.
+A similar feature has been recently added to Linux in the form of the pidfd API\cite{pidfd}.
+These notions of process capability are quite limited, however;
+pidfds and process descriptors only allow sending signals to a process or waiting for its death,
+rather than exercising full control over the process.
+\subsection{Capability-secure libc replacements}
+The Capsicum system provides a set of system calls
+which can be used to provide a capability-secure sandbox.\cite{capsicum}
+Latter efforts\cite{oblivious}, including CloudABI\cite{cloudabi} and WASI\cite{wasi},
+are focused primarily on backwards compatibility with the existing POSIX environment, and ease of porting existing code.
+Programs ported to the capability-secure environment
+are typically launched with an accompanying spawn-style wrapper.
+This means that capability-secure processes are created primarily by an API
+that is distinct from the capability-secure syscall API they have access to,
+which makes it harder for a capability-secure process to spawn further capability-secure processes of its own.
+Direct-style process creation could make it easier to create capability-secure processes 
+from the capability-secure syscall API,
+and reduce the need for a separate spawn-style wrapper.
+
+Also related is the PLASH, ``Principle of Least Authority Shell''\cite{plash}.
+PLASH provides a spawn-style API (in the form of a shell)
+to launch processes in a capability-secure environment.
+PLASH, like all spawn-style APIs, abstracts over the native Linux environment,
+and is therefore limited in what kind of processes it can create.
 \section{Future work}\label{future_work}
@@
 \bibliography{bibliography}
-
 \end{document}

### 2020-05-11 python/rsysapps/miredo.py

@@ -100,6 +100,70 @@ async def start_miredo_internal(thread: Thread, miredo_exec: MiredoExecutables) 
     await inet_sock.bind(await thread.ptr(SockaddrIn(0, 0)))
     # set some miscellaneous additional sockopts that Miredo wants
     await set_miredo_sockopts(thread, inet_sock)
+    ### create dummy thread just for cloning off
+    outer_ns_thread = await thread.clone(CLONE.NEWNET|CLONE.NEWUSER|CLONE.FILES)
+    ### create a thread in a new network namespace
+    netns_thread = await outer_ns_thread.clone(CLONE.FILES) # lol skipping it for now
+    ### create in-network-namespace raw INET6 socket which Miredo will use to relay pings
+    icmp6_fd = await netns_thread.socket(AF.INET6, SOCK.RAW, IPPROTO.ICMPV6)
+    ### create in-network-namespace socket which Miredo will use for unassociated Ifreq ioctls
+    reqsock = await netns_thread.socket(AF.INET, SOCK.STREAM)
+    ### create and set up the TUN interface
+    tun_fd, tun_index = await make_tun(netns_thread, "miredo", reqsock)
+    ### create socketpair which Miredo will use to communicate between privileged process and Teredo client
+    privproc_pair = await netns_thread.socketpair(AF.UNIX, SOCK.STREAM)
+    ### create thread for miredo client
+    # the client process doesn't need to be in the same network namespace, since it is passed all
+    # the resources it needs as fds at startup.
+    client_thread = await thread.clone(CLONE.NEWUSER|CLONE.NEWNET|CLONE.NEWNS|CLONE.NEWPID)
+    # lightly sandbox by unmounting everything except for the executable and its deps (known via package manager)
+    await unmount_everything_except(client_thread, miredo_exec.run_client.executable_path)
+    # a helper function
+    async def pass_fd(fd: FileDescriptor) -> str:
+        await client_thread.inherit_fd(fd).disable_cloexec()
+        return str(int(fd))
+    client_child = await client_thread.execve(miredo_exec.run_client.executable_path, [
+        miredo_exec.run_client.executable_path,
+        await pass_fd(inet_sock), await pass_fd(tun_fd), await pass_fd(reqsock),
+        await pass_fd(icmp6_fd), await pass_fd(privproc_pair.second),
+        "teredo.remlab.net", "teredo.remlab.net"
+    ])
+    ### prepare to exec miredo's privileged helper in the network namespace thread
[…]

### 2020-05-11 python/rsyscall/tests/test_fork.py

@@ -72,25 +72,23 @@ async def prep() -> None:
     return (await local.thread.environ.which("echo")).args("hello world")
 cmd = trio.run(prep)
 echo_path = cmd.executable_path
-async def rsys_run() -> None:
-    count = 500
+async def rsys_run(count: int) -> None:
     for _ in range(count):
         thread = await local.thread.clone()
         child = await thread.execv(cmd.executable_path, [cmd.executable_path, "-n", "hello"])
         await child.waitpid(W.EXITED)
         await thread.close()
-        # print("free list length", len(local.thread.ram.allocator.shared_allocator.arenas[0].free_list))
 import pstats
-def main():
+async def main():
     pr = cProfile.Profile()
-    pr.enable()    
-    trio.run(rsys_run)
+    await rsys_run(100)
+    pr.enable()
+    await rsys_run(500)
     pr.disable()
     pr.print_stats(sort='cumtime')
     # ps = pstats.Stats(pr).strip_dirs().sort_stats('cumulative')
     # ps.print_callees()
-    print("arenas count", len(local.thread.ram.allocator.shared_allocator.arenas))
     return pr
 
 if __name__ == "__main__":
-    pr = main()
+    pr = trio.run(main)

### 2020-05-11 research/paper/clone.c

@@ -1,0 +1,90 @@
+#include <unistd.h>
+#include <stdio.h>
+#include <stdlib.h>
+#include <sys/wait.h>
+#include <sys/time.h>
+#include <sys/mman.h>
+#include <sys/syscall.h>
+#include <sched.h>
+#include <string.h>
+#include <err.h>
+
+/* how to store these timestamps? */
+/* I guess I'll put them in shared memory, and then dump at the end? */
+/* uh hm. */
+/* they need to be in map_shared memory I guess, so that they're inherited across the fork. */
+
+#define NUM_RUNS 1000
+
+struct log {
+        int pid;
+        struct timeval pre_fork;
+        struct timeval in_child;
+        struct timeval in_parent;
+        struct timeval after_exit;
+};
+
+
+void run(struct log *logs) {
+        for (int i = 0; i < NUM_RUNS; i++) {
+                struct log* const log = &logs[i];
+                int rc;
+                gettimeofday(&log->pre_fork, NULL);
+                // rc = syscall(SYS_clone, 0, NULL, NULL, NULL, NULL);
+                rc = syscall(SYS_fork);
+                if (rc == 0) {
+                        gettimeofday(&log->in_child, NULL);
+                        exit(0);
+                }
+                log->pid = rc;
+                gettimeofday(&log->in_parent, NULL);
+                waitpid(rc, NULL, 0);
+                gettimeofday(&log->after_exit, NULL);
+        }
+}
+
+struct options {
+        int num_mappings;
+};
+
+static struct options parse_options(int argc, char **argv) {
+        void usage() {
+                printf("Usage: %s num_mappings\n", argc >= 1 ? argv[0] : "clone_bench");
+                exit(1);
+        }
+        if (argc < 2) usage();
+        return (struct options) { .num_mappings = atoi(argv[1]) };
+}
+
+int main(int argc, char **argv) {
+        struct options opt = parse_options(argc, argv);
+        /* set up memory mappings with a lot of memory */
+        for (int i = 0; i < opt.num_mappings; i++) {
+                char* addr = mmap(NULL, 4096, PROT_READ|PROT_WRITE, MAP_PRIVATE|MAP_ANONYMOUS, -1, 0);
+                // touch the memory
+                *addr = 1;
+        }
+        /* run the benchmark */
[…]

### 2020-05-11 python/ffibuilder.py
sys.resource: add setpriority/getpriority

@@ -31,6 +31,7 @@ ffibuilder.set_source_pkgconfig(
 #include <sys/mount.h>
 #include <sys/prctl.h>
 #include <sys/ptrace.h>
+#include <sys/resource.h>
 #include <sys/signal.h>
 #include <sys/signalfd.h>
 #include <sys/socket.h>
@@ -578,6 +579,15 @@ struct inotify_event {
 typedef struct {
   unsigned long __bits[...];
 } cpu_set_t;
+
+// getpriority/setpriority
+
+#define PRIO_PROCESS ...
+#define PRIO_PGRP ...
+#define PRIO_USER ...
+
+#define SYS_getpriority ...
+#define SYS_setpriority ...
 
 // mount stuff
 #define SYS_mount ...

### 2020-05-11 python/rsyscall/handle/__init__.py

@@ -61,6 +61,7 @@ from rsyscall.unistd.io    import IOFileDescriptor, SeekableFileDescriptor
 from rsyscall.sys.capability import CapabilityTask
 from rsyscall.sys.prctl    import PrctlTask
 from rsyscall.sys.mount    import MountTask
+from rsyscall.sys.resource import ResourceTask
 from rsyscall.sched        import SchedTask
 
 # re-exported
@@ -185,6 +186,7 @@ class Task(
         CredentialsTask,
         ProcessTask,
         SchedTask,
+        ResourceTask,
         SignalTask, rsyscall.far.Task,
 ):
     def __init__(self,

### 2020-05-11 python/rsyscall/handle/process.py

@@ -6,6 +6,7 @@ from rsyscall.handle.pointer import Pointer, WrittenPointer
 from rsyscall.linux.futex import FutexNode
 from rsyscall.sched import Stack, CLONE, _clone, _unshare
 from rsyscall.signal import SIG, Siginfo, _kill
+from rsyscall.sys.resource import PRIO, _setpriority, _getpriority
 from rsyscall.sys.wait import W, ChildState, _waitid
 from rsyscall.unistd.credentials import _getpgid, _setpgid
 import contextlib
@@ -43,6 +44,12 @@ class Process:
 
     async def getpgid(self) -> rsyscall.near.ProcessGroup:
         return (await _getpgid(self.task.sysif, self.near))
+
+    async def setpriority(self, prio: int) -> None:
+        return (await _setpriority(self.task.sysif, PRIO.PROCESS, self.near.id, prio))
+
+    async def getpriority(self) -> int:
+        return (await _getpriority(self.task.sysif, PRIO.PROCESS, self.near.id))
 
 class ChildProcess(Process):
     """A process that is our child, which we can monitor with waitid and safely signal.

### 2020-05-11 python/rsyscall/sys/resource.py

@@ -1,0 +1,52 @@
+from rsyscall._raw import lib # type: ignore
+import enum
+import typing as t
+
+__all__ = ["PRIO"]
+
+class PRIO(enum.IntFlag):
+    NONE = 0
+    PROCESS = lib.PRIO_PROCESS
+    PGRP = lib.PRIO_PGRP
+    USER = lib.PRIO_USER
+
+#### Classes ####
+import rsyscall.far
+
+class ResourceTask(rsyscall.far.Task):
+    async def setpriority(self, which: PRIO, prio: int) -> None:
+        return (await _setpriority(self.sysif, which, 0, prio))
+
+    async def getpriority(self, which: PRIO) -> int:
+        return (await _getpriority(self.sysif, which, 0))
+
+#### Raw syscalls ####
+from rsyscall.near.sysif import SyscallInterface
+from rsyscall.sys.syscall import SYS
+
+async def _setpriority(
+        sysif: SyscallInterface,
+        which: PRIO,
+        who: int,
+        prio: int,
+) -> None:
+    """The raw setpriority syscall
+
+    This syscall is tricky to assign a type to. It's similar to waitid, but has the extra
+    complication that we can pass 0 for the "who" argument to operate on the current process/process
+    group/user, as determined by PRIO.
+
+    I considered making the type of "who", t.Union[Process, ProcessGroup, User, PRIO], and removing
+    the "which" argument; passing PRIO would cause setpriority to operate on the current
+    Process/ProcessGroup/User as determined by the PRIO value. That's the same as how our waitid
+    wrapper works, but that usage of PRIO seemed too weird.
+
+    """
+    await sysif.syscall(SYS.setpriority, which, who, prio)
+
+async def _getpriority(
+        sysif: SyscallInterface,
+        which: PRIO,
+        who: int,
+) -> int:
+    return await sysif.syscall(SYS.getpriority, which, who)

### 2020-05-11 python/rsyscall/sys/syscall.py

@@ -36,6 +36,7 @@ class SYS(enum.IntEnum):
     getgid = lib.SYS_getgid
     getpeername = lib.SYS_getpeername
     getpgid = lib.SYS_getpgid
+    getpriority = lib.SYS_getpriority
     getsockname = lib.SYS_getsockname
     getsockopt = lib.SYS_getsockopt
     getuid = lib.SYS_getuid
@@ -73,6 +74,7 @@ class SYS(enum.IntEnum):
     set_tid_address = lib.SYS_set_tid_address
     setns = lib.SYS_setns
     setpgid = lib.SYS_setpgid
+    setpriority = lib.SYS_setpriority
     setsid = lib.SYS_setsid
     setsockopt = lib.SYS_setsockopt
     shutdown = lib.SYS_shutdown

### 2020-05-11 python/rsyscall/memory/socket_transport.py
socket_transport: check pointer and data size match when writing
If these are mismatched, then we can try to read more data than we
wrote, and deadlock.

@@ -275,6 +275,8 @@ class PrimitiveSocketMemoryTransport(MemoryTransport):
         return PrimitiveSocketMemoryTransport(self.local, task.make_fd_handle(self.remote))
 
     async def write(self, dest: Pointer, data: bytes) -> None:
+        if dest.size() != len(data):
+            raise Exception("mismatched pointer size", dest.size(), "and data size", len(data))
         dest = to_span(dest)
         src = await self.local.ram.ptr(data)
         async def write() -> None:
@@ -381,6 +383,8 @@ class SocketMemoryTransport(MemoryTransport):
                             _, split_rest = await self.remote.read(split_rest)
 
     def _start_single_write(self, dest: Pointer, data: bytes) -> WriteOp:
+        if dest.size() != len(data):
+            raise Exception("mismatched pointer size", dest.size(), "and data size", len(data))
         write = WriteOp(dest, data)
         self.pending_writes.append(write)
         return write

### 2020-05-11 python/rsyscall/thread.py
thread: add some additional methods directly to Thread
These could be called through thread.task, but having them directly on
Thread simplifies the examples in the direct-style paper.

@@ -17,7 +17,8 @@ from rsyscall.linux.dirent import DirentList
 from rsyscall.sched import CLONE
 from rsyscall.sys.mount import MS
 from rsyscall.sys.wait import ChildState, W
-from rsyscall.unistd import Arg
+from rsyscall.sys.socket import Socketpair, AF, SOCK
+from rsyscall.unistd import Arg, Pipe
 
 async def write_user_mappings(thr: RAMThread, uid: int, gid: int,
                               in_namespace_uid: int=None, in_namespace_gid: int=None) -> None:
@@ -130,6 +131,21 @@ class Thread(UnixThread):
             )
         source_ptr, target_ptr, filesystemtype_ptr, data_ptr = await self.ram.perform_batch(op)
         await self.task.mount(source_ptr, target_ptr, filesystemtype_ptr, mountflags, data_ptr)
+
+    async def socket(self, domain: AF, type: SOCK, protocol: int=0) -> FileDescriptor:
+        return await self.task.socket(domain, type, protocol)
+
+    async def pipe(self) -> Pipe:
+        return await (await self.task.pipe(await self.malloc(Pipe))).read()
+
+    async def socketpair(self, domain: AF, type: SOCK, protocol: int=0) -> Socketpair:
+        return await (await self.task.socketpair(domain, type, protocol, await self.malloc(Socketpair))).read()
+
+    async def chroot(self, path: Path) -> None:
+        await self.task.chroot(await self.ptr(path))
+
+    def inherit_fd(self, fd: FileDescriptor) -> FileDescriptor:
+        return self.task.inherit_fd(fd)
 
     async def clone(self, flags: CLONE=CLONE.NONE) -> ChildThread:
         "Create a new child thread"

### 2020-05-11 python/rsyscall/handle/fd.py
fd: support inheritance of implicitly-inherited fds
If a task clones, then clones again, the grandchild will have
inherited its fds, because it inherits everything the child had. But
we didn't have the bookkeeping for that previously; instead the
grandchild would seem to have inherited nothing.

Now we include inherited fds in the list of inherited fds, so when
inheritance is iterated without intervening calls to inherit_fd, it
will work correctly.

@@ -273,6 +273,8 @@ class FDTable(rsyscall.far.FDTable):
             for handles in parent.near_to_handles.values():
                 for handle in handles:
                     self.inherited.add(handle)
+            for handle in parent.inherited:
+                self.inherited.add(handle)
 
     def remove_inherited(self) -> None:
         self.inherited = WeakSet()

### 2020-05-11 python/rsyscall/path.py
path: correct minor type error

@@ -18,7 +18,7 @@ class PurePosixPath(pathlib.PurePosixPath):
     class insulates us from that stuff, so it can be inherited from naively.
 
     """
-    def __new__(cls, *args, **kwargs) -> None:
+    def __new__(cls, *args, **kwargs) -> PurePosixPath:
         """Override pathlib.PurePath.__new__ to restore default behavior
 
         pathlib.PurePath inherits from object, so we just use object's __new__.

### 2020-05-11 python/rsyscall/nix.py
nix: fix lazy creation of local Store object

@@ -257,6 +257,6 @@ def import_nix_dep(name: str) -> StorePath:
     store_path = StorePath._load_without_registering(name)
     # the local store has a root for every StorePath; that's where the
     # paths actually originally are.
-    local_store._add_root(store_path, store_path.path)
+    __getattr__("local_store")._add_root(store_path, store_path.path)
     _imported_store_paths[name] = store_path
     return store_path

### 2020-05-11 python/rsyscall/net/if_.py
net.if: make Ifreq.name field into a string, not bytes
No need for bytes here, the string doesn't have null bytes in it, it's
the same semantics as paths.

@@ -20,15 +20,20 @@ IFF_TUN: int = lib.IFF_TUN
 TUNSETIFF: int = lib.TUNSETIFF
 SIOCGIFINDEX: int = lib.SIOCGIFINDEX
 
-class BytesField:
+class CStringField:
     def __init__(self, name: str) -> None:
         self.name = name
 
-    def __get__(self, instance, owner) -> bytes:
-        return bytes(ffi.buffer(getattr(instance.cffi, self.name)))
+    def __get__(self, instance, owner) -> str:
+        data = bytes(ffi.buffer(getattr(instance.cffi, self.name)))
+        try:
+            valid_data = data[:data.index(b'\0')]
+        except ValueError:
+            valid_data = data
+        return valid_data.decode()
 
-    def __set__(self, instance, value: bytes) -> None:
-        setattr(instance.cffi, self.name, value)
+    def __set__(self, instance, value: str) -> None:
+        setattr(instance.cffi, self.name, value.encode() + b"\0")
 
 class IntField:
     def __init__(self, name: str) -> None:
@@ -63,12 +68,12 @@ class Ifreq(Struct):
     extract some specific field from the union, stored as a cffi type.
 
     """
-    name = BytesField("ifr_name")
+    name = CStringField("ifr_name")
     addr = AddressField("ifr_addr")
     ifindex = IntField("ifr_ifindex")
     flags = IntField("ifr_flags")
     
-    def __init__(self, name: bytes=None, *, flags: int=None, cffi=None) -> None:
+    def __init__(self, name: str=None, *, flags: int=None, cffi=None) -> None:
         if cffi is None:
             cffi = ffi.new('struct ifreq*')
         self.cffi = cffi
@@ -95,13 +100,10 @@ class Ifreq(Struct):
         return ffi.sizeof('struct ifreq')
 
     def __str__(self) -> str:
-        try:
-            strname = self.name[:self.name.index(b'\0')].decode()
-        except ValueError:
-            strname = self.name.decode()
-        if strname:
-            return f"Ifreq({strname}, ...)"
+        if self.name:
+            return f"Ifreq({self.name}, ...)"
         else:
+            # if it's an empty string, indicate that there's no name
             return "Ifreq(<no interface name>, ...)"
 
 
@@ -110,7 +112,7 @@ from unittest import TestCase
 class TestIf(TestCase):
     def test_ifreq(self) -> None:
         initial = Ifreq()
-        initial.name = b"hello"
[…]

### 2020-05-11 python/rsyscall/tests/test_net.py

@@ -26,7 +26,7 @@ class TestNet(TrioTestCase):
 
     async def test_make_tun(self) -> None:
         tun_fd = await self.thr.task.open(await self.thr.ram.ptr(Path("/dev/net/tun")), O.RDWR)
-        ptr = await self.thr.ram.ptr(Ifreq(b'tun0', flags=IFF_TUN))
+        ptr = await self.thr.ram.ptr(Ifreq('tun0', flags=IFF_TUN))
         await tun_fd.ioctl(TUNSETIFF, ptr)
         sock = await self.thr.task.socket(AF.INET, SOCK.STREAM)
         await sock.ioctl(SIOCGIFINDEX, ptr)
@@ -38,7 +38,7 @@ class TestNet(TrioTestCase):
         await netsock.bind(await self.thr.ram.ptr(SockaddrNl(0, RTMGRP.LINK)))
 
         tun_fd = await self.thr.task.open(await self.thr.ram.ptr(Path("/dev/net/tun")), O.RDWR)
-        ptr = await self.thr.ram.ptr(Ifreq(b'tun0', flags=IFF_TUN))
+        ptr = await self.thr.ram.ptr(Ifreq('tun0', flags=IFF_TUN))
         await tun_fd.ioctl(TUNSETIFF, ptr)
         sock = await self.thr.task.socket(AF.INET, SOCK.STREAM)
         await sock.ioctl(SIOCGIFINDEX, ptr)

### 2020-05-11 python/rsyscall/sched.py
sched: fix type of _clone bare syscall
The child stack is optional.

@@ -131,7 +131,7 @@ from rsyscall.near.sysif import SyscallInterface
 from rsyscall.sys.syscall import SYS
 import rsyscall.near as near
 
-async def _clone(sysif: SyscallInterface, flags: int, child_stack: near.Address,
+async def _clone(sysif: SyscallInterface, flags: int, child_stack: t.Optional[near.Address],
                 ptid: t.Optional[near.Address], ctid: t.Optional[near.Address],
                 newtls: t.Optional[near.Address]) -> near.Process:
     # We don't use CLONE_THREAD, so we can say without confusion, that clone returns a Process.

### 2020-05-11 python/rsyscall/environ.py
thread: add execv method, use cached environment arglist when called
Serializing the environment on each call is one of the most expensive
things we're doing; let's stop doing it.

It also doesn't call sigprocmask to clear the signal blocks; that's
another minor optimization.

@@ -13,6 +13,7 @@ from rsyscall.command import Command
 from rsyscall.concurrency import run_all
 from rsyscall.handle import Path, Task, FileDescriptor, WrittenPointer
 from rsyscall.memory.ram import RAM
+from rsyscall.unistd import Arg, ArgList
 import os
 import typing as t
 import functools
@@ -95,11 +96,14 @@ class ExecutablePathCache:
 
 class Environment:
     "A representation of Unix environment variables."
-    def __init__(self, task: Task, ram: RAM, environment: t.Dict[str, str]) -> None:
+    def __init__(self, task: Task, ram: RAM, environment: t.Dict[str, str],
+                 arglist_ptr: WrittenPointer[ArgList]=None,
+    ) -> None:
         self.data = environment
         self.sh = Command(Path("/bin/sh"), ['sh'], {})
         self.tmpdir = Path(self.get("TMPDIR", "/tmp"))
         self.path = ExecutablePathCache(task, ram, self.get("PATH", "").split(":"))
+        self.arglist_ptr = arglist_ptr
 
     def __getitem__(self, key: str) -> str:
         return self.data[key]
@@ -135,6 +139,16 @@ class Environment:
         they're shared between all threads.
 
         """
-        env = Environment(task, ram, dict(self.data))
+        env = Environment(task, ram, dict(self.data), self.arglist_ptr)
         env.path = self.path
         return env
+
+    async def as_arglist(self, ram: RAM) -> WrittenPointer[ArgList]:
+        if self.arglist_ptr is None:
+            envp = [Arg('='.join([key, value])) for key, value in self.data.items()]
+            async def op(sem: RAM) -> WrittenPointer[ArgList]:
+                envp_ptrs = ArgList([await sem.ptr(arg) for arg in envp])
+                return await sem.ptr(envp_ptrs)
+            ptr: WrittenPointer[ArgList] = await ram.perform_batch(op)
+            self.arglist_ptr = ptr
+        return self.arglist_ptr

### 2020-05-11 python/rsyscall/unix_thread.py

@@ -111,6 +111,20 @@ class ChildUnixThread(UnixThread):
         await self.task.execve(filename, argv_ptr, envp_ptr, command=command)
         return self.process
 
+    async def execv(self, path: Path,
+                    argv: t.Sequence[t.Union[str, os.PathLike]],
+                    command: Command=None,
+    ) -> AsyncChildProcess:
+        """Replace the running executable in this thread with another; see execve.
+        """
+        async def op(sem: RAM) -> t.Tuple[WrittenPointer[Path], WrittenPointer[ArgList]]:
+            argv_ptrs = ArgList([await sem.ptr(Arg(arg)) for arg in argv])
+            return (await sem.ptr(path), await sem.ptr(argv_ptrs))
+        filename_ptr, argv_ptr = await self.ram.perform_batch(op)
+        envp_ptr = await self.environ.as_arglist(self.ram)
+        await self.task.execve(filename_ptr, argv_ptr, envp_ptr, command=command)
+        return self.process
+
     async def execve(self, path: Path,
                      argv: t.Sequence[t.Union[str, os.PathLike]],
                      env_updates: t.Mapping[str, t.Union[str, os.PathLike]]={},
@@ -140,6 +154,9 @@ class ChildUnixThread(UnixThread):
         for block in inherited_signal_blocks:
             sigmask = sigmask.union(block.mask)
         await self.task.sigprocmask((HowSIG.SETMASK, await self.ram.ptr(Sigset(sigmask))))
+        if not env_updates:
+            # use execv if we aren't updating the env, as an optimization.
+            return await self.execv(path, argv, command=command)
         envp: t.Dict[str, str] = {**self.environ.data}
         for key, value in env_updates.items():
             envp[key] = os.fsdecode(value)

### 2020-05-11 python/rsyscall/handle/fd.py
handle.fd: optimize construction of inherited WeakSet
This is minor but helpful.

@@ -1,6 +1,7 @@
 from __future__ import annotations
 from dataclasses import dataclass
 from weakref import WeakSet
+import itertools
 import abc
 import gc
 import rsyscall.far
@@ -268,13 +269,13 @@ class FDTable(rsyscall.far.FDTable):
         super().__init__(creator_pid)
         self.near_to_handles: t.Dict[rsyscall.near.FileDescriptor, t.List[BaseFileDescriptor]] = {}
         self.tasks: t.List[FileDescriptorTask] = []
-        self.inherited: WeakSet[BaseFileDescriptor] = WeakSet()
         if parent:
-            for handles in parent.near_to_handles.values():
-                for handle in handles:
-                    self.inherited.add(handle)
-            for handle in parent.inherited:
-                self.inherited.add(handle)
+            self.inherited: WeakSet[BaseFileDescriptor] = WeakSet(
+                itertools.chain(itertools.chain.from_iterable(parent.near_to_handles.values()),
+                                parent.inherited)
+            )
+        else:
+            self.inherited = WeakSet()
 
     def remove_inherited(self) -> None:
         self.inherited = WeakSet()

### 2020-05-11 python/rsyscall/far.py
add some additional nicer __str__ methods

@@ -151,5 +151,8 @@ class Task:
             stack.enter_context(ptr.borrow(self))
             return ptr.near
 
+    def __str__(self) -> str:
+        return f"Task({self.near_process.id})"
+
     def __post_init__(self) -> None:
         pass

### 2020-05-11 python/rsyscall/sys/mman.py

@@ -34,6 +34,9 @@ class MemoryMapping:
             raise rsyscall.far.AddressSpaceMismatchError()
         return MemoryMapping(task, self.near, self.file)
 
+    def __str__(self) -> str:
+        return f"MemoryMapping({str(self.task)}, {str(self.near)})"
+
 from rsyscall.handle.fd import BaseFileDescriptor, FileDescriptorTask
 
 class MemoryMappingTask(FileDescriptorTask):

### 2020-05-11 python/rsyscall/handle/__init__.py
add __slots__ to many basic objects

@@ -137,6 +137,7 @@ class FileDescriptor(
     on-demand when run_fd_table_gc is run.
 
     """
+    __slots__ = ()
     def __init__(self, task: Task, near: rsyscall.near.FileDescriptor) -> None:
         super().__init__(task, near)
         self.task: Task = task

### 2020-05-11 python/rsyscall/handle/fd.py

@@ -73,9 +73,18 @@ class BaseFileDescriptor:
     on-demand when run_fd_table_gc is run.
 
     """
+    __slots__ = ('task', 'near', 'valid')
     task: FileDescriptorTask
     near: rsyscall.near.FileDescriptor
-    valid: bool = True
+    valid: bool
+
+    def __init__(self,
+                 task: FileDescriptorTask,
+                 near: rsyscall.near.FileDescriptor,
+    ) -> None:
+        self.task = task
+        self.near = near
+        self.valid = True
 
     def _make_fd_handle(self: T_fd, near: rsyscall.near.FileDescriptor) -> T_fd:
         return self.task.make_fd_handle(near)

### 2020-05-11 python/rsyscall/handle/pointer.py

@@ -51,11 +51,24 @@ class Pointer(t.Generic[T]):
     See also the inheriting class WrittenPointer
 
     """
+    __slots__ = ('mapping', 'transport', 'serializer', 'allocation', 'valid')
     mapping: MemoryMapping
     transport: MemoryGateway
     serializer: Serializer[T]
     allocation: AllocationInterface
-    valid: bool = True
+    valid: bool
+
+    def __init__(self,
+                 mapping: MemoryMapping,
+                 transport: MemoryGateway,
+                 serializer: Serializer[T],
+                 allocation: AllocationInterface,
+    ) -> None:
+        self.mapping = mapping
+        self.transport = transport
+        self.serializer = serializer
+        self.allocation = allocation
+        self.valid = True
 
     async def write(self, value: T) -> WrittenPointer[T]:
         "Write this value to this pointer, consuming it and returning a new WrittenPointer"
@@ -285,6 +298,7 @@ class WrittenPointer(Pointer[T_co]):
     the WrittenPointer. That's mostly syscalls using Sockbufs...
 
     """
+    __slots__ = ('value')
     def __init__(self,
                  mapping: MemoryMapping,
                  transport: MemoryGateway,

### 2020-05-11 python/rsyscall/near/types.py

@@ -26,6 +26,7 @@ class FileDescriptor:
     object know what file descriptor table that is.
 
     """
+    __slots__ = ('number')
     number: int
 
     def __str__(self) -> str:
@@ -64,6 +65,7 @@ class Address:
     what address space that is.
 
     """
+    __slots__ = ('address')
     address: int
 
     def __add__(self, other: int) -> 'Address':
@@ -117,6 +119,7 @@ class MemoryMapping:
     unexpected page sizes.
 
     """
+    __slots__ = ('address', 'length', 'page_size')
     address: int
     length: int
     page_size: int

### 2020-05-11 python/rsyscall/network/connection.py
network.connection: add comment explaining odd socketpair creation
I went to optimize this by passing NONBLOCK when creating the
socketpair, and found I broke it. This comment will hopefully help my
future self avoid doing that again.

@@ -152,6 +152,7 @@ class FDPassConnection(Connection):
         chans = await self.open_channels(count)
         access_socks, local_socks = zip(*chans)
         async def make_afd(sock: FileDescriptor) -> AsyncFileDescriptor:
+            # have to set NONBLOCK after creation because we want the other end to be blocking
             await sock.fcntl(F.SETFL, O.NONBLOCK)
             return await AsyncFileDescriptor.make(self.access_epoller, self.access_ram, sock)
         async_access_socks = [await make_afd(sock) for sock in access_socks]

### 2020-05-11 research/paper/clone.c
benchmark: add stuff

@@ -1,0 +1,148 @@
+#define _GNU_SOURCE
+#include <unistd.h>
+#include <stdio.h>
+#include <stdlib.h>
+#include <sys/wait.h>
+#include <sys/time.h>
+#include <sys/mman.h>
+#include <sys/syscall.h>
+#include <sched.h>
+#include <string.h>
+#include <err.h>
+#include <sched.h>
+
+/* how to store these timestamps? */
+/* I guess I'll put them in shared memory, and then dump at the end? */
+/* uh hm. */
+/* they need to be in map_shared memory I guess, so that they're inherited across the fork. */
+
+#define NUM_RUNS 500
+
+struct log {
+        int pid;
+        struct timeval pre_fork;
+        struct timeval in_child;
+        struct timeval in_parent;
+        struct timeval after_exit;
+};
+
+enum proc_create {
+        FORK,
+        CLONE_FORK,
+        CLONE,
+};
+
+static int do_fork(struct log *log) {
+        int rc;
+        rc = fork();
+        if (rc == 0) {
+                gettimeofday(&log->in_child, NULL);
+                exit(0);
+        }
+        return rc;
+}
+
+static int do_clone_fork(struct log *log) {
+        int rc;
+        rc = syscall(SYS_clone, 0, NULL, NULL, NULL, NULL);
+        if (rc == 0) {
+                gettimeofday(&log->in_child, NULL);
+                exit(0);
+        }
+        return rc;
+}
+
+static int clone_func(void *arg) {
+        struct log *log = arg;
+        gettimeofday(&log->in_child, NULL);
+        exit(0);
+}
+
+char stack[4096];
+
+static int do_clone(struct log *log) {
+        int rc;
+        rc = clone(clone_func, stack + sizeof(stack), CLONE_VM|SIGCHLD, log);
+        return rc;
+}
+
+void run(enum proc_create mode, struct log *logs) {
+        for (int i = 0; i < NUM_RUNS; i++) {
+                struct log* const log = &logs[i];
+                int rc;
+                gettimeofday(&log->pre_fork, NULL);
+                switch (mode) {
+                case FORK: {
+                        rc = do_fork(log);
+                } break;
+                case CLONE_FORK: {
+                        rc = do_clone_fork(log);
+                } break;
+                case CLONE: {
+                        rc = do_clone(log);
+                } break;
+                }
+                log->pid = rc;
[…]

### 2020-05-15 research/paper/paper.tex
paper stuff

@@
 
+\usepackage{graphicx}
 \usepackage{hyperref}
@@
 % Title portion. Note the short title for running heads
-\title[Direct-style process creation on Linux]{Direct-style process creation on Linux}
+\title[DRAFT: Direct-style process creation on Linux]{DRAFT: Direct-style process creation on Linux}
 
@@
 
-In our implementation, a parent process calls the \texttt{clone} syscall
-to create a child process which shares memory and other resources with the parent,
-and which runs no user code.
-The parent then uses \texttt{rsyscall}
-to call arbitrary syscalls in the child process to initialize it;
-for example, the parent might call \texttt{chdir} to change the child's current working directory.
-After the process is fully initialized,
-the parent can call \texttt{execve} in the child,
-causing the process to start running an arbitrary executable
-and detaching the child from the control of \texttt{rsyscall}.
-
+% TODO this sounds low-content and boring.
+% we should probably remove it?
+% In our implementation, a parent process calls the \texttt{clone} syscall
+% to create a child process which shares memory and other resources with the parent,
+% and which runs no user code.
+% The parent then uses \texttt{rsyscall}
+% to call arbitrary syscalls in the child process to initialize it;
+% for example, the parent might call \texttt{chdir} to change the child's current working directory.
+% After the process is fully initialized,
+% the parent can call \texttt{execve} in the child,
+% causing the process to start running an arbitrary executable
+% and detaching the child from the control of \texttt{rsyscall}.
+
+Our direct-style process creation implementation uses \texttt{clone} to create a child process in an embryonic state
+which can then be manipulated from user code using syscalls with \texttt{rsyscall}.
 A direct-style \texttt{clone} has all the advantages of direct-style process creation
@@
 
-We have used direct-style process creation on Linux to do TODO and TODO,
-and we've benchmarked our implementation as TODO.
+We've implemented a variety of useful applications using direct-style process creation on Linux,
+which can be expressed more concisely and more clearly than with fork-style,
+which we'll showcase in brief in this paper.
 Direct-style \texttt{clone} has minimal performance cost in normal situations,
 and can outperform \texttt{fork} in some cases.
-We have deployed direct-style process creation in production internally at Two Sigma.
+We've benchmarked our Python implementation of direct-style \texttt{clone}
+as comparable to the Python standard library's process creation mechanism.
+Finally, we have deployed direct-style process creation in production internally at Two Sigma.
 
@@
 In section \ref{evaluation}, we evaluate, somehow (TODO).
+In section \ref{related_work},
+we examine other work which also touches on the topics covered by this paper.
 In section \ref{future_work},
@@
 
-The most common complaint about fork is its poor performance \cite{forkroad}.
-It copies many attributes about the parent process when creating the child process,
-including setting up copy-on-write memory-mappings in the child process.
-This becomes slower as the parent process has more memory-mappings,
+Fork can have poor performance when called from processes with many memory mappings.\cite{forkroad}
+Fork copies many attributes about the parent process when creating the child process,
+including setting up copy-on-write memory mappings in the child process.
+This becomes slower as the parent process has more memory mappings,
 eventually taking a significant amount of time.
@@
 Some thread libraries provide partial mitigations for this issue,
-but it's up to user code to make use of those mitigations.
+but it's up to user code to make use of those mitigations.\cite{pthread_atfork}
 In combination with fork's poor performance in large-memory programs,
@@
 In a spawn-style interface,
-all the details about the new process are provided up-front as arguments to some function,
-which creates the new process from a mostly-clean slate, initialized with the provided details.
+details about the new process are provided up-front as arguments to some function
+which creates the new process all at once.
 The most well known spawn-style interfaces are Unix's \texttt{posix\_spawn} \cite{posix_spawn}
@@
 the Unix shell also serves as a spawn-style process creation interface,
-especially when used with chain-loading techniques.[fn:chainloading]
+especially when used with Bernstein chaining.\cite{chainloading}
 
@@
 Spawn-style interfaces lift the constraints on the calling process that fork-style interfaces impose.
-Since spawn-style interfaces don't run user code in the new process during initialization,
-there is no need to copy memory,
-and there are no inherent issues when the interface is used from a multi-threaded program
-(though operating systems which implement spawn-style interfaces on top of \texttt{fork} can still have bugs).
+Spawn-style interfaces don't run user code in the new process during initialization.
+This means there is no need to copy memory,
+so performance is good when a spawn-style interface is used from a process with many memory mappings.
+Likewise, there are no inherent issues when a spawn-style interface is used from a multi-threaded program,
+though operating systems which implement spawn-style interfaces on top of \texttt{fork} can still have threading bugs.
 
@@
 
-Furthermore, spawn-style process creation
-creates a new domain-specific language for specifying the attributes of the process.
+Spawn-style process creation
+essentially creates a new domain-specific language for specifying the attributes of the process.
 Fork-style allows existing system calls,
@@
 
-Spawn-style process creation also does not allow for conditional logic during the setup;
-if the setup of the new process encounters an error at some point,
+Spawn-style process creation also does not allow for conditional logic during the setup.
+If the setup of the new process encounters an error at some point,
 the only option is to return from the entire spawn call with an error.
@@
 Since everything happens directly from the parent process,
-process initialization is compatible with a variety of techniques
-which were otherwise incompatible with process creation.
-For example, a program could provide a UI which directly customizes the new process,
-or might initialize several processes at once and share data between them.
-
-Most significantly,
+direct-style process creation is compatible with a variety of techniques
+which are difficult with fork-style or spawn-style.
+For example, a REPL could be used to interactively create the new process,
+or a program might initialize several processes at once and share data between them.
+
+Significantly,
 since all operations are performed through syscalls called directly from the main program,
@@
 direct-style interfaces have no constraints on the calling process.
-No user code runs in the child process,
-so, unlike fork-style interfaces,
+Also like spawn-style interface,
+no user code runs in the child process, so
 direct-style interfaces can be used from multi-threaded or large-memory processes without issues.
@@
 bypassing libc.
+
 Unlike POSIX C libraries such as glibc or musl,
@@
 
-From the perspective of an program using rsyscall,
-there are multiple processes which are under its control,
-through multiple capabilities.
-We use the term ``thread'' to refer to all such controlled processes,
-including the main "thread" on which the program is running.
-On Linux, the shared-memory "threads" provided by libraries such as \texttt{pthreads}
+We use the term ``thread'' to refer to the processes under the control of rsyscall,
+including the ``local'' process.
+On Linux, the shared-memory ``threads'' provided by libraries such as \texttt{pthreads}
 are implemented as processes,
 and their lifetime and execution is completely controlled by a single program.
-The same is mostly true of the controlled-process "threads" provided by rsyscall.
-The most significant difference is that rsyscall "threads"
-do not run their own code concurrently with the main program;
-nevertheless rsyscall "threads" do provide the opportunity for parallel execution of system calls,
+The same is mostly true of the controlled-process ``threads'' provided by rsyscall.
+The most significant difference is that rsyscall ``threads''
+do not run user code concurrently with the main program;
+nevertheless rsyscall ``threads'' do provide the opportunity for concurrent execution of system calls,
 and so the terminology provides useful intuition.
@@
 
-\begin{lstlisting}[float=*,language=Python,label={basic},caption={Creating a new process, changing CWD, and execing}]
-child = local_thread.clone()
-child.chdir("/dev")
-child.execve("/bin/cat",
-  ["cat", "./null"], local_thread.environ)
-\end{lstlisting}
+\lstinputlisting[float=*,language=Python,label={basic},caption={Creating a new process, changing CWD, and execing}]
+{examples/basic/direct.py}
 Listing \ref{basic} shows this in action in Python.
@@
 
-The API for direct-style process creation on Linux is not new;
-it is the same Linux system call API that systems programmers are already familiar with.
+The behavior of system calls made on a remote process through \texttt{rsyscall}
+is identical to the behavior of system calls made implicitly on the current process through libc.
+This is an API that systems programmers are already familiar with,
+and which is already thoroughly documented through Linux system call manpages.
 The novelty is constrained to the newly explicit process capabilities;
@@
 Since the API is simply normal Linux system calls,
-anything that could be expressed with fork and syscalls
-can also be expressed with direct-style clone.
+direct-style clone is at least as expressive as fork and syscalls;
+and, in fact, as we'll see in section \ref{applications},
+is significantly more expressive.
 \section{Applications}\label{applications}
@@
 
-% TODO this intro is all a bit muddled, and also not true.
-These examples are all conventional applications of Linux features;
-they are all possible using Linux's existing \texttt{fork} system call,
-and in some cases, \verb|posix_spawn|,
-with varying degrees of difficulty and workarounds.
-But direct-style makes these easier to write;
+All of these examples use only existing Linux system calls and features.
+Most are possible, though difficult to implement,
+with Linux's existing \texttt{fork} system call,
+and in some cases, \verb|posix_spawn|.
+
+Direct-style makes these examples substantially easier to write;
 we'll compare our implementations in this section to fork-style implementations in section \ref{evaluation}.
@@
 Such software often allows use of one or two of these techniques, but prevents the use of others.
+For example, a shell allows the creation of pipelines and container systems like Docker allow sandboxing,
+but the two are difficult to combine.
 
@@
 to more closely meet the specific needs of individual real-world applications.
+
+For concision, we assume that we are running with sufficient permissions in these examples,
+but these examples can all be run without privileges with appropriate use of \texttt{CLONE.NEWUSER}.
 \subsection{Abstraction through file descriptor passing}\label{fd_abstraction}
@@
 or a pipe connected to another process,
+or a variety of other resources,
 without the process being aware of what specific resource it is accessing.
@@
 or the hostname to it which the socket is connected,
-or the executable running in the process on the other end of the pipe.
-
-\begin{lstlisting}[float=*,language=Python,label={fds},caption={Passing down FDs}]
-db_fd = local_thread.open("/var/db/database.db", O.RDWR)
-child = local_thread.clone()
-child_fd = child.inherit_fd(db_fd)
-child_fd.fcntl(F.SETFD, 0)
-child.execv("/bin/fooserver",
-  ["fooserver", "--database-fd", str(int(child_fd))])
-\end{lstlisting}
+or the executable running in the process on the other end of the pipe,
+or other abstracted-away details.
+
+\lstinputlisting[float=*,language=Python,label={fds},caption={Passing down FDs}]
+{examples/fds/direct.py}
 In Listing \ref{fds},
@@
 We indicate that we want to operate on the child's copy of the file descriptor with \verb|child.inherit_fd|.
-\verb|inherit_fd| only updates resource-tracking bookkeeping in our program;
-it makes no system calls.
+\verb|inherit_fd| only updates resource-tracking bookkeeping in our program,
+it makes no system calls;
+the file descriptor was already copied at process creation time.
 Then we disable \texttt{CLOEXEC} on the child's file descriptor,
@@
 
-A popular parallel programming environment is the Unix shell,
+One of the most popular parallel programming environments is the Unix shell,
 which obtains its parallelism by running multiple processes connected via pipes.
@@
 
-\begin{lstlisting}[float=*,language=Python,label={pipe},caption={Creating a parallel processing pipeline}]
-def as_argument(child, fd):
-  child_fd = child.inherit_fd(fd)
-  child_fd.fcntl(F.SETFD, 0) # unset CLOEXEC
-  return str(int(child_fd)) # turn fd number to string
-
-audio_pipe = local_thread.pipe()
-video_pipe = local_thread.pipe()
-source = local.thread.clone()
-source.execve('/bin/source', ['source',
-  '--audio-output', as_argument(source, audio_pipe.write),
-  '--video-output', as_argument(source, video_pipe.write)])
-video_sink = local.thread.clone()
-video_sink.execve('/bin/video_sink', ['video_sink',
-  '--video-input', as_argument(video_sink, video_pipe.read)])
-audio_sink = local.thread.clone()
-audio_sink.execve('/bin/audio_sink', ['audio_sink',
-  '--audio-input', as_argument(audio_sink, audio_pipe.read)])
-\end{lstlisting}
+\lstinputlisting[float=*,language=Python,label={pipe},caption={Creating a parallel processing pipeline}]
+{examples/pipe/direct.py}
 In listing \ref{pipe},
@@
 \subsection{Customization without explicit support}
-While, ideally, all resources would be passed by file descriptor,[fn:capsicum]
+While, ideally, all resources would be passed by file descriptor,\cite{capsicum}
 Unix has a number of resources that are global,
@@
 by customizing the environment the process runs in.
-
-\begin{lstlisting}[float=*,language=Python,label={mount},caption={Overriding absolute path using a mount namespace}]
-child = local.thread.clone(CLONE.NEWNS)
-child.mount("/home/foo/custom_foo.conf", "/etc/foo.conf", "", MS.BIND, "")
-child.execv('/bin/fooserver', ['fooserver'])
-\end{lstlisting}
+Other system calls such as \texttt{chroot} can also be used for this.
+
+\lstinputlisting[float=*,language=Python,label={mount},caption={Overriding absolute path using a mount namespace}]
+{examples/mount/direct.py}
 One classic form of customization enabled by namespaces
-is overriding files at paths that were hardcoded into the program.
+is overriding files at paths that were hardcoded into a program we can't change,
+without changing those files for the rest of the system.
 In listing \ref{mount},
@@
 pass an argument to \texttt{clone}.
+
 \verb|CLONE_NEWNS| causes clone to create the process in a new mount namespace,
-which allows us to use the \texttt{mount} system call
-to manipulate the filesystem tree in a way that only this process will see.
-We call mount, passing \texttt{MS.BIND}, to make the file \verb|/home/foo/custom_foo.conf|
-appear at the path \verb|/etc/foo.conf|;
-this is known as a bind mount.
+which allows us to manipulate the filesystem tree in a way that only this process will see
+using the \texttt{mount} system call.
+We call \texttt{mount}, passing \texttt{MS.BIND}, to make the file \verb|/home/foo/custom_foo.conf|
+appear at the path \verb|/etc/foo.conf|.
+This is known as a bind mount.
 Then we execute some executable,
-which will see the \verb|/etc/foo.conf| as we specified.
+which, when it opens \verb|/etc/foo.conf|, will see the contents of our \verb|custom_foo.conf|.
 \subsection{Sandboxing}
@@
 
-Historically, system calls such as \texttt{chroot} were the main mechanism for doing this;
-today, Linux's namespaces system is the most important component.
-
-\begin{lstlisting}[float=*,language=Python,label={unmount},caption={Unmount all and run executable via fexec}]
-thread = local_thread.clone(CLONE.NEWNS)
-executable_fd = child.open("/bin/static_executable", O.RDONLY)
-db_fd = child.open("/var/db/database.db", O.RDWR)
-child.umount("/", MNT.DETACH)
-db_fd.fcntl(F.SETFD, 0)
-child.fexec(executable_fd, ["static_executable", "--database-fd", str(int(db_fd))])
-\end{lstlisting}
+\lstinputlisting[float=*,language=Python,label={unmount},caption={Unmount all and run executable via fexec}]
+{examples/unmount/direct.py}
 In listing \ref{unmount}
@@
 \subsection{Nested clone and pid namespaces}
-Process-local resource can also be used to control the lifetime of resources used by that process.
+Process-local resources can also be used to control the lifetime of resources used by that process.
 Many Linux resources are not automatically cleaned up on process exit;
@@
 and not ensure that they are cleaned up,
-leaving behind unuseds litter on the system.
+leaving behind unused garbage on the system.
 
@@
 
-We can use a pid namespace, in conjunction with nested cloning, to solve this issue.
+We can use a pid namespace to solve this issue.
 The lifetime of all processes in a pid namespace is tied to the first process created in it,
@@
 
-\begin{lstlisting}[float=*,language=Python,label={pidns},caption={Nested clone and pid namespace}]
-init = local_thread.clone(CLONE.NEWPID)
-grandchild = init.clone()
-grandchild.execv('/bin/fooserver', ['fooserver'])
-\end{lstlisting}
+\lstinputlisting[float=*,language=Python,label={pidns},caption={Nested clone and pid namespace}]
+{examples/pidns/direct.py}
 In listing \ref{pidns},
@@
 passing \verb|CLONE_NEWPID| to create it in a new pid namespace.
+
 To create another process in the pid namespace,
-we clone again, this time from \texttt{init};
-note that this is the first time we've cloned from one of our child processes.
+we clone again, this time from \texttt{init}.
+This is the first example in which we've cloned from one of our child threads;
+as normal, this produces a new process, a grandchild,
+which can be used exactly like a direct child thread.
 We exec in the grandchild,
@@
 and \texttt{grandchild} will be cleaned up.
-\subsection{Shared file descriptor tables}
+\subsection{Shared file descriptor tables}\label{shared_fd_table}
 A library can create a child process to perform a useful service.
@@
 
-\begin{lstlisting}[float=*,language=Python,label={fuse},caption={Shared file descriptor tables}]
-ns_child = local_thread.clone(CLONE.FILES|CLONE.NEWNS)
-server_child = ns_child.clone()
-await server_child.execve('/bin/foofs', ['foofs', "--mount-at", "/"])
-fd = ns_child.open("/foo/bar", O.RDONLY)
-parent_fd = local_thread.use_fd(fd)
-\end{lstlisting}
+\lstinputlisting[float=*,language=Python,label={fuse},caption={Shared file descriptor tables}]
+{examples/fuse/direct.py}
 In listing \ref{fuse},
@@
 through the shared file descriptor table.
+
+File descriptor passing over a Unix domain socket
+can allow similar behavior without sharing the file descriptor table,
+but is substantially more complex for the user.
+Nevertheless, for fork-style and spawn-style process creation,
+file descriptor passing over a Unix domain socket
+is the only option to do something like this.
 \section{Implementation}\label{implementation}
 \subsection{Basics about rsyscall}
-Our main need for implementing direct-style process creation
-is a robust system for cross-process syscalls.
-The rsyscall project provides this.
-
-The rsyscall project develops language-specific libraries
-for the Linux syscall interface, with minimal abstraction.
-The rsyscall project develops a collection of language-specific libraries tightly coupled to Linux,
-which provide access to Linux syscalls with minimal abstraction,
-bypassing libc.
-
-Unlike POSIX C libraries, an rsyscall library is organized in a capability-oriented style;
-the capability to make syscalls in a process is explicicitly passed around as a language object.
-This organization allows an rsyscall program to be process-independent,
-and straightforwardly make system calls in any of the processes under its control,
-so-called ``cross-process'' system calls.
-
-for 
-These 
-for the is an alternative, low-abstraction to the Linux
-The rsyscall project is a toolkit for cross-process syscalls on Linux,
-with several language-specific library implementations.
-
-In this section, we'll give a brief overview of rsyscall,
-and focus on implementation issues specific to process creation.
+In this section, we'll give a brief overview of the implementation of rsyscall,
+and then focus on implementation issues specific to process creation.
 
@@
 the basic cross-process syscall primitive,
-and a language-specific library built on top
-to handle the complexities of manipulating resources across multiple processes.
+and a language-specific library built on top.
 The Python language-specific library has already been demonstrated above.
@@
 they are, for the most part, agnostic to how the cross-process syscall is implemented.
-
-Using the Python library as an example,
-it provides Python wrappers for Linux system calls and structs
-which are type-safe using Python 3 type annotations and runtime checks
-while still providing low-abstraction access to a large subset of native Linux functionality.
-It also provides garbage collection for remote file descriptors, memory and other resources.
-Such features are independent of the precise implementation of the cross-process syscall primitive.
 
@@
 
+We use pipes (or other file descriptors) for transport of data rather than a shared memory segment
+to support straightforward integration with event loops.
+
+ptrace provides an alternative means to perform arbitrary actions on other processes,
+including system calls.
+However, among other issues, it has the unavoidable substantial disadvantage of not permitting multiple ptracers.
+A ptrace-based implementation would prevent using strace or gdb on rsyscall-controlled processes,
+which is an unacceptable limitation for a general-purpose utility.
+
 Many syscalls either take or return pointers to memory,
@@
 Therefore, an rsyscall library needs a way to access memory in the target process.
-We implement this through another set of pipes,
+
+The most simple way to access memory is for the main thread to be in the same address space as the target process.
+This is the case most of the time; we pass \texttt{CLONE.VM} by default in the Python wrapper.
+However, the target process may be in another address.
+For example, if the target process is at a different privilege level,
+it must be in a different address space,
+as sharing an address space with a process at a different privilege level is unsafe.
+
+We implement access to the memory of processes in different address spaces through another set of pipes,
 by explicitly copying memory into and out of those pipes using the \texttt{read} and \texttt{write} system calls.
@@
 then perform a \verb|read(memory_pipe, A, N)| in the target process to copy that data from the pipe into memory.
-
-ptrace provides an alternative means to perform arbitrary actions on other processes.
-However, among other issues, it has the unavoidable substantial disadvantage of not permitting multiple ptracers.
-A ptrace-based implementation would prevent using strace or gdb on rsyscall-controlled processes,
-which is an unacceptable limitation for a general-purpose utility.
 
@@
 \subsection{\texttt{clone}}\label{clone}
-Now that we've established the basic operations which rsyscall provides,
-let's consider the specific issues related to process creation and initialization.
+Now that we've established the basic implementation details of rsyscall,
+we'll consider specific issues related to process creation and initialization.
 
@@
 \texttt{fork}, \texttt{vfork} and \texttt{clone}.
-\texttt{clone} provides a superset of the functionality of the other two,
+\texttt{vfork} suspends execution of the parent process
+while waiting for the child process to exit or call \texttt{execve},
+which is immediately unsuitable for direct-style process creation.
+\texttt{clone} provides a superset of the functionality of \texttt{fork},
 so we focused our attention on \texttt{clone}.
@@
 
-\texttt{clone} only lets us change the stack register for the new process.
+\texttt{clone} only lets us change the stack for the new process;
+the new process begins executing, after \texttt{clone}, at the same instruction as the parent process.
 We would like to be able to set arbitrary registers for the new process,
 so that we can control where it begins executing and the stack it executes on.
+
 Fortunately, changing the stack is sufficient.
-
-We ensure that the next instruction executed after any syscall
-is (in x86 terms) a \texttt{RET};
-this is always the case, so we have no need to special case the execution of \texttt{clone}.
-Since we control the stack of the new process,
+We ensure that the next instruction executed after any syscall is a \texttt{RET}.
+Since we control the state of the stack in the new process after a \texttt{clone} syscall,
 the \texttt{RET} will jump to a code address that we control.
@@
 we can now call \texttt{clone} so that it launches a process running our infinite syscall loop,
-which is implented in C and, as described in the previous section,
+which is implented in freestanding C and, as described in the previous section,
 uses two pipes passed as arguments to receive syscall requests and respond with syscall results.
-
-% TODO is this section necessary?
-The addresses of these functions and trampolines are discovered through a linking procedure.
-When the process being created is in the same address space as the main process which is running user code,
-the location of the rsyscall library in memory, and the addresses of code within it,
-are known through normal language-specific linking mechanisms.
-However, when a process is created with a different address space,
-such as when we establish a connection to a process after it's been started,
-we need to perform linking to learn the addresses of functions.
-This linking procedure is performed while bootstrapping the connection,
-and involves the target process sending a table of important addresses to the connecting process.
 
@@
 most system calls can be called as normal.
-The new process can be modified freely through chdir, dup2, and other system calls.
+The new process can be modified freely through \texttt{chdir}, \texttt{dup2}, and other system calls.
 Out of system calls related to process creation,
@@
 \subsection{\texttt{exec}}
-Eventually, most programs will want to call \texttt{execve} in the processes they create.
 \texttt{execve} is unusual and requires careful design,
@@
 
-One traditional means to detect a successful \texttt{execve} is to create a pipe before forking,
+There is a traditional technique to detect a successful \texttt{execve} using a pipe.
+We create a pipe before forking,
 ensure both ends are marked \verb|O_CLOEXEC|,
@@
 Child processes can be created with the \verb|CLONE_FILES| flag passed to \texttt{clone},
-which causes the parent process and child process to share a single fd table.
+which causes the parent process and child process to share a single fd table;
+we showed an example of this in section \ref{shared_fd_table}.
 This means that when the parent process closes the write end of the pipe,
@@
 and set up a process to wait on that futex.
-Then, when we call \texttt{execve},
-we wait for either the \texttt{execve} to return an error or the futex process to exit,
+Then, when we call any syscall,
+we wait for either the syscall to return an error or the futex process to exit,
 whichever comes first.
-If the futex process exits,
-and the child process doesn't itself exit,
-we know that the child has successfully called \texttt{execve}.
+Either will normally indicate that the syscall fails,
+but if the futex process exits,
+without the child process itself exiting,
+then we can catch that in \texttt{execve} and know that the child has successfully called \texttt{execve}.
 
@@
 kqueue, on FreeBSD, allows waiting for exec in arbitrary processes through kqueue's \verb|EVFILT_PROC|.
-One approach for Linux would be to add a new \texttt{clone} flag to opt-in to receiving \texttt{WEXECED} events through \texttt{waitid};
+
+One approach for Linux would be to add a new \texttt{clone} flag to
+opt-in to receiving \texttt{WEXECED} events through \texttt{waitid};
 note that a \texttt{waitid} flag alone is not sufficient,
 since it's necessary to receive \texttt{SIGCHLD} signals for the \texttt{WEXECED} event if waiting for it from an event loop.
+
 Alternatively, some way to wait for futex wakeups through a file descriptor could be added,
@@
 \subsection{Managing file descriptor tables}\label{fdtables}
-As mentioned in the previous section,
-the \verb|CLONE_FILES| flag can be passed to \texttt{clone}.
-When this flag is passed,
-the file descriptor table is shared between the parent process and child process.
+As demonstrated in section \ref{shared_fd_table},
+the \verb|CLONE_FILES| flag can be passed to \texttt{clone},
+causing the file descriptor table to be shared between the parent process and child process.
 The same file descriptors are open in both processes at the same numbers,
@@
 they are also visible in the other process.
-This is simple to model,
-and convenient for many purposes;
-for example, the child process might be in a different network namespace from the parent,
-and the shared file descriptor table would allow the child to bind a socket
-and the parent to use it;
-the file descriptor table is unshared once the child calls \texttt{execve}.
+This is simple to model, and convenient for many purposes.
+The file descriptor table is unshared once the child calls \texttt{execve}.
 
@@
 containing copies of all the file descriptors existing in the parent at the time of the system call.
-This same behavior can also be triggered after process creation by calling \verb|unshare(CLONE_FILES)| or \texttt{execve};
-if \verb|unshare(CLONE_FILES)| or \texttt{execve} (ignoring \texttt{CLOEXEC}, which we'll discuss later),
-are called in a process currently sharing its file descriptor table with another process,
+This is known as file descriptor inheritance.
+
+This same behavior can also be triggered after process creation by calling \verb|unshare(CLONE_FILES)| or \texttt{execve}.
+If \verb|unshare(CLONE_FILES)|
+is called in a process currently sharing its file descriptor table with another process,
 then after the call that process will have a new, private file descriptor table,
@@
 
+If \texttt{execve} is called,
+the treatment of the file descriptor table is the same as \verb|unshare(CLONE_FILES)|,
+except that file descriptors with the \texttt{CLOEXEC} flag set on them
+are not copied into the newly created file descriptor table.
+In standard Unix practice, \texttt{CLOEXEC} is set habitually by users on all file descriptors;
+we'll discuss the semantics of \texttt{CLOEXEC} more in section \ref{cloexec}.
+\subsubsection{Tracking file descriptors through handles}
 After a system call has copied all the file descriptors in the old table into the new table,
@@
 and which file descriptors should be closed.
-Keeping some file descriptors from the old table in the new table
-is referred to as "file descriptor inheritance".
-\subsubsection{Inheriting file descriptors}
-The rule about which file descriptors stay open is simple:
-We want to keep a file descriptor open in the new table
-if there is a process using that file descriptor.
-We track which processes are using which file descriptors as part of our file descriptor garbage collection system.
-File descriptors are used through garbage-collected handles,
-each of which is associated with a process.
-If there is an existing handle for a (process, file descriptor number) combination,
-this means that the file descriptor with that number in that process's file descriptor table is in use by that process.
-
-This makes inheritance simple in the case of \verb|unshare(CLONE_FILES)|.
-Every file descriptor in use by a process
-is referenced by a (process, file descriptor number) combination;
-these handles are still valid after a call to \verb|unshare(CLONE_FILES)|,
-they simply now refer to the new file descriptor in the process's new file descriptor table.
-So, these file descriptors will automatically be kept open;
-that is, they will be automatically inherited into the new table.
-This same mechanism can work across \texttt{execve},
-if control over the process is re-established after the \texttt{execve},
-and if we carefully manage \texttt{CLOEXEC}.
+
+This is a question of resource management.
+In a language with manual resource management,
+such as C,
+this would be delegated to the user,
+and the responsibilities of the \texttt{rsyscall} library would end there.
+Manual resource management in C has led to many bugs,
+many of which with significant security implications.
+
+Python, however, is a garbage collected language,
+and therefore we determine which file descriptors should stay open through garbage collection:
+if a certain file descriptor is in use by the program,
+through a certain process,
+that file descriptor should stay open in that process's file descriptor table.
+
+To achieve this, file descriptors are used through garbage-collected handles;
+each handle is associated with a process and contains a specific file descriptor number.
+For a handle to remain valid,
+that file descriptor number must continue to refer to an open file descriptor in that process.
+We determine which handles are live (in use by the program)
+by recourse to the native Python garbage collection system.
+We keep open exactly what file descriptors are necessary to keep the live handles valid.
+
+When a process changes file descriptor table through \verb|unshare(CLONE_FILES)|,
+we need to make sure all the live file descriptor handles associated with that process continue to be valid.
+This is straightforward:
+we enumerate all the handles associated with the process,
+and don't close the file descriptors that those handles refer to.
+
+Likewise, if a process changes file descriptor table through \texttt{execve}
+in a way that allows the program to keep control of the process,
+we make sure to keep all the file descriptors with live handles open.
+This requires careful management of \texttt{CLOEXEC}.
 
 When a process is newly created,
-no file descriptor handles yet exist which involve that process.
-Once a handle is created with that process and some file descriptor number,
-the referenced file descriptor will be inherited into (that is, kept open in) the new file descriptor table.
-These handles can be created through the \verb|inherit_fd| function.
-
-At the time of creating a new file descriptor table,
-we perform some bookkeeping:
-We make a list of all the file descriptors that existed in the old table
-at the time of the creation of the new table.
-These file descriptors are the ones which were copied into the new table.
-
-The \verb|inherit_fd| function uses this list.
-It takes as arguments a new process and a file descriptor handle from another process,
-and checks (with the list) that that file descriptor was copied into the new process's file descriptor table
-and hasn't since been closed.
-If so, it creates a new handle for that (process, file descriptor number) combination.
-This causes the file descriptor to stay open in the new file descriptor table,
-and it can be used in the new process through the handle.
-\subsubsection{Closing file descriptors after inheritance}
-After file descriptor inheritance is complete,
-we must promptly close other file descriptors we don't want to inherit.
-Leaving these file descriptors open in the new table is a form of resource leakage.
+no file descriptor handles yet exist;
+New handles can be created by opening files in the process,
+or by using the \verb|inherit_fd| function to create handles for inherited file descriptors.
+\verb|inherit_fd| takes a file descriptor handle associated with the parent process,
+and, if that file descriptor was open at the time the child process was created,
+creates a new handle associated with the child process
+for the child process's copy of that file descriptor.
+Once the inherited file descriptors are closed as described in section \ref{cloexec},
+\verb|inherit_fd| can't be used again until the process next changes file descriptor table.
+\subsubsection{Closing file descriptors after inheritance}\label{cloexec}
+After creating a new file descriptor table,
+and establishing which file descriptors should be kept open,
+we must promptly close other file descriptors.
+Leaving unwanted file descriptors open in the new table is a form of resource leakage.
 It can also cause erroneous behavior.
@@
 
-However, we can't simply close all other file descriptors.
+However, we can't simply close all other file descriptors,
+because we must preserve the semantics of implicit file descriptor inheritance.
 The possibility of implicit inheritance of file descriptors is a traditional Unix feature,
 which is useful in a wide variety of situations,
-in much the same way as implicit inheritance of environment variables;
-it can allow a resource to be passed down a process hierarchy without intervening programs being aware.
-
-Here is where \texttt{CLOEXEC} becomes relevant:
-\texttt{CLOEXEC}, in practice,
-is a tag, set by userspace, for file descriptors which should not be implicitly inherited.
-If \texttt{CLOEXEC} is set on a file descriptor,
-we should close that file descriptor if we don't explicitly want to inherit it;
-if \texttt{CLOEXEC} is not set, we should not close the file descriptor,
-but instead we should allow it to be implicitly inherited.
-
-This interpretation of \texttt{CLOEXEC} is a consequence of \texttt{CLOEXEC}'s primary purpose:
-Managing inheritance of file descriptors over \texttt{execve},
-in programs where the caller of \texttt{execve}
-doesn't know all the file descriptors that may have been opened by the rest of the program;
-this describes, in practice, all programs which call \texttt{execve}.
-When \texttt{CLOEXEC} is set on a file descriptor,
-it will not be copied into the new file descriptor table created after \texttt{execve}.
-Thus \texttt{CLOEXEC} is a way of saying, before an \texttt{execve},
-that this file descriptor will not be used by the new program after an \texttt{execve},
-and therefore should not be inherited.
-Since, in a general purpose program, \texttt{execve} may be called at any time,
-libraries must have \texttt{CLOEXEC} set correctly at all times.
-So we can check \texttt{CLOEXEC} at any time to see whether a file descriptor should be inherited or not,
-even if the file descriptor is used by a library unrelated to rsyscall.
-
-The implementation of closing non-inherited file descriptors is then simple:
-We close all file descriptors which have \texttt{CLOEXEC} set
-and which aren't referenced by an rsyscall file descriptor handle.
-We also clear the list of copied file descriptors which \verb|inherit_fd| uses.
-We do this closing operation in userspace;
-a syscall to perform this would be a useful addition to Linux.
-
-\texttt{CLOEXEC} is set by default on all file descriptors opened by rsyscall,
-though it may be unset by a user program.
-Many user programs will unset \texttt{CLOEXEC} on some file descriptors immediately before calling \texttt{execve}
-so that the executable they run will inherit those file descriptors.
-The new program run by the \texttt{execve} will typically immediately set \texttt{CLOEXEC} again.
-
-An additional argument to the \texttt{execve} syscall
-which allows specifying an explicit list of file descriptor numbers to inherit despite the \texttt{CLOEXEC} flag being set
-would allow programs to avoid this pointless behavior of un-setting and then immediately re-setting \texttt{CLOEXEC}.
-It would also, more significantly, allow programs to inherit file descriptors across \texttt{execve}
-while the file descriptor table is shared;
-unsetting \texttt{CLOEXEC} while the file descriptor table is shared will cause race conditions,
-because the other processes sharing the file descriptor table might call \texttt{execve} at any time,
-and so the file descriptor table must first be unshared before unsetting \texttt{CLOEXEC} and calling \texttt{execve}.
-\section{Evaluation}\label{evaluation}
+in much the same way as environment variables, which are also implicitly inherited.
+As with environment variables,
+implicitly inherited file descriptors allow a resource to be passed down a process hierarchy
+without intervening programs being aware;
+this is a form of dynamic scope, which, though notorious, is often useful.
+
+Fortunately, file descriptors which should be implicitly inherited are marked:
+they have \texttt{CLOEXEC} turned off.
+For implicit inheritance to function, \texttt{CLOEXEC} must not be set;
+otherwise, the file descriptor will be closed, rather than inherited,
+the next time the process calls \texttt{execve}.
+And since \texttt{execve} can be called at any time in a general purpose program,
+any library that does not want its file descriptors to be inherited across \texttt{execve}
+must keep \texttt{CLOEXEC} turned on at all times.
+So, at any time, we know that the file descriptors with \texttt{CLOEXEC} off
+are exactly the file descriptors which should be implicitly inherited.
+
+Therefore, we simply need to close all file descriptors that have \texttt{CLOEXEC} set,
+and which are not referenced by a live file descriptor handle;
+we refer to this operation as \verb|close_cloexec|.
+
+For \verb|unshare(CLONE_FILES)|, we perform \verb|close_cloexec| automatically in userspace.
+For \texttt{clone}, we perform \verb|close_cloexec| closing operation only on user request,
+so that the user can control when \verb|inherit_fd| stops working.
+Even if the user never requests that unused file descriptors be closed,
+as long as they call \texttt{execve}, the table will be cleaned at that point.
+In both of these cases,
+it would be beneficial for efficiency to delegate \verb|close_cloexec|
+to a new system call which takes a list of file descriptors
+and closes all \texttt{CLOEXEC} file descriptors except for those on the list.
+
+For \texttt{execve}, this oepration is, in some sense, done automatically for us by the kernel;
+except we don't get the chance to preserve live file descriptors handles,
+without first unsetting \texttt{CLOEXEC} on them.
+Again an extension to the kernel would be useful:
+A new argument to \texttt{execve} which allows specifying a list of file descriptor numbers to inherit
+despite the \texttt{CLOEXEC} flag.
+This would allow programs to avoid the pattern
+of unsetting \texttt{CLOEXEC}, calling \texttt{execve}, and then immediately re-setting \texttt{CLOEXEC},
+just to inherit a file descriptor across a single exec.
+This might seem minor,
+but unsetting \texttt{CLOEXEC} before \texttt{execve} is not possible
+when the process calling \texttt{execve} has a shared file descriptor table,
+because some other process might call \texttt{execve} as well,
+and erroneously inherit the file descriptor.
+\section{Evaluation \textbf{TODO}}\label{evaluation}
+
 We quantitatively answer the following questions:
-
-    - Does direct-style process creation on Linux allow for simpler programs than the alternatives?
-      We implement and compare similar functionality using both direct-style and fork-style.\ref{ease}
-
-    - Does a clone-based implementation avoid the well-known performance overheads of fork?
-      We evaluate clone against fork on several microbenchmarks.\ref{microbench}
-
-    - Does a high-level Python direct-style process creation interface
+\begin{itemize}
+\item
+  Does direct-style process creation on Linux allow for simpler programs than the alternatives?
+  We implement and compare similar functionality using both direct-style and fork-style in \ref{ease}.
+\item
+  Does a clone-based implementation avoid the well-known performance overheads of fork?
+  We evaluate clone against fork on several microbenchmarks in \ref{microbench}.
+\item
+  Does a high-level Python direct-style process creation interface
   have acceptable performance relative to the Python standard library's spawn-style interface?
-      We benchmark direct-style process creation against spawn-style.\ref{subprocess_run}
+  We benchmark direct-style process creation against spawn-style in \ref{subprocess_bench}.
+% TODO have an item about use at TS? hm.
+% could be qualitative
+\end{itemize}
 \subsection{Ease of programming with direct-style}\label{ease}
-Implementing Python's popen/subprocess with rsyscall?
-But that's not fair, theirs is portable, we'll be way simpler for free.
-
-Rewrite some existing system with rsyscall?
-Write a shell?
+\begin{table}
+\input{ease.tex}
+\caption{Line counts with direct-style vs fork-style}
+\label{tab:ease}
+\end{table}
+We implement a variety of simple but powerful techniques with processes
+using both direct-style and fork.
+Our results are summarized in table \ref{tab:ease}.
+Direct-style consistently takes under half the lines the express the same thing.
+\lstinputlisting[float=*,language=Python,label={basic_fork},caption={Fork-style: Creating a new process, changing CWD, and execing}]
+{examples/basic/fork.py}
+An example of one of the fork-style implementations is in listing \ref{basic_fork}.
+It corresponds to the direct-style example in listing \ref{basic}.
+
+In our fork-style implementations,
+we assume a great deal of infrastructure is available to make fork-style simpler.
+We assume the existence of a robust IPC system,
+with an already-defined protocol which can cover all errors from all system calls.
+Despite this, fork-style implementations are still significantly more verbose than direct-style.
 \subsection{Clone versus fork}\label{microbench}
-Can mmap a bunch of stuff in a process and then create some children,
-and show that without having to copy page tables (due to \verb|CLONE_VM|), it's faster.
-
-Can make the classic chart of fork vs glibc's \verb|posix_spawn| vs direct-style.
-\subsection{High-level direct-style versus spawn-style}\label{subprocess_run}
-Compare clone/exec with rsyscall against Python's subprocess.run.
+\begin{figure}[h!]
+\centering
+ \includegraphics[width=0.5\textwidth]{microbench}
+ \caption{fork and clone performance under varying memory usage}
+ \label{fig:microbench}
+\end{figure}
+
+We run a benchmark which times how long process creation takes
+with \texttt{fork}, \texttt{clone}, and \verb|clone(CLONE_VM)|
+with varying amounts of anonymous memory mapped in the parent process
+before running the benchmark.
+\texttt{clone} without \verb|CLONE_VM| functions identically to fork.
+The results are in figure \ref{fig:microbench}.
+Without \verb|CLONE_VM|,
+process creation time is linear in the amount of memory mapped in the parent process.
+At the upper end of our benchmark, with 4 GB of memory mapped ($10^6$ pages),
+\texttt{fork} takes an average of 42.6 milliseconds to create a process,
+while \verb|clone(CLONE_VM)| stays flat at 21.1 microseconds.
+
+This is because \texttt{fork}, and \texttt{clone} without \verb|CLONE_VM|,
+must set up memory mappings in the child process for all the memory mapped in the parent process.
+\verb|clone(CLONE_VM)| does not have to do this bookkeeping,
+so its process creation time is constant in the amount of memory used in the parent process.
+
+\subsection{High-level direct-style versus spawn-style}\label{subprocess_bench}
+\begin{figure}[h!]
+\centering
+ \includegraphics[width=0.5\textwidth]{subprocess_bench}
+ \caption{Python subprocess and direct-style clone performance under varying memory usage}
+ \label{fig:subprocess_bench}
+\end{figure}
+
+We benchmark Python's \texttt{subprocess.run} against direct-style \texttt{clone} with \texttt{rsyscall}
+with varying memory usage, as in section \ref{microbench}.
+The results are in Figure \ref{fig:subprocess_bench}.
+TODO generate again to get the higher end where it crosses over.
 \section{Related work}\label{related_work}
@@
 \subsection{Remote system calls}
-There are a large number of Unix-based systems which do something called "remote system calls".
+There are many Unix-based systems which do something called "remote system calls".
 Most such systems do not allow a single program to manipulate multiple processes.

### 2020-05-15 research/paper/paper.tex
paper

@@
 With a fork-style interface,
-a process calls \texttt{fork}, and a copy of that process is made,
-with the return value of fork being the only differentiation between the process and its copy.
-The same program continues executing in both processes.
-Typically, the program will case on the return value to determine if it's running in the parent or the child,
-and run process setup and initialization code if the program is running in the child,
-calling various syscalls to set up the state of the child.
+a process, called the parent process, calls \texttt{fork},
+and a copy of that process is made,
+called the child process,
+differing only in a few attributes such as process id.
+The same program continues executing in both processes,
+with a different return value from fork to allow it to distinguish whether it is running in the parent or the child.
+Typically, the program will case on the return value
+and, if running in the child,
+call various system calls to modify the state of the new process.
 
@@
 In direct-style, a process is created by its parent in a non-running state,
-and then supplied with various resources,
+then supplied with various resources,
 and then started running once it is fully set up.
@@
 the syscalls that can mutate a process
-take explicit arguments to indicate which process they should operate on.
+take explicit arguments to indicate the process on which they should operate.
 In this way, the same syscalls that can mutate a process while it is running,

### 2020-05-15 research/paper/paper.tex
paper

@@
 and consider their advantages and disadvantages in more depth.
-In section \ref{overview} we discuss the design of the API,
+In section \ref{overview}, we discuss the design of the API,
 including our usage of \texttt{rsyscall}.
-In section \ref{applications} we show examples of usage of direct-style process creation in Python on Linux,
+In section \ref{applications}, we show examples of usage of direct-style process creation in Python on Linux,
 including use of process features which are difficult to use on Linux with \texttt{fork} or \texttt{posix\_spawn}.
@@
 Such a technique can also be used when a full sandbox is not relevant,
-to ease reasoning about the behavior of the program being run,
+to ease reasoning about the behavior of the program being run
 and protect against bugs.
@@
 \texttt{init} in a pid namespace has some special powers and responsibilities.
-We can handle these responsibilities ourselves, directly from this program,
+We can handle these responsibilities ourselves directly from our program,
 or we can continue on by execing an init daemon in \texttt{init} to handle it for us.

### 2020-05-15 research/paper/paper.tex
paper

@@
 over Linux's fork-style \texttt{fork} syscall and spawn-style \texttt{posix\_spawn} libc function.
+\begin{itemize}
+\item
 Direct-style \texttt{clone} can be used correctly and performantly from any process.
+\item
 Cross-process syscalls made with rsyscall individually report errors, as normal, and can be easily branched on.
+\item
 Arbitrary processes involving complex namespaces and initialization can be created
 using techniques that are impractical with \texttt{fork}.
+\item
 The performance cost from high-level code varies, but the underlying primitives are substantially faster.
+\end{itemize}
 
@@
 we list several directions for future work related to direct-style process creation and to \texttt{rsyscall}.
+% TODO
 In section \ref{conclusions}, we briefly and optimistically conclude.
@@
 \section{Background}\label{background}
-Processes are a widespread feature of operating systems,
-with substantial variation in characteristics between systems.
-One of the areas of variation is in process creation mechanisms.
-
 On most systems today,
@@
 take explicit arguments to indicate the process on which they should operate.
-In this way, the same syscalls that can mutate a process while it is running,
+In this way, the same syscalls that can mutate a process while it is running
 are called by the parent process to mutate the process while it is being set up.
@@
 In this section,
-we'll discuss several more advanced uses of processes,
+we'll discuss several more advanced uses of processes
 using more sophisticated Linux features,
@@
 In listing \ref{fuse},
-we first create a child process, \verb|ns_child|, in a new mount namespace with \verb|CLONE_NEWNS|,
-and for the first time, also pass \verb|CLONE_FILES|,
+we first create a child process \verb|ns_child| in a new mount namespace with \verb|CLONE_NEWNS|,
+and, for the first time, also pass \verb|CLONE_FILES|,
 which causes the file descriptor table to be shared between the parent process and the child process.
-We create a grandchild off of \verb|ns_child|,
+We create a grandchild off of \verb|ns_child|
 and use it to exec a FUSE filesystem,
@@
 they are also visible in the other process.
-This is simple to model, and convenient for many purposes.
+This is simple to model and convenient for many purposes.
 The file descriptor table is unshared once the child calls \texttt{execve}.

### 2020-05-16 research/paper/paper.tex
paper

@@
 
-Fork-style (such as \texttt{fork}) and spawn-style (such as \texttt{posix\_spawn}) each have advantages and disadvantages.
-Correct and performant usage of fork-style places significant constraints on the calling process.
-Spawn-style lifts many of these constraints,
+Fork-style (such as \texttt{fork}) and spawn-style (such as \texttt{posix\_spawn}) are the most popular,
+but have a number of disadvantages.
+Correct, performant use of fork-style places significant constraints on the calling process,
+and also has a complex programming model.
+Spawn-style lifts many of these constraints, and is simpler to program against,
 but is fundamentally limited in the kinds of processes it can create.
@@
 and robust indication of errors in spawn-style requires substantial growth in the spawn interface.
-Additionally, modern operating systems such as Linux have features
-which are difficult or impossible to fully exploit through fork-style or spawn-style.
-
-Direct-style is rare in general, but relatively common on academic operating systems.
-\cite{keykos}\cite{sel4}\cite{exokernel}\cite{fuschia}\cite{singularity}
+
+Direct-style is more rarely seen, but has a number of advantages.
 Direct-style involves creating a new child process in a non-running state,
-and then calling individual syscalls from the parent to initialize the child.
+and then calling individual system calls from the parent to initialize the child.
 Direct-style does not constrain the calling process,
@@
 Our contribution is an implementation of direct-style process creation for Linux,
-making use of our work on the rsyscall project.
-The rsyscall project develops language-specific ``process-independent'' syscall libraries for Linux,
-which replace libc syscall wrappers and allow a program to perform cross-process syscalls.
+making use of our work on the \texttt{rsyscall} project.
+The \texttt{rsyscall} project develops language-specific ``process-independent'' system call libraries for Linux,
+which replace libc system call wrappers, and allow a program to perform cross-process system calls.
 rsyscall currently has best support for Python,
@@
 
-% TODO this sounds low-content and boring.
-% we should probably remove it?
-% In our implementation, a parent process calls the \texttt{clone} syscall
-% to create a child process which shares memory and other resources with the parent,
-% and which runs no user code.
-% The parent then uses \texttt{rsyscall}
-% to call arbitrary syscalls in the child process to initialize it;
-% for example, the parent might call \texttt{chdir} to change the child's current working directory.
-% After the process is fully initialized,
-% the parent can call \texttt{execve} in the child,
-% causing the process to start running an arbitrary executable
-% and detaching the child from the control of \texttt{rsyscall}.
-
 Our direct-style process creation implementation uses \texttt{clone} to create a child process in an embryonic state
@@
 A direct-style \texttt{clone} has all the advantages of direct-style process creation
-over Linux's fork-style \texttt{fork} syscall and spawn-style \texttt{posix\_spawn} libc function.
+over Linux's fork-style \texttt{fork} and spawn-style \texttt{posix\_spawn}:
 \begin{itemize}
 \item
-Direct-style \texttt{clone} can be used correctly and performantly from any process.
+  Processes can be created correctly and performantly from any parent process,
+  with a straightforward, familiar API.
 \item
-Cross-process syscalls made with rsyscall individually report errors, as normal, and can be easily branched on.
+  Processes can be created with the full set of Linux features.
 \item
-Arbitrary processes involving complex namespaces and initialization can be created
-using techniques that are impractical with \texttt{fork}.
-\item
-The performance cost from high-level code varies, but the underlying primitives are substantially faster.
+  Syscall results are reported individually to the parent process,
+  rather than batched together as with spawn-style or manually reported via IPC as with fork-style.
 \end{itemize}
@@
 We've implemented a variety of useful applications using direct-style process creation on Linux,
-which can be expressed more concisely and more clearly than with fork-style,
 which we'll showcase in brief in this paper.
@@
 and can outperform \texttt{fork} in some cases.
-We've benchmarked our Python implementation of direct-style \texttt{clone}
-as comparable to the Python standard library's process creation mechanism.
-Finally, we have deployed direct-style process creation in production internally at Two Sigma.
+We have deployed direct-style process creation in production internally at Two Sigma.
 
@@
 and is immediately deployable on recent Linux kernels.
-rsyscall is open source and available from \url{https://github.com/catern/rsyscall}.
+\texttt{rsyscall} is open source and available from \url{https://github.com/catern/rsyscall}.
 
 This paper is organized in several sections.
-In section \ref{background}, we provide background on fork-style, spawn-style, and direct-style process creation,
+\begin{itemize}
+\item In section \ref{background}, we provide background on fork-style, spawn-style, and direct-style process creation,
 and consider their advantages and disadvantages in more depth.
-In section \ref{overview}, we discuss the design of the API,
+\item In section \ref{overview}, we discuss the design of the API,
 including our usage of \texttt{rsyscall}.
-In section \ref{applications}, we show examples of usage of direct-style process creation in Python on Linux,
+\item In section \ref{applications}, we show examples of usage of direct-style process creation in Python on Linux,
 including use of process features which are difficult to use on Linux with \texttt{fork} or \texttt{posix\_spawn}.
-In section \ref{implementation}, we explore our implementation in depth,
+\item In section \ref{implementation}, we explore our implementation in depth,
 and discuss several difficult process-creation related details of \texttt{rsyscall}'s implementation.
-In section \ref{evaluation}, we evaluate, somehow (TODO).
-In section \ref{related_work},
+\item In section \ref{evaluation}, we evaluate, somehow (TODO).
+\item In section \ref{related_work},
 we examine other work which also touches on the topics covered by this paper.
-In section \ref{future_work},
+\item In section \ref{future_work},
 we list several directions for future work related to direct-style process creation and to \texttt{rsyscall}.
-% TODO
-In section \ref{conclusions}, we briefly and optimistically conclude.
-
+\item In section \ref{conclusions}, we briefly outline our conclusions about direct-style and the future of process creation.
+\end{itemize}
 \section{Background}\label{background}
-On most systems today,
-the interfaces for process creation
-can be divided into "fork-style" and "spawn-style".
-A few systems use a third style, which we refer to as "direct-style".
+As mentioned in the introduction,
+on most operating systems today,
+the available interfaces for process creation
+are either "fork-style" or "spawn-style".
+A few operating systems have a third style, which we refer to as "direct-style".
 \subsection{Fork-style}
@@
 the Unix shell also serves as a spawn-style process creation interface,
-especially when used with Bernstein chaining.\cite{chainloading}
+especially when used with Bernstein chaining\cite{chainloading}.
 
@@
 Likewise, there are no inherent issues when a spawn-style interface is used from a multi-threaded program,
-though operating systems which implement spawn-style interfaces on top of \texttt{fork} can still have threading bugs.
+though operating systems, such as Linux,
+which implement spawn-style interfaces on top of \texttt{fork} can still have threading bugs.
 
@@
 This is the kind of direct-style interface that we contribute in this paper.
-
 \section{Overview}\label{overview}
@@
 To perform direct-style process creation on Linux,
-we need a Linux API where we instead explicitly specify in which process we want to make a syscall.
-
-\texttt{rsyscall} provides this API for Linux.
+we need an API where we instead explicitly specify in which process we want to make a syscall.
+
+\texttt{rsyscall} provides this API.
 The \texttt{rsyscall} project develops
@@
 Unlike POSIX C libraries such as glibc or musl,
-an rsyscall library is organized based on the object-capability model.
+an \texttt{rsyscall} library is organized based on the object-capability model.
 The capabilities to make syscalls in any given process are reified as language objects.
@@
 The ``local'' process is the one which hosts the runtime for the running program,
-and in which a legacy libc would implicitly make syscalls;
-every rsyscall program starts with the ``local'' process
+and in which a legacy libc would implicitly make syscalls.
+Every \texttt{rsyscall} program starts with the ``local'' process
 and uses it to bootstrap the initial capabilities for other processes.
 
-We use the term ``thread'' to refer to the processes under the control of rsyscall,
+We use the term ``thread'' to refer to the processes under the control of \texttt{rsyscall},
 including the ``local'' process.
@@
 and their lifetime and execution is completely controlled by a single program.
-The same is mostly true of the controlled-process ``threads'' provided by rsyscall.
-The most significant difference is that rsyscall ``threads''
+The same is mostly true of the controlled-process ``threads'' provided by \texttt{rsyscall}.
+The most significant difference is that \texttt{rsyscall} ``threads''
 do not run user code concurrently with the main program;
-nevertheless rsyscall ``threads'' do provide the opportunity for concurrent execution of system calls,
+nevertheless \texttt{rsyscall} ``threads'' do provide the opportunity for concurrent execution of system calls,
 and so the terminology provides useful intuition.
@@
 \section{Implementation}\label{implementation}
-\subsection{Basics about rsyscall}
-In this section, we'll give a brief overview of the implementation of rsyscall,
+\subsection{Basics about \texttt{rsyscall}}
+In this section, we'll give a brief overview of the implementation of \texttt{rsyscall},
 and then focus on implementation issues specific to process creation.
 
-rsyscall can be conceptually divided in two parts:
+\texttt{rsyscall} can be conceptually divided in two parts:
 the basic cross-process syscall primitive,
@@
 a syscall returns one register-sized value.
-rsyscall's default implementation of cross-process syscalls sends those seven integers over a pipe,
+\texttt{rsyscall}'s default implementation of cross-process syscalls sends those seven integers over a pipe,
 and waits for a response on another pipe.
@@
 However, among other issues, it has the unavoidable substantial disadvantage of not permitting multiple ptracers.
-A ptrace-based implementation would prevent using strace or gdb on rsyscall-controlled processes,
+A ptrace-based implementation would prevent using strace or gdb on \texttt{rsyscall}-controlled processes,
 which is an unacceptable limitation for a general-purpose utility.
@@
 and require the caller to read or write that memory to provide arguments or receive results.
-Therefore, an rsyscall library needs a way to access memory in the target process.
+Therefore, an \texttt{rsyscall} library needs a way to access memory in the target process.
 
@@
 which is a niche but possible system configuration.
-To ensure that rsyscall can be used for arbitrary purposes and on arbitrary systems, we avoided these calls.
+To ensure that \texttt{rsyscall} can be used for arbitrary purposes and on arbitrary systems, we avoided these calls.
 \subsection{\texttt{clone}}\label{clone}
-Now that we've established the basic implementation details of rsyscall,
+Now that we've established the basic implementation details of \texttt{rsyscall},
 we'll consider specific issues related to process creation and initialization.
@@
 with its registers in generally the same state as the parent process.
-\footnote{Note that =glibc= defines a wrapper for the clone syscall;
+\footnote{Note that glibc defines a wrapper for the clone syscall;
 we are talking about the raw kernel syscall.}
@@
 
-We typically cause the new process to jump to a trampoline provided by the rsyscall library
+We typically cause the new process to jump to a trampoline provided by the \texttt{rsyscall} library
 which sets all registers to values found on the stack
@@
 \subsection{Process capabilities}
-We mentioned the concept of process capabilities while discussing the design of rsyscall in section \ref{overview}.
+We mentioned the concept of process capabilities while discussing the design of \texttt{rsyscall} in section \ref{overview}.
 The Capsicum system provides something called process descriptors\cite{capsicum},
@@
 \section{Future work}\label{future_work}
-\subsection{Other applications of rsyscall}
-Most avenues of future work focus on rsyscall.
-rsyscall was not developed solely for the purpose of this paper,
+\subsection{Other applications of \texttt{rsyscall}}
+Most avenues of future work focus on \texttt{rsyscall}.
+\texttt{rsyscall} was not developed solely for the purpose of this paper,
 and it has many uses unrelated to direct-style process creation,
@@
 We are actively exploring such applications,
-as well as broadening rsyscall's language support.
+as well as broadening \texttt{rsyscall}'s language support.
 \subsection{Kernel support}
-rsyscall's cross-process syscalls can be performed entirely in userspace,
+\texttt{rsyscall}'s cross-process syscalls can be performed entirely in userspace,
 which has substantial benefits for deployability.
@@
 to implement them.
-We hope that this work will help encourage more use of the process abstraction,
+
+We believe that direct-style process creation is superior to fork-style and spawn-style,
+including on Linux.
+Our technique is immediately deployable and can be used for a wide variety of complex applications.
+
+We hope that a better process creation mechanism
+will help encourage more use of the process abstraction,
 which, though widespread,

### 2020-05-16 research/paper/paper.tex
paper: remove mention of thread

@@
 
-We use the term ``thread'' to refer to the processes under the control of \texttt{rsyscall},
-including the ``local'' process.
-On Linux, the shared-memory ``threads'' provided by libraries such as \texttt{pthreads}
-are implemented as processes,
-and their lifetime and execution is completely controlled by a single program.
-The same is mostly true of the controlled-process ``threads'' provided by \texttt{rsyscall}.
-The most significant difference is that \texttt{rsyscall} ``threads''
-do not run user code concurrently with the main program;
-nevertheless \texttt{rsyscall} ``threads'' do provide the opportunity for concurrent execution of system calls,
-and so the terminology provides useful intuition.
-
 To perform direct-style process creation,
-we use \texttt{rsyscall} to call \texttt{clone} on the ``local'' thread to create a new child thread.
-We call various syscalls in the child thread to mutate it until it reaches the desired state,
-at which point we can call \texttt{execve} in the child thread to start it running on its own.
-After that, the process is no longer under our control, and it functions like any other child process
-(hence we say ``process'' rather than ``thread'').
-The child process can be monitored using normal Linux child monitoring syscalls from our main thread,
-such as \texttt{waitid},
-which we can also do with \texttt{rsyscall}.
+we use \texttt{rsyscall} to call \texttt{clone} on the ``local'' process to create a new child process.
+We call various syscalls in the child process to mutate it until it reaches the desired state,
+at which point we can call \texttt{execve} in the child process to start it running on its own.
+After that, the process is no longer under our control, and it functions like any other child process.
+The child process can be monitored using normal Linux child monitoring syscalls,
+such as \texttt{waitid}.
 
@@
 Listing \ref{basic} shows this in action in Python.
-We create a new child thread using \texttt{clone} on \verb|local_thread|,
+We create a new child process using \texttt{clone} on the local process \verb|local|,
 and receive back a capability to control it.
-We pass no flags to use \texttt{clone}'s defaults
-(which essentially match \texttt{fork}) to create the process.
-We call \texttt{chdir} in the thread to change its working directory to \texttt{/dev}.
+We call \texttt{chdir} in the child process to change its working directory to \texttt{/dev}.
 Then we call \texttt{execve},
@@
 Calling \texttt{execve} consumes the capability,
-and later use of this child thread capability will fail with an exception.
+and later use of this child process capability will fail with an exception.
 
@@
 we first open a file with read-write permission.
-Then we create a child thread,
+Then we create a child process,
 which inherits all file descriptors from its parent.
@@
 In listing \ref{mount},
-we create a new child thread as normal,
+we create a new child process as normal,
 but for the first time,
@@
 In listing \ref{unmount}
-we create a new child thread,
+we create a new child process,
 again in a new mount namespace using \texttt{CLONE.NEWNS}.
-We open the executable that we'll later exec in this child thread,
+We open the executable that we'll later exec in this child process,
 as well as another file as in listing \ref{fds},
@@
 we clone again, this time from \texttt{init}.
-This is the first example in which we've cloned from one of our child threads;
-as normal, this produces a new process, a grandchild,
-which can be used exactly like a direct child thread.
+This is the first example in which we've cloned from one of our child processes;
+as normal, this gives us a capability for a new process, a grandchild,
+which can be used exactly like a direct child process.
 We exec in the grandchild,
@@
 
-The most simple way to access memory is for the main thread to be in the same address space as the target process.
-This is the case most of the time; we pass \texttt{CLONE.VM} by default in the Python wrapper.
+The most simple way to access memory is for the local process to be in the same address space as the target process.
+This is the case most of the time; we pass \texttt{CLONE.VM} to \texttt{clone} by default in the Python wrapper.
 However, the target process may be in another address.

### 2020-05-16 research/paper/ease.py
examples: add
We use O_RDWR instead of O.RDWR (and so on) because that's more
familiar for people.

@@ -1,0 +1,41 @@
+from pathlib import Path
+import typing as t
+
+dirname = Path(__file__).parent
+examples = dirname/"examples"
+
+def extract_label(line: str) -> str:
+    _, rest = line.split('label={')
+    label, _ = rest.split("},caption")
+    return label
+
+listing_labels = [extract_label(line) for line in open(dirname/"paper.tex") if line.startswith("\\lstinputlisting")]
+
+headings = (
+    "Name",
+    "Listing",
+    "Direct-style",
+    "Fork-style",
+)
+rows = []
+for label in listing_labels:
+    dir = examples/label
+    direct_wc = len(list((dir/"direct.py").open()))
+    fork_path = dir/"fork.py"
+    fork_wc = len(list(fork_path.open())) if fork_path.exists() else "n/a"
+    rows.append((
+        label,
+        "\\ref{" + label + "}",
+        direct_wc,
+        fork_wc,
+    ))
+def line(row: t.List) -> str:
+    return " & ".join(str(x) for x in row) + " \\\\"
+print('\\begin{tabular}{' + ''.join('r' for _ in rows[0]) + '}')
+print('\\hline')
+print(line(headings))
+print('\\hline')
+for row in rows:
+    print(line(row))
+print('\\hline')
+print('\\end{tabular}')

### 2020-05-16 research/paper/examples/basic/direct.py

@@ -1,0 +1,3 @@
+child = local.clone()
+child.chdir("/dev")
+child.execv("/bin/cat", ["cat", "./null"])

### 2020-05-16 research/paper/examples/basic/fork.py

@@ -1,0 +1,14 @@
+pid = os.fork()
+if pid == 0:
+    try:
+        os.chdir("/dev")
+        os.execv("/bin/cat", ["cat", "./null"])
+    except OSError as e:
+        ipc.send(e)
+        os.exit(1)
+else:
+    result = ipc.recv()
+    if result.is_eof:
+        pass # success
+    elif result.is_exception:
+        raise result.exception

### 2020-05-16 research/paper/examples/fds/direct.py

@@ -1,0 +1,6 @@
+db_fd = local.open("/var/db/database.db", O_RDWR)
+child = local.clone()
+child_fd = child.inherit_fd(db_fd)
+child_fd.fcntl(F_SETFD, 0)
+child.execv("/bin/fooserver",
+  ["fooserver", "--database-fd", str(int(child_fd))])

### 2020-05-16 research/paper/examples/fds/fork.py

@@ -1,0 +1,16 @@
+db_fd = local.open("/var/db/database.db", O_RDWR)
+pid = os.fork()
+if pid == 0:
+    try:
+        fcntl(db_fd, F_SETFD, 0)
+        os.execv("/bin/fooserver",
+                 ["fooserver", "--database-fd", str(int(db_fd))])
+    except OSError as e:
+        await ipc.send(e)
+        os.exit(1)
+else:
+    result = await ipc.recv()
+    if result.is_eof:
+        print("successfully exec'd")
+    elif result.is_exception:
+        raise result.exception

### 2020-05-16 research/paper/examples/fuse/direct.py

@@ -1,0 +1,6 @@
+ns_child = local.clone(CLONE_FILES|CLONE_NEWNS)
+server_child = ns_child.clone()
+server_child.execve('/bin/foofs',
+  ['foofs', "--mount-at", "/"])
+fd = ns_child.open("/foo/bar", O_RDONLY)
+parent_fd = local.use_fd(fd)

### 2020-05-16 research/paper/examples/fuse/fork.py

@@ -1,0 +1,27 @@
+init_pid = clone(CLONE_NEWPID)
+if init_pid == 0:
+    try:
+        grandchild_pid = clone()
+        if grandchild_pid == 0:
+            try:
+                execve('/bin/foofs',
+                       ['foofs', "--mount-at", "/"])
+            except OSError as e:
+                await ipc.send(e)
+                os.exit(1)
+        else:
+            pass
+    except OSError as e:
+        await ipc.send(e)
+        os.exit(1)
+    ipc.send("success")
+    fd = open("/foo/bar", O_RDONLY)
+    ipc.send_fd(fd) # use a Unix socket
+else:
+    result = await ipc.recv()
+    if result.is_eof:
+        pass # failure
+    elif result.is_exception:
+        raise result.exception
+    elif result.value == "success":
+        parent_fd = ipc.recv_fd()

### 2020-05-16 research/paper/examples/mount/direct.py

@@ -1,0 +1,4 @@
+child = local.clone(CLONE_NEWNS)
+child.mount("/home/foo/custom_foo.conf",
+  "/etc/foo.conf", "", MS_BIND, "")
+child.execv('/bin/fooserver', ['fooserver'])

### 2020-05-16 research/paper/examples/mount/fork.py

@@ -1,0 +1,15 @@
+pid = clone(CLONE_NEWNS)
+if pid == 0:
+    try:
+        mount("/home/foo/custom_foo.conf",
+              "/etc/foo.conf", "", MS_BIND, "")
+        child.execv('/bin/fooserver', ['fooserver'])
+    except OSError as e:
+        await ipc.send(e)
+        os.exit(1)
+else:
+    result = await ipc.recv()
+    if result.is_eof:
+        print("successfully exec'd")
+    elif result.is_exception:
+        raise result.exception

### 2020-05-16 research/paper/examples/pidns/direct.py

@@ -1,0 +1,3 @@
+init = local.clone(CLONE_NEWPID)
+grandchild = init.clone()
+grandchild.execv('/bin/fooserver', ['fooserver'])

### 2020-05-16 research/paper/examples/pidns/fork.py

@@ -1,0 +1,23 @@
+init_pid = clone(CLONE_NEWPID)
+if init_pid == 0:
+    try:
+        grandchild_pid = clone()
+        if grandchild_pid == 0:
+            try:
+                os.execv('/bin/fooserver', ['fooserver'])
+            except OSError as e:
+                await ipc.send(e)
+                os.exit(1)
+        else:
+            pass
+    except OSError as e:
+        await ipc.send(e)
+        os.exit(1)
+else:
+    result = await ipc.recv()
+    if result.is_eof:
+        pass # failure
+    elif result.is_exception:
+        raise result.exception
+    elif result.value == "success":
+        pass # success

### 2020-05-16 research/paper/examples/pipe/direct.py

@@ -1,0 +1,17 @@
+def argfd(child, fd):
+  child_fd = child.inherit_fd(fd)
+  child_fd.fcntl(F_SETFD, 0) # unset CLOEXEC
+  return str(int(child_fd))
+
+audio_pipe = local.pipe()
+video_pipe = local.pipe()
+source = local.clone()
+source.execv('/bin/source', ['source',
+  '--audio-out', argfd(source, audio_pipe.write),
+  '--video-out', argfd(source, video_pipe.write)])
+video_sink = local.clone()
+video_sink.execv('/bin/video_sink', ['video_sink',
+  '--video-in', argfd(video_sink, video_pipe.read)])
+audio_sink = local.clone()
+audio_sink.execv('/bin/audio_sink', ['audio_sink',
+  '--audio-in', argfd(audio_sink, audio_pipe.read)])

### 2020-05-16 research/paper/examples/pipe/fork.py

@@ -1,0 +1,49 @@
+def as_argument(fd):
+    fcntl(fd, F_SETFD, 0) # unset CLOEXEC
+    return str(int(fd)) # turn fd number to string
+
+audio_pipe = local.pipe()
+video_pipe = local.pipe()
+source_pid = os.fork()
+if source_pid == 0:
+    try:
+        os.execv('/bin/source', ['source',
+          '--audio-output', as_argument(audio_pipe.write),
+          '--video-output', as_argument(video_pipe.write)])
+    except OSError as e:
+        await ipc.send(e)
+        os.exit(1)
+else:
+    result = await ipc.recv()
+    if result.is_eof:
+        print("successfully exec'd")
+    elif result.is_exception:
+        raise result.exception
+video_sink_pid = os.fork()
+if video_sink_pid == 0:
+    try:
+        os.execv('/bin/video_sink', ['video_sink',
+          '--video-input', as_argument(video_sink, video_pipe.read)])
+    except OSError as e:
+        await ipc.send(e)
+        os.exit(1)
+else:
+    result = await ipc.recv()
+    if result.is_eof:
+        print("successfully exec'd")
+    elif result.is_exception:
+        raise result.exception
+audio_sink_pid = os.fork()
+if audio_sink_pid == 0:
+    try:
+        os.execv('/bin/audio_sink', ['audio_sink',
+          '--audio-input', as_argument(audio_sink, audio_pipe.read)])
+    except OSError as e:
+        await ipc.send(e)
+        os.exit(1)
+else:
+    result = await ipc.recv()
+    if result.is_eof:
+        print("successfully exec'd")
+    elif result.is_exception:
+        raise result.exception

### 2020-05-16 research/paper/examples/unmount/direct.py

@@ -1,0 +1,7 @@
+child = local.clone(CLONE_NEWNS)
+exec_fd = child.open("/bin/foo_static", O_RDONLY)
+db_fd = child.open("/var/db/database.db", O_RDWR)
+child.umount("/", MNT_DETACH)
+db_fd.fcntl(F_SETFD, 0)
+child.fexec(exec,
+  ["foo_static", "--database-fd", str(int(db_fd))])

### 2020-05-16 research/paper/examples/unmount/fork.py

@@ -1,0 +1,18 @@
+pid = os.fork()
+if pid == 0:
+    try:
+        exec_fd = os.open("/bin/foo_static", O_RDONLY)
+        db_fd = os.open("/var/db/database.db", O_RDWR)
+        os.umount("/", MNT_DETACH)
+        fcntl(db_fd, F_SETFD, 0)
+        os.fexecve(exec_fd,
+          ["foo_static", "--database-fd", str(int(db_fd))])
+    except OSError as e:
+        await ipc.send(e)
+        os.exit(1)
+else:
+    result = await ipc.recv()
+    if result.is_eof:
+        print("successfully exec'd")
+    elif result.is_exception:
+        raise result.exception

### 2020-05-16 research/paper/paper.tex
paper stuff

@@
 and which is already thoroughly documented through Linux system call manpages.
-The novelty is constrained to the newly explicit process capabilities;
-once that is understood, the rest of the API is mundane,
-and the representation of these programs is obvious in any language.
-
-The familiar and mundane nature of the API is important not just for the learning curve,
-but also for expressivity.
-Since the API is simply normal Linux system calls,
-direct-style clone is at least as expressive as fork and syscalls;
-and, in fact, as we'll see in section \ref{applications},
-is significantly more expressive.
+The novelty is constrained to the newly explicit process capabilities.
+
+The lack of novelty in the API is important not only for the learning curve,
+but also for power.
+Since we are using conventional Linux system calls,
+just in a cross-process way,
+direct-style \texttt{clone} will always be at least as powerful as \texttt{fork}.
 \section{Applications}\label{applications}
@@
 and accompany them with Python examples using direct-style process creation.
-
-All of these examples use only existing Linux system calls and features.
-Most are possible, though difficult to implement,
-with Linux's existing \texttt{fork} system call,
-and in some cases, \verb|posix_spawn|.
-
-Direct-style makes these examples substantially easier to write;
-we'll compare our implementations in this section to fork-style implementations in section \ref{evaluation}.
-Furthermore, since the examples are using the Linux system call API directly,
-they can be translated straightforwardly into any language with an \texttt{rsyscall} API;
-Python is chosen simply for convenience.
-
-Some of these examples showcase functionality that can be done today by specialized software and services.
+All of these examples use only conventional Linux system calls,
+and can therefore translated straightforwardly into any language with an \texttt{rsyscall} API.
+For concision, we assume that we are running with sufficient permissions in these examples,
+but these examples can all be run without privileges with appropriate use of \texttt{CLONE.NEWUSER}.
+
+Implementing these examples with \texttt{fork} or \verb|posix_spawn| requires substantially more boilerplate,
+or even a complete change in implementation approach,
+such as in section \ref{shared_fd_table}.
+We'll compare our direct-style implementations in this section
+to fork-style implementations
+in section \ref{evaluation}.
+
+Several of these examples demonstrate functionality that can be done today by specialized software and services.
 Such software often allows use of one or two of these techniques, but prevents the use of others.
@@
 
-By improving the process creation interface,
-all these techniques can be used together, flexibly,
-to more closely meet the specific needs of individual real-world applications.
-
-For concision, we assume that we are running with sufficient permissions in these examples,
-but these examples can all be run without privileges with appropriate use of \texttt{CLONE.NEWUSER}.
+One of the key advantages of direct-style process creation is that the techniques we demonstrate here
+can be easily used in combination,
+to meet the specific needs of individual real-world applications.
 \subsection{Abstraction through file descriptor passing}\label{fd_abstraction}
 In Linux (and Unix in general),
-a program can use file descriptor inheritance
-to provide file descriptors created by the parent process to child processes.
-Put simply,
-a new Linux child process receives copies of all file descriptors in the parent process.
-When calling \texttt{execve}, these file descriptors are normally closed automatically,
-but selected file descriptors can be preserved over \texttt{execve}
-by disabling the \texttt{CLOEXEC} flag on individual file descriptors.
-We will discuss file descriptor inheritance in more detail in section \label{fdtables},
-as part of our discussion of resource management in the implementation of \texttt{rsyscall}.
+a program can pass file descriptors from the parent process to a child process
+by using file descriptor inheritance.
+All open file descriptors are copied at process creation time from the parent process to the child process;
+the child process can then make independent use of the file descriptors.
+We'll discuss this important feature further in section \ref{fdtables}.
 
 Most resources in Linux are managed through file descriptors,
-so this allows the parent process to pass a wide variety of resources to the child process.
+so this feature allows the parent process to pass a wide variety of resources to the child process.
 An executable program can be provided a file descriptor,
@@
 which inherits all file descriptors from its parent.
-We indicate that we want to operate on the child's copy of the file descriptor with \verb|child.inherit_fd|.
-\verb|inherit_fd| only updates resource-tracking bookkeeping in our program,
-it makes no system calls;
-the file descriptor was already copied at process creation time.
+We indicate that we want to operate on the child's inherited copy of the file descriptor with \verb|child.inherit_fd|.
 Then we disable \texttt{CLOEXEC} on the child's file descriptor,
@@
 passing the file descriptor number as an argument;
-when the executable starts up, that file descriptor number will already be pointing to this file descriptor.
-We call the \texttt{execv} wrapper rather than \texttt{execve} as in listing \ref{basic}
-so that we don't need to pass the (unmodified) environment to every exec call in these examples.
+since we disabled \texttt{CLOEXEC} on the file descriptor,
+the executable will be able to use that file descriptor through that file descriptor number.
+Unlike in listing \ref{basic}, we call the \texttt{execv} wrapper rather than \texttt{execve},
+so that we don't need to pass the (unmodified) environment to every \texttt{exec} call in these examples.
 \subsection{Non-shared-memory concurrency}
@@
 \subsection{Customization without explicit support}
-While, ideally, all resources would be passed by file descriptor,\cite{capsicum}
+While, ideally, all resources would be passed by file descriptor\cite{capsicum},
 Unix has a number of resources that are global,
@@
 we can not only pass resources and customize the process's environment,
-we can also deny the process access to resources that it otherwise gets by default.
+we can also deny the process access to resources that it otherwise receives by default.
 This is a key part of creating a secure sandbox for potentially malicious code.
@@
 as we do here with the database file descriptor.
-This is the essence of capability-based security,
-which Linux can increasingly closely approximate.
+This allows us to approximate capability-based security\cite{capsicum}.
 \subsection{Nested clone and pid namespaces}
@@
 file descriptor passing over a Unix domain socket
-is the only option to do something like this.
+is our only option to do something like this.
 \section{Implementation}\label{implementation}
@@
 and waits for a response on another pipe.
-Processes are created running an infinite loop which, at each iteration,
+Processes are created running a loop which, at each iteration,
 reads a syscall request off the pipe,
@@
 including system calls.
-However, among other issues, it has the unavoidable substantial disadvantage of not permitting multiple ptracers.
-A ptrace-based implementation would prevent using strace or gdb on \texttt{rsyscall}-controlled processes,
+However, among other issues, it has the substantial disadvantage of not permitting multiple ptracers.
+A ptrace-based implementation would prevent using \texttt{strace} or \texttt{gdb} on \texttt{rsyscall}-controlled processes,
 which is an unacceptable limitation for a general-purpose utility.
@@
 We implement access to the memory of processes in different address spaces through another set of pipes,
-by explicitly copying memory into and out of those pipes using the \texttt{read} and \texttt{write} system calls.
-When we wish to read \texttt{N} bytes of memory at address \texttt{A} in the target process,
-we first perform a \verb|write(memory_pipe, A, N)| in the target process,
-and then read that data off the other end of the pipe in the parent process.
-When we wish to write \texttt{N} bytes of data at address \texttt{A} in the target process,
-we first write that data to the pipe in the parent process,
-then perform a \verb|read(memory_pipe, A, N)| in the target process to copy that data from the pipe into memory.
+by copying memory into and out of those pipes using \texttt{read} and \texttt{write} system calls in the target process.
+A \texttt{read} syscall corresponds to a write to memory,
+and a \texttt{write} syscall corresponds to a read from memory.
 
@@
 
-\texttt{clone} (along with \texttt{fork}) creates a new process
+\texttt{clone}
+(along with \texttt{fork}) creates a new process
 which immediately starts executing at the next instruction after the syscall instruction,
@@
 with its registers in generally the same state as the parent process.
-\footnote{Note that glibc defines a wrapper for the clone syscall;
-we are talking about the raw kernel syscall.}
-In the style of Plan 9's \texttt{fork} syscall\cite{rfork}, which inspired \texttt{clone},
-\texttt{clone} takes a mask of flags which determines whether several attributes of the new process
-are either shared with, or copied from, the parent process.
-
-\texttt{clone} only lets us change the stack for the new process;
-the new process begins executing, after \texttt{clone}, at the same instruction as the parent process.
-We would like to be able to set arbitrary registers for the new process,
-so that we can control where it begins executing and the stack it executes on.
-
-Fortunately, changing the stack is sufficient.
-We ensure that the next instruction executed after any syscall is a \texttt{RET}.
-Since we control the state of the stack in the new process after a \texttt{clone} syscall,
-the \texttt{RET} will jump to a code address that we control.
-We can then supply additional arguments to this code
-by putting them on the stack.
-
-We typically cause the new process to jump to a trampoline provided by the \texttt{rsyscall} library
-which sets all registers to values found on the stack
-and then jumps to another address.
-\footnote{This is also a generally useful utility for hackers performing return-oriented-programming attacks;
-but similar functionality exists in any standards-compliant C library,
-so there is no increase in attack surface.}
-With this trampoline,
-we can provide a helper Python function that,
-when given a function pointer following C calling conventions, and some arguments,
-will prepare a stack for a call to clone such that the new process will call that function with those arguments.
-
-With our new ability to call arbitrary C-compatible functions,
-we can now call \texttt{clone} so that it launches a process running our infinite syscall loop,
-which is implented in freestanding C and, as described in the previous section,
-uses two pipes passed as arguments to receive syscall requests and respond with syscall results.
-
-After using \texttt{clone} to create a new process running our syscall loop,
+
+Common Linux libcs provide a wrapper for \texttt{clone}
+which allow executing an arbitrary function with an arbitrary argument,
+but since we want to support remote execution,
+we can't use the libc wrapper.
+
+Instead, since \texttt{clone} lets us change the stack for the new process,
+we use this to return, after the system call to an arbitrary function address stored on the stack.
+Further arguments can also be passed on the stack,
+with the aid of a trampoline to match C calling conventions if necessary.
+
+We use this to create processes running the syscall request-response loop described in the previous section.
+After this,
 most system calls can be called as normal.
@@
 and then, crucially, performs a futex wakeup on.
-More specifically,
-the kernel clears and does a futex wakeup on \texttt{ctid} when the child process leaves its current address space;
-this precisely coincides with exiting or execing,
-since those are the only way to change address space in Linux as of this writing.
-
-A futex is a Linux-specific feature,
-which is generally used for the implementation of userspace shared-memory synchronization constructs,
-such as mutexes and condition variables.
-The relevant detail for us here is that we can wait on an address
-until a futex wakeup is performed on that address;
-that means we can wait on \texttt{ctid} until the futex wakeup is performed,
-and in this way get notified of the child process calling \texttt{execve}.
+
+Futexes are usually used for the implementation of userspace shared-memory synchronization constructs,
+but the relevant detail for us here is that we can wait on an address
+until a futex wakeup is performed on that address.
+This means we can wait on \texttt{ctid} until the futex wakeup is performed,
+and in this way get notified of the child process calling \texttt{execve} or exiting.
 
@@
 One approach for Linux would be to add a new \texttt{clone} flag to
-opt-in to receiving \texttt{WEXECED} events through \texttt{waitid};
-note that a \texttt{waitid} flag alone is not sufficient,
-since it's necessary to receive \texttt{SIGCHLD} signals for the \texttt{WEXECED} event if waiting for it from an event loop.
-
-Alternatively, some way to wait for futex wakeups through a file descriptor could be added,
-so we can use file-descriptor-monitoring syscalls to wait for the \texttt{ctid} futex;
-such a feature used to exist in the form of \verb|FUTEX_FD|,
-but was removed from Linux long ago due to race conditions in its design.
+opt-in to receiving \texttt{WEXECED} events through \texttt{waitid}.
+A new \texttt{waitid} flag alone is not sufficient,
+since, to integrate this feature into an event loop,
+it's necessary to receive \texttt{SIGCHLD} signals or readability notifications on a pidfd
+when the \texttt{WEXECED} event happens.
 \subsection{Managing file descriptor tables}\label{fdtables}
@@
 they are also visible in the other process.
+The file descriptor table is unshared once the child calls \texttt{execve}.
 This is simple to model and convenient for many purposes.
-The file descriptor table is unshared once the child calls \texttt{execve}.
 
@@
 are not copied into the newly created file descriptor table.
-In standard Unix practice, \texttt{CLOEXEC} is set habitually by users on all file descriptors;
-we'll discuss the semantics of \texttt{CLOEXEC} more in section \ref{cloexec}.
+We'll discuss the semantics of \texttt{CLOEXEC} more in section \ref{cloexec}.
 \subsubsection{Tracking file descriptors through handles}
@@
 
-Python, however, is a garbage collected language,
-and therefore we determine which file descriptors should stay open through garbage collection:
+Python, instead, has automatic resource management,
+and therefore we determine which file descriptors should stay open automatically:
 if a certain file descriptor is in use by the program,
@@
 by recourse to the native Python garbage collection system.
-We keep open exactly what file descriptors are necessary to keep the live handles valid.
+The file descriptors we keep open are then those which are necessary to keep the live handles valid.
 
@@
 When a process is newly created,
-no file descriptor handles yet exist;
+no file descriptor handles yet exist.
 New handles can be created by opening files in the process,
@@
 we must promptly close other file descriptors.
-Leaving unwanted file descriptors open in the new table is a form of resource leakage.
-It can also cause erroneous behavior.
-For example, it's a common practice to close the write end of a pipe
+Leaving unwanted file descriptors open in the new table wastes resources,
+and can also cause bugs.
+For example, it's common practice to close the write end of a pipe
 and expect an EOF on the read end;
-if the write end is copied into the new file descriptor table before being closed,
+if the write end is copied into a new file descriptor table between creating and closing,
 and the write end is never closed in the new table,
@@
 because we must preserve the semantics of implicit file descriptor inheritance.
-The possibility of implicit inheritance of file descriptors is a traditional Unix feature,
+The possibility of implicit inheritance of file descriptors is an old Unix feature,
 which is useful in a wide variety of situations,
@@
 For \verb|unshare(CLONE_FILES)|, we perform \verb|close_cloexec| automatically in userspace.
-For \texttt{clone}, we perform \verb|close_cloexec| closing operation only on user request,
+For \texttt{clone}, we perform \verb|close_cloexec| only on user request,
 so that the user can control when \verb|inherit_fd| stops working.
@@
 In both of these cases,
-it would be beneficial for efficiency to delegate \verb|close_cloexec|
-to a new system call which takes a list of file descriptors
-and closes all \texttt{CLOEXEC} file descriptors except for those on the list.
-
-For \texttt{execve}, this oepration is, in some sense, done automatically for us by the kernel;
+we would benefit from a new system call which takes a list of file descriptors
+and closes all \texttt{CLOEXEC} file descriptors except for those on the list;
+this would replace our userspace implementation of \verb|close_cloexec|.
+
+For \texttt{execve}, \verb|close_cloexec| is, in some sense, done automatically for us by the kernel;
 except we don't get the chance to preserve live file descriptors handles,
@@
 Again an extension to the kernel would be useful:
-A new argument to \texttt{execve} which allows specifying a list of file descriptor numbers to inherit
+A new argument to \texttt{execve} which allows specifying a list of file descriptors to inherit
 despite the \texttt{CLOEXEC} flag.

### 2020-05-16 research/paper/paper.tex
paper: format for OSDI

@@
-\documentclass[sigplan]{acmart}
-% \documentclass{acmart}
+\documentclass[letterpaper,twocolumn,10pt]{article}
+\usepackage{usenix2019_v3}
+
+\usepackage{tikz}
+\usepackage{amsmath}
 
@@
 \usepackage{listings}
-\usepackage{booktabs} % For formal tables
-
-% Copyright
-%\setcopyright{none}
-%\setcopyright{acmcopyright}
-%\setcopyright{acmlicensed}
-\setcopyright{rightsretained}
-%\setcopyright{usgov}
-%\setcopyright{usgovmixed}
-%\setcopyright{cagov}
-%\setcopyright{cagovmixed}
-
-% DOI
-%% \acmDOI{10.475/123_4}
-
-% ISBN
-%% \acmISBN{123-4567-24-567/08/06}
-
-%Conference
-\acmConference[OSDI 2020]{SIGBOVIK}{November 2020}{Banff, Alberta, Canada}
-\acmYear{2020}
-\copyrightyear{2020}
-
-%\acmBadgeL[http://ctuning.org/ae/ppopp2016.html]{ae-logo}
-%\acmBadgeR[http://ctuning.org/ae/ppopp2016.html]{ae-logo}
-
-% Document starts
+\usepackage{booktabs}
+\usepackage{paralist}
+\usepackage{titlesec}
+
+\titleformat*{\subsection}{\normalsize\bfseries}
+\titleformat*{\subsubsection}{\normalsize\bfseries}
+\titlespacing{\subsection}{0pt}{1ex}{0.5ex}
+\titlespacing{\subsubsection}{0pt}{0.5ex}{0ex}
+
 \begin{document}
-
-% Title portion. Note the short title for running heads
-\title[DRAFT: Direct-style process creation on Linux]{DRAFT: Direct-style process creation on Linux}
-
-\author{Spencer Baugh}
-\affiliation{%
-  \institution{Two Sigma}
-  \streetaddress{100 6th Avenue}
-  \city{New York City, New York}
-  \country{USA}}
-\email{sbaugh@twosigma.com}
-
+\date{}
+\title{\Large \bf DRAFT: Direct-style process creation on Linux}
+% \author{
+% {\rm Spencer Baugh}\\
+% Two Sigma
+% }
+\maketitle
 \begin{abstract}
@@
 
-%
-% The code below should be generated by the tool at
-% http://dl.acm.org/ccs.cfm
-% Please copy and paste the code instead of the example below.
-%
-\begin{CCSXML}
-<ccs2012>
-   <concept>
-       <concept_id>10011007.10010940.10010941.10010949.10010957</concept_id>
-       <concept_desc>Software and its engineering~Process management</concept_desc>
-       <concept_significance>500</concept_significance>
-       </concept>
-   <concept>
-       <concept_id>10011007.10010940.10010941.10010949.10010950.10010951</concept_id>
-       <concept_desc>Software and its engineering~Virtual memory</concept_desc>
-       <concept_significance>300</concept_significance>
-       </concept>
-   <concept>
-       <concept_id>10010147.10010169</concept_id>
-       <concept_desc>Computing methodologies~Parallel computing methodologies</concept_desc>
-       <concept_significance>500</concept_significance>
-       </concept>
- </ccs2012>
-\end{CCSXML}
-
-\ccsdesc[500]{Software and its engineering~Process management}
-\ccsdesc[300]{Software and its engineering~Virtual memory}
-\ccsdesc[500]{Computing methodologies~Parallel computing methodologies}
-
-\keywords{Process creation, Linux, fork, spawn}
-
-\maketitle
-
 \section{Introduction}\label{introduction}
@@
 over Linux's fork-style \texttt{fork} and spawn-style \texttt{posix\_spawn}:
-\begin{itemize}
+\begin{compactitem}
 \item
@@
   rather than batched together as with spawn-style or manually reported via IPC as with fork-style.
-\end{itemize}
+\end{compactitem}
 
@@
 This paper is organized in several sections.
-\begin{itemize}
+\begin{compactitem}
 \item In section \ref{background}, we provide background on fork-style, spawn-style, and direct-style process creation,
@@
 \item In section \ref{conclusions}, we briefly outline our conclusions about direct-style and the future of process creation.
-\end{itemize}
+\end{compactitem}
 \section{Background}\label{background}
@@
 
-\lstinputlisting[float=*,language=Python,label={basic},caption={Creating a new process, changing CWD, and execing}]
+\lstinputlisting[float=*ht,language=Python,label={basic},caption={Creating a new process, changing CWD, and execing}]
 {examples/basic/direct.py}
@@
 
-\lstinputlisting[float=*,language=Python,label={fds},caption={Passing down FDs}]
+\lstinputlisting[float=*ht,language=Python,label={fds},caption={Passing down FDs}]
 {examples/fds/direct.py}
@@
 
-\lstinputlisting[float=*,language=Python,label={pipe},caption={Creating a parallel processing pipeline}]
+\lstinputlisting[float=*ht,language=Python,label={pipe},caption={Creating a parallel processing pipeline}]
 {examples/pipe/direct.py}
@@
 
-\lstinputlisting[float=*,language=Python,label={mount},caption={Overriding absolute path using a mount namespace}]
+\lstinputlisting[float=*ht,language=Python,label={mount},caption={Overriding absolute path using a mount namespace}]
 {examples/mount/direct.py}
@@
 
-\lstinputlisting[float=*,language=Python,label={unmount},caption={Unmount all and run executable via fexec}]
+\lstinputlisting[float=*ht,language=Python,label={unmount},caption={Unmount all and run executable via fexec}]
 {examples/unmount/direct.py}
@@
 
-\lstinputlisting[float=*,language=Python,label={pidns},caption={Nested clone and pid namespace}]
+\lstinputlisting[float=*ht,language=Python,label={pidns},caption={Nested clone and pid namespace}]
 {examples/pidns/direct.py}
@@
 
-\lstinputlisting[float=*,language=Python,label={fuse},caption={Shared file descriptor tables}]
+\lstinputlisting[float=*ht,language=Python,label={fuse},caption={Shared file descriptor tables}]
 {examples/fuse/direct.py}
@@
 We quantitatively answer the following questions:
-\begin{itemize}
+\begin{compactitem}
 \item
@@
 % could be qualitative
-\end{itemize}
+\end{compactitem}
 \subsection{Ease of programming with direct-style}\label{ease}
@@
 Direct-style consistently takes under half the lines the express the same thing.
-\lstinputlisting[float=*,language=Python,label={basic_fork},caption={Fork-style: Creating a new process, changing CWD, and execing}]
+\lstinputlisting[float=*ht,language=Python,label={basic_fork},caption={Fork-style: Creating a new process, changing CWD, and execing}]
 {examples/basic/fork.py}
@@
 
-\bibliographystyle{ACM-Reference-Format}
+\bibliographystyle{plain}
 \bibliography{bibliography}

### 2020-05-16 python/subprocess_bench.py
update benchmark

@@ -114,7 +114,7 @@ async def main() -> None:
             times.append(time)
         return(mean(times))
     async def get_data(mode: str) -> t.List[t.Tuple[int, int]]:
-        return [(i, await run_many(mode, 10**i)) for i in range(5)]
+        return [(i, await run_many(mode, 10**i)) for i in range(7)]
     subprocess = [(0, 1482.3333333333333), (1, 1530.3333333333333), (2, 1504.6666666666667), (3, 1665.3333333333333), (4, 1873.3333333333333)]
     rsyscall = [(0, 2249.3333333333335), (1, 2234.3333333333335), (2, 2261.6666666666665), (3, 2241.3333333333335), (4, 2321.3333333333335)]
     subprocess = await get_data("subprocess")

### 2020-05-17 python/benchmark.py
fix benchmark

@@ -83,7 +83,7 @@ async def main() -> None:
             times.append(time)
         return(mean(times))
     async def get_data(mode: str) -> t.List[t.Tuple[int, int]]:
-        return [(i, await run_many(mode, 10**i)) for i in range(5)]
+        return [(i, await run_many(mode, 10**i)) for i in range(6)]
     # fork = [(0, 76.55446666666667), (1, 77.37133333333334), (2, 82.745), (3, 132.49946666666668), (4, 471.80173333333335), (5, 4062.4732), (6, 42670.6794)]
     # clone = [(0, 18.5436), (1, 20.663266666666665), (2, 18.799533333333333), (3, 19.606266666666667), (4, 20.892733333333332), (5, 20.747866666666667), (6, 21.152866666666668)]
     # clone_fork = [(0, 42.185066666666664), (1, 43.29793333333333), (2, 44.864733333333334), (3, 66.6312), (4, 212.4854), (5, 2751.410533333333), (6, 22280.86373333333)]

### 2020-05-17 research/paper/clone.c

@@ -8,6 +8,7 @@
 #include <sys/syscall.h>
 #include <sched.h>
 #include <string.h>
+#include <signal.h>
 #include <err.h>
 #include <sched.h>
 
@@ -44,7 +45,7 @@ static int do_fork(struct log *log) {
 
 static int do_clone_fork(struct log *log) {
         int rc;
-        rc = syscall(SYS_clone, 0, NULL, NULL, NULL, NULL);
+        rc = syscall(SYS_clone, SIGCHLD, NULL, NULL, NULL, NULL);
         if (rc == 0) {
                 gettimeofday(&log->in_child, NULL);
                 exit(0);

### 2020-05-17 research/paper/paper.tex
paper

@@
 \begin{document}
+\newcommand{\twosigma}{[tech company]} 
 \date{}
@@
 We've implemented a variety of useful applications using direct-style process creation on Linux,
-which we'll showcase in brief in this paper.
+which we'll demonstrate in this paper.
 Direct-style \texttt{clone} has minimal performance cost in normal situations,
 and can outperform \texttt{fork} in some cases.
-We have deployed direct-style process creation in production internally at Two Sigma.
+We have deployed direct-style process creation in production internally at \twosigma.
 
@@
 and is immediately deployable on recent Linux kernels.
-\texttt{rsyscall} is open source and available from \url{https://github.com/catern/rsyscall}.
+\texttt{rsyscall} is open source and available from
+[Github url].
+% \url{https://github.com/catern/rsyscall}.
 
@@
 only \texttt{execve} and \texttt{unshare} need substantial further attention.
-\subsection{\texttt{exec}}
+\subsection{\texttt{exec}}\label{execve}
 \texttt{execve} is unusual and requires careful design,
@@
 Despite this, fork-style implementations are still significantly more verbose than direct-style.
+
+We ensure that errors are reported back to the main program,
+even without handling them;
+this is a significant source of boilerplate in fork,
+because reporting an error back to the main program requires IPC.
+Even if there's no desire to handle errors,
+proper error reporting is necessary for debugging.
+And for serious users, proper error reporting and handling
+is essential for a robust application.
+% TODO hmm talk about... some stuff.
 \subsection{Clone versus fork}\label{microbench}
@@
 before running the benchmark.
-\texttt{clone} without \verb|CLONE_VM| functions identically to fork.
+Note that \texttt{clone} without \verb|CLONE_VM| is expected to function identically to fork.
+
+All benchmarks are run pinned to an isolated core,
+running on an Intel i9-9900K CPU with hyperthreading enabled.
+Our benchmarks are intended to give a sense of relative, not absolute, performance,
+and we have no reason to believe a more controlled benchmarking setup
+would have differential effects on the performance of \texttt{fork} and \texttt{clone}.
+
 The results are in figure \ref{fig:microbench}.
@@
 so its process creation time is constant in the amount of memory used in the parent process.
-
 \subsection{High-level direct-style versus spawn-style}\label{subprocess_bench}
@@
 
-We benchmark Python's \texttt{subprocess.run} against direct-style \texttt{clone} with \texttt{rsyscall}
-with varying memory usage, as in section \ref{microbench}.
+We create a child process running \texttt{true} and wait for it to exit
+using both the Python standard library's \texttt{subprocess.run},
+and direct-style \texttt{clone} with \texttt{rsyscall}.
+We benchmark the time each implementation takes
+with varying amounts of mapped memory, as in section \ref{microbench}.
+Both are running under CPython 3.7.7,
+again pinned to an isolated core as with section \ref{microbench}.
 The results are in Figure \ref{fig:subprocess_bench}.
-TODO generate again to get the higher end where it crosses over.
+
+The direct-style \texttt{clone} implementation stays around 2.2 milliseconds regardless of memory usage.
+\texttt{subprocess.run}, which uses \texttt{fork},
+starts at 1.4 milliseconds,
+and then scales linearly with the amount of memory mapped, as expected.
+Direct-style \texttt{clone} starts being faster than \texttt{subprocess.run}
+at about 40 megabytes of memory mapped.
+
+These results are acceptable.
+Process creation in Python is already slow, taking milliseconds of time,
+so it is not expected to be on the fast path.
+In that context, \texttt{rsyscall} has a reasonable cost compared to \texttt{subprocess.run},
+and avoids the bad scaling of \texttt{fork} which might be unexpected by the naive programmer.
+
+Furthermore,
+our implementation of \texttt{rsyscall} is barely-optimized pure-Python,
+all the way down the syscall primitive,
+whereas \texttt{subprocess.run} is entirely in C.
+If we need to speed up our implementation,
+we can perform some targeted optimizations,
+reimplement some parts of the Python \texttt{rsyscall} library in native code,
+or run our code using a faster Python interpreter such as PyPy.
 \section{Related work}\label{related_work}
-% TODO all this is just slapped onto the page without much thought, needs revising
 \subsection{Direct-style process creation}
@@
 and serviced by another process.
-Also relevant is seccomp system call interception,
-which can be used to intercept system calls and emulate them in another process.
 
@@
 FlexSC\cite{flexsc} is one example.
-System calls were forwarded to dedicated system call execution threads.
+System calls in FlexSC are forwarded to a pool of dedicated system call execution threads.
 The recently added Linux feature \verb|io_uring| is, in some sense, another example;
@@
 which is capable of operating on multiple processes at once.
-Unfortunately, as discussed in section \ref{implementation},
-ptrace is unsuitable as a mechanism for a general purpose system since it is not re-entrant;
-a process which is already being ptraced cannot be ptraced again,
-so using ptrace as a cross-process system call mechanism would disallow use of strace or debuggers.
+As discussed in section \ref{implementation},
+a ptrace-based has approach has limitations which make it unsuitable for \texttt{rsyscall}.
 \subsection{Process capabilities}
-We mentioned the concept of process capabilities while discussing the design of \texttt{rsyscall} in section \ref{overview}.
 The Capsicum system provides something called process descriptors\cite{capsicum},
@@
 Latter efforts\cite{oblivious}, including CloudABI\cite{cloudabi} and WASI\cite{wasi},
-are focused primarily on backwards compatibility with the existing POSIX environment, and ease of porting existing code.
-Programs ported to the capability-secure environment
-are typically launched with an accompanying spawn-style wrapper.
-This means that capability-secure processes are created primarily by an API
-that is distinct from the capability-secure syscall API they have access to,
-which makes it harder for a capability-secure process to spawn further capability-secure processes of its own.
-Direct-style process creation could make it easier to create capability-secure processes 
-from the capability-secure syscall API,
-and reduce the need for a separate spawn-style wrapper.
-
-Also related is the PLASH, ``Principle of Least Authority Shell''\cite{plash}.
-PLASH provides a spawn-style API (in the form of a shell)
-to launch processes in a capability-secure environment.
+developed this into a partial or full replacement for a POSIX libc.
+Like all other Unix libcs that we know of,
+these libc-replacements implicitly make system calls on the current process,
+rather than using an explicit process capability as \texttt{rsyscall} does.
+
+PLASH provides a spawn-style API, in the form of a shell,
+to launch processes in a capability-secure environment.\cite{plash}
 PLASH, like all spawn-style APIs, abstracts over the native Linux environment,
@@
 \texttt{rsyscall} was not developed solely for the purpose of this paper,
-and it has many uses unrelated to direct-style process creation,
+and it has other uses unrelated to direct-style process creation,
 such as asynchronous system calls, exceptionless system calls\cite{flexsc}, cross-host operations, among others.
@@
 which has substantial benefits for deployability.
-Nevertheless, direct kernel support for creating a stub process and performing syscalls in the context of that process
+Nevertheless, direct support in the Linux kernel
+for creating a stub process and performing syscalls in the context of that process
 may provide efficiency benefits, as well as reducing userspace-visible complexity.
 
-Several other aspects of our implementation would be improved by kernel support.
-We discussed these in the implementation section;
-in brief, we would most benefit from kernel support for
-detecting when a child process finishes =execve=,
-closing all =CLOEXEC= file descriptors except for an explicitly specified list,
-and explicitly specifying a list of =CLOEXEC= file descriptors to inherit over =excve=.
-Implementing these features in the kernel in a generally useful way, and upstreaming them,
-is an important direction for our future work.
+Several other aspects of our implementation would also be improved by new features in the Linux kernel,
+as discussed in sections \ref{execve} and \ref{cloexec}.
 \subsection{Portability to other Unix systems}
@@
 We have made use of the techniques described in this paper
-in proprietary software at Two Sigma.
+in proprietary software at \twosigma.
 While this gives us personally greater confidence in these techniques,
-it would be better to use them in a publicly available, open source system.
+it would be beneficial to use them in a publicly available, open source system.
+Likely the best 
 Either porting an existing system to use these techniques,
 or using these techniques to create a substantial new system from scratch,
-would provide a meaningful demonstration of the viability of these techniques.
-% \subsection{file descriptor lifetime management}
-% % TODO this is hard to explain, probably best to drop this
-% Keeping file descriptors open if and only if there is a specific process using that file descriptor
-% is not the only possibility.
-% Keep in mind that "a process using the file descriptor" is a slight abuse of terminology;
-% processes don't use file descriptors, programs do, and in our system there is only one program.
-% To reflect this, we could instead more directly use file descriptors in a process-agnostic way;
-% this would support the creation of objects
-% which work transparently across multiple processes which share a file descriptor table.
-% Such objects would automatically use the relevant process to perform the syscalls for any specific operation.
-% A process-centric view of file descriptors instead forces each object to be associated with one process.
-% Nevertheless, we found that the process-centric perspective better matches the existing intuitions of users,
-% especially those with prior experience in programming with processes.
-% We hope that future systems for multi-process programming
-% might explore an object-centric approach for managing resources.
+would provide a substantial demonstration of the viability of these techniques.
 \section{Conclusions}\label{conclusions}

### 2020-05-17 research/paper/paper.tex
paper

@@
 and discuss several difficult process-creation related details of \texttt{rsyscall}'s implementation.
-\item In section \ref{evaluation}, we evaluate, somehow (TODO).
+\item In section \ref{evaluation}, we evaluate using both quantitative and qualitative approaches.
 \item In section \ref{related_work},
@@
 and erroneously inherit the file descriptor.
-\section{Evaluation \textbf{TODO}}\label{evaluation}
-
-We quantitatively answer the following questions:
+\section{Evaluation}\label{evaluation}
+We answer the following questions:
 \begin{compactitem}
@@
   We benchmark direct-style process creation against spawn-style in \ref{subprocess_bench}.
-% TODO have an item about use at TS? hm.
-% could be qualitative
+\item
+  Does direct-style process creation work in the ``real world''?
+  We discuss our experience with using direct-style process creation at \twosigma in \ref{realworld}.
 \end{compactitem}
@@
 is essential for a robust application.
-% TODO hmm talk about... some stuff.
 \subsection{Clone versus fork}\label{microbench}
@@
 or run our code using a faster Python interpreter such as PyPy.
+\subsection{Usage in the real world}\label{realworld}
+At \twosigma we have used \texttt{rsyscall} and direct-style process creation
+as a component of a distributed system deployment library.
+This library is in use in production,
+and is extensively used as part of testing and development.
+
+With direct-style process creation,
+we can robustly and easily write custom code to start up each program,
+passing down file descriptors and initializing process attributes according to each program's needs,
+all of which can be done directly from a single parent process.
+This has allowed us to make more use of file descriptor abstraction;
+for example, connections over our shared memory transport are bootstrapped over passed-down file descriptors,
+and our management interfaces listen on passed-in sockets,
+allowing them to be easily swapped between network-accessible TCP sockets and host-local Unix domain sockets.
+We can also flexibly use namespaces where appropriate;
+for example, for legacy applications which directly start subprocesses,
+we prevent process leaking with pid namespaces.
+
+All this has allowed us to run our systems in a much more dynamic way than before;
+we can freely start up arbitrary subsets of our systems on-demand,
+spread across one or more hosts,
+for development, testing, or production,
+regardless of the configuration of that host,
+without worrying about interference between instances,
+and without a dependency on any external privileged services,
+such as container runtimes.
+This has become a key part of our new development workflow.
 \section{Related work}\label{related_work}
@@
 \subsection{Large scale open source usage}
-We have made use of the techniques described in this paper
-in proprietary software at \twosigma.
-While this gives us personally greater confidence in these techniques,
-it would be beneficial to use them in a publicly available, open source system.
-Likely the best 
-Either porting an existing system to use these techniques,
-or using these techniques to create a substantial new system from scratch,
-would provide a substantial demonstration of the viability of these techniques.
+We would like to open source libraries built on top of direct-style process creation.
+From our experience using such libraries at \twosigma,
+we believe this would greatly ease the development of complex systems involving processes,
+and would provide a substantial demonstration of the power of direct-style process creation.
 \section{Conclusions}\label{conclusions}
@@
 We hope that a better process creation mechanism
-will help encourage more use of the process abstraction,
+will help encourage more creative use of processes,
 which, though widespread,
-is still not used to its full potential.
+are still not used to their full potential.
 

### 2020-05-20 research/paper/paper.tex
paper

@@
 \usepackage{amsmath}
+\usepackage{amssymb}
+\usepackage{pifont}
+\usepackage{comment}
+
+\newcommand{\cmark}{\ding{51}}
+\newcommand{\xmark}{\ding{55}}
 
@@
 fork-style requires IPC if it wishes to communicate errors back to the rest of the program,
-and robust indication of errors in spawn-style requires substantial growth in the spawn interface.
-
-Direct-style is more rarely seen, but has a number of advantages.
+and robust indication of errors in spawn-style requires substantial expansion of the spawn interface.
+
+Direct-style is more rarely seen, but gives a number of advantages.
 Direct-style involves creating a new child process in a non-running state,
@@
 which replace libc system call wrappers, and allow a program to perform cross-process system calls.
-rsyscall currently has best support for Python,
+\texttt{rsyscall} can work entirely in userspace,
+and can be deployed on existing Linux systems without kernel modifications.
+\texttt{rsyscall} is currently focused on supporting, and is implemented in, Python,
 but can be ported to other languages;
@@
 Our direct-style process creation implementation uses \texttt{clone} to create a child process in an embryonic state
-which can then be manipulated from user code using syscalls with \texttt{rsyscall}.
+which can then be manipulated from user code in the parent process using syscalls with \texttt{rsyscall}.
 A direct-style \texttt{clone} has all the advantages of direct-style process creation
@@
 \item
-  Syscall results are reported individually to the parent process,
-  rather than batched together as with spawn-style or manually reported via IPC as with fork-style.
+  Syscall results are reported individually to the parent process.
+  % rather than batched together as with spawn-style or manually reported via IPC as with fork-style.
 \end{compactitem}
@@
 Our implementation is distributed as a feature of \texttt{rsyscall},
-and is immediately deployable on recent Linux kernels.
+and is immediately deployable, entirely in userspace, on recent Linux kernels.
 \texttt{rsyscall} is open source and available from
@@
 \begin{compactitem}
-\item In section \ref{background}, we provide background on fork-style, spawn-style, and direct-style process creation,
-and consider their advantages and disadvantages in more depth.
-\item In section \ref{overview}, we discuss the design of the API,
+\item Section \ref{background} provides background on fork-style, spawn-style, and direct-style process creation,
+and considers their advantages and disadvantages in more depth.
+\item Section \ref{overview} discusses the design of the API,
 including our usage of \texttt{rsyscall}.
-\item In section \ref{applications}, we show examples of usage of direct-style process creation in Python on Linux,
+\item Section \ref{examples} shows examples of direct-style process creation in Python on Linux,
 including use of process features which are difficult to use on Linux with \texttt{fork} or \texttt{posix\_spawn}.
-\item In section \ref{implementation}, we explore our implementation in depth,
-and discuss several difficult process-creation related details of \texttt{rsyscall}'s implementation.
-\item In section \ref{evaluation}, we evaluate using both quantitative and qualitative approaches.
-\item In section \ref{related_work},
-we examine other work which also touches on the topics covered by this paper.
-\item In section \ref{future_work},
-we list several directions for future work related to direct-style process creation and to \texttt{rsyscall}.
-\item In section \ref{conclusions}, we briefly outline our conclusions about direct-style and the future of process creation.
+\item Section \ref{implementation} explores our implementation in depth,
+and discusses several difficult process-creation related details of \texttt{rsyscall}.'s implementation.
+\item Section \ref{evaluation} evaluates our results using both quantitative and qualitative approaches.
+\item Section \ref{related_work} examines other work which also touches on the topics covered by this paper.
+\item Section \ref{future_work} lists several directions for future work
+related to direct-style process creation and to \texttt{rsyscall}.
+\item Section \ref{conclusions} briefly outlines our conclusions about direct-style and the future of process creation.
 \end{compactitem}
-\section{Background}\label{background}
+\section{Background on process-creation}\label{background}
 As mentioned in the introduction,
@@
 A few operating systems have a third style, which we refer to as "direct-style".
+
+% Ugh, these aren't the same size right now.
+\newcommand{\tbad}{\xmark:}
+\newcommand{\tgood}{\cmark:}
+% \newcommand{\tbad}{\text{\sffamily X} :}
+% \newcommand{\tgood}{\checkmark:}
+
+\begin{table*}
+\begin{tabular}{l|l|l|l}
+% BEGIN RECEIVE ORGTBL compare
+ & fork-style & spawn-style & direct-style\\
+\hline
+Works with any caller & \tbad No threads or large memory & \tgood Yes & \tgood Yes\\
+Programming model & \tbad Complex (returns twice) & \tgood Simple (single call) & \tgood Simple (imperative)\\
+Maximally powerful & \tgood Yes, can call any syscall & \tbad No, limited interface & \tgood Can call any syscall\\
+Error handling & \tbad Poor, requires IPC & \tbad Poor, not fine-grained & \tgood Reports individual errors\\
+Non-copied attributes & Mutated by code in child & Set by arguments & Mutated by code in parent\\
+% END RECEIVE ORGTBL compare
+\end{tabular}
+\caption{Features of fork-style vs spawn-style vs direct-style}
+\label{tab:ease}
+\end{table*}
+\begin{comment}
+#+ORGTBL: SEND compare orgtbl-to-latex :splice t
+|                       | fork-style                       | spawn-style                  | direct-style                     |
+|-----------------------+----------------------------------+------------------------------+----------------------------------|
+| Works with any caller | \tbad No threads or large memory | \tgood Yes                   | \tgood Yes                       |
+| Programming model     | \tbad Complex (returns twice)    | \tgood Simple (single call)  | \tgood Simple (imperative)       |
+| Maximally powerful    | \tgood Yes, can call any syscall | \tbad No, limited interface  | \tgood Can call any syscall      |
+| Error handling        | \tbad Poor, requires IPC         | \tbad Poor, not fine-grained | \tgood Reports individual errors |
+| Non-copied attributes | Mutated by code in child         | Set by arguments             | Mutated by code in parent        |
+% $
+\end{comment}
 \subsection{Fork-style}
@@
 with a different return value from fork to allow it to distinguish whether it is running in the parent or the child.
-Typically, the program will case on the return value
+Typically, the program will branch based on the return value
 and, if running in the child,
@@
 and much ink has been spilled detailing them.
-We'll cover the most prominent briefly here,
+We'll cover the most egregious briefly here,
 but see Baumann 2019 \cite{forkroad} for a more detailed evaluation.
@@
 
-However, the arguments that can be provided to a spawn-style process creation syscall
+The arguments that can be provided to a spawn-style process creation syscall
 do not cover all the possible attributes that one might want to set for the new process.
@@
 Spawn-style requires some form of declarative description of the new process,
-which must be learned in additional to the normal system call interface.
+which must be learned in addition to the normal system call interface.
 
@@
 and some others \cite{exokernel} \cite{fuschia} \cite{singularity},
-use another style of process creation, which we refer to as "direct-style".
+use another style of process creation.
+We use the new term "direct-style" to refer to this,
+since we know of no pre-existing concise and generic term for this style of process creation.
 
@@
 This is the kind of direct-style interface that we contribute in this paper.
-\section{Overview}\label{overview}
+\section{Overview of API}\label{overview}
 In most operating system APIs, including the ones available on Linux,
@@
 direct-style \texttt{clone} will always be at least as powerful as \texttt{fork}.
-\section{Applications}\label{applications}
+\section{Examples}\label{examples}
 In this section,
-we'll discuss several more advanced uses of processes
-using more sophisticated Linux features,
-and accompany them with Python examples using direct-style process creation.
-All of these examples use only conventional Linux system calls,
-and can therefore translated straightforwardly into any language with an \texttt{rsyscall} API.
-For concision, we assume that we are running with sufficient permissions in these examples,
+we'll demonstrate direct-style process creation
+by demonstrating several advanced uses of processes
+using more sophisticated Linux features.
+For concision, we assume in these examples that we are running with sufficient permissions in these examples,
 but these examples can all be run without privileges with appropriate use of \texttt{CLONE.NEWUSER}.
 
-Implementing these examples with \texttt{fork} or \verb|posix_spawn| requires substantially more boilerplate,
+These examples are not novel, as such, as they use only conventional Linux system calls;
+but implementing them with \texttt{fork} or \verb|posix_spawn| requires substantially more code,
 or even a complete change in implementation approach,
-such as in section \ref{shared_fd_table}.
+as described in section \ref{shared_fd_table}.
 We'll compare our direct-style implementations in this section
@@
 
-Several of these examples demonstrate functionality that can be done today by specialized software and services.
-Such software often allows use of one or two of these techniques, but prevents the use of others.
+Several of these examples provide functionality that can be done today by specialized software and services.
+Such software often enables one or two of these uses of processes, but prevents others.
 For example, a shell allows the creation of pipelines and container systems like Docker allow sandboxing,
-but the two are difficult to combine.
-
-One of the key advantages of direct-style process creation is that the techniques we demonstrate here
+but the two are difficult to combine.\cite{docker_pipe}
+
+One of the key advantages of direct-style process creation is that the features we demonstrate here
 can be easily used in combination,
@@
 
+% TODO need to clarify this more, this is v2, Larry didn't like v1 beyond first sentence
 Most resources in Linux are managed through file descriptors,
-so this feature allows the parent process to pass a wide variety of resources to the child process.
-An executable program can be provided a file descriptor,
-which can correspond to any file in any filesystem,
-or to a network connection\cite{ucspi},
-or a pipe connected to another process,
-or a variety of other resources,
-without the process being aware of what specific resource it is accessing.
-In this way, we immediately know "for free"\cite{theoremsforfree}
-that the behavior of the program does not depend on the name of the file,
-or the hostname to it which the socket is connected,
-or the executable running in the process on the other end of the pipe,
-or other abstracted-away details.
+so this feature allows the parent process to pass a variety of resources to the child process,
+such as files, network connections, pipes, or other resources.\cite{ucspi}
+Since the resource is created outside the child process and passed down as an opqaue file descriptor,
+the precise details and type of the resource are not available to the child.
+Thus, this is a form of abstraction,
+and we can know ``for free''\cite{theoremsforfree}
+that the program running in the child process
+will not rely on the details of how the file descriptor was created,
+such as the filename.
 
@@
 can be used to customize a program's behavior
-in ways that were not anticipated by the author
+in ways that were not anticipated at development time
 by customizing the environment the process runs in.
@@
 \section{Implementation}\label{implementation}
-\subsection{Basics about \texttt{rsyscall}}
 In this section, we'll give a brief overview of the implementation of \texttt{rsyscall},
 and then focus on implementation issues specific to process creation.
-
+\subsection{Cross-process syscalls}
 \texttt{rsyscall} can be conceptually divided in two parts:
@@
 they are, for the most part, agnostic to how the cross-process syscall is implemented.
-
-On Linux \verb|x86_64|, a syscall is specified by a syscall number plus six register-sized arguments;
-a syscall returns one register-sized value.
-\texttt{rsyscall}'s default implementation of cross-process syscalls sends those seven integers over a pipe,
-and waits for a response on another pipe.
-Processes are created running a loop which, at each iteration,
-reads a syscall request off the pipe,
-performs that syscall,
-and writes the return value back over the return pipe.
-In this way, a cross-process syscall works much like a very primitive remote procedure call.
-
-We use pipes (or other file descriptors) for transport of data rather than a shared memory segment
+The cross-process syscall primitive takes a syscall number, and some number of register-sized integer arguments,
+and arranges to call that syscall in the target process,
+returning the response as a single register-sized integer.
+
+\texttt{rsyscall}'s default userspace cross-process syscall primitive sends the syscall request
+to a userspace ``syscall server'' running in the target process,
+which performs the syscall and sends the response back.
+Communication to the syscall server happens over file descriptors,
+typically a pair of pipes.
+We use file descriptors for transport of data rather than shared memory
 to support straightforward integration with event loops.
 
-ptrace provides an alternative means to perform arbitrary actions on other processes,
-including system calls.
-However, among other issues, it has the substantial disadvantage of not permitting multiple ptracers.
+ptrace provides a way to perform arbitrary actions on other processes,
+including system calls,
+but is not suitable for us.
+Among other issues, it has the substantial disadvantage of not permitting multiple ptracers.
 A ptrace-based implementation would prevent using \texttt{strace} or \texttt{gdb} on \texttt{rsyscall}-controlled processes,
@@
 This is the case most of the time; we pass \texttt{CLONE.VM} to \texttt{clone} by default in the Python wrapper.
-However, the target process may be in another address.
-For example, if the target process is at a different privilege level,
-it must be in a different address space,
-as sharing an address space with a process at a different privilege level is unsafe.
-
-We implement access to the memory of processes in different address spaces through another set of pipes,
-by copying memory into and out of those pipes using \texttt{read} and \texttt{write} system calls in the target process.
-A \texttt{read} syscall corresponds to a write to memory,
-and a \texttt{write} syscall corresponds to a read from memory.
-
-The \verb|process_vm_readv| and \verb|process_vm_writev| system calls
-allow the caller to read and write memory from the virtual address space of other processes.
-However, they require that the caller have specific credentials relative to the process being accessed,
-which may not always be the case.
-Additionally, these system calls are disabled if ptrace is disabled system-wide,
-which is a niche but possible system configuration.
-To ensure that \texttt{rsyscall} can be used for arbitrary purposes and on arbitrary systems, we avoided these calls.
+
+Sometimes, the target process may be in another address space;
+for example, if the target process is at a different privilege level,
+it must be in a different address space.
+There are a number of available techniques in that scenario;
+we choose to copy memory over a pair of pipes,
+again using file descriptors for the sake of easy event-loop integration.
 \subsection{\texttt{clone}}\label{clone}
@@
 
-There are three Linux system calls which create processes:
-\texttt{fork}, \texttt{vfork} and \texttt{clone}.
+Besides \texttt{clone}, \texttt{vfork} and \texttt{fork} also create processes,
+but they are not suitable for flexible direct-style process creation.
 \texttt{vfork} suspends execution of the parent process
 while waiting for the child process to exit or call \texttt{execve},
-which is immediately unsuitable for direct-style process creation.
-\texttt{clone} provides a superset of the functionality of \texttt{fork},
-so we focused our attention on \texttt{clone}.
-
-\texttt{clone}
-(along with \texttt{fork}) creates a new process
+which is immediately unsuitable.
+\texttt{fork} lacks many features which are restricted to \texttt{clone}.
+For example, with \texttt{fork},
+we could not create child processes which share the parent's address space,
+which would complicate memory access.
+
+\texttt{clone} creates a new process
 which immediately starts executing at the next instruction after the syscall instruction,
@@
 
-Common Linux libcs provide a wrapper for \texttt{clone}
-which allow executing an arbitrary function with an arbitrary argument,
-but since we want to support remote execution,
-we can't use the libc wrapper.
-
-Instead, since \texttt{clone} lets us change the stack for the new process,
-we use this to return, after the system call to an arbitrary function address stored on the stack.
+\texttt{clone} lets us change the stack for the new process;
+we use this to call an arbitrary function address stored on the stack.
 Further arguments can also be passed on the stack,
@@
 
-We use this to create processes running the syscall request-response loop described in the previous section.
+We use this to create processes running the syscall server.
 After this,
-most system calls can be called as normal.
-The new process can be modified freely through \texttt{chdir}, \texttt{dup2}, and other system calls.
-Out of system calls related to process creation,
+the parent process can be begin to call system calls in the child process.
+Most system calls work as normal;
+the new process can be modified freely through \texttt{chdir}, \texttt{dup2}, and other system calls.
+From the system calls related to process creation,
 only \texttt{execve} and \texttt{unshare} need substantial further attention.
@@
 because when it is successful, it does not return.
-Therefore we need a way to determine if \texttt{execve} is successful;
+We need a way to determine if \texttt{execve} is successful;
 naively waiting for a response to the syscall request will leave us waiting forever.
 
-There is a traditional technique to detect a successful \texttt{execve} using a pipe.
-We create a pipe before forking,
-ensure both ends are marked \verb|O_CLOEXEC|,
-perform the fork,
-call \texttt{execve} in the child,
-close the write end of the pipe,
-and wait for EOF on the read end.
-If the child process has neither successfully called \texttt{execve}, nor exited for some other reason,
-then the write end of the pipe will still be open in the child process's fd table,
-and the read end of the pipe will not return EOF.
-But once the child process calls \texttt{execve} successfully,
-\verb|O_CLOEXEC| will cause the write end of the pipe to be closed,
-and the read end of the pipe will return EOF.
-
-This trick works well with \texttt{fork};
+There is a traditional technique used with \texttt{fork}
+to detect a successful \texttt{execve} using a pipe,
+which unfortunately won't work for us.
+In this technique, the parent process creates a pipe before forking,
+ensures both ends are marked \verb|O_CLOEXEC|,
+performs the fork,
+closes the write end of the pipe,
+and \texttt{read}s the read end of the pipe.
+The child process either successfully calls \texttt{execve} or exits;
+either way, the last copy of the write end of the pipe will be closed,
+which causes the \texttt{read} in the parent process to return EOF.
+The parent can then check that the child process hasn't exited;
+if the child hasn't exited, then it must have successfully called exec.
+
+This trick works with \texttt{fork},
 but it's not general enough to work with \texttt{clone}.
@@
 and the read end of the pipe will immediately return EOF,
-regardless of whether the child has called \texttt{execve} or exited.
+before the child has called \texttt{execve} or exited.
 
 Fortunately, there is an alternative solution, which does work with \verb|CLONE_FILES|.
-The \texttt{ctid} argument to \texttt{clone} specifies a memory address which,
-when the \verb|CLONE_CHILD_CLEARTID| flag is set,
-the kernel sets to zero when the child exits or execs,
-and then, crucially, performs a futex wakeup on.
+\texttt{clone} has an argument, \texttt{ctid}, which specifies a memory address.
+If the \verb|CLONE_CHILD_CLEARTID| \texttt{clone} flag is set,
+then when the child exits or execs,
+the kernel will set \texttt{ctid} to zero and then,
+crucially, perform a futex wakeup on it.
 
@@
 
-Unfortunately, futexes in current Linux integrate poorly:
-There is no way for a single process to wait for more than one futex at a time,
-and no way to monitor a futex with file-descriptor-monitoring syscalls such as \texttt{poll}.
-The best we can do is create a dedicated child process for each futex we want to wait on,
+A process can only wait on one futex at a time;
+to monitor multiple futexes from a single event loop,
+we need to create a dedicated child process for each futex we want to wait on,
 and have this child process exit when the futex has a wakeup.
-Monitoring child processes can be straightforwardly integrated into an event loop.
-
-While slightly complex to implement, this solution works well.
-We provide \texttt{ctid} whenever we call \texttt{clone},
+We can then monitor these child processes from an event loop using standard techniques.\cite{signalfd}\cite{pidfd}
+
+So, we pass \texttt{ctid} whenever we call \texttt{clone},
 and set up a process to wait on that futex.
@@
 this ambiguity is unfortunate, but it is also present in the pipe-based approach.
-This is, we believe, the best solution currently available.
-
-We would prefer for Linux to natively provide functionality to wait for a child's \texttt{execve}.
-Some other Unix-like systems provide this;
-kqueue, on FreeBSD, allows waiting for exec in arbitrary processes through kqueue's \verb|EVFILT_PROC|.
-
-One approach for Linux would be to add a new \texttt{clone} flag to
+
+Some other Unix-like systems natively provide the ability to wait for a child's \texttt{execve};
+our implementation would be simplified, and the ambiguity eliminated,
+if we had this ability on Linux.\cite{freebsd_kqueue}
+One approach would be to add a new \texttt{clone} flag to
 opt-in to receiving \texttt{WEXECED} events through \texttt{waitid}.
@@
 \subsection{Managing file descriptor tables}\label{fdtables}
-As demonstrated in section \ref{shared_fd_table},
-the \verb|CLONE_FILES| flag can be passed to \texttt{clone},
-causing the file descriptor table to be shared between the parent process and child process.
-The same file descriptors are open in both processes at the same numbers,
-and if new file descriptors are opened in either process,
-they are also visible in the other process.
-The file descriptor table is unshared once the child calls \texttt{execve}.
-This is simple to model and convenient for many purposes.
-
-If \verb|CLONE_FILES| is not passed to \texttt{clone},
-then \texttt{clone} has the same behavior as \texttt{fork}:
-The new process has a new file descriptor table,
+\texttt{clone} (without \verb|CLONE_FILES|) and \texttt{fork}
+deal with file descriptor tables in the same way:
+The new child process has a new file descriptor table,
 containing copies of all the file descriptors existing in the parent at the time of the system call.
@@
 
-This same behavior can also be triggered after process creation by calling \verb|unshare(CLONE_FILES)| or \texttt{execve}.
-If \verb|unshare(CLONE_FILES)|
-is called in a process currently sharing its file descriptor table with another process,
-then after the call that process will have a new, private file descriptor table,
-again with a copy of all the file descriptors existing at the time of the system call.
-
-If \texttt{execve} is called,
-the treatment of the file descriptor table is the same as \verb|unshare(CLONE_FILES)|,
-except that file descriptors with the \texttt{CLOEXEC} flag set on them
-are not copied into the newly created file descriptor table.
-We'll discuss the semantics of \texttt{CLOEXEC} more in section \ref{cloexec}.
+As demonstrated in section \ref{shared_fd_table},
+when the \verb|CLONE_FILES| flag is passed to \texttt{clone},
+the parent process and the child process will share a single file descriptor table,
+so changes to the file descriptor table made in one process will affect the other.
+
+The file descriptor table can be unshared by calling \verb|unshare(CLONE_FILES)| or \texttt{execve}.
+After either system call, the process will have a new file descriptor table
+which is no longer shared with any other process.
+All the file descriptors existing in the old table at the time of the system call
+will be copied into the new file descriptor table,
+except for, in the case of \texttt{execve}, file descriptors with the \texttt{CLOEXEC} flag set.
 \subsubsection{Tracking file descriptors through handles}
-After a system call has copied all the file descriptors in the old table into the new table,
+After \texttt{clone}, \verb|unshare(CLONE_FILES)|, or \texttt{execve}
+has copied all the file descriptors from some old table into a new table,
 we need to decide which file descriptors we want to keep open in the new table,
@@
 
-This is a question of resource management.
-In a language with manual resource management,
-such as C,
-this would be delegated to the user,
-and the responsibilities of the \texttt{rsyscall} library would end there.
-Manual resource management in C has led to many bugs,
-many of which with significant security implications.
-
-Python, instead, has automatic resource management,
-and therefore we determine which file descriptors should stay open automatically:
-if a certain file descriptor is in use by the program,
+In general, if a certain file descriptor is in use by the program,
 through a certain process,
@@
 each handle is associated with a process and contains a specific file descriptor number.
-For a handle to remain valid,
-that file descriptor number must continue to refer to an open file descriptor in that process.
-We determine which handles are live (in use by the program)
-by recourse to the native Python garbage collection system.
-The file descriptors we keep open are then those which are necessary to keep the live handles valid.
+A handle is ``valid''
+when that file descriptor number refers to an open file descriptor in that process.
+A handle is ``live''
+when the handle is in use by the program, which we can determine
+from Python's garbage collection system.
+We then keep open exactly those file descriptors which are necessary to keep all live handles, valid.
 
@@
 
-When a process is newly created,
-no file descriptor handles yet exist.
-New handles can be created by opening files in the process,
-or by using the \verb|inherit_fd| function to create handles for inherited file descriptors.
+When a child process is created,
+there are no file descriptor handles associated with it.
+We can use the \verb|inherit_fd| function to create handles for inherited file descriptors.
 \verb|inherit_fd| takes a file descriptor handle associated with the parent process,
@@
 Once the inherited file descriptors are closed as described in section \ref{cloexec},
-\verb|inherit_fd| can't be used again until the process next changes file descriptor table.
+\verb|inherit_fd| can no longer be used.
 \subsubsection{Closing file descriptors after inheritance}\label{cloexec}
 After creating a new file descriptor table,
-and establishing which file descriptors should be kept open,
+and determining which file descriptors should be kept open,
 we must promptly close other file descriptors.
@@
 
-We ensure that errors are reported back to the main program,
-even without handling them;
+We ensure that errors are reported back to the main program;
 this is a significant source of boilerplate in fork,
 because reporting an error back to the main program requires IPC.
-Even if there's no desire to handle errors,
-proper error reporting is necessary for debugging.
-And for serious users, proper error reporting and handling
-is essential for a robust application.
 \subsection{Clone versus fork}\label{microbench}
@@
 while \verb|clone(CLONE_VM)| stays flat at 21.1 microseconds.
+\texttt{clone} without \verb|CLONE_VM| is comparable to fork.
 
@@
  \includegraphics[width=0.5\textwidth]{subprocess_bench}
- \caption{Python subprocess and direct-style clone performance under varying memory usage}
+ \caption{Python spawn-style vs direct-style performance under varying memory usage}
  \label{fig:subprocess_bench}
@@
 
-We create a child process running \texttt{true} and wait for it to exit
+We create a child process, exec the \texttt{true} binary, and wait for it to exit
 using both the Python standard library's \texttt{subprocess.run},
-and direct-style \texttt{clone} with \texttt{rsyscall}.
+and direct-style with \texttt{rsyscall}.
 We benchmark the time each implementation takes
@@
 
-The direct-style \texttt{clone} implementation stays around 2.2 milliseconds regardless of memory usage.
+The direct-style implementation stays around 2.2 milliseconds regardless of memory usage.
 \texttt{subprocess.run}, which uses \texttt{fork},
@@
 and then scales linearly with the amount of memory mapped, as expected.
-Direct-style \texttt{clone} starts being faster than \texttt{subprocess.run}
+Direct-style starts being faster than \texttt{subprocess.run}
 at about 40 megabytes of memory mapped.
@@
 \subsection{Direct-style process creation}
+Baumann 2019\cite{forkroad} briefly suggests cross-process operations as a possible replacement for fork on Unix,
+positing many of the same advantages we find in practice.
+This work, though developed independently and largely completed before that paper,
+is in many ways an elaboration on and implementation of that idea.
+
 As we discuss in section \ref{background},
@@
 but others may wish to explore porting these techniques to other operating systems.
+% TODO maybe we could get rid of this paragraph?
 \subsection{Large scale open source usage}

### 2020-05-21 python/rsyscall/sys/syscall.py
unistd.credentials: add getpid syscall

@@ -36,6 +36,7 @@ class SYS(enum.IntEnum):
     getgid = lib.SYS_getgid
     getpeername = lib.SYS_getpeername
     getpgid = lib.SYS_getpgid
+    getpid = lib.SYS_getpid
     getpriority = lib.SYS_getpriority
     getsockname = lib.SYS_getsockname
     getsockopt = lib.SYS_getsockopt

### 2020-05-21 python/rsyscall/unistd/credentials.py

@@ -15,6 +15,9 @@ class CredentialsTask(rsyscall.far.Task):
     async def getgid(self) -> int:
         return (await _getgid(self.sysif))
 
+    async def getpid(self) -> near.Process:
+        return (await _getpid(self.sysif))
+
     async def getpgid(self) -> near.ProcessGroup:
         return (await _getpgid(self.sysif, None))
 
@@ -40,6 +43,9 @@ async def _getuid(sysif: SyscallInterface) -> int:
 
 async def _getgid(sysif: SyscallInterface) -> int:
     return (await sysif.syscall(SYS.getgid))
+
+async def _getpid(sysif: SyscallInterface) -> near.Process:
+    return near.Process(await sysif.syscall(SYS.getpid))
 
 async def _getpgid(sysif: SyscallInterface, pid: t.Optional[near.Process]) -> near.ProcessGroup:
     if pid is None:

### 2020-05-21 research/paper/paper.tex
paper

@@
 The \texttt{rsyscall} project develops language-specific ``process-independent'' system call libraries for Linux,
-which replace libc system call wrappers, and allow a program to perform cross-process system calls.
-\texttt{rsyscall} can work entirely in userspace,
+which replace libc system call wrappers,
+and allow a program to perform cross-process system calls.
+\texttt{rsyscall} works entirely in userspace
 and can be deployed on existing Linux systems without kernel modifications.
@@
 
-\lstinputlisting[float=*ht,language=Python,label={basic},caption={Creating a new process, changing CWD, and execing}]
+\lstinputlisting[float=ht,language=Python,label={basic},caption={Creating a new process, changing CWD, and execing}]
 {examples/basic/direct.py}
@@
 We call \texttt{chdir} in the child process to change its working directory to \texttt{/dev}.
-Then we call \texttt{execve},
+Then we call \texttt{execve} in the child,
 passing the path of the \texttt{cat} executable,
@@
 For concision, we assume in these examples that we are running with sufficient permissions in these examples,
-but these examples can all be run without privileges with appropriate use of \texttt{CLONE.NEWUSER}.
+but these examples can all be run without privileges with appropriate use of \verb|CLONE_NEWUSER|.
 
@@
 the precise details and type of the resource are not available to the child.
-Thus, this is a form of abstraction,
+This is a form of abstraction,
 and we can know ``for free''\cite{theoremsforfree}
@@
 will not rely on the details of how the file descriptor was created,
-such as the filename.
+such as the filename that was opened or the hostname that was connected to.
 
@@
 In Listing \ref{fds},
-we first open a file with read-write permission.
+we first open a file with read-write permission in the local process.
 Then we create a child process,
@@
 which on Linux we can do by clearing the FD flags with \verb|fcntl(fd, F_SETFD, 0)|.
-Finally, we execute some executable,
+Finally, we execute some executable in the child process,
 passing the file descriptor number as an argument;
@@
 connected by pipes.
-We first create two pipes,
-then inherit them down to several processes using a helper function, \verb|as_argument|,
+We first create two pipes in the local process,
+then inherit them down to several child processes using a helper function, \verb|as_argument|,
 which uses \verb|inherit_fd| and \texttt{fcntl} as described in section \ref{fd_abstraction}.
-We exec each program passing some file descriptors as arguments.
+We call exec in each child process, passing some file descriptors as arguments to the new programs.
 The whole system then runs in parallel,
@@
 we create a new child process,
-again in a new mount namespace using \texttt{CLONE.NEWNS}.
+again in a new mount namespace using \verb|CLONE_NEWNS|.
 We open the executable that we'll later exec in this child process,
@@
 file descriptor passing over a Unix domain socket
-is our only option to do something like this.
+is our only option.
 \section{Implementation}\label{implementation}
@@
 and then focus on implementation issues specific to process creation.
-\subsection{Cross-process syscalls}
+
 \texttt{rsyscall} can be conceptually divided in two parts:
@@
 The most simple way to access memory is for the local process to be in the same address space as the target process.
-This is the case most of the time; we pass \texttt{CLONE.VM} to \texttt{clone} by default in the Python wrapper.
+This is the case most of the time; we pass \verb|CLONE_VM| to \texttt{clone} by default in the Python wrapper.
 
@@
 
-\texttt{clone} creates a new process
+The raw \texttt{clone} system call creates a new process
 which immediately starts executing at the next instruction after the syscall instruction,
@@
 
-\texttt{clone} lets us change the stack for the new process;
-we use this to call an arbitrary function address stored on the stack.
+\texttt{clone} lets us change the stack for the new process.
+We can use this to make the new process call an arbitrary function,
+by storing its address on the new stack.
 Further arguments can also be passed on the stack,
@@
 Most system calls work as normal;
-the new process can be modified freely through \texttt{chdir}, \texttt{dup2}, and other system calls.
+the new child process can be modified freely through \texttt{chdir}, \texttt{dup2}, and other system calls.
 From the system calls related to process creation,
@@
 We need a way to determine if \texttt{execve} is successful;
-naively waiting for a response to the syscall request will leave us waiting forever.
+naively waiting for a response to the syscall request may leave us waiting forever.
 
@@
 and in this way get notified of the child process calling \texttt{execve} or exiting.
-
 A process can only wait on one futex at a time;
@@
 \section{Evaluation}\label{evaluation}
-We answer the following questions:
+Our goal is a powerful approach to process creation that is easy to use correctly.
+Performance of our implementation is secondary,
+but the overhead of our implementation must be low
+relative to alternative approaches.
+
+We evaluate the degree to which we meet this goal by answering the following questions:
 \begin{compactitem}
@@
   Does direct-style process creation on Linux allow for simpler programs than the alternatives?
-  We implement and compare similar functionality using both direct-style and fork-style in \ref{ease}.
+  We compare implementations of similar functionality using both direct-style and fork-style in \ref{ease},
+  and find substantial benefits for direct-style.
 \item
-  Does a clone-based implementation avoid the well-known performance overheads of fork?
-  We evaluate clone against fork on several microbenchmarks in \ref{microbench}.
+  Does our \texttt{rsyscall}-based Python direct-style process creation interface
+  have acceptable performance relative to the Python standard library's spawn-style \texttt{subprocess.run}?
+  We benchmark direct-style process creation against spawn-style in \ref{subprocess_bench},
+  and find an acceptable level of overhead.
 \item
-  Does a high-level Python direct-style process creation interface
-  have acceptable performance relative to the Python standard library's spawn-style interface?
-  We benchmark direct-style process creation against spawn-style in \ref{subprocess_bench}.
-\item
-  Does direct-style process creation work in the ``real world''?
-  We discuss our experience with using direct-style process creation at \twosigma in \ref{realworld}.
+  Does direct-style process creation perform well in the ``real world''?
+  We discuss our positive experience with using direct-style process creation at \twosigma in \ref{realworld}.
 \end{compactitem}
@@
 \end{table}
-We implement a variety of simple but powerful techniques with processes
-using both direct-style and fork.
-Our results are summarized in table \ref{tab:ease}.
-Direct-style consistently takes under half the lines the express the same thing.
-\lstinputlisting[float=*ht,language=Python,label={basic_fork},caption={Fork-style: Creating a new process, changing CWD, and execing}]
+Does direct-style process creation on Linux allow for simpler programs than the alternatives?
+To answer this, we compare the direct-style examples shown in section \ref{examples}
+to programs with the same behavior implemented in fork-style.
+Most of the examples cannot be implemented with typical Linux spawn-style interfaces such as \verb|posix_spawn|,
+so we compare only to fork-style.
+
+The line counts of the direct-style and fork-style implementations are listed in table \ref{tab:ease}.
+Direct-style consistently takes under half the lines to express the same functionality.
+
+\lstinputlisting[float=ht,language=Python,label={basic_fork},caption={Fork-style: Creating a new process, changing CWD, and execing}]
 {examples/basic/fork.py}
-An example of one of the fork-style implementations is in listing \ref{basic_fork}.
-It corresponds to the direct-style example in listing \ref{basic}.
+Listing \ref{basic_fork} gives an example of one of our fork-style implementations,
+corresponding to the direct-style example in listing \ref{basic}.
 
@@
 
-We ensure that errors are reported back to the main program;
-this is a significant source of boilerplate in fork,
-because reporting an error back to the main program requires IPC.
-\subsection{Clone versus fork}\label{microbench}
-\begin{figure}[h!]
-\centering
- \includegraphics[width=0.5\textwidth]{microbench}
- \caption{fork and clone performance under varying memory usage}
- \label{fig:microbench}
-\end{figure}
-
-We run a benchmark which times how long process creation takes
-with \texttt{fork}, \texttt{clone}, and \verb|clone(CLONE_VM)|
-with varying amounts of anonymous memory mapped in the parent process
-before running the benchmark.
-Note that \texttt{clone} without \verb|CLONE_VM| is expected to function identically to fork.
-
-All benchmarks are run pinned to an isolated core,
-running on an Intel i9-9900K CPU with hyperthreading enabled.
-Our benchmarks are intended to give a sense of relative, not absolute, performance,
-and we have no reason to believe a more controlled benchmarking setup
-would have differential effects on the performance of \texttt{fork} and \texttt{clone}.
-
-The results are in figure \ref{fig:microbench}.
-Without \verb|CLONE_VM|,
-process creation time is linear in the amount of memory mapped in the parent process.
-At the upper end of our benchmark, with 4 GB of memory mapped ($10^6$ pages),
-\texttt{fork} takes an average of 42.6 milliseconds to create a process,
-while \verb|clone(CLONE_VM)| stays flat at 21.1 microseconds.
-\texttt{clone} without \verb|CLONE_VM| is comparable to fork.
-
-This is because \texttt{fork}, and \texttt{clone} without \verb|CLONE_VM|,
-must set up memory mappings in the child process for all the memory mapped in the parent process.
-\verb|clone(CLONE_VM)| does not have to do this bookkeeping,
-so its process creation time is constant in the amount of memory used in the parent process.
-\subsection{High-level direct-style versus spawn-style}\label{subprocess_bench}
+The main source of additional code in the fork-style implementation
+is reporting errors back to the main program using IPC.
+If a program that is already reporting errors to some other destination,
+such code is not necessary;
+for example, a process supervisor or a shell might write process startup errors
+as textual logs which are read by a human.
+If a program does not already have such a destination for errors,
+then robustly dealing with errors in \texttt{fork} requires this additional IPC code.
+
+Direct-style does not require any extra work to report errors;
+errors are reported just like any other system call.
+\subsection{Microbenchmark results}\label{subprocess_bench}
 \begin{figure}[h!]
@@
 \end{figure}
-
-We create a child process, exec the \texttt{true} binary, and wait for it to exit
-using both the Python standard library's \texttt{subprocess.run},
-and direct-style with \texttt{rsyscall}.
-We benchmark the time each implementation takes
-with varying amounts of mapped memory, as in section \ref{microbench}.
-Both are running under CPython 3.7.7,
-again pinned to an isolated core as with section \ref{microbench}.
+Does our \texttt{rsyscall}-based Python direct-style \texttt{clone}
+have acceptable performance relative to the Python standard library's spawn-style \texttt{subprocess.run}?
+
+We first evaluate a simple process creation workload:
+create a child process, exec the \texttt{true} binary, and wait for it to exit.
+We implement this using both direct-style \texttt{clone}
+and the standard library's \texttt{subprocess.run}.
+We benchmark both implementations under CPython 3.7.7, on Linux 4.19.93,
+pinned to an isolated single core on an Intel i9-9900K CPU with 60GB of RAM.
+We vary the amount of anonymous memory mapped in the parent process
+to demonstrate how each implementation scales with memory usage.
 The results are in Figure \ref{fig:subprocess_bench}.
@@
 starts at 1.4 milliseconds,
-and then scales linearly with the amount of memory mapped, as expected.
-Direct-style starts being faster than \texttt{subprocess.run}
-at about 40 megabytes of memory mapped.
-
-These results are acceptable.
+and then scales linearly with the amount of memory mapped, as expected,
+up to 48.5 milliseconds at 4 GBs of mapped memory.
+The cross-over point for this micro-benchmark is about 40 megabytes of memory mapped.
+
+We perform another microbenchmark to evaluate the overhead of modifying the child process.
+We simply run \texttt{getpid} in the child process on the same benchmark setup;
+the average time per \texttt{getpid} call is .561 milliseconds.
+The \texttt{getpid} syscall returns essentially instantly,
+so we can regard this time as the amount of overhead incurred
+by performing our child process setup system calls through \texttt{rsyscall}.
+Through profiling, we've found that most of this per-syscall overhead is spent in Python,
+rather than waiting for syscall responses from the child task.
+
+These slowdowns in process creation and modification are substantial,
+but we found that this overhead is acceptable in practice.
 Process creation in Python is already slow, taking milliseconds of time,
@@
 
-Furthermore,
-our implementation of \texttt{rsyscall} is barely-optimized pure-Python,
-all the way down the syscall primitive,
-whereas \texttt{subprocess.run} is entirely in C.
-If we need to speed up our implementation,
-we can perform some targeted optimizations,
-reimplement some parts of the Python \texttt{rsyscall} library in native code,
-or run our code using a faster Python interpreter such as PyPy.
+Furthermore, as we'll discuss in section \ref{realworld},
+we've found that the greater expressivity of direct-style
+provides for programs that are more efficient on a large scale,
+which makes up for the performance cost in micro-benchmarks.
+We've also found that, for many interesting applications,
+the performance overhead of direct-style process creation (or Python, for that matter)
+is dwarfed by the execution time of the native-code programs we ultimately run.
+
+As a result, though implementations in native code and in the kernel would likely remove most of this overhead,
+we have chosen to not invest effort into optimizing process creation at this micro-level,
+to preserve implementation simplicity;
+such optimization is reserved for future work.
 \subsection{Usage in the real world}\label{realworld}
 At \twosigma we have used \texttt{rsyscall} and direct-style process creation
-as a component of a distributed system deployment library.
+as a component of an internal distributed system deployment library, written in Python.
 This library is in use in production,
@@
 
-With direct-style process creation,
-we can robustly and easily write custom code to start up each program,
-passing down file descriptors and initializing process attributes according to each program's needs,
-all of which can be done directly from a single parent process.
-This has allowed us to make more use of file descriptor abstraction;
-for example, connections over our shared memory transport are bootstrapped over passed-down file descriptors,
-and our management interfaces listen on passed-in sockets,
-allowing them to be easily swapped between network-accessible TCP sockets and host-local Unix domain sockets.
-We can also flexibly use namespaces where appropriate;
-for example, for legacy applications which directly start subprocesses,
-we prevent process leaking with pid namespaces.
-
-All this has allowed us to run our systems in a much more dynamic way than before;
-we can freely start up arbitrary subsets of our systems on-demand,
+Each service deployed by this library is started with custom code,
+written using direct-style process creation.
+This allows us to pass down file descriptors and initialize process attributes
+according to each program's needs;
+which in turn has let us make optimizations which previously required too much program-specific customization.
+
+For example,
+we are able to bootstrap connections over our internal shared memory transports over passed-down file descriptors,
+and our management interfaces listen on passed-down sockets.
+This allows substantial parallelism in the startup of services.\cite{socketactivation}
+Taking one representative example,
+one test built with an old library took an average of 221 seconds to run,
+mostly spent in system startup time as processes started up serially.
+With parallelized system startup, the same test takes an average of 16 seconds.
+
+We've found that programmers
+are able to use file descriptor passing as a normal feature of their development process with relative ease.
+Programmers without substantial prior experience with Unix programming
+are able to write code to support new services
+nd pass down file descriptors to the program without encountering significant issues.
+
+We also flexibly use namespaces where appropriate.
+For example, for legacy applications which directly start subprocesses,
+we prevent process leaking with a pid namespace.
+
+All this has allowed us to run our systems in a much more dynamic way than before.
+We can freely start up arbitrary subsets of our systems on-demand,
 spread across one or more hosts,
@@
 such as container runtimes.
-This has become a key part of our new development workflow.
+This has become a key part of our development workflow.
 \section{Related work}\label{related_work}
@@
 positing many of the same advantages we find in practice.
-This work, though developed independently and largely completed before that paper,
-is in many ways an elaboration on and implementation of that idea.
+Our work, though developed independently and largely completed before that paper,
+is in many ways an elaboration on and implementation of that idea,
+to which we owe a great debt.
 
@@
 \bibliographystyle{plain}
-\bibliography{bibliography}
+% \bibliography{bibliography}
 \end{document}

### 2020-05-22 research/paper/paper.tex
paper

@@
 Performance of our implementation is secondary,
-but the overhead of our implementation must be low
-relative to alternative approaches.
+but the overhead of our implementation relative to alternative approaches
+must be low.
 
@@
 \item
-  Does our \texttt{rsyscall}-based Python direct-style process creation interface
-  have acceptable performance relative to the Python standard library's spawn-style \texttt{subprocess.run}?
+  In Python, does our \texttt{rsyscall}-based direct-style process creation interface
+  have acceptable performance relative to the standard library's spawn-style \texttt{subprocess.run}?
   We benchmark direct-style process creation against spawn-style in \ref{subprocess_bench},
@@
 to programs with the same behavior implemented in fork-style.
+
 Most of the examples cannot be implemented with typical Linux spawn-style interfaces such as \verb|posix_spawn|,
@@
 errors are reported just like any other system call.
+This is the main immediately-visible simplification of using direct-style process creation.
+
+Direct-style also removes the need to be concerned about the state of the calling process,
+though this simplification is not immediately visible in a small example.
+The fork-style examples may break if there are multiple threads in the calling process,
+and may be slow if the calling process has large amounts of memory mapped, as we'll see in section \ref{subprocess_bench}.
+The direct-style examples do not have these limitations.
 \subsection{Microbenchmark results}\label{subprocess_bench}
@@
 \end{figure}
-Does our \texttt{rsyscall}-based Python direct-style \texttt{clone}
-have acceptable performance relative to the Python standard library's spawn-style \texttt{subprocess.run}?
+In Python, does our \texttt{rsyscall}-based direct-style process creation interface
+have acceptable performance relative to the standard library's spawn-style \texttt{subprocess.run}?
 
@@
 The direct-style implementation stays around 2.2 milliseconds regardless of memory usage.
-\texttt{subprocess.run}, which uses \texttt{fork},
+\texttt{subprocess.run}, which is implemented with \texttt{fork},
 starts at 1.4 milliseconds,
@@
 We perform another microbenchmark to evaluate the overhead of modifying the child process.
-We simply run \texttt{getpid} in the child process on the same benchmark setup;
-the average time per \texttt{getpid} call is .561 milliseconds.
-The \texttt{getpid} syscall returns essentially instantly,
-so we can regard this time as the amount of overhead incurred
+We simply call \texttt{getpid} from Python
+in both the child process and the parent process on the same benchmark setup.
+The average time per \texttt{getpid} call is
+3 microseconds when called in the parent process without going through \texttt{rsyscall},
+and 561 miroseconds when called through \texttt{rsyscall} in the child process.
+The difference, 558 microseconds,
+is the amount of overhead incurred
 by performing our child process setup system calls through \texttt{rsyscall}.
@@
 positing many of the same advantages we find in practice.
-Our work, though developed independently and largely completed before that paper,
+Our work, though developed independently,
 is in many ways an elaboration on and implementation of that idea,
@@
 \bibliographystyle{plain}
-% \bibliography{bibliography}
+\bibliography{bibliography}
 \end{document}

### 2020-05-23 research/paper/paper.tex
paper

@@
 such optimization is reserved for future work.
+
+\subsection{Microbenchmarks for nested processes}\label{clone_bench}
+\begin{figure}[h!]
+\centering
+ \includegraphics[width=0.5\textwidth]{clone_bench}
+ \caption{Time to create different processes, by variety of parent process}
+ \label{fig:clone_bench}
+\end{figure}
+
+As demonstrated in sections \ref{pidns} and \ref{fuse},
+\texttt{rsyscall} supports nested operation.
+That is, we can spawn a child from one of our children;
+this can impact performance.
+\texttt{clone} also supports specifying a variety of flags to create new namespaces when creating the child,
+which can also impact performance.
+
+We've benchmarked process creation using the same benchmark as in section \ref{subprocess_bench}
+from a variety of parent processes: the local process,
+a child (created with \texttt{clone()}),
+a grandchild (created with \texttt{clone().clone()}),
+or a child in a different namespace (created with \verb=clone(CLONE_NEWNS|CLONE_NEWPID)=).
+We've also specified different combinations of flags for the process being created in the benchmark.
+The results are in figure \ref{fig:clone_bench}.
+We don't include the Python standard library in these benchmarks
+as there's no way to specify \texttt{CLONE} flags in the Python stdlib.
+
+Process creation in a child has a significant performance impact,
+primarily because of the overhead of performing syscalls in a child.
+Performance of process creation in a grandchild is the same as in a child,
+because there is no additional overhead for syscalls in a grandchild relative to a child.
+Benchmarking \texttt{getpid} in a grandchild
+confirms that the average time per \texttt{getpid} call is still 561 microseconds on our benchmark setup.
+
+The use of namespaces also has a significant impact,
+both when the parent process is in a namespace, and when the child process is created in a namespace.
+We perform the benchmark with no clone flags, and with the common clone flags
+\verb|CLONE_NEWNS| and \verb|CLONE_NEWPID|, first on their own and then together.
+This benchmark essentially measures the performance of the Linux namespaces system, not the work in this paper,
+but it is illustrative to compare it to the performance of \texttt{rsyscall}.
+Note that the cost of creating a new child process from a parent process that is already inside a namespace
+(the difference between the \texttt{clone()} and \texttt{clone(NS|PID)} parent processes)
+is around 2 milliseconds, comparable to the cost of running with \texttt{rsyscall}.
+Note also that the overhead of creating a new process inside a new namespace
+(the difference between the \texttt{clone()} and \texttt{clone(NS|PID)} child processes)
+is around 15 milliseconds; subtantially greater than the overhead of \texttt{rsyscall}.
 \subsection{Usage in the real world}\label{realworld}
 At \twosigma we have used \texttt{rsyscall} and direct-style process creation
-as a component of an internal distributed system deployment library, written in Python.
-This library is in use in production,
+to implement an internal distributed system deployment library, written in Python,
+which we'll refer to here as Toplevel.
+Toplevel is in use in production,
 and is extensively used as part of testing and development.
-
-Each service deployed by this library is started with custom code,
-written using direct-style process creation.
-This allows us to pass down file descriptors and initialize process attributes
-according to each program's needs;
-which in turn has let us make optimizations which previously required too much program-specific customization.
-
-For example,
-we are able to bootstrap connections over our internal shared memory transports over passed-down file descriptors,
-and our management interfaces listen on passed-down sockets.
-This allows substantial parallelism in the startup of services.\cite{socketactivation}
-Taking one representative example,
-one test built with an old library took an average of 221 seconds to run,
-mostly spent in system startup time as processes started up serially.
-With parallelized system startup, the same test takes an average of 16 seconds.
-
-We've found that programmers
-are able to use file descriptor passing as a normal feature of their development process with relative ease.
-Programmers without substantial prior experience with Unix programming
-are able to write code to support new services
-nd pass down file descriptors to the program without encountering significant issues.
-
-We also flexibly use namespaces where appropriate.
-For example, for legacy applications which directly start subprocesses,
+Toplevel consists of a collection of functions and modules
+for each component in our system,
+which start up components using direct-style process creation,
+all from a single Python parent process.
+
+One of the major benefits of direct-style process creation has been easy use of file descriptor inheritance.
+Our experience has been that with fork-style or spawn-style interfaces,
+all file descriptor logic must be centralized,
+whereas file descriptors in a new process can be built up over time using direct-style,
+simplifying the implementation.
+We've found that even programmers without substantial prior experience with Unix programming
+are able to use file descriptor passing as a normal feature of their development process with relative ease,
+writing code to support new programs and passing down file descriptors to those programs without issue.
+
+Easy file descriptor passing has, in turn, allowed us to heavily use socket activation techniques
+to start up services in parallel,
+substantially speeding up system startup.\cite{socketactivation}
+We are able to bootstrap connections over our internal shared memory transports
+over pre-created, passed-down file descriptors,
+and our management interfaces listen on passed-down sockets;
+this allows starting up services in parallel, rather than in dependency order.
+Benchmarking one test built with an old library and its equivalent built with Toplevel,
+the old test took an average of 221 seconds to run,
+mostly spent in system startup,
+while the new test takes an average of 16 seconds.
+
+Easy file descriptor inheritance has also allowed us to treat file descriptors as a uniform interface in Toplevel.
+We pass around file descriptors freely inside Toplevel.
+A given component can implemented with a separate process which has the file descriptor passed down,
+running a program written in native code for performance,
+or implemented in-process with a Python implementation for flexibility;
+either is possible behind a uniform file descriptor interface.
+
+We've also been able to create complicated pipelines in Toplevel,
+as in section \ref{pipe}.
+
+We've also made use of namespaces,
+chiefly user namespaces and pid namespaces.
+For legacy applications which directly start subprocesses,
 we prevent process leaking with a pid namespace.
+This is part of strategy of noninterference and cleanup
+that ensures that multiple users of Toplevel
+can coexist simultaneously on the same host,
+even if running the same system.
+
+Toplevel also makes uses of other features of \texttt{rsyscall} which are not covered in this paper.
+These include the ability to operate on remote hosts through capabilities for remote processes,
+and the ability to perform system calls in parallel across a pool of processes.
+These have significant synergy with direct-style process creation
+in forming an overall powerful system.
 

### 2020-05-23 research/paper/paper.tex
paper!!

@@
 \usepackage{tikz}
+\usepackage{dblfloatfix}
 \usepackage{amsmath}
@@
 \usepackage{titlesec}
+
+% Ugh, these aren't the same size right now.
+\newcommand{\tbad}{\xmark:}
+\newcommand{\tgood}{\cmark:}
+% \newcommand{\tbad}{\text{\sffamily X} :}
+% \newcommand{\tgood}{\checkmark:}
 
@@
 \section{Introduction}\label{introduction}
-Existing process creation interfaces can be divided into three categories:
-Fork-style, spawn-style, and direct-style.
-
-Fork-style (such as \texttt{fork}) and spawn-style (such as \texttt{posix\_spawn}) are the most popular,
-but have a number of disadvantages.
-Correct, performant use of fork-style places significant constraints on the calling process,
-and also has a complex programming model.
-Spawn-style lifts many of these constraints, and is simpler to program against,
-but is fundamentally limited in the kinds of processes it can create.
-With both fork-style and spawn-style,
-it is difficult to
-branch based on results or errors encountered while initializing the new process:
-fork-style requires IPC if it wishes to communicate errors back to the rest of the program,
-and robust indication of errors in spawn-style requires substantial expansion of the spawn interface.
+\begin{table*}
+\begin{tabular}{l|l|l|l}
+% BEGIN RECEIVE ORGTBL compare
+ & fork & spawn & direct-style\\
+\hline
+Requirements on parent process & \tbad Single-threaded, small memory & \tgood None & \tgood None\\
+Programming model & \tbad Complex (returns twice) & \tgood Simple (single call) & \tgood Simple (imperative)\\
+Maximally powerful & \tgood Yes, can call any syscall & \tbad No, limited interface & \tgood Yes, can call any syscall\\
+Reporting of results & \tbad Requires IPC & \tbad Not fine-grained & \tgood From individual child syscalls\\
+Non-inherited attributes & Mutated by code in child & Set by arguments & Mutated by code in parent\\
+% END RECEIVE ORGTBL compare
+\end{tabular}
+\caption{Features of fork-style vs spawn-style vs direct-style}
+\label{tab:styles}
+\end{table*}
+\begin{comment}
+#+ORGTBL: SEND compare orgtbl-to-latex :splice t
+|                                | fork                                | spawn                       | direct-style                          |
+|--------------------------------+-------------------------------------+-----------------------------+---------------------------------------|
+| Requirements on parent process | \tbad Single-threaded, small memory | \tgood None                 | \tgood None                           |
+| Programming model              | \tbad Complex (returns twice)       | \tgood Simple (single call) | \tgood Simple (imperative)            |
+| Maximally powerful             | \tgood Yes, can call any syscall    | \tbad No, limited interface | \tgood Yes, can call any syscall      |
+| Reporting of results           | \tbad Requires IPC                  | \tbad Not fine-grained      | \tgood From individual child syscalls |
+| Non-inherited attributes       | Mutated by code in child            | Set by arguments            | Mutated by code in parent             |
+% $
+\end{comment}
+The most well-known process creation interface on Linux, and Unix in general, is \texttt{fork}.
+When a parent process creates a child process through \texttt{fork},
+or other similar fork-style interface,
+both processes execute concurrently,
+and the child process typically inherits all but a few attributes from the parent process.\cite{tlpi}
+The child process can call arbitrary system calls
+to mutate the new process until it reaches the desired state,
+typically ultimately calling \texttt{exec}.
+
+A user of fork must think carefully about the characteristics of the process from which fork is being called.
+Fork can have poor performance when called from processes with many memory mappings.\cite{forkroad}
+This can be mitigated to some degree, but inflicts further complexity on the caller.\cite{vfork}\cite{vfork_dangerous}
+Multi-threaded programs generally cannot use fork safely.\cite{forkroad}\cite{tlpi}
+Some thread libraries provide partial mitigations for multi-threading issues,
+but it's up to user code to make use of those mitigations.\cite{pthread_atfork}
+
+Since the child process runs concurrently with the parent process,
+communicating events in the child back to the parent requires some form of IPC,
+complicating process creation.\cite{posix_spawn_error_pipe}
+Even the basic interface of fork - one function call which returns twice - is unusual;
+such functions are generally considered to be complex, advanced topics in other settings.
+\cite{callcc_fork}\cite{continuations}\cite{bashcc}
+
+In a spawn-style interface,
+details about the new process are provided up-front as arguments to some function
+which creates the new process all at once;
+\texttt{posix\_spawn} is a typical example of this style.\cite{posix_spawn}\cite{create_process}\cite{chainloading}
+Other details, such as security context, are inherited from the parent process.\cite{tlpi}\cite{create_process}
+Spawn-style APIs
+can be called from any parent process without concerns about memory usage or multi-threading,
+when used correctly.
+\cite{forkroad}\cite{posix_spawn_error_pipe}
+
+Unlike \texttt{fork}, a spawn-style interface provides an API for process creation
+which is independent from existing system calls.\cite{spawn_h}\cite{subprocess_run}
+Spawn-style interfaces therefore often lack support for some features,
+and they also require learning a new interface distinct from the normal system call API,
+both of which encourage systems programmers to avoid them.\cite{tlpi}\cite{posix_spawn_lwn}
+Also, spawn-style interfaces, by their nature as a single call,
+don't allow programs to branch on the results of individual modifications to the new process,
+and have complex, or substantially less information, error reporting for such modifications.
+\cite{posix_spawn}\cite{posix_spawn_error_pipe}
+
+Table \ref{tab:styles} summarizes the differences between fork-style and spawn-style.
+Each has its own advantages and disadvantages.
+
+A few academic operating systems - KeyKOS, seL4, and some others, for example -
+use another style of process creation.\cite{keykos}\cite{sel4}\cite{exokernel}\cite{fuschia}\cite{singularity}
+We use the new term "direct-style" to refer to this,
+since we know of no pre-existing generic term for this style of process creation.
 
@@
 
-Our contribution is an implementation of direct-style process creation for Linux,
-making use of our work on the \texttt{rsyscall} project.
-The \texttt{rsyscall} project develops language-specific ``process-independent'' system call libraries for Linux,
-which replace libc system call wrappers,
-and allow a program to perform cross-process system calls.
-\texttt{rsyscall} works entirely in userspace
-and can be deployed on existing Linux systems without kernel modifications.
-\texttt{rsyscall} is currently focused on supporting, and is implemented in, Python,
+We'll discuss each style in more detail in section \ref{background}.
+
+To perform direct-style process creation on Linux,
+we need an API where we explicitly specify in which process we want to make a syscall.
+This differs from the normal mode of operation on Linux,
+where programs making syscalls implicitly operate on the current process.
+
+Our contribution in this paper is this API, through the \texttt{rsyscall} library.
+The \texttt{rsyscall} project develops
+language-specific, object-capability-model, low-abstraction libraries for the Linux syscall interface,
+bypassing libc.
+\texttt{rsyscall} is currently focused on supporting Python,
 but can be ported to other languages;
@@
 but generalize easily to other languages.
-
-Our direct-style process creation implementation uses \texttt{clone} to create a child process in an embryonic state
-which can then be manipulated from user code in the parent process using syscalls with \texttt{rsyscall}.
-A direct-style \texttt{clone} has all the advantages of direct-style process creation
-over Linux's fork-style \texttt{fork} and spawn-style \texttt{posix\_spawn}:
-\begin{compactitem}
-\item
-  Processes can be created correctly and performantly from any parent process,
-  with a straightforward, familiar API.
-\item
-  Processes can be created with the full set of Linux features.
-\item
-  Syscall results are reported individually to the parent process.
-  % rather than batched together as with spawn-style or manually reported via IPC as with fork-style.
-\end{compactitem}
-
-We've implemented a variety of useful applications using direct-style process creation on Linux,
-which we'll demonstrate in this paper.
-Direct-style \texttt{clone} has minimal performance cost in normal situations,
-and can outperform \texttt{fork} in some cases.
-We have deployed direct-style process creation in production internally at \twosigma.
-
-Our implementation is distributed as a feature of \texttt{rsyscall},
-and is immediately deployable, entirely in userspace, on recent Linux kernels.
 \texttt{rsyscall} is open source and available from
 [Github url].
-% \url{https://github.com/catern/rsyscall}.
-
-This paper is organized in several sections.
-\begin{compactitem}
-\item Section \ref{background} provides background on fork-style, spawn-style, and direct-style process creation,
-and considers their advantages and disadvantages in more depth.
-\item Section \ref{overview} discusses the design of the API,
-including our usage of \texttt{rsyscall}.
-\item Section \ref{examples} shows examples of direct-style process creation in Python on Linux,
+
+Unlike POSIX C libraries such as glibc or musl,
+an \texttt{rsyscall} library is organized based on the object-capability model.
+The capabilities to make syscalls in any given process are reified as language objects.
+If the capability to make syscalls in a specific process is not passed (in some way) to a function,
+then that function cannot make syscalls in that process.
+
+Due to this design, an \texttt{rsyscall} program can make use of multiple processes at once,
+by manipulating capabilities for multiple processes.
+The relevant two capabilities for this paper are the initial capability to the ``local'' process,
+and capabilities to child processes.
+The ``local'' process is the one which hosts the runtime for the running program,
+and in which a legacy libc would implicitly make syscalls.
+Every \texttt{rsyscall} program starts with the ``local'' process
+and uses it to bootstrap the initial capabilities for other processes.
+
+To perform direct-style process creation,
+we use \texttt{rsyscall} to call \texttt{clone} on the ``local'' process to create a new child process.
+We call various syscalls in the child process to mutate it until it reaches the desired state,
+at which point we can call \texttt{exec} in the child process to start it running on its own.
+After that, the process is no longer under our control, and it functions like any other child process.
+The child process can be monitored using normal Linux child monitoring syscalls,
+such as \texttt{waitid}.
+
+These child process capabilities are created and managed in userspace,
+by launching new processes running a syscall server,
+which receives syscall requests, performs the syscall, and returns the response.
+No kernel modifications are required, and \texttt{rsyscall} is immediately deployable on recent Linux kernel.
+We discuss the implementation in depth in section \ref{implementation},
+and examine several difficult process-creation related details.
+
+\lstinputlisting[float=h,language=Python,label={basic},caption={Create new process, change CWD, and exec}]
+{examples/basic/direct.py}
+Listing \ref{basic} shows a program using \texttt{rsyscall}.
+We create a new child process using \texttt{clone} on the local process \verb|local|,
+and receive back a capability to control it.
+We call \texttt{chdir} in the child process to change its working directory to \texttt{/dev}.
+Then we call \texttt{execv} in the child,
+passing the path of the \texttt{cat} executable
+and the arguments to the new executable.
+The name of the executable is passed as the first argument,
+as is typical in Unix.
+Calling \texttt{execv} consumes the capability,
+and later use of this child process capability will fail with an exception.
+
+The behavior of system calls made on a remote process through \texttt{rsyscall}
+is identical to the behavior of system calls made implicitly on the current process through libc.
+This is an API that systems programmers are already familiar with,
+and which is already thoroughly documented through Linux system call manpages.
+The novelty is constrained to the newly explicit process capabilities.
+Besides flattening the learning curve,
+this also ensures that direct-style \texttt{clone} will always be at least as powerful as \texttt{fork}.
+
+We show more examples of direct-style process creation in section \ref{examples},
 including use of process features which are difficult to use on Linux with \texttt{fork} or \texttt{posix\_spawn}.
-\item Section \ref{implementation} explores our implementation in depth,
-and discusses several difficult process-creation related details of \texttt{rsyscall}.'s implementation.
-\item Section \ref{evaluation} evaluates our results using both quantitative and qualitative approaches.
-\item Section \ref{related_work} examines other work which also touches on the topics covered by this paper.
-\item Section \ref{future_work} lists several directions for future work
-related to direct-style process creation and to \texttt{rsyscall}.
-\item Section \ref{conclusions} briefly outlines our conclusions about direct-style and the future of process creation.
-\end{compactitem}
+
+Direct-style \texttt{clone} has acceptable performance cost,
+and can outperform \texttt{fork} in some cases.
+The performance overhead of creating processes with direct-style clone
+is comparable to the performance cost of creating processes while inside a Linux namespace.
+At \twosigma, we have deployed direct-style process creation in production
+inside a library for distributed system deployment, Toplevel.
+We evaluate in further detail in section \ref{evaluation}.
 \section{Background on process-creation}\label{background}
@@
 
-% Ugh, these aren't the same size right now.
-\newcommand{\tbad}{\xmark:}
-\newcommand{\tgood}{\cmark:}
-% \newcommand{\tbad}{\text{\sffamily X} :}
-% \newcommand{\tgood}{\checkmark:}
-
-\begin{table*}
-\begin{tabular}{l|l|l|l}
-% BEGIN RECEIVE ORGTBL compare
- & fork-style & spawn-style & direct-style\\
-\hline
-Works with any caller & \tbad No threads or large memory & \tgood Yes & \tgood Yes\\
-Programming model & \tbad Complex (returns twice) & \tgood Simple (single call) & \tgood Simple (imperative)\\
-Maximally powerful & \tgood Yes, can call any syscall & \tbad No, limited interface & \tgood Can call any syscall\\
-Error handling & \tbad Poor, requires IPC & \tbad Poor, not fine-grained & \tgood Reports individual errors\\
-Non-copied attributes & Mutated by code in child & Set by arguments & Mutated by code in parent\\
-% END RECEIVE ORGTBL compare
-\end{tabular}
-\caption{Features of fork-style vs spawn-style vs direct-style}
-\label{tab:ease}
-\end{table*}
-\begin{comment}
-#+ORGTBL: SEND compare orgtbl-to-latex :splice t
-|                       | fork-style                       | spawn-style                  | direct-style                     |
-|-----------------------+----------------------------------+------------------------------+----------------------------------|
-| Works with any caller | \tbad No threads or large memory | \tgood Yes                   | \tgood Yes                       |
-| Programming model     | \tbad Complex (returns twice)    | \tgood Simple (single call)  | \tgood Simple (imperative)       |
-| Maximally powerful    | \tgood Yes, can call any syscall | \tbad No, limited interface  | \tgood Can call any syscall      |
-| Error handling        | \tbad Poor, requires IPC         | \tbad Poor, not fine-grained | \tgood Reports individual errors |
-| Non-copied attributes | Mutated by code in child         | Set by arguments             | Mutated by code in parent        |
-% $
-\end{comment}
 \subsection{Fork-style}
 By far the most well-known fork-style interface is Unix's fork.\cite{forkhist}
-With a fork-style interface,
-a process, called the parent process, calls \texttt{fork},
-and a copy of that process is made,
-called the child process,
-differing only in a few attributes such as process id.
-The same program continues executing in both processes,
-with a different return value from fork to allow it to distinguish whether it is running in the parent or the child.
-Typically, the program will branch based on the return value
-and, if running in the child,
-call various system calls to modify the state of the new process.
-
-Fork has a number of flaws,
-and much ink has been spilled detailing them.
-We'll cover the most egregious briefly here,
-but see Baumann 2019 \cite{forkroad} for a more detailed evaluation.
-
-Fork, when called successfully, returns twice,
-creating two different processes which execute concurrently.
-This immediately makes fork difficult to integrate;
-for example, errors can't be easily communicated between the two processes.
-If there is an error in the initialization of the child process,
-the part of the program running in the child process
-needs to use some form of inter-process communication (IPC) to communicate it back to the parent process.
-The child process can run arbitrary code to make decisions about its initialization,
-but that code needs to use IPC if it wishes to communicate with the rest of the program,
-which is still in the parent process.
-
+When a parent process creates a child process by fork,
+both processes execute concurrently,
+and the child process typically inherits all but a few attributes from the parent process.\cite{tlpi}
+The child process can call arbitrary system calls
+to mutate the new process until it reaches the desired state,
+typically ultimately calling \texttt{exec}.
+
+Since the child process runs concurrently with the parent process,
+communicating events in the child back to the parent requires some form of IPC,
+complicating process creation.\cite{posix_spawn_error_pipe}
+Even the basic interface of fork - one function call which returns twice - is unusual;
+such functions are generally considered to be complex, advanced topics in other settings.
+\cite{callcc_fork}\cite{continuations}\cite{bashcc}
+
+A user of fork must think carefully about the characteristics of the process from which fork is being called.
 Fork can have poor performance when called from processes with many memory mappings.\cite{forkroad}
-Fork copies many attributes about the parent process when creating the child process,
-including setting up copy-on-write memory mappings in the child process.
-This becomes slower as the parent process has more memory mappings,
-eventually taking a significant amount of time.
-This copying is required to robustly implement fork's model,
-where the same program continues executing without changes in both the parent process and child process;
-the copy of the program running in the child process might access any memory at any time.
-
-Multi-threaded programs generally cannot use fork safely.
-Typical Unix fork implementations duplicate only the thread calling fork from the parent process to the child process.
-In a multi-threaded program, this can cause various issues;
-for example, another thread might be holding a memory allocator lock at the time of the fork,
-which in the child process will never be unlocked,
-causing the child process to deadlock if it tries to allocate memory.
-Some thread libraries provide partial mitigations for this issue,
+This can be mitigated to some degree, but inflicts further complexity on the caller.\cite{vfork}\cite{vfork_dangerous}
+Multi-threaded programs generally cannot use fork safely.\cite{forkroad}\cite{tlpi}
+Some thread libraries provide partial mitigations for multi-threading issues,
 but it's up to user code to make use of those mitigations.\cite{pthread_atfork}
-In combination with fork's poor performance in large-memory programs,
-a user of fork must think carefully
-about the characteristics of the process from which fork is being called.
 \subsection{Spawn-style}
@@
 which creates the new process all at once.
-The most well known spawn-style interfaces are Unix's \texttt{posix\_spawn} \cite{posix_spawn}
-and Windows \texttt{CreateProcess} \cite{create_process};
+Other details, such as security context, are inherited from the parent process.\cite{tlpi}\cite{create_process}
+The most well known spawn-style interfaces are Unix's \texttt{posix\_spawn}
+and Windows \texttt{CreateProcess};
 the Unix shell also serves as a spawn-style process creation interface,
-especially when used with Bernstein chaining\cite{chainloading}.
-
-Spawn-style interfaces typically still transparently copy some details from the parent process;
-for example, various security contexts,
-and any other process attribute which is not explicitly specified in the arguments to the spawn interface.
-The key difference between fork-style and spawn-style is not how much they copy,
-it is how the attributes which are not copied are specified:
-by mutation from code running in the new process in the case of fork-style,
-or by explicit argument passing in the case of spawn-style.
-
-Spawn-style interfaces lift the constraints on the calling process that fork-style interfaces impose.
-Spawn-style interfaces don't run user code in the new process during initialization.
-This means there is no need to copy memory,
-so performance is good when a spawn-style interface is used from a process with many memory mappings.
-Likewise, there are no inherent issues when a spawn-style interface is used from a multi-threaded program,
-though operating systems, such as Linux,
-which implement spawn-style interfaces on top of \texttt{fork} can still have threading bugs.
-
-The arguments that can be provided to a spawn-style process creation syscall
-do not cover all the possible attributes that one might want to set for the new process.
-Most systems have a large number of syscalls which can mutate the state of a process during its lifetime;
-for a spawn-style interface to work in all scenarios,
-all those possible mutations must be reproduced in the interface.
-For example, the \texttt{posix\_spawn} function provided by glibc does not support creating processes in new namespaces,
-as is required for container functionality on Linux.
-
-Spawn-style process creation
-essentially creates a new domain-specific language for specifying the attributes of the process.
-Fork-style allows existing system calls,
-with which programmers are already familiar,
-to be used to set up the new process.
-Spawn-style requires some form of declarative description of the new process,
-which must be learned in addition to the normal system call interface.
-
-Spawn-style process creation also does not allow for conditional logic during the setup.
-If the setup of the new process encounters an error at some point,
-the only option is to return from the entire spawn call with an error.
-Such errors returned from spawn-style calls
-are typically much less informative
-than the errors returned by the syscalls which directly mutate the process attributes.
-Other forms of conditional logic are also impossible in spawn-style;
-one modification to the process cannot depend on the result of some other modification.
+especially when used with Bernstein chaining.\cite{posix_spawn}\cite{create_process}\cite{chainloading}
+
+Spawn-style APIs
+can be called from any parent process without concerns about memory usage or multi-threading,
+when used correctly.
+\cite{forkroad}\cite{posix_spawn_error_pipe}
+
+Unlike \texttt{fork}, a spawn-style interface provides an API for process creation
+which is independent from existing system calls.\cite{spawn_h}\cite{subprocess_run}
+Spawn-style interfaces therefore often lack support for some features,
+and they also require learning a new interface distinct from the normal system call API,
+both of which encourage systems programmers to avoid them.\cite{tlpi}\cite{posix_spawn_lwn}
+
+Spawn-style interfaces, by their nature as a single call,
+don't allow programs to branch on the results of individual modifications to the new process,
+and have complex, or substantially less information, error reporting for such modifications.
+\cite{posix_spawn}\cite{posix_spawn_error_pipe}
+This further reduces the power of spawn-style interfaces.
 \subsection{Direct-style}
@@
 This is the kind of direct-style interface that we contribute in this paper.
-\section{Overview of API}\label{overview}
-In most operating system APIs, including the ones available on Linux,
-when a program makes a syscall, it implicitly operates on the current process.
-To perform direct-style process creation on Linux,
-we need an API where we instead explicitly specify in which process we want to make a syscall.
-
-\texttt{rsyscall} provides this API.
-The \texttt{rsyscall} project develops
-language-specific, object-capability-model, low-abstraction libraries for the Linux syscall interface,
-bypassing libc.
-
-Unlike POSIX C libraries such as glibc or musl,
-an \texttt{rsyscall} library is organized based on the object-capability model.
-The capabilities to make syscalls in any given process are reified as language objects.
-If the capability to make syscalls in a specific process is not passed (in some way) to a function,
-then that function cannot make syscalls in that process.
-
-Due to this design, an \texttt{rsyscall} program can make use of multiple processes at once,
-by manipulating capabilities for multiple processes.
-The relevant two capabilities for this paper are the initial capability to the ``local'' process,
-and capabilities to child processes.
-The ``local'' process is the one which hosts the runtime for the running program,
-and in which a legacy libc would implicitly make syscalls.
-Every \texttt{rsyscall} program starts with the ``local'' process
-and uses it to bootstrap the initial capabilities for other processes.
-
-To perform direct-style process creation,
-we use \texttt{rsyscall} to call \texttt{clone} on the ``local'' process to create a new child process.
-We call various syscalls in the child process to mutate it until it reaches the desired state,
-at which point we can call \texttt{execve} in the child process to start it running on its own.
-After that, the process is no longer under our control, and it functions like any other child process.
-The child process can be monitored using normal Linux child monitoring syscalls,
-such as \texttt{waitid}.
-
-\lstinputlisting[float=ht,language=Python,label={basic},caption={Creating a new process, changing CWD, and execing}]
-{examples/basic/direct.py}
-Listing \ref{basic} shows this in action in Python.
-We create a new child process using \texttt{clone} on the local process \verb|local|,
-and receive back a capability to control it.
-We call \texttt{chdir} in the child process to change its working directory to \texttt{/dev}.
-Then we call \texttt{execve} in the child,
-passing the path of the \texttt{cat} executable,
-the arguments to the new executable,
-and the unchanged set of environment variables.
-The name of the executable is passed as the first argument,
-as is typical in Unix.
-Calling \texttt{execve} consumes the capability,
-and later use of this child process capability will fail with an exception.
-
-The behavior of system calls made on a remote process through \texttt{rsyscall}
-is identical to the behavior of system calls made implicitly on the current process through libc.
-This is an API that systems programmers are already familiar with,
-and which is already thoroughly documented through Linux system call manpages.
-The novelty is constrained to the newly explicit process capabilities.
-
-The lack of novelty in the API is important not only for the learning curve,
-but also for power.
-Since we are using conventional Linux system calls,
-just in a cross-process way,
-direct-style \texttt{clone} will always be at least as powerful as \texttt{fork}.
 \section{Examples}\label{examples}
@@
 the executable will be able to use that file descriptor through that file descriptor number.
-Unlike in listing \ref{basic}, we call the \texttt{execv} wrapper rather than \texttt{execve},
-so that we don't need to pass the (unmodified) environment to every \texttt{exec} call in these examples.
 \subsection{Non-shared-memory concurrency}

### 2020-05-23 research/paper/paper.tex
paper

@@
 \hline
-Requirements on parent process & \tbad Single-threaded, small memory & \tgood None & \tgood None\\
+Parent process requirements & \tbad Single thread, small memory & \tgood None & \tgood None\\
 Programming model & \tbad Complex (returns twice) & \tgood Simple (single call) & \tgood Simple (imperative)\\
@@
 |                             | fork                              | spawn                       | direct-style                          |
-|--------------------------------+-------------------------------------+-----------------------------+---------------------------------------|
-| Requirements on parent process | \tbad Single-threaded, small memory | \tgood None                 | \tgood None                           |
+|-----------------------------+-----------------------------------+-----------------------------+---------------------------------------|
+| Parent process requirements | \tbad Single thread, small memory | \tgood None                 | \tgood None                           |
 | Programming model           | \tbad Complex (returns twice)     | \tgood Simple (single call) | \tgood Simple (imperative)            |
@@
 When a parent process creates a child process through \texttt{fork},
-or other similar fork-style interface,
+or other similar fork-style interfaces,
 both processes execute concurrently,
-and the child process typically inherits all but a few attributes from the parent process.\cite{tlpi}
+and the child process typically inherits all but a few attributes from the parent process.\cite{manfork}
 The child process can call arbitrary system calls
@@
 and they also require learning a new interface distinct from the normal system call API,
-both of which encourage systems programmers to avoid them.\cite{tlpi}\cite{posix_spawn_lwn}
+both of which encourage programmers to avoid them.\cite{tlpi}\cite{posix_spawn_lwn}
 Also, spawn-style interfaces, by their nature as a single call,
@@
 
-Direct-style is more rarely seen, but gives a number of advantages.
-Direct-style involves creating a new child process in a non-running state,
-and then calling individual system calls from the parent to initialize the child.
-Direct-style does not constrain the calling process,
-easily branches at any step of the initialization,
-and can create any kind of process supported by the underlying system.
-
-We'll discuss each style in more detail in section \ref{background}.
+In direct-style, a process is created by its parent in a non-running state,
+then supplied with various resources,
+and then started running once it is fully set up.
+In operating systems with true direct-style process creation,
+the syscalls that can mutate a process
+take explicit arguments to indicate the process on which they should operate.\cite{keykos}\cite{sel4}
+In this way, the same syscalls that can mutate a process while it is running
+are called by the parent process to mutate the process while it is being set up.
 
@@
 such as \texttt{waitid}.
+
+Like a spawn-style interface,
+we can call \texttt{clone} through \texttt{rsyscall} from any parent process
+without concerns about memory usage or multi-threading.
 
@@
 is identical to the behavior of system calls made implicitly on the current process through libc.
-This is an API that systems programmers are already familiar with,
-and which is already thoroughly documented through Linux system call manpages.
 The novelty is constrained to the newly explicit process capabilities.
-Besides flattening the learning curve,
-this also ensures that direct-style \texttt{clone} will always be at least as powerful as \texttt{fork}.
+This ensures that direct-style \texttt{clone} will always be at least as powerful as \texttt{fork}.
+Furthermore, this is an API that systems programmers are already familiar with,
+and which is thoroughly documented.\cite{manpages}\cite{tlpi}
+
+User code can easily branch on the results of system calls,
+and errors are reported in the same as any other system call called by Python:
+An exception thrown at the point of the call.
 
@@
 We evaluate in further detail in section \ref{evaluation}.
-\section{Background on process-creation}\label{background}
-As mentioned in the introduction,
-on most operating systems today,
-the available interfaces for process creation
-are either "fork-style" or "spawn-style".
-A few operating systems have a third style, which we refer to as "direct-style".
-
-\subsection{Fork-style}
-By far the most well-known fork-style interface is Unix's fork.\cite{forkhist}
-When a parent process creates a child process by fork,
-both processes execute concurrently,
-and the child process typically inherits all but a few attributes from the parent process.\cite{tlpi}
-The child process can call arbitrary system calls
-to mutate the new process until it reaches the desired state,
-typically ultimately calling \texttt{exec}.
-
-Since the child process runs concurrently with the parent process,
-communicating events in the child back to the parent requires some form of IPC,
-complicating process creation.\cite{posix_spawn_error_pipe}
-Even the basic interface of fork - one function call which returns twice - is unusual;
-such functions are generally considered to be complex, advanced topics in other settings.
-\cite{callcc_fork}\cite{continuations}\cite{bashcc}
-
-A user of fork must think carefully about the characteristics of the process from which fork is being called.
-Fork can have poor performance when called from processes with many memory mappings.\cite{forkroad}
-This can be mitigated to some degree, but inflicts further complexity on the caller.\cite{vfork}\cite{vfork_dangerous}
-Multi-threaded programs generally cannot use fork safely.\cite{forkroad}\cite{tlpi}
-Some thread libraries provide partial mitigations for multi-threading issues,
-but it's up to user code to make use of those mitigations.\cite{pthread_atfork}
-\subsection{Spawn-style}
-In a spawn-style interface,
-details about the new process are provided up-front as arguments to some function
-which creates the new process all at once.
-Other details, such as security context, are inherited from the parent process.\cite{tlpi}\cite{create_process}
-The most well known spawn-style interfaces are Unix's \texttt{posix\_spawn}
-and Windows \texttt{CreateProcess};
-the Unix shell also serves as a spawn-style process creation interface,
-especially when used with Bernstein chaining.\cite{posix_spawn}\cite{create_process}\cite{chainloading}
-
-Spawn-style APIs
-can be called from any parent process without concerns about memory usage or multi-threading,
-when used correctly.
-\cite{forkroad}\cite{posix_spawn_error_pipe}
-
-Unlike \texttt{fork}, a spawn-style interface provides an API for process creation
-which is independent from existing system calls.\cite{spawn_h}\cite{subprocess_run}
-Spawn-style interfaces therefore often lack support for some features,
-and they also require learning a new interface distinct from the normal system call API,
-both of which encourage systems programmers to avoid them.\cite{tlpi}\cite{posix_spawn_lwn}
-
-Spawn-style interfaces, by their nature as a single call,
-don't allow programs to branch on the results of individual modifications to the new process,
-and have complex, or substantially less information, error reporting for such modifications.
-\cite{posix_spawn}\cite{posix_spawn_error_pipe}
-This further reduces the power of spawn-style interfaces.
-\subsection{Direct-style}
-A few academic operating systems, such as KeyKOS \cite{keykos}, seL4 \cite{sel4}
-and some others \cite{exokernel} \cite{fuschia} \cite{singularity},
-use another style of process creation.
-We use the new term "direct-style" to refer to this,
-since we know of no pre-existing concise and generic term for this style of process creation.
-
-In direct-style, a process is created by its parent in a non-running state,
-then supplied with various resources,
-and then started running once it is fully set up.
-In operating systems with true direct-style process creation,
-the syscalls that can mutate a process
-take explicit arguments to indicate the process on which they should operate.
-In this way, the same syscalls that can mutate a process while it is running
-are called by the parent process to mutate the process while it is being set up.
-
-We refer to this as "direct-style" process creation,
-because the parent creating the process operates on it directly and imperatively
-rather than dispatching a distinct unit of code to perform setup from inside the context of the new process,
-as in fork-style,
-or building up a declarative specification of what the new process should look like ahead of time,
-as in spawn-style.
-
-Since everything happens directly from the parent process,
-direct-style process creation is compatible with a variety of techniques
-which are difficult with fork-style or spawn-style.
-For example, a REPL could be used to interactively create the new process,
-or a program might initialize several processes at once and share data between them.
-
-Significantly,
-since all operations are performed through syscalls called directly from the main program,
-errors are indicated in the same way as any other syscall error.
-This is unlike fork-style, which needs IPC to indicate errors,
-and unlike spawn-style, which typically indicates errors at a very coarse-grained level.
-
-Like fork-style interfaces,
-direct-style interfaces can set up arbitrary attributes in the new process.
-Any attribute that can be changed by a syscall
-can be manipulated by a direct-style interface,
-just as with fork-style interfaces.
-This is unlike spawn-style interaces,
-which only can change attributes that are supported by the interface.
-
-Like spawn-style interfaces,
-direct-style interfaces have no constraints on the calling process.
-Also like spawn-style interface,
-no user code runs in the child process, so
-direct-style interfaces can be used from multi-threaded or large-memory processes without issues.
-
-Direct-style can be more complex to use;
-it is most typically used in capability-oriented operating systems,
-where a great deal of resources and information must be provided explicitly to initialize the new process.
-In a truly capability-oriented operating system,
-nothing is copied implicitly to the new process;
-everything must be explicitly specified.
-This can appear more complex
-when compared to process creation on Unix systems or Windows.
-
-However, we believe that a natural port of direct-style process creation to Linux
-provides an interface that is just as simple to use as fork-style or spawn-style on Linux,
-without their disadvantages.
-Such an interface, like fork-style and spawn-style,
-would implicitly copy some attributes to the new process,
-rather than being capability-oriented and requiring all attributes to be specified explicitly.
-This is the kind of direct-style interface that we contribute in this paper.
 \section{Examples}\label{examples}
@@
 
-As we discuss in section \ref{background},
+As we discuss in section \ref{introduction},
 many academic operating systems have direct-style process creation.

### 2020-05-24 research/paper/paper.tex
paper

@@
 \titlespacing{\subsubsection}{0pt}{0.5ex}{0ex}
+\lstset{
+language=Python,
+basicstyle=\small\ttfamily,
+frame=tb,
+columns=fullflexible,
+showstringspaces=false,
+belowskip=0pt,
+}
 
@@
 has better error handling,
-is more efficient,
+is comparably efficient,
 and is more amenable to abstraction.
@@
 The most well-known process creation interface on Linux, and Unix in general, is \texttt{fork}.
-When a parent process creates a child process through \texttt{fork},
-or other similar fork-style interfaces,
+When a parent process creates a child process through \texttt{fork} or other fork-style interfaces,
 both processes execute concurrently,
@@
 This can be mitigated to some degree, but inflicts further complexity on the caller.\cite{vfork}\cite{vfork_dangerous}
-Multi-threaded programs generally cannot use fork safely.\cite{forkroad}\cite{tlpi}
+Processes which are multi-threaded can encounter deadlocks and other issues if they call fork.
+\cite{forkroad}\cite{tlpi}\cite{posix_spawn_error_pipe}
 Some thread libraries provide partial mitigations for multi-threading issues,
@@
 communicating events in the child back to the parent requires some form of IPC,
-complicating process creation.\cite{posix_spawn_error_pipe}
+adding complexity.\cite{posix_spawn_error_pipe}
 Even the basic interface of fork - one function call which returns twice - is unusual;
@@
 Spawn-style interfaces therefore often lack support for some features,
-and they also require learning a new interface distinct from the normal system call API,
-both of which encourage programmers to avoid them.\cite{tlpi}\cite{posix_spawn_lwn}
+and they also require learning a new interface distinct from the normal system call API.\cite{tlpi}\cite{posix_spawn_lwn}
 Also, spawn-style interfaces, by their nature as a single call,
 don't allow programs to branch on the results of individual modifications to the new process,
-and have complex, or substantially less information, error reporting for such modifications.
+and have worse error reporting for such modifications.
 \cite{posix_spawn}\cite{posix_spawn_error_pipe}
@@
 
-Our contribution in this paper is this API, through the \texttt{rsyscall} library.
-The \texttt{rsyscall} project develops
-language-specific, object-capability-model, low-abstraction libraries for the Linux syscall interface,
-bypassing libc.
-\texttt{rsyscall} is currently focused on supporting Python,
-but can be ported to other languages;
-the examples we show in this paper will be in Python,
-but generalize easily to other languages.
+Our contribution in this paper is such an API, in the form of the \texttt{rsyscall} library.
+\texttt{rsyscall} is a
+language-specific, object-capability-model, low-abstraction library for the Linux system call interface,
+bypassing libc.\cite{capmyths}
+The \texttt{rsyscall} project is currently focused on supporting Python,
+but the library can be ported to other languages.
+The examples we show in this paper will be in Python,
+but generalize easily.
 \texttt{rsyscall} is open source and available from
@@
 
-Unlike POSIX C libraries such as glibc or musl,
-an \texttt{rsyscall} library is organized based on the object-capability model.
+Unlike typical C libraries such as glibc or musl,
+the \texttt{rsyscall} library is organized based on the object-capability model.\cite{capmyths}
 The capabilities to make syscalls in any given process are reified as language objects.
@@
 Every \texttt{rsyscall} program starts with the ``local'' process
-and uses it to bootstrap the initial capabilities for other processes.
-
-To perform direct-style process creation,
-we use \texttt{rsyscall} to call \texttt{clone} on the ``local'' process to create a new child process.
-We call various syscalls in the child process to mutate it until it reaches the desired state,
-at which point we can call \texttt{exec} in the child process to start it running on its own.
-After that, the process is no longer under our control, and it functions like any other child process.
-The child process can be monitored using normal Linux child monitoring syscalls,
+and uses it to bootstrap capabilities for other processes.
+
+\lstinputlisting[float,language=Python,label={basic},caption={Create new process, change CWD, and exec}]
+{examples/basic/direct.py}
+Listing \ref{basic} shows a Python program using \texttt{rsyscall}.
+We create a new child process using direct-style \texttt{clone} in the local process \verb|local|.
+Here, \texttt{local.clone()} does not create another task running the same user program,
+as with the usual \texttt{clone} system call,
+but instead returns a capability through which we can control the new child process.
+We can then call various syscalls in the child process to mutate it until it reachs the desired state.
+In this example,
+we call \texttt{chdir} in the child process to change its working directory,
+then call \texttt{execv} in the child to execute a new executable.
+Calling \texttt{execv} consumes the capability,
+releasing the process from our control;
+later use of this child process capability will fail with an exception.
+The child process can now be monitored using normal Linux child monitoring syscalls,
 such as \texttt{waitid}.
-
-Like a spawn-style interface,
-we can call \texttt{clone} through \texttt{rsyscall} from any parent process
-without concerns about memory usage or multi-threading.
 
@@
 which receives syscall requests, performs the syscall, and returns the response.
-No kernel modifications are required, and \texttt{rsyscall} is immediately deployable on recent Linux kernel.
+No kernel modifications are required, and \texttt{rsyscall} is immediately deployable on recent Linux kernels.
 We discuss the implementation in depth in section \ref{implementation},
-and examine several difficult process-creation related details.
-
-\lstinputlisting[float=h,language=Python,label={basic},caption={Create new process, change CWD, and exec}]
-{examples/basic/direct.py}
-Listing \ref{basic} shows a program using \texttt{rsyscall}.
-We create a new child process using \texttt{clone} on the local process \verb|local|,
-and receive back a capability to control it.
-We call \texttt{chdir} in the child process to change its working directory to \texttt{/dev}.
-Then we call \texttt{execv} in the child,
-passing the path of the \texttt{cat} executable
-and the arguments to the new executable.
-The name of the executable is passed as the first argument,
-as is typical in Unix.
-Calling \texttt{execv} consumes the capability,
-and later use of this child process capability will fail with an exception.
-
-The behavior of system calls made on a remote process through \texttt{rsyscall}
-is identical to the behavior of system calls made implicitly on the current process through libc.
-The novelty is constrained to the newly explicit process capabilities.
-This ensures that direct-style \texttt{clone} will always be at least as powerful as \texttt{fork}.
-Furthermore, this is an API that systems programmers are already familiar with,
-and which is thoroughly documented.\cite{manpages}\cite{tlpi}
-
-User code can easily branch on the results of system calls,
-and errors are reported in the same as any other system call called by Python:
-An exception thrown at the point of the call.
-
-We show more examples of direct-style process creation in section \ref{examples},
-including use of process features which are difficult to use on Linux with \texttt{fork} or \texttt{posix\_spawn}.
+and examine several difficult process-creation details.
+
+System calls called in a child process through \texttt{rsyscall}
+behave identically to system calls called implicitly in the current process through libc.
+A system call returning an error is reported in the standard way for system calls in Python:
+An exception is thrown at the point of the call.
+User code can branch as normal on the results of system calls,
+and implement fallbacks or other error handling logic.
+
+Since we can call any syscall,
+we can access any feature available in Linux; we are therefore at least as powerful as \texttt{fork}.
+As we'll show in section \ref{examples},
+we can in fact use Linux features in ways that are impractical with the usual fork-style or spawn-style interfaces.
 
@@
 and can outperform \texttt{fork} in some cases.
+Like a spawn-style interface,
+we can call \texttt{clone}
+without worse performance in the presence of large memory usage,
+and without the possibility of bugs in the presence of multi-threading.
 The performance overhead of creating processes with direct-style clone
 is comparable to the performance cost of creating processes while inside a Linux namespace.
-At \twosigma, we have deployed direct-style process creation in production
-inside a library for distributed system deployment, Toplevel.
+At \twosigma, we have deployed direct-style process creation
+as part of a library for distributed system deployment, used in production.
 We evaluate in further detail in section \ref{evaluation}.
@@
 we'll demonstrate direct-style process creation
-by demonstrating several advanced uses of processes
+with several advanced uses of processes
 using more sophisticated Linux features.
-For concision, we assume in these examples that we are running with sufficient permissions in these examples,
+For concision, we assume that we are running with sufficient privileges,
 but these examples can all be run without privileges with appropriate use of \verb|CLONE_NEWUSER|.
+\cite{user_namespaces}\cite{unpriv_fuse}
 
@@
 but implementing them with \texttt{fork} or \verb|posix_spawn| requires substantially more code,
-or even a complete change in implementation approach,
-as described in section \ref{shared_fd_table}.
+or even a complete change in implementation approach as in section \ref{shared_fd_table}.
 We'll compare our direct-style implementations in this section
@@
 Most resources in Linux are managed through file descriptors,
-so this feature allows the parent process to pass a variety of resources to the child process,
+so this allows the parent process to pass a variety of resources to the child process,
 such as files, network connections, pipes, or other resources.\cite{ucspi}
@@
 This is a form of abstraction,
-and we can know ``for free''\cite{theoremsforfree}
+and so we know ``for free''
 that the program running in the child process
 will not rely on the details of how the file descriptor was created,
-such as the filename that was opened or the hostname that was connected to.
-
-\lstinputlisting[float=*ht,language=Python,label={fds},caption={Passing down FDs}]
+such as the filename that we opened or the hostname to which we connected.\cite{theoremsforfree}
+
+\lstinputlisting[float,language=Python,label={fds},caption={Passing down FDs}]
 {examples/fds/direct.py}
@@
 
-\lstinputlisting[float=*ht,language=Python,label={mount},caption={Overriding absolute path using a mount namespace}]
+\lstinputlisting[float,language=Python,label={mount},caption={Overriding absolute path using a mount namespace}]
 {examples/mount/direct.py}
@@
 \item
-  In Python, does our \texttt{rsyscall}-based direct-style process creation interface
-  have acceptable performance relative to the standard library's spawn-style \texttt{subprocess.run}?
-  We benchmark direct-style process creation against spawn-style in \ref{subprocess_bench},
+  Does our \texttt{rsyscall}-based direct-style process creation interface
+  have acceptable overhead relative to the alternatives?
+  We evaluate a number of microbenchmarks in \ref{subprocess_bench},
   and find an acceptable level of overhead.
@@
 \subsection{Microbenchmark results}\label{subprocess_bench}
+\subsubsection{Basic process creation}
 \begin{figure}[h!]
@@
 \end{figure}
-In Python, does our \texttt{rsyscall}-based direct-style process creation interface
-have acceptable performance relative to the standard library's spawn-style \texttt{subprocess.run}?
-
-We first evaluate a simple process creation workload:
+We evaluate a simple process creation workload:
 create a child process, exec the \texttt{true} binary, and wait for it to exit.
-We implement this using both direct-style \texttt{clone}
-and the standard library's \texttt{subprocess.run}.
-We benchmark both implementations under CPython 3.7.7, on Linux 4.19.93,
-pinned to an isolated single core on an Intel i9-9900K CPU with 60GB of RAM.
+We implement this using Python \texttt{rsyscall} direct-style \texttt{clone},
+the Python standard library's \texttt{subprocess.run} (which is implemented primarily with \texttt{fork}),
+\texttt{fork} from C,
+and \texttt{posix\_spawn} from C.
+We run on CPython 3.7.7, glibc 2.30, Linux 4.19.93,
+pinned to an isolated single core on an Intel i9-9900K CPU at standard clock speed,
+with 60GB of RAM.
 We vary the amount of anonymous memory mapped in the parent process
 to demonstrate how each implementation scales with memory usage.
-The results are in Figure \ref{fig:subprocess_bench}.
-
-The direct-style implementation stays around 2.2 milliseconds regardless of memory usage.
-\texttt{subprocess.run}, which is implemented with \texttt{fork},
-starts at 1.4 milliseconds,
-and then scales linearly with the amount of memory mapped, as expected,
-up to 48.5 milliseconds at 4 GBs of mapped memory.
-The cross-over point for this micro-benchmark is about 40 megabytes of memory mapped.
-
-We perform another microbenchmark to evaluate the overhead of modifying the child process.
-We simply call \texttt{getpid} from Python
+The results are summarized in figure \ref{fig:subprocess_bench}.
+
+Our true baseline for performance is the Python standard library's \texttt{subprocess.run};
+this takes an average of 1.4 milliseconds at low memory,
+while \texttt{rsyscall}'s \texttt{clone} takes an average of 2.2 milliseconds at any memory usage.
+As expected, the fork-based implementations scale poorly with memory usage,
+and the C implementations vastly outperform the Python implementations.
+
+We perform another microbenchmark to evaluate the overhead of performing additional modifications of the child process.
+We call \texttt{getpid} from Python
 in both the child process and the parent process on the same benchmark setup.
@@
 by performing our child process setup system calls through \texttt{rsyscall}.
-Through profiling, we've found that most of this per-syscall overhead is spent in Python,
-rather than waiting for syscall responses from the child task.
 
@@
 such optimization is reserved for future work.
-
-\subsection{Microbenchmarks for nested processes}\label{clone_bench}
-\begin{figure}[h!]
+\subsection{Nested processes, \texttt{CLONE} flags}\label{clone_bench}
+\begin{figure}
 \centering
  \includegraphics[width=0.5\textwidth]{clone_bench}
- \caption{Time to create different processes, by variety of parent process}
+ \caption{Time to create different processes from \texttt{rsyscall}, by variety of parent process and child process}
  \label{fig:clone_bench}
 \end{figure}
-
 As demonstrated in sections \ref{pidns} and \ref{fuse},
@@
 We've also specified different combinations of flags for the process being created in the benchmark.
+We don't include the Python standard library in these benchmarks
+as there's no way to create nested child processes or specify \texttt{CLONE} flags in the Python standard library.
 The results are in figure \ref{fig:clone_bench}.
-We don't include the Python standard library in these benchmarks
-as there's no way to specify \texttt{CLONE} flags in the Python stdlib.
 
@@
 but it is illustrative to compare it to the performance of \texttt{rsyscall}.
+
 Note that the cost of creating a new child process from a parent process that is already inside a namespace
@@
 Our experience has been that with fork-style or spawn-style interfaces,
-all file descriptor logic must be centralized,
+file descriptor logic must ultimately be centralized,
 whereas file descriptors in a new process can be built up over time using direct-style,
@@
 this allows starting up services in parallel, rather than in dependency order.
-Benchmarking one test built with an old library and its equivalent built with Toplevel,
-the old test took an average of 221 seconds to run,
+Benchmarking a representative example built with an old library
+and its equivalent built with Toplevel,
+the old example took an average of 221 seconds to run,
 mostly spent in system startup,
-while the new test takes an average of 16 seconds.
+while the new example takes an average of 16 seconds.
 
@@
 We pass around file descriptors freely inside Toplevel.
-A given component can implemented with a separate process which has the file descriptor passed down,
+A given component can be implemented with a separate process which has the file descriptor passed down,
 running a program written in native code for performance,
@@
 either is possible behind a uniform file descriptor interface.
-
-We've also been able to create complicated pipelines in Toplevel,
-as in section \ref{pipe}.
 
@@
 
-Toplevel also makes uses of other features of \texttt{rsyscall} which are not covered in this paper.
+Toplevel also makes uses of other features of \texttt{rsyscall} which are not covered in this paper,
+and will be covered in future work.
 These include the ability to operate on remote hosts through capabilities for remote processes,

### 2020-05-25 research/paper/paper.tex
paper

@@
 
+\usepackage{microtype}
 \usepackage{tikz}
@@
 basicstyle=\small\ttfamily,
+keywordstyle=,
+stringstyle=\color{red},
 frame=tb,
@@
 \begin{document}
-\newcommand{\twosigma}{[tech company]} 
+\newcommand{\twosigma}{{[tech company]}}
+\newcommand{\githuburl}{[Github url]}
 \date{}
@@
 \texttt{rsyscall} is open source and available from
-[Github url].
+\githuburl{}.
 
@@
 is comparable to the performance cost of creating processes while inside a Linux namespace.
-At \twosigma, we have deployed direct-style process creation
+At \twosigma{}, we have deployed direct-style process creation
 as part of a library for distributed system deployment, used in production.
@@
 a program can pass file descriptors from the parent process to a child process
-by using file descriptor inheritance.
+by using file descriptor inheritance.\cite{tlpi}
 All open file descriptors are copied at process creation time from the parent process to the child process;
@@
 
-% TODO need to clarify this more, this is v2, Larry didn't like v1 beyond first sentence
 Most resources in Linux are managed through file descriptors,
@@
 We indicate that we want to operate on the child's inherited copy of the file descriptor with \verb|child.inherit_fd|.
+\texttt{inherit\_fd} performs no system calls,
+it just updates bookkeeping to create a new handle for the inherited file descriptor;
+we'll discuss it in more depth in section \ref{handles}.
 Then we disable \texttt{CLOEXEC} on the child's file descriptor,
@@
 Processes run concurrently,
-which allows exploiting the parallelism of the underlying hardware.
+which enables more modular programs
+and allows exploiting the parallelism of the underlying hardware.
 Since processes don't share memory,
 they can provide a less complex parallel programming environment
-than shared-memory thread-based approaches.
-
-One of the most popular parallel programming environments is the Unix shell,
-which obtains its parallelism by running multiple processes connected via pipes.
-The Unix shell has a relatively constrained form of parallel processing,
-but it's also possible to create more complex webs of parallel processes,
-where, for example, one process might take multiple inputs over multiple pipes,
-or produce multiple outputs.
-
-\lstinputlisting[float=*ht,language=Python,label={pipe},caption={Creating a parallel processing pipeline}]
+than shared-memory thread-based approaches.\cite{threads}
+
+\lstinputlisting[float,language=Python,label={pipe},caption={Creating a parallel processing pipeline}]
 {examples/pipe/direct.py}
@@
 connected by pipes.
+The \texttt{source} program generates two outputs, which go to \texttt{video\_sink} and \texttt{audio\_sink}.
 We first create two pipes in the local process,
-then inherit them down to several child processes using a helper function, \verb|as_argument|,
+then inherit them down to several child processes using a helper function, \verb|argfd|,
 which uses \verb|inherit_fd| and \texttt{fcntl} as described in section \ref{fd_abstraction}.
@@
 \subsection{Customization without explicit support}
-While, ideally, all resources would be passed by file descriptor\cite{capsicum},
+While, ideally, all resources would be passed by file descriptor,
 Unix has a number of resources that are global,
 such as the pid space or the mount table,
-and which cannot be changed on a per-process basis.
-
-In Linux, most global resources have been made process-local
-through the namespaces system.
+and which cannot be changed on a per-process basis.\cite{capsicum}
+
+In Linux, many global resources have been made process-local
+through the namespaces system.\cite{lwn_namespaces}
 Like other process-local resources such as the current working directory,
@@
 
-The most popular usage of namespaces is for large-scale container systems such as Docker;
-however, this is far from the only use.
+One key use of namespaces is to implement large-scale container systems such as Docker.\cite{lwn_namespaces}
+However, this is far from the only use.
 With an adaquate process creation primitive,
@@
 in ways that were not anticipated at development time
-by customizing the environment the process runs in.
-Other system calls such as \texttt{chroot} can also be used for this.
+by customizing the environment the process runs in.\cite{plan9ns}
+Other system calls such as \texttt{chdir} and \texttt{chroot} can also be used for this,
+but namespaces allow new ways of customization.\cite{mount_namespaces}
 
@@
 {examples/mount/direct.py}
-One classic form of customization enabled by namespaces
-is overriding files at paths that were hardcoded into a program we can't change,
+One classic form of customization
+is overriding files at paths that were hardcoded into a program,
 without changing those files for the rest of the system.
@@
 which allows us to manipulate the filesystem tree in a way that only this process will see
-using the \texttt{mount} system call.
+using the \texttt{mount} system call.\cite{mount_namespaces}\cite{clone}
 We call \texttt{mount}, passing \texttt{MS.BIND}, to make the file \verb|/home/foo/custom_foo.conf|
-appear at the path \verb|/etc/foo.conf|.
-This is known as a bind mount.
+appear at the path \verb|/etc/foo.conf|;
+this is known as a bind mount.\cite{mount}
 Then we execute some executable,
@@
 we can also deny the process access to resources that it otherwise receives by default.
-This is a key part of creating a secure sandbox for potentially malicious code.
-
-\lstinputlisting[float=*ht,language=Python,label={unmount},caption={Unmount all and run executable via fexec}]
+This is a key part of creating a secure sandbox for potentially malicious code.\cite{seccomp}\cite{firejail}\cite{gvisor}
+
+\lstinputlisting[float,language=Python,label={unmount},caption={Unmount all and run executable via fexec}]
 {examples/unmount/direct.py}
@@
 as in listing \ref{fds}.
-(The underlying system call for \texttt{fexec} is \texttt{execveat};
-this \texttt{fexec} is a helper function.)
+The underlying system call for \texttt{fexec} is \texttt{execveat};
+this \texttt{fexec} is a helper function.\cite{execveat}
 Note that this executable must be statically linked,
@@
 that it won't be able to tamper with the rest of the system.
-A sandbox which is truly robust against malicious or compromised programs requires some additional steps,
-but this is a substantial start.
+A sandbox which is truly robust against malicious or compromised programs requires additional steps,
+but this is a substantial start.\cite{firejail}\cite{gvisor}
 Such a technique can also be used when a full sandbox is not relevant,
@@
 as we do here with the database file descriptor.
-This allows us to approximate capability-based security\cite{capsicum}.
+This allows us to approximate capability-based security.\cite{capsicum}
 \subsection{Nested clone and pid namespaces}
 Process-local resources can also be used to control the lifetime of resources used by that process.
-Many Linux resources are not automatically cleaned up on process exit;
+Some Linux resources are not automatically cleaned up on process exit;
 a poorly coded program may allocate global resources
@@
 those subprocesses may unintentionally leak,
-and be left running on the system even after the program itself has stopped.
+and be left running on the system even after the program itself has stopped.\cite{caternfork}
 
@@
 When the init process dies,
-all other processes in the pid namespace are destroyed.
-
-\lstinputlisting[float=*ht,language=Python,label={pidns},caption={Nested clone and pid namespace}]
+all other processes in the pid namespace are destroyed.\cite{pid_namespaces}
+
+\lstinputlisting[float,language=Python,label={pidns},caption={Nested clone and pid namespace}]
 {examples/pidns/direct.py}
@@
 we create a child process which we call \texttt{init},
-passing \verb|CLONE_NEWPID| to create it in a new pid namespace.
+passing \verb|CLONE_NEWPID| to create it in a new pid namespace.\cite{clone}
 
@@
 or we can continue on by execing an init daemon in \texttt{init} to handle it for us.
+\cite{pid_namespaces}\cite{tini}\cite{dumb_init}
 
@@
 \subsection{Shared file descriptor tables}\label{shared_fd_table}
-A library can create a child process to perform a useful service.
-This can sometimes be done more easily when the child process shares file descriptors with the parent.
-
-\lstinputlisting[float=*ht,language=Python,label={fuse},caption={Shared file descriptor tables}]
+We can create a child process which shares its file descriptor table with the parent process.
+This can be useful for a variety of purposes,
+such as accessing resources in other namespaces, or in multiple namespaces at once.
+We'll give one example using FUSE.
+
+The Filesystem in Userspace (FUSE) Linux subsystem allows a userspace program
+to implement a filesystem.\cite{fuse}
+Many interesting filesystems have been implemented using FUSE,
+representing a variety of resources as files.\cite{sshfs}\cite{encfs}\cite{mp3fs}
+
+For improved modularity, we might want to mount and use a FUSE filesystem in our program,
+in a way that no other process can see it,
+without entering a new namespace ourselves.
+We can do this by creating a child process that shares its file descriptor tables with the parent process.
+
+\lstinputlisting[float,language=Python,label={fuse},caption={Shared file descriptor tables}]
 {examples/fuse/direct.py}
@@
 we first create a child process \verb|ns_child| in a new mount namespace with \verb|CLONE_NEWNS|,
-and, for the first time, also pass \verb|CLONE_FILES|,
-which causes the file descriptor table to be shared between the parent process and the child process.
-We create a grandchild off of \verb|ns_child|
+and this time also pass \verb|CLONE_FILES|,
+which causes the file descriptor table to be shared between the parent process and the child process.\cite{clone}
+We create another child from \verb|ns_child|
 and use it to exec a FUSE filesystem,
@@
 through the shared file descriptor table.
+We call \texttt{local.use\_fd} which, like \texttt{inherit\_fd},
+merely updates bookkeeping to create a new handle for the file descriptor existing in the shared fd table;
+we'll discuss \texttt{use\_fd} in section \ref{handles}.
 
@@
 
-ptrace provides a way to perform arbitrary actions on other processes,
+We chose to implement in userspace first,
+rather than immediately implementing this in the kernel,
+to aid deployability
+and to allow fast iteration while developing userspace code using these features.
+In the past, attempts to upstream kernel support for novel features without extensive userspace usage
+has had a poor reception in the Linux kernel community.\cite{lwn_checkpoint}\cite{first_class_address_space}
+We believe proving viability in userspace first will be better in the long term.
+
+\texttt{ptrace} provides a way to perform arbitrary actions on a target process,
 including system calls,
-but is not suitable for us.
-Among other issues, it has the substantial disadvantage of not permitting multiple ptracers.
-A ptrace-based implementation would prevent using \texttt{strace} or \texttt{gdb} on \texttt{rsyscall}-controlled processes,
+but is not suitable for us.\cite{ptrace}
+As we discuss in section \ref{related}, a number of systems, such as \texttt{gdb} and \texttt{strace},
+use \texttt{ptrace} to implement debuggers.
+But multiple processes cannot use \texttt{ptrace} on a single target process at the same time;
+thus, if we used \texttt{ptrace} to implement \texttt{rsyscall},
+such debugging tools would not work on \texttt{rsyscall}-controlled processes,
 which is an unacceptable limitation for a general-purpose utility.
@@
 The most simple way to access memory is for the local process to be in the same address space as the target process.
-This is the case most of the time; we pass \verb|CLONE_VM| to \texttt{clone} by default in the Python wrapper.
+This is the case most of the time; we pass \verb|CLONE_VM| to \texttt{clone} by default
+in the \texttt{rsyscall} wrapper for \texttt{clone}.
 
@@
 for example, if the target process is at a different privilege level,
-it must be in a different address space.
+we will want it to be in a different address space for security reasons.\cite{vfork_dangerous}
 There are a number of available techniques in that scenario;
@@
 while waiting for the child process to exit or call \texttt{execve},
-which is immediately unsuitable.
+which is immediately unsuitable.\cite{vfork}
 \texttt{fork} lacks many features which are restricted to \texttt{clone}.
@@
 which unfortunately won't work for us.
-In this technique, the parent process creates a pipe before forking,
+With this technique, the parent process creates a pipe before forking,
 ensures both ends are marked \verb|O_CLOEXEC|,
@@
 but the relevant detail for us here is that we can wait on an address
-until a futex wakeup is performed on that address.
+until a futex wakeup is performed on that address.\cite{futex}
 This means we can wait on \texttt{ctid} until the futex wakeup is performed,
@@
 without the child process itself exiting,
-then we can catch that in \texttt{execve} and know that the child has successfully called \texttt{execve}.
+then we can catch that in \texttt{execve}
+and know that the child has successfully completed an \texttt{execve} call.
 
@@
 
-The file descriptor table can be unshared by calling \verb|unshare(CLONE_FILES)| or \texttt{execve}.
+The file descriptor table can then be unshared by calling \verb|unshare(CLONE_FILES)| or \texttt{execve}.
 After either system call, the process will have a new file descriptor table
@@
 except for, in the case of \texttt{execve}, file descriptors with the \texttt{CLOEXEC} flag set.
-\subsubsection{Tracking file descriptors through handles}
+\subsubsection{Tracking file descriptors through handles}\label{handles}
 After \texttt{clone}, \verb|unshare(CLONE_FILES)|, or \texttt{execve}
@@
 there are no file descriptor handles associated with it.
+We can make file descriptor handles by opening file descriptors in the child,
+or by creating new handles for file descriptors opened in other processes.
+
 We can use the \verb|inherit_fd| function to create handles for inherited file descriptors.
@@
 \verb|inherit_fd| can no longer be used.
+
+If the process shares its file descriptor table with any other processes,
+we can use the \verb|use_fd| function to create handles for file descriptors
+in the shared file descriptor table.
+\verb|use_fd| takes as arguments a process $A$
+and a file descriptor handle associated with any process sharing that process's file descriptor table;
+it returns a new handle associated with process $A$ for that file descriptor.
 \subsubsection{Closing file descriptors after inheritance}\label{cloexec}
@@
 
+% TODO should definitely cite this
 However, we can't simply close all other file descriptors,
@@
 except we don't get the chance to preserve live file descriptors handles,
-without first unsetting \texttt{CLOEXEC} on them.
+other than by unsetting \texttt{CLOEXEC} on them before calling \texttt{execve}.
 Again an extension to the kernel would be useful:
-A new argument to \texttt{execve} which allows specifying a list of file descriptors to inherit
-despite the \texttt{CLOEXEC} flag.
-This would allow programs to avoid the pattern
+A new argument to \texttt{execve} which allows specifying a list of file descriptors
+which should be inherited across \texttt{execve} even if \texttt{CLOEXEC} is set on them.
+This new argument would allow programs to avoid the pattern
 of unsetting \texttt{CLOEXEC}, calling \texttt{execve}, and then immediately re-setting \texttt{CLOEXEC},
-just to inherit a file descriptor across a single exec.
-This might seem minor,
-but unsetting \texttt{CLOEXEC} before \texttt{execve} is not possible
-when the process calling \texttt{execve} has a shared file descriptor table,
-because some other process might call \texttt{execve} as well,
-and erroneously inherit the file descriptor.
+just to inherit a file descriptor across a single \texttt{execve}.
+This is primarily useful when the process calling \texttt{execve} has a shared file descriptor table;
+in that case, unsetting \texttt{CLOEXEC} before \texttt{execve} is not possible,
+because some other process might also call \texttt{execve}
+and erroneously inherit the file descriptor from the shared file descriptor table.
 \section{Evaluation}\label{evaluation}
@@
   Does our \texttt{rsyscall}-based direct-style process creation interface
-  have acceptable overhead relative to the alternatives?
+  have acceptable performance overhead relative to the alternatives?
   We evaluate a number of microbenchmarks in \ref{subprocess_bench},
@@
   Does direct-style process creation perform well in the ``real world''?
-  We discuss our positive experience with using direct-style process creation at \twosigma in \ref{realworld}.
+  We discuss our positive experience with using direct-style process creation at \twosigma{} in \ref{realworld}.
 \end{compactitem}
@@
 
-\lstinputlisting[float=ht,language=Python,label={basic_fork},caption={Fork-style: Creating a new process, changing CWD, and execing}]
+\lstinputlisting[float,language=Python,label={basic_fork},caption={Fork-style: Creating a new process, changing CWD, and execing}]
 {examples/basic/fork.py}
-Listing \ref{basic_fork} gives an example of one of our fork-style implementations,
-corresponding to the direct-style example in listing \ref{basic}.
+Listing \ref{basic_fork} shows one of our fork-style implementations,
+corresponding to the direct-style implementation in listing \ref{basic}.
 
 In our fork-style implementations,
-we assume a great deal of infrastructure is available to make fork-style simpler.
+we assume substantial infrastructure is available to make fork-style simpler;
+despite this, fork-style implementations are still significantly more verbose than direct-style.
 We assume the existence of a robust IPC system,
+with communication channels set up automatically between the parent process and the child process,
 with an already-defined protocol which can cover all errors from all system calls.
-Despite this, fork-style implementations are still significantly more verbose than direct-style.
-
-The main source of additional code in the fork-style implementation
+
+The main source of additional code in fork-style implementations
 is reporting errors back to the main program using IPC.
-If a program that is already reporting errors to some other destination,
-such code is not necessary;
-for example, a process supervisor or a shell might write process startup errors
-as textual logs which are read by a human.
-If a program does not already have such a destination for errors,
-then robustly dealing with errors in \texttt{fork} requires this additional IPC code.
-
-Direct-style does not require any extra work to report errors;
+This pattern is common in complex usage of fork.\cite{posix_spawn_error_pipe}\cite{python_subprocess_errpipe}
+Many users of fork avoid complex IPC
+by encoding a subset of the error information into the exit code of the child process
+when an error is encountered during child setup.
+Such code still must be explicitly written, like the IPC code we show,
+and code in the parent process must detect the error by decoding the exit code.
+
+Direct-style does not require any extra work to report errors from the child process;
 errors are reported just like any other system call.
@@
 though this simplification is not immediately visible in a small example.
-The fork-style examples may break if there are multiple threads in the calling process,
+The fork-style implementations may break if there are multiple threads in the calling process,
 and may be slow if the calling process has large amounts of memory mapped, as we'll see in section \ref{subprocess_bench}.
-The direct-style examples do not have these limitations.
+The direct-style implementations do not have these limitations.
 \subsection{Microbenchmark results}\label{subprocess_bench}
@@
 and the C implementations vastly outperform the Python implementations.
+\texttt{posix\_spawn} takes an average of 440 microseconds to start a process,
+regardless of the memory usage of the calling process.
 
@@
 or a child in a different namespace (created with \verb=clone(CLONE_NEWNS|CLONE_NEWPID)=).
-We've also specified different combinations of flags for the process being created in the benchmark.
+We've also specified different combinations of flags for the child process being created in the benchmark.
 We don't include the Python standard library in these benchmarks
@@
 is around 15 milliseconds; subtantially greater than the overhead of \texttt{rsyscall}.
+As namespaces are widely used,
+this suggests our overhead is acceptable.
 \subsection{Usage in the real world}\label{realworld}
-At \twosigma we have used \texttt{rsyscall} and direct-style process creation
+At \twosigma{} we have used \texttt{rsyscall} and direct-style process creation
 to implement an internal distributed system deployment library, written in Python,
@@
 We pass around file descriptors freely inside Toplevel.
-A given component can be implemented with a separate process which has the file descriptor passed down,
-running a program written in native code for performance,
-or implemented in-process with a Python implementation for flexibility;
-either is possible behind a uniform file descriptor interface.
+This uniform interface allows us to flexibly swap implementations between in-process and out-of-process,
+allowing us to decide on a case-by-case basis
+whether we prefer the flexibility of an in-process Python implementation,
+or the performance of a native code implementation running in a child process.
 
@@
 we prevent process leaking with a pid namespace.
-This is part of strategy of noninterference and cleanup
+This is part of a strategy of noninterference and cleanup
 that ensures that multiple users of Toplevel
@@
 \subsection{Direct-style process creation}
-Baumann 2019\cite{forkroad} briefly suggests cross-process operations as a possible replacement for fork on Unix,
-positing many of the same advantages we find in practice.
+Baumann (2019) briefly suggests cross-process operations as a possible replacement for fork on Unix,
+positing many of the same advantages we find in practice.\cite{forkroad}
 Our work, though developed independently,
-is in many ways an elaboration on and implementation of that idea,
-to which we owe a great debt.
+is in many ways an elaboration on and implementation of that idea.
 
@@
 many academic operating systems have direct-style process creation.
-We are not aware of any previous instances of direct-style process creation in a Unix-like environment.
+We are not aware of any other instances of direct-style process creation in a Unix-like environment.
 
@@
 in such an environment.
-\subsection{Remote system calls}
-There are many Unix-based systems which do something called "remote system calls".
-Most such systems do not allow a single program to manipulate multiple processes.
+\subsection{Remote system calls}\label{related}
+Many Unix-based systems have features described as ``remote system calls''.
+Most such systems do not allow a single program to manipulate multiple processes;
+rather, a program implicitly makes system calls in a single remote process
+which is distinct from the process the program's code runs in.
 Those systems that do allow manipulating multiple processes
@@
 
-Many systems use system call forwarding to implement migration in a computing cluster.
-HTCondor\cite{condor} and Popcorn\cite{popcorn} are two examples.
-In both systems, processes can be live-migrated between hosts;
-when this occurs, the system will automatically forward IO-related system calls
+% probably we should rework this whole section to say "X and Y, among other systems, do Z."
+HTCondor\cite{condor} and Popcorn\cite{popcorn}, among others,
+use system call forwarding to implement migration between hosts in a computing cluster.
+In these systems, processes can be live-migrated between hosts;
+when this occurs, the system will transparently forward IO-related system calls
 back to the original host.
 
-Some systems intercept system calls and forward them to another process to be implemented
-as part of a virtualization system.
-gvisor\cite{gvisor} and ViewOS\cite{viewos} are two examples.
-In both systems,
-system calls made in one process are intercepted,
-and serviced by another process.
-
-Some systems forward system calls elsewhere for asynchronous execution.
-FlexSC\cite{flexsc} is one example.
-System calls in FlexSC are forwarded to a pool of dedicated system call execution threads.
-The recently added Linux feature \verb|io_uring| is, in some sense, another example;
-it supports sending a subset of system calls to the kernel,
-which executes them asynchronously,
-rather than blocking the thread.
-
-Some systems call system calls in other processes for debugging or introspection purposes.
-strace, gdb, and CRIU\cite{criu} are examples of this.
-These systems typically uses ptrace,
-which is capable of operating on multiple processes at once.
-As discussed in section \ref{implementation},
-a ptrace-based has approach has limitations which make it unsuitable for \texttt{rsyscall}.
+gvisor\cite{gvisor} and ViewOS\cite{viewos}, among others,
+use system call interception as a means of virtualization.
+In these systems,
+system calls made in one process are intercepted and forwarded to another process,
+which performs the specified system call and returns the results back to the original process.
+
+Many other systems have unusual system call invocation patterns,
+such as enclave systems like SCONE\cite{scone}\cite{eleos}
+or systems for exception-less system calls like FlexSC\cite{flexsc}.
+In these systems, for a variety of reasons,
+a system call in one process is not evaluated by directly entering the kernel,
+but instead is sent to some other process to be evaluated,
+and the results received from that process.
+
+A number of debugging or instrospection systems have the ability to perform or monitor system calls in other processes,
+typically using ptrace.
+Systems like gdb and CRIU
+retrieve information about the target process by forcing it to call various system calls to dump information.
+These systems typicaly use \texttt{ptrace},
+which can be used by a single ptracer process to operate on multiple target processes at once.
+Unfortunately, as discussed in section \ref{implementation},
+multiple ptracer processes cannot use \texttt{ptrace} on a single target process simultaneously,
+which means only one of these systems can be used on a process at a time.
+
+% maybe mention the notion of a formal interface definition for the linux syscall interface
+% that might go in future work even.
+% yeah it's not yet something developed.
+% a formal interface definition so that remote system calls can be employed without building a new system every time.
 \subsection{Process capabilities}
-The Capsicum system provides something called process descriptors\cite{capsicum},
-which is a file descriptor handle for a process.
-A similar feature has been recently added to Linux in the form of the pidfd API\cite{pidfd}.
+Many capability-oriented operating systems, such as KeyKOS and seL4,
+have process capabilities which allow one process to operate on another process.\cite{keykos}\cite{sel4}\cite{fuschia}
+These systems are our main inspiration for this work.
+
+On several Unix systems,
+Capsicum provides something called process descriptors,
+which is a file descriptor handle for a process.\cite{capsicum}
+A similar feature has been recently added to Linux in the form of the pidfd API.\cite{pidfd}
 These notions of process capability are quite limited, however;
@@
 \subsection{Capability-secure libc replacements}
-The Capsicum system provides a set of system calls
+Capsicum provides a set of system calls
 which can be used to provide a capability-secure sandbox.\cite{capsicum}
-Latter efforts\cite{oblivious}, including CloudABI\cite{cloudabi} and WASI\cite{wasi},
-developed this into a partial or full replacement for a POSIX libc.
+Latter efforts, including CloudABI and WASI,
+developed this into a partial or full replacement for a POSIX libc.\cite{oblivious}\cite{cloudabi}\cite{wasi}
 Like all other Unix libcs that we know of,
@@
 \subsection{Other applications of \texttt{rsyscall}}
-Most avenues of future work focus on \texttt{rsyscall}.
 \texttt{rsyscall} was not developed solely for the purpose of this paper,
 and it has other uses unrelated to direct-style process creation,
-such as asynchronous system calls, exceptionless system calls\cite{flexsc}, cross-host operations, among others.
+such as asynchronous system calls, exceptionless system calls\cite{flexsc}, and cross-host operations, among others.
 We are actively exploring such applications,
@@
 \texttt{rsyscall}'s cross-process syscalls can be performed entirely in userspace,
-which has substantial benefits for deployability.
+which has substantial benefits for deployability and flexibility.
 Nevertheless, direct support in the Linux kernel
 for creating a stub process and performing syscalls in the context of that process
-may provide efficiency benefits, as well as reducing userspace-visible complexity.
+may provide efficiency benefits, as well as reducing overall complexity.
 
@@
 We would like to open source libraries built on top of direct-style process creation.
-From our experience using such libraries at \twosigma,
-we believe this would greatly ease the development of complex systems involving processes,
-and would provide a substantial demonstration of the power of direct-style process creation.
+From our experience using such libraries, discussed in section \ref{realworld},
+we believe this would make it easier to develop complex systems involving processes.
 \section{Conclusions}\label{conclusions}
-Direct-style process creation is much less known and much less used than fork-style and spawn-style.
-We have implemented direct-style process creation for Linux.
-Our implementation is immediately deployable on today's Linux systems.
-We have discussed various applications of processes,
-and demonstrated the use of Linux direct-style process creation
-to implement them.
-
-We believe that direct-style process creation is superior to fork-style and spawn-style,
-including on Linux.
-Our technique is immediately deployable and can be used for a wide variety of complex applications.
+We have introduced direct-style process creation on Linux for the first time,
+through \texttt{rsyscall}.
+This style of process creation is common on academic operating systems,
+but was previously not usable on Linux.
+We've provided a number of examples which demonstrate the usefulness
+of direct-style process creation in section \ref{examples}.
+Our current implementation of \texttt{rsyscall} is focused on supporting Python,
+works entirely in userspace,
+and is immediately deployable on today's Linux systems.
+\texttt{rsyscall} is open source and available from \githuburl{}.
+
+We've found that direct-style process creation is more expressive than fork-style and spawn-style,
+and even with an unoptimized implementation, has comparable performance.
+We've used direct-style process creation successfully in production at \twosigma{}.
+Direct-style process creation is our technique of choice for new applications involving process management,
+including in container runtimes and distributed systems.
+Future work can make direct-style process creation more efficient
+and decrease the complexity of its implementation.
 
 We hope that a better process creation mechanism
-will help encourage more creative use of processes,
-which, though widespread,
-are still not used to their full potential.
-
+will help encourage more creative use of processes and the features available in Linux.
+Though processes are a long-standing, widespread feature,
+we believe there is still much to be learned about how to use processes to their full potential.
 \bibliographystyle{plain}

### 2020-05-26 research/paper/paper.tex
paper

@@
 \date{}
-\title{\Large \bf DRAFT: Direct-style process creation on Linux}
+\title{\Large \bf Direct-style process creation on Linux}
 % \author{
@@
 We develop a new process creation interface for Linux
-which allows a program to create a child process in an non-running state
+which allows a program to create a child process in a non-running state
 and initialize the new process by operating on it from the outside.
@@
 such functions are generally considered to be complex, advanced topics in other settings.
-\cite{callcc_fork}\cite{continuations}\cite{bashcc}
-
-In a spawn-style interface,
+\cite{callcc_fork}\cite{continuations}
+
+On the other hand, in a spawn-style interface,
 details about the new process are provided up-front as arguments to some function
-which creates the new process all at once;
+which creates the new process all at once.
 \texttt{posix\_spawn} is a typical example of this style.\cite{posix_spawn}\cite{create_process}\cite{chainloading}
@@
 Unlike \texttt{fork}, a spawn-style interface provides an API for process creation
-which is independent from existing system calls.\cite{spawn_h}\cite{subprocess_run}
+which is independent from normal system calls.\cite{spawn_h}\cite{subprocess_run}
 Spawn-style interfaces therefore often lack support for some features,
-and they also require learning a new interface distinct from the normal system call API.\cite{tlpi}\cite{posix_spawn_lwn}
+and they also require learning a new interface distinct from the existing system call API.\cite{tlpi}\cite{posix_spawn_lwn}
 Also, spawn-style interfaces, by their nature as a single call,
@@
 language-specific, object-capability-model, low-abstraction library for the Linux system call interface,
-bypassing libc.\cite{capmyths}
+bypassing libc.\cite{capmyths}\cite{robust_composition}
 The \texttt{rsyscall} project is currently focused on supporting Python,
@@
 the \texttt{rsyscall} library is organized based on the object-capability model.\cite{capmyths}
-The capabilities to make syscalls in any given process are reified as language objects.
+The capability to make syscalls in any specific process is reified as a language object.
 If the capability to make syscalls in a specific process is not passed (in some way) to a function,
@@
 by manipulating capabilities for multiple processes.
-The relevant two capabilities for this paper are the initial capability to the ``local'' process,
-and capabilities to child processes.
+The relevant two types of capabilities for this paper are the initial capability for the ``local'' process
+and capabilities for child processes.
 The ``local'' process is the one which hosts the runtime for the running program,
 and in which a legacy libc would implicitly make syscalls.
-Every \texttt{rsyscall} program starts with the ``local'' process
+Every \texttt{rsyscall} program starts with the capability for the ``local'' process
 and uses it to bootstrap capabilities for other processes.
@@
 is comparable to the performance cost of creating processes while inside a Linux namespace.
-At \twosigma{}, we have deployed direct-style process creation
-as part of a library for distributed system deployment, used in production.
+At \twosigma{}, we've used direct-style process creation
+to implement a production library for distributed system deployment.
 We evaluate in further detail in section \ref{evaluation}.
@@
 we'll demonstrate direct-style process creation
-with several advanced uses of processes
-using more sophisticated Linux features.
+by using several sophisticated features of Linux processes.
 For concision, we assume that we are running with sufficient privileges,
@@
 
-Several of these examples provide functionality that can be done today by specialized software and services.
-Such software often enables one or two of these uses of processes, but prevents others.
+Several of these examples can be implemented by configuring and invoking existing software;
+such software effectively provides a specialized spawn-style interface.
+But such tools are often inflexible.
 For example, a shell allows the creation of pipelines and container systems like Docker allow sandboxing,
@@
 
-One of the key advantages of direct-style process creation is that the features we demonstrate here
-can be easily used in combination,
-to meet the specific needs of individual real-world applications.
+With improved process creation techniques,
+these features can be used directly by programmers
+instead of by configuring stand-alone software.
+In this way,
+individual real-world applications can pick and choose the features that are useful for them,
+as we describe in section \ref{realworld}.
 \subsection{Abstraction through file descriptor passing}\label{fd_abstraction}
@@
 by using file descriptor inheritance.\cite{tlpi}
-All open file descriptors are copied at process creation time from the parent process to the child process;
-the child process can then make independent use of the file descriptors.
+When a process is created, all open file descriptors are copied from the parent process to the child process.
+The child process can then make independent use of the file descriptors.
 We'll discuss this important feature further in section \ref{fdtables}.
@@
 Then we disable \texttt{CLOEXEC} on the child's file descriptor,
-which on Linux we can do by clearing the FD flags with \verb|fcntl(fd, F_SETFD, 0)|.
-Finally, we execute some executable in the child process,
-passing the file descriptor number as an argument;
-since we disabled \texttt{CLOEXEC} on the file descriptor,
-the executable will be able to use that file descriptor through that file descriptor number.
+which on Linux we can do by clearing the FD flags with \verb|fcntl(fd, F_SETFD, 0)|;
+this ensures that the file descriptor will be usable by the new program after we call \texttt{exec}.
+Finally, we execute a new program in the child process,
+passing the file descriptor number as an argument.
+The new program will be able to use the file descriptor we opened through that file descriptor number.
 \subsection{Non-shared-memory concurrency}
 Processes run concurrently,
-which enables more modular programs
+which enables modularity
 and allows exploiting the parallelism of the underlying hardware.
@@
 
-\lstinputlisting[float,language=Python,label={pipe},caption={Creating a parallel processing pipeline}]
+\lstinputlisting[float,language=Python,label={pipe},caption={Creating a concurrent processing pipeline}]
 {examples/pipe/direct.py}
 In listing \ref{pipe},
-we execute a few programs in parallel,
+we execute a few programs concurrently,
 connected by pipes.
@@
 which uses \verb|inherit_fd| and \texttt{fcntl} as described in section \ref{fd_abstraction}.
-We call exec in each child process, passing some file descriptors as arguments to the new programs.
-The whole system then runs in parallel,
-as normal child processes.
+We call exec in each child process to run the desired programs,
+passing some file descriptors as arguments to each.
+The whole system of connected processes then runs concurrently.
 \subsection{Customization without explicit support}
-While, ideally, all resources would be passed by file descriptor,
+While ideally all resources would be passed by file descriptor,
 Unix has a number of resources that are global,
-such as the pid space or the mount table,
-and which cannot be changed on a per-process basis.\cite{capsicum}
+such as the process identifier space or the mount table,
+which cannot be changed on a per-process basis.\cite{capsicum}
 
@@
 
-\verb|CLONE_NEWNS| causes clone to create the process in a new mount namespace,
-which allows us to manipulate the filesystem tree in a way that only this process will see
+\verb|CLONE_NEWNS| causes \texttt{clone} to create the process in a new mount namespace,
+which allows us to manipulate the filesystem tree in a way that only this process will see,
 using the \texttt{mount} system call.\cite{mount_namespaces}\cite{clone}
@@
 this is known as a bind mount.\cite{mount}
-Then we execute some executable,
+Then we execute some program,
 which, when it opens \verb|/etc/foo.conf|, will see the contents of our \verb|custom_foo.conf|.
@@
 again in a new mount namespace using \verb|CLONE_NEWNS|.
-We open the executable that we'll later exec in this child process,
-as well as another file as in listing \ref{fds},
-since we won't be able to do exec or open files normally after the next step.
+Since we won't be able to open files or use \texttt{execv} in the child process after the next step,
+we open several files in the child up-front.
 Then we use \texttt{umount},
@@
 removing it all from the view of this process.
-As in listing \ref{fds}, we unset the \texttt{CLOEXEC} flag from the database file descriptor,
-so that it can be inherited across \texttt{exec};
-we don't need to call \verb|inherit_fd| since, in this case, the database fd was opened directly from the child.
-
-We then exec the file we opened earlier, using \texttt{fexec},
+
+Then, as in listing \ref{fds}, we prepare to pass \texttt{db\_fd} to the new program by unsetting \texttt{CLOEXC}.
+In this case we don't need to call \verb|inherit_fd|,
+since the file descriptor was opened directly from the child.
+We then exec the executable we opened earlier using \texttt{fexec},
 which allows execing a file descriptor,
-and also pass the database file descriptor as an argument,
-as in listing \ref{fds}.
-The underlying system call for \texttt{fexec} is \texttt{execveat};
-this \texttt{fexec} is a helper function.\cite{execveat}
+and pass the database file descriptor number as an argument.\cite{execveat}
 Note that this executable must be statically linked,
-or it wouldn't work in an empty filesystem namespace,
+or it wouldn't work in an empty filesystem namespace
 with no libraries to dynamically link against.
@@
 In listing \ref{pidns},
-we create a child process which we call \texttt{init},
+we create a child process which we'll name \texttt{init},
 passing \verb|CLONE_NEWPID| to create it in a new pid namespace.\cite{clone}
@@
 
-The Filesystem in Userspace (FUSE) Linux subsystem allows a userspace program
-to implement a filesystem.\cite{fuse}
+The Filesystem in Userspace (FUSE) Linux subsystem
+allows a filesystem to be implemented with a userspace program.\cite{fuse}
 Many interesting filesystems have been implemented using FUSE,
@@
 without entering a new namespace ourselves.
-We can do this by creating a child process that shares its file descriptor tables with the parent process.
+We can do this by creating a child process that shares its file descriptor table with the parent process.
 
@@
 Then we can open FUSE files in \verb|ns_child|
-and use those files in the parent process,
+and use those files in \texttt{local},
 through the shared file descriptor table.
-We call \texttt{local.use\_fd} which, like \texttt{inherit\_fd},
-merely updates bookkeeping to create a new handle for the file descriptor existing in the shared fd table;
+We call \texttt{use\_fd} which, like \texttt{inherit\_fd},
+updates bookkeeping to create a new handle for the file descriptor existing in the shared fd table;
 we'll discuss \texttt{use\_fd} in section \ref{handles}.
@@
 can allow similar behavior without sharing the file descriptor table,
-but is substantially more complex for the user.
+but is substantially more complex.\cite{scm_rights}
 Nevertheless, for fork-style and spawn-style process creation,
@@
 
-We chose to implement in userspace first,
+We have chosen to implement in userspace first,
 rather than immediately implementing this in the kernel,
@@
 the kernel will set \texttt{ctid} to zero and then,
-crucially, perform a futex wakeup on it.
+crucially, perform a futex wakeup on it.\cite{clone}
 
@@
 and, if that file descriptor was open at the time the child process was created,
-creates a new handle associated with the child process
+returns a new handle associated with the child process
 for the child process's copy of that file descriptor.
@@
 \verb|use_fd| takes as arguments a process $A$
-and a file descriptor handle associated with any process sharing that process's file descriptor table;
+and a file descriptor handle associated with any process sharing process $A$'s file descriptor table;
 it returns a new handle associated with process $A$ for that file descriptor.
@@
 without intervening programs being aware;
-this is a form of dynamic scope, which, though notorious, is often useful.
+this is a form of dynamic scope, which, though notorious, is sometimes useful.\cite{implicit_params}
 
@@
 just to inherit a file descriptor across a single \texttt{execve}.
-This is primarily useful when the process calling \texttt{execve} has a shared file descriptor table;
+Avoiding that pattern would primarily be useful
+when the process calling \texttt{execve} has a shared file descriptor table;
 in that case, unsetting \texttt{CLOEXEC} before \texttt{execve} is not possible,
@@
 when an error is encountered during child setup.
-Such code still must be explicitly written, like the IPC code we show,
-and code in the parent process must detect the error by decoding the exit code.
+Such an approach removes the need for IPC,
+but still requires similar code for catching errors, encoding them into the exit code,
+and detecting the error in the parent process by decoding the exit code.
 
@@
 errors are reported just like any other system call.
-This is the main immediately-visible simplification of using direct-style process creation.
+This is the main immediately visible simplification of using direct-style process creation.
 
@@
 Our true baseline for performance is the Python standard library's \texttt{subprocess.run};
-this takes an average of 1.4 milliseconds at low memory,
+this takes an average of 1.4 milliseconds at low memory usage,
 while \texttt{rsyscall}'s \texttt{clone} takes an average of 2.2 milliseconds at any memory usage.
@@
 The closest related work known to us
-is efforts to build Unix compatibility environments, including fork-style process creation,
-on academic operating systems which natively use direct-style process creation.\cite{exokernel}
-Such compatibility environments could theoretically be bypassed to perform direct-style process creation
+is efforts to build Unix compatibility environments
+on academic operating systems which natively use direct-style process creation;
+examples include Exokernel and Fuschia.\cite{exokernel}\cite{fuschia_libc}
+Such compatibility environments could be bypassed to perform direct-style process creation
 using the underlying primitives,
@@
 a system call in one process is not evaluated by directly entering the kernel,
-but instead is sent to some other process to be evaluated,
-and the results received from that process.
+but instead is sent to some other process or thread to be evaluated,
+and the results received from that process or thread.
 
@@
 of direct-style process creation in section \ref{examples}.
-Our current implementation of \texttt{rsyscall} is focused on supporting Python,
+Our current implementation of \texttt{rsyscall} has excellent support for Python,
 works entirely in userspace,

### 2020-05-26 research/paper/paper.tex
paper...

@@
 \date{}
-\title{\Large \bf Direct-style process creation on Linux}
+\title{\Large \bf DRAFT: Direct-style process creation on Linux}
 % \author{
@@
 Fork can have poor performance when called from processes with many memory mappings.\cite{forkroad}
-This can be mitigated to some degree, but inflicts further complexity on the caller.\cite{vfork}\cite{vfork_dangerous}
+This can be mitigated to some degree through sytem calls like \texttt{vfork},
+but inflicts further complexity on the caller.\cite{vfork}\cite{vfork_dangerous}
 Processes which are multi-threaded can encounter deadlocks and other issues if they call fork.
@@
 or even a complete change in implementation approach as in section \ref{shared_fd_table}.
-We'll compare our direct-style implementations in this section
-to fork-style implementations
-in section \ref{evaluation}.
+We demonstrate this by comparison to fork-style implementations of these examples in section \ref{ease}.
 
@@
 \texttt{fork} from C,
+\texttt{vfork} from C,
 and \texttt{posix\_spawn} from C.
@@
 \texttt{posix\_spawn} takes an average of 440 microseconds to start a process,
-regardless of the memory usage of the calling process.
+and \texttt{vfork} takes 400 microseconds,
+regardless of the memory usage of the calling process,
 

### 2020-05-26 research/paper/paper.tex
paper

@@
 \usepackage{microtype}
+\usepackage{calc}
 \usepackage{tikz}
@@
 \usepackage{comment}
-
-\newcommand{\cmark}{\ding{51}}
-\newcommand{\xmark}{\ding{55}}
 
@@
 
-% Ugh, these aren't the same size right now.
-\newcommand{\tbad}{\xmark:}
-\newcommand{\tgood}{\cmark:}
-% \newcommand{\tbad}{\text{\sffamily X} :}
-% \newcommand{\tgood}{\checkmark:}
+\newcommand{\tbad}{\makebox[\widthof{\ding{51}}]{\ding{55}}:}
+\newcommand{\tgood}{\ding{51}:}
 
@@
 
-A user of fork must think carefully about the characteristics of the process from which fork is being called.
-Fork can have poor performance when called from processes with many memory mappings.\cite{forkroad}
-This can be mitigated to some degree through sytem calls like \texttt{vfork},
-but inflicts further complexity on the caller.\cite{vfork}\cite{vfork_dangerous}
-Processes which are multi-threaded can encounter deadlocks and other issues if they call fork.
+A user of \texttt{fork} must think carefully about the characteristics of the process
+from which \texttt{fork} is being called.
+\texttt{fork} can have poor performance when called from processes with many memory mappings.\cite{forkroad}
+This can be mitigated to some degree by using \texttt{vfork} or \texttt{CLONE\_VM},
+but those inflict further complexity on the caller.\cite{vfork_dangerous}
+Processes which are multi-threaded can encounter deadlocks and other issues if they call \texttt{fork}.
 \cite{forkroad}\cite{tlpi}\cite{posix_spawn_error_pipe}
@@
 adding complexity.\cite{posix_spawn_error_pipe}
-Even the basic interface of fork - one function call which returns twice - is unusual;
+Even the basic interface of \texttt{fork} --- one function call which returns twice --- is unusual;
 such functions are generally considered to be complex, advanced topics in other settings.
@@
 A few academic operating systems - KeyKOS, seL4, and some others, for example -
-use another style of process creation.\cite{keykos}\cite{sel4}\cite{exokernel}\cite{fuschia}\cite{singularity}
+use another style of process creation.\cite{keykos}\cite{sel4}\cite{exokernel}\cite{fuchsia}\cite{singularity}
 We use the new term "direct-style" to refer to this,
@@
 \texttt{inherit\_fd} performs no system calls,
-it just updates bookkeeping to create a new handle for the inherited file descriptor;
+it just updates bookkeeping to return a new handle for the inherited file descriptor;
 we'll discuss it in more depth in section \ref{handles}.
@@
 using the \texttt{mount} system call.\cite{mount_namespaces}\cite{clone}
-We call \texttt{mount}, passing \texttt{MS.BIND}, to make the file \verb|/home/foo/custom_foo.conf|
+We call \texttt{mount}, passing \texttt{MS\_BIND}, to make the file \verb|/home/foo/custom_foo.conf|
 appear at the path \verb|/etc/foo.conf|;
@@
 such as accessing resources in other namespaces, or in multiple namespaces at once.
-We'll give one example using FUSE.
-
-The Filesystem in Userspace (FUSE) Linux subsystem
-allows a filesystem to be implemented with a userspace program.\cite{fuse}
+We'll give one example using the Filesystem in Userspace (FUSE) Linux subsystem.
+
+FUSE allows a filesystem to be implemented with a userspace program.\cite{fuse}
 Many interesting filesystems have been implemented using FUSE,
@@
 ensures both ends are marked \verb|O_CLOEXEC|,
-performs the fork,
+performs the \texttt{fork},
 closes the write end of the pipe,
@@
 
-For \verb|unshare(CLONE_FILES)|, we perform \verb|close_cloexec| automatically in userspace.
+For \texttt{unshare(CLONE\_FILES)}, we perform \texttt{close\_cloexec} automatically in userspace.
 For \texttt{clone}, we perform \verb|close_cloexec| only on user request,
@@
 is reporting errors back to the main program using IPC.
-This pattern is common in complex usage of fork.\cite{posix_spawn_error_pipe}\cite{python_subprocess_errpipe}
-Many users of fork avoid complex IPC
+This pattern is common in complex usage of \texttt{fork}.\cite{posix_spawn_error_pipe}\cite{python_subprocess_errpipe}
+Many users of \texttt{fork} avoid complex IPC
 by encoding a subset of the error information into the exit code of the child process
@@
 while \texttt{rsyscall}'s \texttt{clone} takes an average of 2.2 milliseconds at any memory usage.
-As expected, the fork-based implementations scale poorly with memory usage,
+As expected, the \texttt{fork}-based implementations scale linearly with memory usage,
 and the C implementations vastly outperform the Python implementations.
@@
 \subsection{Direct-style process creation}
-Baumann (2019) briefly suggests cross-process operations as a possible replacement for fork on Unix,
+Baumann (2019) briefly suggests cross-process operations as a possible replacement for \texttt{fork} on Unix,
 positing many of the same advantages we find in practice.\cite{forkroad}
@@
 on academic operating systems which natively use direct-style process creation;
-examples include Exokernel and Fuschia.\cite{exokernel}\cite{fuschia_libc}
+examples include Exokernel and Fuchsia.\cite{exokernel}\cite{fuchsia_libc}
 Such compatibility environments could be bypassed to perform direct-style process creation
@@
 Many capability-oriented operating systems, such as KeyKOS and seL4,
-have process capabilities which allow one process to operate on another process.\cite{keykos}\cite{sel4}\cite{fuschia}
+have process capabilities which allow one process to operate on another process.\cite{keykos}\cite{sel4}\cite{fuchsia}
 These systems are our main inspiration for this work.

### 2020-05-27 research/paper/paper.tex
paper

@@
 
+Closely related to \texttt{fork}
+are process creation interfaces where the new process is launched running a function of the caller's choice,
+which can then call arbitrary system calls to mutate the new process.
+\cite{clone}\cite{sfork}
+Such interfaces have similar issues as \texttt{fork},
+and we class them together as ``fork-style'' interfaces.
+
 On the other hand, in a spawn-style interface,
@@
 
-A few academic operating systems - KeyKOS, seL4, and some others, for example -
+A few academic operating systems --- KeyKOS, seL4, and some others, for example ---
 use another style of process creation.\cite{keykos}\cite{sel4}\cite{exokernel}\cite{fuchsia}\cite{singularity}
-We use the new term "direct-style" to refer to this,
+We use the new term ``direct-style'' to refer to this,
 since we know of no pre-existing generic term for this style of process creation.
@@
 This allows us to approximate capability-based security.\cite{capsicum}
-\subsection{Nested clone and pid namespaces}
+\subsection{Nested clone and pid namespaces}\label{pidns}
 Process-local resources can also be used to control the lifetime of resources used by that process.
@@
 
-\lstinputlisting[float,language=Python,label={pidns},caption={Nested clone and pid namespace}]
+\lstinputlisting[float,language=Python,label={lst:pidns},caption={Nested clone and pid namespace}]
 {examples/pidns/direct.py}
-In listing \ref{pidns},
+In listing \ref{lst:pidns},
 we create a child process which we'll name \texttt{init},
@@
 
-\lstinputlisting[float,language=Python,label={fuse},caption={Shared file descriptor tables}]
+\lstinputlisting[float,language=Python,label={lst:fuse},caption={Shared file descriptor tables}]
 {examples/fuse/direct.py}
-In listing \ref{fuse},
+In listing \ref{lst:fuse},
 we first create a child process \verb|ns_child| in a new mount namespace with \verb|CLONE_NEWNS|,
@@
 \texttt{rsyscall} can be conceptually divided in two parts:
-the basic cross-process syscall primitive,
+the basic syscall primitive,
 and a language-specific library built on top.
@@
 Such libraries only need to be able to call syscalls and explicitly specify a process in some way;
-they are, for the most part, agnostic to how the cross-process syscall is implemented.
-The cross-process syscall primitive takes a syscall number, and some number of register-sized integer arguments,
-and arranges to call that syscall in the target process,
+they are, for the most part, agnostic to how the syscall primitive is implemented.
+The syscall primitive takes a syscall number, and some number of register-sized integer arguments,
+and arranges to call that syscall in the specified process,
 returning the response as a single register-sized integer.
 
+When making a syscall in the local process, the syscall is performed normally,
+directly on the local thread.
+
+When making a syscall in another process,
 \texttt{rsyscall}'s default userspace cross-process syscall primitive sends the syscall request
@@
 typically a pair of pipes.
+
 We use file descriptors for transport of data rather than shared memory
-to support straightforward integration with event loops.
+to support straightforward integration with existing event loops;
+this is a key design constraint.
+While we only show synchronous programs in this paper,
+calling a syscall in another process may block at any time,
+and a complex program will likely have other requests to service concurrently.
+Even a relatively simple program may want to monitor multiple child processes at a time.
+To support this,
+our implementation fully supports asynchronous usage with a user-provided event loop,
+including the Python \texttt{async/await} coroutine support.\cite{python_coroutines}
 
@@
 
-In general, if a certain file descriptor is in use by the program,
+In general, if a certain file descriptor is in use by our program,
 through a certain process,
@@
 This is the main immediately visible simplification of using direct-style process creation.
+
+It is possible to build an abstracted wrapper for fork-style process creation
+which abstracts away this error reporting code.
+We have built several such wrappers,
+but have ultimately discarded them in favor of direct-style process creation.\cite{sfork}
+
+We found that such wrappers, besides their complexity,
+don't provide any fundamental improvement
+for the issue of communication between the concurrently executing child process and parent process.
+As a result, while such wrappers can remove some boilerplate in easy cases,
+they remain complex in difficult cases like in sections \ref{pidns} and \ref{shared_fd_table},
+which rely on a single program being able to coordinate multiple processes.
+We believe only direct-style process creation is able to easily express such functionality.
 
@@
 \end{figure}
-As demonstrated in sections \ref{pidns} and \ref{fuse},
+As demonstrated in sections \ref{lst:pidns} and \ref{lst:fuse},
 \texttt{rsyscall} supports nested operation.

### 2020-05-27 research/paper/paper.tex
paper

@@
 
-\lstinputlisting[float,language=Python,label={basic},caption={Create new process, change CWD, and exec}]
+\lstinputlisting[float,language=Python,label={lst:basic},caption={Create new process, change CWD, and exec}]
 {examples/basic/direct.py}
-Listing \ref{basic} shows a Python program using \texttt{rsyscall}.
+Listing \ref{lst:basic} shows a Python program using \texttt{rsyscall}.
 We create a new child process using direct-style \texttt{clone} in the local process \verb|local|.
@@
 
-\lstinputlisting[float,language=Python,label={fds},caption={Passing down FDs}]
+\lstinputlisting[float,language=Python,label={lst:fds},caption={Passing down FDs}]
 {examples/fds/direct.py}
-In Listing \ref{fds},
+In Listing \ref{lst:fds},
 we first open a file with read-write permission in the local process.
@@
 
-\lstinputlisting[float,language=Python,label={pipe},caption={Creating a concurrent processing pipeline}]
+\lstinputlisting[float,language=Python,label={lst:pipe},caption={Creating a concurrent processing pipeline}]
 {examples/pipe/direct.py}
-In listing \ref{pipe},
+In listing \ref{lst:pipe},
 we execute a few programs concurrently,
@@
 
-\lstinputlisting[float,language=Python,label={mount},caption={Overriding absolute path using a mount namespace}]
+\lstinputlisting[float,language=Python,label={lst:mount},caption={Overriding absolute path using a mount namespace}]
 {examples/mount/direct.py}
@@
 without changing those files for the rest of the system.
-In listing \ref{mount},
+In listing \ref{lst:mount},
 we create a new child process as normal,
@@
 
-\lstinputlisting[float,language=Python,label={unmount},caption={Unmount all and run executable via fexec}]
+\lstinputlisting[float,language=Python,label={lst:unmount},caption={Unmount all and run executable via fexec}]
 {examples/unmount/direct.py}
-In listing \ref{unmount}
+In listing \ref{lst:unmount}
 we create a new child process,
@@
 
-Then, as in listing \ref{fds}, we prepare to pass \texttt{db\_fd} to the new program by unsetting \texttt{CLOEXC}.
+Then, as in listing \ref{lst:fds}, we prepare to pass \texttt{db\_fd} to the new program by unsetting \texttt{CLOEXC}.
 In this case we don't need to call \verb|inherit_fd|,
@@
 Listing \ref{basic_fork} shows one of our fork-style implementations,
-corresponding to the direct-style implementation in listing \ref{basic}.
+corresponding to the direct-style implementation in listing \ref{lst:basic}.
 
@@
 \end{figure}
-As demonstrated in sections \ref{lst:pidns} and \ref{lst:fuse},
+As demonstrated in sections \ref{pidns} and \ref{shared_fd_table},
 \texttt{rsyscall} supports nested operation.

### 2020-05-27 research/paper/ease.py

@@ -25,7 +25,7 @@ for label in listing_labels:
     fork_wc = len(list(fork_path.open())) if fork_path.exists() else "n/a"
     rows.append((
         label,
-        "\\ref{" + label + "}",
+        "\\ref{lst:" + label + "}",
         direct_wc,
         fork_wc,
     ))

### 2020-05-27 research/paper/paper.tex
paper stufff

@@
 \newcommand{\twosigma}{{[tech company]}}
-\newcommand{\githuburl}{[Github url]}
+\newcommand{\githuburl}{[Github url removed]}
 \date{}
@@
 Traditional process creation interfaces,
-such as fork and spawn,
+such as \texttt{fork} and \texttt{spawn},
 are complex to use, limited in power, and difficult to abstract over.
@@
 
-Our contribution in this paper is such an API, in the form of the \texttt{rsyscall} library.
+Our main contribution in this paper is such an API, in the form of the \texttt{rsyscall} library.
 \texttt{rsyscall} is a
@@
 We evaluate in further detail in section \ref{evaluation}.
+
+In summary, this paper makes the following contributions:
+\begin{compactitem}
+\item We coin the term ``direct-style process creation''
+  to refer to a previously-unnamed style of process creation which is common on academic operating systems.
+\item We built \texttt{rsyscall}, a library for the Linux system call interface following the object-capability model.
+\item Using \texttt{rsyscall}, we built the first implementation of direct-style process creation on a Unix-like kernel.
+\end{compactitem}
+\texttt{rsyscall} is open source and available from \githuburl{}.
 \section{Examples}\label{examples}
@@
 as we describe in section \ref{realworld}.
+
+In these examples and in this paper,
+we'll have multiple processes,
+all of which are under the control of a single program with a single flow of control.
+Because of the absence of multiple concurrent control flows,
+these examples could be executed interactively by a human operator,
+typing them in line by line in a REPL.
+To emphasize this point, we use the first person plural --- ``we'' ---
+when describing the actions of the single program in these examples and in later sections.
 \subsection{Abstraction through file descriptor passing}\label{fd_abstraction}
@@
 
-For \texttt{unshare(CLONE\_FILES)}, we perform \texttt{close\_cloexec} automatically in userspace.
+When we call \texttt{unshare(CLONE\_FILES)},
+we perform \texttt{close\_cloexec} automatically in userspace.
 For \texttt{clone}, we perform \verb|close_cloexec| only on user request,
@@
 such optimization is reserved for future work.
-\subsection{Nested processes, \texttt{CLONE} flags}\label{clone_bench}
+\subsubsection{Nested processes, \texttt{CLONE} flags}\label{clone_bench}
 \begin{figure}

### 2020-05-27 research/paper/paper.tex
paper

@@
 typing them in line by line in a REPL.
-To emphasize this point, we use the first person plural --- ``we'' ---
-when describing the actions of the single program in these examples and in later sections.
+We'll always be working from the perspective of this single program;
+never from the perspective of a child process, because (until we call \texttt{exec} and give up control)
+there is no meaningful program running in the child processes.
 \subsection{Abstraction through file descriptor passing}\label{fd_abstraction}
@@
 which uses \verb|inherit_fd| and \texttt{fcntl} as described in section \ref{fd_abstraction}.
-We call exec in each child process to run the desired programs,
+We call \texttt{execv} in each child process to run the desired programs,
 passing some file descriptors as arguments to each.
@@
 since the file descriptor was opened directly from the child.
-We then exec the executable we opened earlier using \texttt{fexec},
-which allows execing a file descriptor,
+We then run the executable we opened earlier using \texttt{fexec},
+which allows executing a file descriptor,
 and pass the database file descriptor number as an argument.\cite{execveat}
@@
 which can be used exactly like a direct child process.
-We exec in the grandchild,
+We \texttt{exec} in the grandchild,
 and we can monitor the grandchild process from \texttt{init},
@@
 We can handle these responsibilities ourselves directly from our program,
-or we can continue on by execing an init daemon in \texttt{init} to handle it for us.
+or we can continue on by \texttt{exec}ing an init daemon in \texttt{init} to handle it for us.
 \cite{pid_namespaces}\cite{tini}\cite{dumb_init}
@@
 We create another child from \verb|ns_child|
-and use it to exec a FUSE filesystem,
+and use it to \texttt{exec} a FUSE filesystem,
 which will appear only in the mount namespace of \verb|ns_child| and its descendents.

### 2020-05-27 research/paper/paper.tex
paper

@@
 
-\titleformat*{\subsection}{\normalsize\bfseries}
-\titleformat*{\subsubsection}{\normalsize\bfseries}
-\titlespacing{\subsection}{0pt}{1ex}{0.5ex}
-\titlespacing{\subsubsection}{0pt}{0.5ex}{0ex}
+% \titleformat*{\subsection}{\normalsize\bfseries}
+% \titleformat*{\subsubsection}{\normalsize\bfseries}
+% \titlespacing{\subsection}{0pt}{1ex}{0.5ex}
+% \titlespacing{\subsubsection}{0pt}{0.5ex}{0ex}
 \lstset{
@@
 In these examples and in this paper,
-we'll have multiple processes,
+we will have multiple processes,
 all of which are under the control of a single program with a single flow of control.
@@
 there is no meaningful program running in the child processes.
-\subsection{Abstraction through file descriptor passing}\label{fd_abstraction}
+We believe this absence of concurrency simplifies reasoning.
+\subsection{Abstraction through FD passing}\label{fd_abstraction}
 In Linux (and Unix in general),
-a program can pass file descriptors from the parent process to a child process
+a program can pass file descriptors (FDs) from the parent process to a child process
 by using file descriptor inheritance.\cite{tlpi}
@@
 The child process can then make independent use of the file descriptors.
-We'll discuss this important feature further in section \ref{fdtables}.
+We'll discuss this important feature further in section \ref{inheritance}.
 
@@
 it just updates bookkeeping to return a new handle for the inherited file descriptor;
-we'll discuss it in more depth in section \ref{handles}.
+we'll discuss it in more depth in section \ref{inheritance}.
 Then we disable \texttt{CLOEXEC} on the child's file descriptor,
@@
 updates bookkeeping to create a new handle for the file descriptor existing in the shared fd table;
-we'll discuss \texttt{use\_fd} in section \ref{handles}.
+we'll discuss \texttt{use\_fd} in section \ref{inheritance}.
 
@@
 From the system calls related to process creation,
-only \texttt{execve} and \texttt{unshare} need substantial further attention.
+only \texttt{execve} needs substantial further attention.
 \subsection{\texttt{exec}}\label{execve}
@@
 when the \texttt{WEXECED} event happens.
-\subsection{Managing file descriptor tables}\label{fdtables}
-\texttt{clone} (without \verb|CLONE_FILES|) and \texttt{fork}
-deal with file descriptor tables in the same way:
-The new child process has a new file descriptor table,
-containing copies of all the file descriptors existing in the parent at the time of the system call.
-This is known as file descriptor inheritance.
-
-As demonstrated in section \ref{shared_fd_table},
-when the \verb|CLONE_FILES| flag is passed to \texttt{clone},
-the parent process and the child process will share a single file descriptor table,
-so changes to the file descriptor table made in one process will affect the other.
-
-The file descriptor table can then be unshared by calling \verb|unshare(CLONE_FILES)| or \texttt{execve}.
-After either system call, the process will have a new file descriptor table
-which is no longer shared with any other process.
-All the file descriptors existing in the old table at the time of the system call
-will be copied into the new file descriptor table,
-except for, in the case of \texttt{execve}, file descriptors with the \texttt{CLOEXEC} flag set.
-\subsubsection{Tracking file descriptors through handles}\label{handles}
-After \texttt{clone}, \verb|unshare(CLONE_FILES)|, or \texttt{execve}
-has copied all the file descriptors from some old table into a new table,
-we need to decide which file descriptors we want to keep open in the new table,
-and which file descriptors should be closed.
-
-In general, if a certain file descriptor is in use by our program,
-through a certain process,
-that file descriptor should stay open in that process's file descriptor table.
-
-To achieve this, file descriptors are used through garbage-collected handles;
-each handle is associated with a process and contains a specific file descriptor number.
-A handle is ``valid''
-when that file descriptor number refers to an open file descriptor in that process.
-A handle is ``live''
-when the handle is in use by the program, which we can determine
-from Python's garbage collection system.
-We then keep open exactly those file descriptors which are necessary to keep all live handles, valid.
-
-When a process changes file descriptor table through \verb|unshare(CLONE_FILES)|,
-we need to make sure all the live file descriptor handles associated with that process continue to be valid.
-This is straightforward:
-we enumerate all the handles associated with the process,
-and don't close the file descriptors that those handles refer to.
-
-Likewise, if a process changes file descriptor table through \texttt{execve}
-in a way that allows the program to keep control of the process,
-we make sure to keep all the file descriptors with live handles open.
-This requires careful management of \texttt{CLOEXEC}.
-
-When a child process is created,
-there are no file descriptor handles associated with it.
-We can make file descriptor handles by opening file descriptors in the child,
-or by creating new handles for file descriptors opened in other processes.
-
-We can use the \verb|inherit_fd| function to create handles for inherited file descriptors.
-\verb|inherit_fd| takes a file descriptor handle associated with the parent process,
-and, if that file descriptor was open at the time the child process was created,
-returns a new handle associated with the child process
-for the child process's copy of that file descriptor.
-Once the inherited file descriptors are closed as described in section \ref{cloexec},
-\verb|inherit_fd| can no longer be used.
-
-If the process shares its file descriptor table with any other processes,
-we can use the \verb|use_fd| function to create handles for file descriptors
-in the shared file descriptor table.
-\verb|use_fd| takes as arguments a process $A$
-and a file descriptor handle associated with any process sharing process $A$'s file descriptor table;
-it returns a new handle associated with process $A$ for that file descriptor.
-\subsubsection{Closing file descriptors after inheritance}\label{cloexec}
-After creating a new file descriptor table,
-and determining which file descriptors should be kept open,
-we must promptly close other file descriptors.
-Leaving unwanted file descriptors open in the new table wastes resources,
-and can also cause bugs.
-For example, it's common practice to close the write end of a pipe
-and expect an EOF on the read end;
-if the write end is copied into a new file descriptor table between creating and closing,
-and the write end is never closed in the new table,
-the read end will never get an EOF.
-
-% TODO should definitely cite this
-However, we can't simply close all other file descriptors,
-because we must preserve the semantics of implicit file descriptor inheritance.
-The possibility of implicit inheritance of file descriptors is an old Unix feature,
-which is useful in a wide variety of situations,
-in much the same way as environment variables, which are also implicitly inherited.
-As with environment variables,
-implicitly inherited file descriptors allow a resource to be passed down a process hierarchy
-without intervening programs being aware;
-this is a form of dynamic scope, which, though notorious, is sometimes useful.\cite{implicit_params}
-
-Fortunately, file descriptors which should be implicitly inherited are marked:
-they have \texttt{CLOEXEC} turned off.
-For implicit inheritance to function, \texttt{CLOEXEC} must not be set;
-otherwise, the file descriptor will be closed, rather than inherited,
-the next time the process calls \texttt{execve}.
-And since \texttt{execve} can be called at any time in a general purpose program,
-any library that does not want its file descriptors to be inherited across \texttt{execve}
-must keep \texttt{CLOEXEC} turned on at all times.
-So, at any time, we know that the file descriptors with \texttt{CLOEXEC} off
-are exactly the file descriptors which should be implicitly inherited.
-
-Therefore, we simply need to close all file descriptors that have \texttt{CLOEXEC} set,
-and which are not referenced by a live file descriptor handle;
-we refer to this operation as \verb|close_cloexec|.
-
-When we call \texttt{unshare(CLONE\_FILES)},
-we perform \texttt{close\_cloexec} automatically in userspace.
-For \texttt{clone}, we perform \verb|close_cloexec| only on user request,
-so that the user can control when \verb|inherit_fd| stops working.
-Even if the user never requests that unused file descriptors be closed,
-as long as they call \texttt{execve}, the table will be cleaned at that point.
-In both of these cases,
-we would benefit from a new system call which takes a list of file descriptors
-and closes all \texttt{CLOEXEC} file descriptors except for those on the list;
-this would replace our userspace implementation of \verb|close_cloexec|.
-
-For \texttt{execve}, \verb|close_cloexec| is, in some sense, done automatically for us by the kernel;
-except we don't get the chance to preserve live file descriptors handles,
-other than by unsetting \texttt{CLOEXEC} on them before calling \texttt{execve}.
-Again an extension to the kernel would be useful:
-A new argument to \texttt{execve} which allows specifying a list of file descriptors
-which should be inherited across \texttt{execve} even if \texttt{CLOEXEC} is set on them.
-This new argument would allow programs to avoid the pattern
-of unsetting \texttt{CLOEXEC}, calling \texttt{execve}, and then immediately re-setting \texttt{CLOEXEC},
-just to inherit a file descriptor across a single \texttt{execve}.
-Avoiding that pattern would primarily be useful
-when the process calling \texttt{execve} has a shared file descriptor table;
-in that case, unsetting \texttt{CLOEXEC} before \texttt{execve} is not possible,
-because some other process might also call \texttt{execve}
-and erroneously inherit the file descriptor from the shared file descriptor table.
+\subsection{Handling file descriptors}\label{fd_handles}
+In Unix, there are many types of resources maintained inside the kernel,
+such as file descriptors or memory mappings,
+which are referred to by identifiers which are valid only within a certain context\cite{tlpi}.
+For example, file descriptors are referred to by file descriptor numbers
+which are valid only in a certain file descriptor table;
+memory and memory mappings are referred to by memory addresses
+which are valid only in a certain address space;
+the same pattern is followed for a number of other resources.
+Each process has one file descriptor table, one address space, and one of each of the other contexts.
+Two or more processes can share individual contexts;
+for example, two processes might share an address space, but not a file descriptor table.
+
+The core issues here are the same for all such resources,
+so we'll focus on file descriptors.
+For concision and clarity, we'll abbreviate ``file descriptor'' as ``FD'' in this section.
+
+A typical POSIX program operates within only one process,
+and so only needs to concern itself with one of each kind of context.
+To identify and operate on an FD, it can use just an FD number,
+which is valid within the global implicit process's FD table.
+That's sufficient to make syscalls involving that FD.
+
+A program using \texttt{rsyscall}, however,
+operates in multiple processes,
+and has to deal with resources across multiple contexts.
+To identify an FD,
+it needs not just a FD number,
+but also some kind of identifier for an FD table.
+To then operate on that FD,
+it needs a capability for a process with that FD table,
+so that it can make syscalls in that process.
+
+To manage this, we add a new concept alongside FD, FD number, and FD table:
+The FD handle.
+An FD handle is a pair of an FD number and a process capability.
+This both precisely identifies an FD, and allows operating on it.
+We say an FD handle is associated with a process
+if that FD handle contains that process capability.
+
+All system calls which would otherwise return FD numbers or take FD numbers as an argument,
+now instead return FD handles or take FD handles as an argument.
+For convenience, as shown in section \ref{examples} for \texttt{fcntl},
+helper methods on the FD handle object are provided for many system calls dealing with a single FD;
+these methods simply call the corresponding underlying system call
+using the process capability contained in the FD handle.
+
+In this way, we can work with FDs without the ambiguity of dealing with raw FD numbers in a multi-process context.
+\subsubsection{File descriptor inheritance and table sharing}\label{inheritance}
+The picture is complicated by the need to support some FD behaviors which implicitly cross between processes.
+FDs are copied between processes at process creation time through FD inheritance,
+and two processes can share the same FDs through a shared FD table.
+
+FD inheritance is a behavior exhibited by \texttt{clone} (without \verb|CLONE_FILES|) and by \texttt{fork}.
+When these system calls are called,
+they create a new child process with a new FD table.
+The new FD table contains copies of all the FDs existing in the parent process's FD table at the time of the system call,
+at the same FD numbers.
+
+FD table sharing occurs when \texttt{clone} \textit{with} \texttt{CLONE\_FILES} is called;
+this creates a new child process that shares its FD table with its parent.
+As a result, either process can operate on any FD in the table,
+through the same FD numbers.
+
+These behaviors are unusual in that they operate on all open FDs at once,
+rather than individual FDs explicitly denoted by system call arguments.
+There are other ways to pass FDs between processes,
+such as \texttt{SCM\_RIGHTS}\cite{scm_rights},
+but those operate on individual FDs through explicit arguments to system calls.
+One or more FD handles are explicitly passed to a system call in one process,
+and one or more FD handles are explicitly returned from another system call in another process.
+The behaviors of FD inheritance and FD table sharing happen for all FDs, implicitly,
+and so they must be dealt with differently to be fully supported.
+
+\texttt{rsyscall} supports both behaviors in the same way,
+by allowing us to create new FD handles on-demand
+referring to already-open FDs created by FD inheritance or FD table sharing;
+by creating new FD handles in new processes, we can call syscalls in those processes on those FDs.
+\texttt{rsyscall} provides two functions
+\texttt{inherit\_fd} and \texttt{use\_fd}, which support FD inheritance and FD table sharing, respectively.
+Both functions take as arguments a process capability $A$ and an FD handle $H$ associated with some other process $B$,
+and return a new FD handle associated with process $A$.
+For \texttt{inherit\_fd},
+this new handle refers to an \textit{inherited copy} of the FD which $H$ refers to;
+\texttt{inherit\_fd} will fail if $B$ is not $A$'s parent process,
+or if $H$ was not open at the time $A$ was created.
+For \texttt{use\_fd},
+this new handle refers to the \textit{same} FD that $H$ refers to, through the shared FD table;
+\texttt{use\_fd} will fail if $A$ and $B$ don't share the same FD table.
 \section{Evaluation}\label{evaluation}
@@
 
-Several other aspects of our implementation would also be improved by new features in the Linux kernel,
-as discussed in sections \ref{execve} and \ref{cloexec}.
+Other aspects of our implementation would also be improved by new features in the Linux kernel;
+one examples is discussed in section \ref{execve}.
 \subsection{Portability to other Unix systems}

### 2020-05-27 research/paper/paper.tex
paper

@@
 \usepackage{graphicx}
-\usepackage{hyperref}
 \usepackage{float}
@@
 % BEGIN RECEIVE ORGTBL compare
- & fork & spawn & direct-style\\
+ & fork-style & spawn-style & direct-style\\
 \hline
@@
 #+ORGTBL: SEND compare orgtbl-to-latex :splice t
-|                             | fork                              | spawn                       | direct-style                          |
+|                             | fork-style                        | spawn-style                 | direct-style                          |
 |-----------------------------+-----------------------------------+-----------------------------+---------------------------------------|
@@
 both processes execute concurrently,
-and the child process typically inherits all but a few attributes from the parent process.\cite{manfork}
+and the child process typically inherits all but a few attributes from the parent process\cite{manfork}.
 The child process can call arbitrary system calls
@@
 from which \texttt{fork} is being called.
-\texttt{fork} can have poor performance when called from processes with many memory mappings.\cite{forkroad}
+\texttt{fork} can have poor performance when called from processes with many memory mappings\cite{forkroad}.
 This can be mitigated to some degree by using \texttt{vfork} or \texttt{CLONE\_VM},
-but those inflict further complexity on the caller.\cite{vfork_dangerous}
-Processes which are multi-threaded can encounter deadlocks and other issues if they call \texttt{fork}.
-\cite{forkroad}\cite{tlpi}\cite{posix_spawn_error_pipe}
+but those inflict further complexity on the caller\cite{vfork_dangerous}.
+Processes which are multi-threaded can encounter deadlocks and other issues if they call \texttt{fork}
+\cite{forkroad}\cite{tlpi}\cite{posix_spawn_error_pipe}.
+
 Some thread libraries provide partial mitigations for multi-threading issues,
-but it's up to user code to make use of those mitigations.\cite{pthread_atfork}
+but it's up to user code to make use of those mitigations\cite{pthread_atfork}.
 
@@
 communicating events in the child back to the parent requires some form of IPC,
-adding complexity.\cite{posix_spawn_error_pipe}
+adding complexity\cite{posix_spawn_error_pipe}.
 Even the basic interface of \texttt{fork} --- one function call which returns twice --- is unusual;
-such functions are generally considered to be complex, advanced topics in other settings.
-\cite{callcc_fork}\cite{continuations}
+such functions are generally considered to be complex, advanced topics in other settings
+\cite{callcc_fork}\cite{continuations}.
 
@@
 are process creation interfaces where the new process is launched running a function of the caller's choice,
-which can then call arbitrary system calls to mutate the new process.
-\cite{clone}\cite{sfork}
+which can then call arbitrary system calls to mutate the new process
+\cite{clone}\cite{sfork}.
 Such interfaces have similar issues as \texttt{fork},
@@
 which creates the new process all at once.
-\texttt{posix\_spawn} is a typical example of this style.\cite{posix_spawn}\cite{create_process}\cite{chainloading}
-Other details, such as security context, are inherited from the parent process.\cite{tlpi}\cite{create_process}
+\texttt{posix\_spawn} is a typical example of this style
+\cite{posix_spawn}\cite{create_process}\cite{chainloading}.
+Other details, such as security context, are inherited from the parent process\cite{tlpi}\cite{create_process}.
 Spawn-style APIs
 can be called from any parent process without concerns about memory usage or multi-threading,
-when used correctly.
-\cite{forkroad}\cite{posix_spawn_error_pipe}
+when used correctly
+\cite{forkroad}\cite{posix_spawn_error_pipe}.
 
 Unlike \texttt{fork}, a spawn-style interface provides an API for process creation
-which is independent from normal system calls.\cite{spawn_h}\cite{subprocess_run}
+which is independent from normal system calls\cite{spawn_h}\cite{subprocess_run}.
 Spawn-style interfaces therefore often lack support for some features,
-and they also require learning a new interface distinct from the existing system call API.\cite{tlpi}\cite{posix_spawn_lwn}
+and they also require learning a new interface distinct from the existing system call API
+\cite{tlpi}\cite{posix_spawn_lwn}.
 Also, spawn-style interfaces, by their nature as a single call,
 don't allow programs to branch on the results of individual modifications to the new process,
-and have worse error reporting for such modifications.
-\cite{posix_spawn}\cite{posix_spawn_error_pipe}
+and have worse error reporting for such modifications
+\cite{posix_spawn}\cite{posix_spawn_error_pipe}.
 
@@
 
-A few academic operating systems --- KeyKOS, seL4, and some others, for example ---
-use another style of process creation.\cite{keykos}\cite{sel4}\cite{exokernel}\cite{fuchsia}\cite{singularity}
+A few academic operating systems
+--- KeyKOS\cite{keykos} and seL4\cite{sel4}, for example ---
+use another style of process creation\cite{exokernel}\cite{fuchsia}\cite{singularity}.
 We use the new term ``direct-style'' to refer to this,
@@
 the syscalls that can mutate a process
-take explicit arguments to indicate the process on which they should operate.\cite{keykos}\cite{sel4}
+take explicit arguments to indicate the process on which they should operate\cite{keykos}\cite{sel4}.
 In this way, the same syscalls that can mutate a process while it is running
@@
 language-specific, object-capability-model, low-abstraction library for the Linux system call interface,
-bypassing libc.\cite{capmyths}\cite{robust_composition}
+bypassing libc\cite{capmyths}\cite{robust_composition}.
 The \texttt{rsyscall} project is currently focused on supporting Python,
@@
 Unlike typical C libraries such as glibc or musl,
-the \texttt{rsyscall} library is organized based on the object-capability model.\cite{capmyths}
+the \texttt{rsyscall} library is organized based on the object-capability model\cite{capmyths}\cite{robust_composition}.
 The capability to make syscalls in any specific process is reified as a language object.
@@
 For concision, we assume that we are running with sufficient privileges,
-but these examples can all be run without privileges with appropriate use of \verb|CLONE_NEWUSER|.
-\cite{user_namespaces}\cite{unpriv_fuse}
+but these examples can all be run without privileges with appropriate use of \verb|CLONE_NEWUSER|
+\cite{user_namespaces}\cite{unpriv_fuse}.
 
@@
 For example, a shell allows the creation of pipelines and container systems like Docker allow sandboxing,
-but the two are difficult to combine.\cite{docker_pipe}
+but the two are difficult to combine\cite{docker_pipe}.
 
@@
 a program can pass file descriptors (FDs) from the parent process to a child process
-by using file descriptor inheritance.\cite{tlpi}
+by using file descriptor inheritance\cite{tlpi}.
 When a process is created, all open file descriptors are copied from the parent process to the child process.
@@
 so this allows the parent process to pass a variety of resources to the child process,
-such as files, network connections, pipes, or other resources.\cite{ucspi}
+such as files, network connections, pipes, or other resources\cite{ucspi}.
 Since the resource is created outside the child process and passed down as an opqaue file descriptor,
@@
 This is a form of abstraction,
-and so we know ``for free''
+and so we know ``for free''\cite{theoremsforfree}
 that the program running in the child process
 will not rely on the details of how the file descriptor was created,
-such as the filename that we opened or the hostname to which we connected.\cite{theoremsforfree}
+such as the filename that we opened or the hostname to which we connected.
 
@@
 they can provide a less complex parallel programming environment
-than shared-memory thread-based approaches.\cite{threads}
+than shared-memory thread-based approaches\cite{threads}.
 
@@
 such as the process identifier space or the mount table,
-which cannot be changed on a per-process basis.\cite{capsicum}
+which cannot be changed on a per-process basis\cite{capsicum}.
 
 In Linux, many global resources have been made process-local
-through the namespaces system.\cite{lwn_namespaces}
+through the namespaces system\cite{lwn_namespaces}.
 Like other process-local resources such as the current working directory,
@@
 
-One key use of namespaces is to implement large-scale container systems such as Docker.\cite{lwn_namespaces}
+One key use of namespaces is to implement large-scale container systems such as Docker\cite{lwn_namespaces}.
 However, this is far from the only use.
@@
 in ways that were not anticipated at development time
-by customizing the environment the process runs in.\cite{plan9ns}
+by customizing the environment the process runs in\cite{plan9ns}.
 Other system calls such as \texttt{chdir} and \texttt{chroot} can also be used for this,
-but namespaces allow new ways of customization.\cite{mount_namespaces}
+but namespaces allow new ways of customization\cite{mount_namespaces}.
 
@@
 which allows us to manipulate the filesystem tree in a way that only this process will see,
-using the \texttt{mount} system call.\cite{mount_namespaces}\cite{clone}
+using the \texttt{mount} system call\cite{mount_namespaces}\cite{clone}.
 We call \texttt{mount}, passing \texttt{MS\_BIND}, to make the file \verb|/home/foo/custom_foo.conf|
 appear at the path \verb|/etc/foo.conf|;
-this is known as a bind mount.\cite{mount}
+this is known as a bind mount\cite{mount}.
 Then we execute some program,
@@
 we can also deny the process access to resources that it otherwise receives by default.
-This is a key part of creating a secure sandbox for potentially malicious code.\cite{seccomp}\cite{firejail}\cite{gvisor}
+This is a key part of creating a secure sandbox for potentially malicious code\cite{seccomp}\cite{firejail}\cite{gvisor}.
 
@@
 which allows executing a file descriptor,
-and pass the database file descriptor number as an argument.\cite{execveat}
+and pass the database file descriptor number as an argument\cite{execveat}.
 Note that this executable must be statically linked,
@@
 A sandbox which is truly robust against malicious or compromised programs requires additional steps,
-but this is a substantial start.\cite{firejail}\cite{gvisor}
+but this is a substantial start\cite{firejail}\cite{gvisor}.
 Such a technique can also be used when a full sandbox is not relevant,
@@
 as we do here with the database file descriptor.
-This allows us to approximate capability-based security.\cite{capsicum}
+This allows us to approximate capability-based security\cite{capsicum}.
 \subsection{Nested clone and pid namespaces}\label{pidns}
@@
 those subprocesses may unintentionally leak,
-and be left running on the system even after the program itself has stopped.\cite{caternfork}
+and be left running on the system even after the program itself has stopped\cite{caternfork}.
 
@@
 When the init process dies,
-all other processes in the pid namespace are destroyed.\cite{pid_namespaces}
+all other processes in the pid namespace are destroyed\cite{pid_namespaces}.
 
@@
 we create a child process which we'll name \texttt{init},
-passing \verb|CLONE_NEWPID| to create it in a new pid namespace.\cite{clone}
+passing \verb|CLONE_NEWPID| to create it in a new pid namespace\cite{clone}.
 
@@
 We can handle these responsibilities ourselves directly from our program,
-or we can continue on by \texttt{exec}ing an init daemon in \texttt{init} to handle it for us.
-\cite{pid_namespaces}\cite{tini}\cite{dumb_init}
+or we can continue on by \texttt{exec}ing an init daemon in \texttt{init} to handle it for us
+\cite{pid_namespaces}\cite{tini}\cite{dumb_init}.
 
@@
 
-FUSE allows a filesystem to be implemented with a userspace program.\cite{fuse}
-Many interesting filesystems have been implemented using FUSE,
-representing a variety of resources as files.\cite{sshfs}\cite{encfs}\cite{mp3fs}
+FUSE\cite{fuse} allows a filesystem to be implemented with a userspace program.
+Many interesting filesystems\cite{sshfs}\cite{encfs}\cite{mp3fs} have been implemented using FUSE,
+representing a variety of resources as files.
 
@@
 and this time also pass \verb|CLONE_FILES|,
-which causes the file descriptor table to be shared between the parent process and the child process.\cite{clone}
+which causes the file descriptor table to be shared between the parent process and the child process\cite{clone}.
 We create another child from \verb|ns_child|
@@
 can allow similar behavior without sharing the file descriptor table,
-but is substantially more complex.\cite{scm_rights}
+but is substantially more complex\cite{scm_rights}.
 Nevertheless, for fork-style and spawn-style process creation,
@@
 our implementation fully supports asynchronous usage with a user-provided event loop,
-including the Python \texttt{async/await} coroutine support.\cite{python_coroutines}
+including the Python \texttt{async/await} coroutine support\cite{python_coroutines}.
 
@@
 In the past, attempts to upstream kernel support for novel features without extensive userspace usage
-has had a poor reception in the Linux kernel community.\cite{lwn_checkpoint}\cite{first_class_address_space}
+has had a poor reception in the Linux kernel community\cite{lwn_checkpoint}\cite{first_class_address_space}.
 We believe proving viability in userspace first will be better in the long term.
@@
 including system calls,
-but is not suitable for us.\cite{ptrace}
+but is not suitable for us\cite{ptrace}.
 As we discuss in section \ref{related}, a number of systems, such as \texttt{gdb} and \texttt{strace},
@@
 for example, if the target process is at a different privilege level,
-we will want it to be in a different address space for security reasons.\cite{vfork_dangerous}
+we will want it to be in a different address space for security reasons\cite{vfork_dangerous}.
 There are a number of available techniques in that scenario;
@@
 while waiting for the child process to exit or call \texttt{execve},
-which is immediately unsuitable.\cite{vfork}
+which is immediately unsuitable\cite{vfork}.
 \texttt{fork} lacks many features which are restricted to \texttt{clone}.
@@
 the kernel will set \texttt{ctid} to zero and then,
-crucially, perform a futex wakeup on it.\cite{clone}
+crucially, perform a futex wakeup on it\cite{clone}.
 
@@
 but the relevant detail for us here is that we can wait on an address
-until a futex wakeup is performed on that address.\cite{futex}
+until a futex wakeup is performed on that address\cite{futex}.
 This means we can wait on \texttt{ctid} until the futex wakeup is performed,
@@
 and have this child process exit when the futex has a wakeup.
-We can then monitor these child processes from an event loop using standard techniques.\cite{signalfd}\cite{pidfd}
+We can then monitor these child processes from an event loop using standard techniques\cite{signalfd}\cite{pidfd}.
 
@@
 our implementation would be simplified, and the ambiguity eliminated,
-if we had this ability on Linux.\cite{freebsd_kqueue}
+if we had this ability on Linux\cite{freebsd_kqueue}.
 One approach would be to add a new \texttt{clone} flag to
@@
 \texttt{rsyscall} provides two functions
-\texttt{inherit\_fd} and \texttt{use\_fd}, which support FD inheritance and FD table sharing, respectively.
+\texttt{inherit\_fd} (used in listings \ref{lst:fds} and \ref{lst:pipe})
+and \texttt{use\_fd} (used in listing \ref{lst:fuse}), which support FD inheritance and FD table sharing, respectively.
 Both functions take as arguments a process capability $A$ and an FD handle $H$ associated with some other process $B$,
@@
 is reporting errors back to the main program using IPC.
-This pattern is common in complex usage of \texttt{fork}.\cite{posix_spawn_error_pipe}\cite{python_subprocess_errpipe}
+This pattern is common in complex usage of \texttt{fork}\cite{posix_spawn_error_pipe}\cite{python_subprocess_errpipe}.
 Many users of \texttt{fork} avoid complex IPC
@@
 We have built several such wrappers,
-but have ultimately discarded them in favor of direct-style process creation.\cite{sfork}
+but have ultimately discarded them in favor of direct-style process creation\cite{sfork}.
 
@@
 to start up services in parallel,
-substantially speeding up system startup.\cite{socketactivation}
+substantially speeding up system startup\cite{socketactivation}.
 We are able to bootstrap connections over our internal shared memory transports
@@
 Baumann (2019) briefly suggests cross-process operations as a possible replacement for \texttt{fork} on Unix,
-positing many of the same advantages we find in practice.\cite{forkroad}
+positing many of the same advantages we find in practice\cite{forkroad}.
 Our work, though developed independently,
@@
 on academic operating systems which natively use direct-style process creation;
-examples include Exokernel and Fuchsia.\cite{exokernel}\cite{fuchsia_libc}
+examples include Exokernel and Fuchsia\cite{exokernel}\cite{fuchsia_libc}.
 Such compatibility environments could be bypassed to perform direct-style process creation
@@
 \subsection{Process capabilities}
-Many capability-oriented operating systems, such as KeyKOS and seL4,
-have process capabilities which allow one process to operate on another process.\cite{keykos}\cite{sel4}\cite{fuchsia}
+Many capability-oriented operating systems, such as KeyKOS\cite{keykos}, seL4\cite{sel4}, and others\cite{fuchsia},
+have process capabilities which allow one process to operate on another process.
 These systems are our main inspiration for this work.
@@
 On several Unix systems,
-Capsicum provides something called process descriptors,
-which is a file descriptor handle for a process.\cite{capsicum}
-A similar feature has been recently added to Linux in the form of the pidfd API.\cite{pidfd}
+Capsicum\cite{capsicum} provides something called process descriptors,
+which is a file descriptor handle for a process.
+A similar feature has been recently added to Linux in the form of the pidfd API\cite{pidfd}.
 These notions of process capability are quite limited, however;
@@
 \subsection{Capability-secure libc replacements}
-Capsicum provides a set of system calls
-which can be used to provide a capability-secure sandbox.\cite{capsicum}
-Latter efforts, including CloudABI and WASI,
-developed this into a partial or full replacement for a POSIX libc.\cite{oblivious}\cite{cloudabi}\cite{wasi}
+Capsicum\cite{capsicum} provides a set of system calls
+which can be used to provide a capability-secure sandbox.
+Latter efforts\cite{oblivious}, including CloudABI\cite{cloudabi} and WASI\cite{wasi},
+developed this into a partial or full replacement for a POSIX libc.
 Like all other Unix libcs that we know of,
@@
 
-PLASH provides a spawn-style API, in the form of a shell,
-to launch processes in a capability-secure environment.\cite{plash}
+PLASH\cite{plash} provides a spawn-style API, in the form of a shell,
+to launch processes in a capability-secure environment.
 PLASH, like all spawn-style APIs, abstracts over the native Linux environment,

### 2020-05-27 research/paper/paper.tex
paper

@@
 \cite{forkroad}\cite{tlpi}\cite{posix_spawn_error_pipe}.
-
-Some thread libraries provide partial mitigations for multi-threading issues,
-but it's up to user code to make use of those mitigations\cite{pthread_atfork}.
+Some thread libraries provide partial mitigations for multi-threading issues\cite{pthread_atfork},
+but it's up to user code to make use of those mitigations.
 
 Since the child process runs concurrently with the parent process,
-communicating events in the child back to the parent requires some form of IPC,
-adding complexity\cite{posix_spawn_error_pipe}.
+communicating events in the child back to the parent requires some form of IPC\cite{posix_spawn_error_pipe},
+adding complexity.
 Even the basic interface of \texttt{fork} --- one function call which returns twice --- is unusual;
@@
 which creates the new process all at once.
-\texttt{posix\_spawn} is a typical example of this style
-\cite{posix_spawn}\cite{create_process}\cite{chainloading}.
+\texttt{posix\_spawn}\cite{posix_spawn} is a typical example of this style
+\cite{chainloading}\cite{create_process}.
 Other details, such as security context, are inherited from the parent process\cite{tlpi}\cite{create_process}.
@@
 when used correctly
-\cite{forkroad}\cite{posix_spawn_error_pipe}.
+\cite{posix_spawn_error_pipe}\cite{forkroad}.
 
@@
 and they also require learning a new interface distinct from the existing system call API
-\cite{tlpi}\cite{posix_spawn_lwn}.
+\cite{posix_spawn_lwn}\cite{tlpi}.
 Also, spawn-style interfaces, by their nature as a single call,
@@
 A few academic operating systems
---- KeyKOS\cite{keykos} and seL4\cite{sel4}, for example ---
-use another style of process creation\cite{exokernel}\cite{fuchsia}\cite{singularity}.
+--- KeyKOS\cite{keykos} and seL4\cite{sel4}, among others\cite{exokernel}\cite{fuchsia}\cite{singularity} ---
+use another style of process creation.
 We use the new term ``direct-style'' to refer to this,
@@
 by manipulating capabilities for multiple processes.
-The relevant two types of capabilities for this paper are the initial capability for the ``local'' process
+The relevant two types of capabilities for this paper are the initial capability for the ``local'' process,
 and capabilities for child processes.
@@
 \item We built \texttt{rsyscall}, a library for the Linux system call interface following the object-capability model.
-\item Using \texttt{rsyscall}, we built the first implementation of direct-style process creation on a Unix-like kernel.
+\item As part of \texttt{rsyscall}, we built the first implementation of direct-style process creation on a Unix-like kernel.
 \end{compactitem}
@@
 
+In these examples and in this paper,
+a single program with a single flow of control is operating across multiple processes.
+We'll always describe operations from the perspective of this single synchronous program,
+never from the perspective of an individual parent or child process.
+
 These examples are not novel, as such, as they use only conventional Linux system calls;
@@
 as we describe in section \ref{realworld}.
-
-In these examples and in this paper,
-we will have multiple processes,
-all of which are under the control of a single program with a single flow of control.
-Because of the absence of multiple concurrent control flows,
-these examples could be executed interactively by a human operator,
-typing them in line by line in a REPL.
-We'll always be working from the perspective of this single program;
-never from the perspective of a child process, because (until we call \texttt{exec} and give up control)
-there is no meaningful program running in the child processes.
-We believe this absence of concurrency simplifies reasoning.
 \subsection{Abstraction through FD passing}\label{fd_abstraction}
@@
 which on Linux we can do by clearing the FD flags with \verb|fcntl(fd, F_SETFD, 0)|;
-this ensures that the file descriptor will be usable by the new program after we call \texttt{exec}.
+this ensures that the file descriptor will be usable by the new program after we call \texttt{exec}\cite{fcntl}.
 Finally, we execute a new program in the child process,
@@
 The new program will be able to use the file descriptor we opened through that file descriptor number.
-\subsection{Non-shared-memory concurrency}
+\subsection{Non-shared-memory concurrency}\label{pipe}
 Processes run concurrently,
@@
 While ideally all resources would be passed by file descriptor,
-Unix has a number of resources that are global,
+traditional Unix has a number of resources that are global,
 such as the process identifier space or the mount table,
@@
 Then, as in listing \ref{lst:fds}, we prepare to pass \texttt{db\_fd} to the new program by unsetting \texttt{CLOEXC}.
-In this case we don't need to call \verb|inherit_fd|,
-since the file descriptor was opened directly from the child.
+Unlike in listing \ref{lst:fds}, we don't need to call \verb|inherit_fd|,
+since the file descriptor was opened directly from the child process.
 We then run the executable we opened earlier using \texttt{fexec},
@@
 a poorly coded program may allocate global resources
-and not ensure that they are cleaned up,
-leaving behind unused garbage on the system.
-
-One resource which is not automatically cleaned up is processes themselves;
-if we run a program which itself spawns subprocesses,
+without ensuring that they will be cleaned up later,
+leaving behind unused garbage on the system when it exits.
+
+One resource which is not automatically cleaned up is processes themselves.
+If we run a program which itself spawns subprocesses,
 those subprocesses may unintentionally leak,
-and be left running on the system even after the program itself has stopped\cite{caternfork}.
+and be left running on the system even after the original program has stopped\cite{caternfork}.
 
@@
 
-\texttt{init} in a pid namespace has some special powers and responsibilities.
+The first process in a pid namespace (usually referred to as \texttt{init})
+has some special powers and responsibilities\cite{pid_namespaces}.
 We can handle these responsibilities ourselves directly from our program,
 or we can continue on by \texttt{exec}ing an init daemon in \texttt{init} to handle it for us
-\cite{pid_namespaces}\cite{tini}\cite{dumb_init}.
+\cite{tini}\cite{dumb_init}.
 
@@
 \subsection{Shared file descriptor tables}\label{shared_fd_table}
-We can create a child process which shares its file descriptor table with the parent process.
+For our final example,
+we'll create a child process which shares its file descriptor table with the parent process.
 This can be useful for a variety of purposes,
-such as accessing resources in other namespaces, or in multiple namespaces at once.
-We'll give one example using the Filesystem in Userspace (FUSE) Linux subsystem.
-
+such as accessing resources in other namespaces, or accessing multiple namespaces at once.
+
+Our example will use the Filesystem in Userspace (FUSE) Linux subsystem.
 FUSE\cite{fuse} allows a filesystem to be implemented with a userspace program.
@@
 file descriptor passing over a Unix domain socket
-is our only option.
+is our only option to implement this kind of sharing.
 \section{Implementation}\label{implementation}
@@
 When making a syscall in the local process, the syscall is performed normally,
-directly on the local thread.
+directly on the local thread, as one might expect.
 
@@
 typically a pair of pipes.
+The syscall server runs on the main and only thread of the target process,
+and is the only userspace code running in an rsyscall-controlled child process;
+when not executing a syscall or writing out the response,
+an rsyscall-controlled child process spends all its time blocked in \texttt{read},
+waiting for new syscall requests.
 
@@
 our implementation fully supports asynchronous usage with a user-provided event loop,
-including the Python \texttt{async/await} coroutine support\cite{python_coroutines}.
-
-We have chosen to implement in userspace first,
+including Python \texttt{async/await} coroutine support\cite{python_coroutines}.
+
+We have chosen to implement our cross-process syscall primitive in userspace,
 rather than immediately implementing this in the kernel,
@@
 In the past, attempts to upstream kernel support for novel features without extensive userspace usage
-has had a poor reception in the Linux kernel community\cite{lwn_checkpoint}\cite{first_class_address_space}.
+have had a poor reception in the Linux kernel community\cite{lwn_checkpoint}\cite{first_class_address_space}.
 We believe proving viability in userspace first will be better in the long term.
 
-\texttt{ptrace} provides a way to perform arbitrary actions on a target process,
+\texttt{ptrace} already provides an in-kernel way to perform arbitrary actions on a target process,
 including system calls,
@@
 we choose to copy memory over a pair of pipes,
-again using file descriptors for the sake of easy event-loop integration.
+again using file descriptors for the sake of easy event loop integration.
 \subsection{\texttt{clone}}\label{clone}
@@
 but they are not suitable for flexible direct-style process creation.
-\texttt{vfork} suspends execution of the parent process
+\texttt{vfork}\cite{vfork} suspends execution of the parent process
 while waiting for the child process to exit or call \texttt{execve},
-which is immediately unsuitable\cite{vfork}.
+which is immediately unsuitable.
 \texttt{fork} lacks many features which are restricted to \texttt{clone}.
@@
 After this,
-the parent process can be begin to call system calls in the child process.
+the parent process can begin to call system calls in the child process.
 Most system calls work as normal;
@@
 With this technique, the parent process creates a pipe before forking,
-ensures both ends are marked \verb|O_CLOEXEC|,
+ensures both ends are marked \texttt{CLOEXEC},
 performs the \texttt{fork},
@@
 Fortunately, there is an alternative solution, which does work with \verb|CLONE_FILES|.
-\texttt{clone} has an argument, \texttt{ctid}, which specifies a memory address.
-If the \verb|CLONE_CHILD_CLEARTID| \texttt{clone} flag is set,
-then when the child exits or execs,
+\texttt{clone} has an argument, \texttt{ctid}, which specifies a memory address\cite{clone}.
+If the \verb|CLONE_CHILD_CLEARTID| flag is set,
+then when the child exits or successfully calls \texttt{exec},
 the kernel will set \texttt{ctid} to zero and then,
-crucially, perform a futex wakeup on it\cite{clone}.
-
-Futexes are usually used for the implementation of userspace shared-memory synchronization constructs,
+crucially, perform a futex wakeup on it.
+
+Futexes are usually used for the implementation of userspace shared-memory synchronization constructs\cite{futex},
 but the relevant detail for us here is that we can wait on an address
-until a futex wakeup is performed on that address\cite{futex}.
+until a futex wakeup is performed on that address.
 This means we can wait on \texttt{ctid} until the futex wakeup is performed,
-and in this way get notified of the child process calling \texttt{execve} or exiting.
+and in this way receive a notification when the child process has exited or successfully called \texttt{exec}.
+
 A process can only wait on one futex at a time;
@@
 whichever comes first.
-Either will normally indicate that the syscall fails,
-but if the futex process exits,
+For other syscalls, either result will indicate that the syscall fails,
+but for \texttt{execve}, if the futex process exits,
 without the child process itself exiting,
-then we can catch that in \texttt{execve}
-and know that the child has successfully completed an \texttt{execve} call.
+then we know that the child has successfully completed the \texttt{execve} call.
 
@@
 this ambiguity is unfortunate, but it is also present in the pipe-based approach.
-
-Some other Unix-like systems natively provide the ability to wait for a child's \texttt{execve};
+This ambiguous situation will only happen
+if the child receives a fatal signal while calling \texttt{execve},
+which we believe will be rare.
+
+Some other Unix-like systems natively provide the ability to wait for a child's \texttt{execve}\cite{freebsd_kqueue};
 our implementation would be simplified, and the ambiguity eliminated,
-if we had this ability on Linux\cite{freebsd_kqueue}.
+if we had this ability on Linux.
 One approach would be to add a new \texttt{clone} flag to
@@
 \subsection{Handling file descriptors}\label{fd_handles}
-In Unix, there are many types of resources maintained inside the kernel,
-such as file descriptors or memory mappings,
-which are referred to by identifiers which are valid only within a certain context\cite{tlpi}.
+In Linux, there are many types of resources maintained inside the kernel
+which are referred to by identifiers which are valid only within a certain context.
 For example, file descriptors are referred to by file descriptor numbers
@@
 which are valid only in a certain address space;
-the same pattern is followed for a number of other resources.
+the same pattern is followed for a number of other resources\cite{tlpi}.
 Each process has one file descriptor table, one address space, and one of each of the other contexts.
@@
 \subsubsection{File descriptor inheritance and table sharing}\label{inheritance}
-The picture is complicated by the need to support some FD behaviors which implicitly cross between processes.
-FDs are copied between processes at process creation time through FD inheritance,
-and two processes can share the same FDs through a shared FD table.
+The picture is complicated by the need to support some FD behaviors which implicitly cross between processes;
+specifically, FD inheritance and FD table sharing.
 
@@
 at the same FD numbers.
+We show examples of this in sections \ref{fd_abstraction} and \ref{pipe}.
 
@@
 through the same FD numbers.
-
-These behaviors are unusual in that they operate on all open FDs at once,
-rather than individual FDs explicitly denoted by system call arguments.
+We show an example of this in section \ref{shared_fd_table}.
+
+These behaviors are unusual in that they operate on all open FDs at once.
 There are other ways to pass FDs between processes,
 such as \texttt{SCM\_RIGHTS}\cite{scm_rights},
-but those operate on individual FDs through explicit arguments to system calls.
-One or more FD handles are explicitly passed to a system call in one process,
-and one or more FD handles are explicitly returned from another system call in another process.
-The behaviors of FD inheritance and FD table sharing happen for all FDs, implicitly,
-and so they must be dealt with differently to be fully supported.
-
-\texttt{rsyscall} supports both behaviors in the same way,
-by allowing us to create new FD handles on-demand
-referring to already-open FDs created by FD inheritance or FD table sharing;
-by creating new FD handles in new processes, we can call syscalls in those processes on those FDs.
+but those interfaces operate on individual FDs through explicit system call arguments and return values.
+In those interfaces,
+one or more FDs are explicitly passed as arguments to a system call in one process,
+and one or more FDs are returned from another system call in another process.
+This differs from FD inheritance and FD table sharing, which happen for all FDs, implicitly;
+these behavior must be dealt with differently to be fully supported.
+
+\texttt{rsyscall} supports both behaviors in the same way:
+by allowing us to create new FD handles
+referring to the already-open FDs created by FD inheritance or FD table sharing.
+By creating new FD handles associated with new processes, we can call syscalls in those processes on those FDs.
+
 \texttt{rsyscall} provides two functions
@@
 and \texttt{use\_fd} (used in listing \ref{lst:fuse}), which support FD inheritance and FD table sharing, respectively.
-Both functions take as arguments a process capability $A$ and an FD handle $H$ associated with some other process $B$,
+Both functions take as arguments a process capability $A$ and an FD handle $F$ associated with some other process $B$,
 and return a new FD handle associated with process $A$.
+
 For \texttt{inherit\_fd},
-this new handle refers to an \textit{inherited copy} of the FD which $H$ refers to;
+the returned handle has the same FD number as $F$,
+and refers to the copy of $F$'s FD that was inherited at process creation time.
 \texttt{inherit\_fd} will fail if $B$ is not $A$'s parent process,
-or if $H$ was not open at the time $A$ was created.
+or if $F$ was not open at the time $A$ was created.
+
 For \texttt{use\_fd},
-this new handle refers to the \textit{same} FD that $H$ refers to, through the shared FD table;
+the returned handle again has the same FD number as $F$,
+and refers to the \textit{same} FD as $F$, through the shared FD table.
 \texttt{use\_fd} will fail if $A$ and $B$ don't share the same FD table.
+
+\texttt{rsyscall}'s support for these behaviors is intended to expose the native functionality of the Linux kernel,
+while still being relatively straightforward to use.
+In this way,
+we hope to ensure that direct-style process creation with \texttt{rsyscall}
+supports everything that the underlying Linux system call interface can support.
 \section{Evaluation}\label{evaluation}
@@
 which abstracts away this error reporting code.
-We have built several such wrappers,
-but have ultimately discarded them in favor of direct-style process creation\cite{sfork}.
+We have built several such wrappers\cite{sfork},
+but have ultimately discarded them in favor of direct-style process creation.
 
@@
 and the C implementations vastly outperform the Python implementations.
-\texttt{posix\_spawn} takes an average of 440 microseconds to start a process,
+glibc's \texttt{posix\_spawn} takes an average of 440 microseconds to start a process,
 and \texttt{vfork} takes 400 microseconds,
@@
 is the amount of overhead incurred
-by performing our child process setup system calls through \texttt{rsyscall}.
+for each child process setup system call performed through \texttt{rsyscall}.
 
@@
 
-We've benchmarked process creation using the same benchmark as in section \ref{subprocess_bench}
-from a variety of parent processes: the local process,
+We've benchmarked process creation using the same simple process creation workload as section \ref{subprocess_bench},
+performed through \texttt{rsyscall} from a variety of parent processes:
+the local process,
 a child (created with \texttt{clone()}),
@@
 file descriptor logic must ultimately be centralized,
-whereas file descriptors in a new process can be built up over time using direct-style,
+while with direct-style,
+file descriptors in a new process can be built up over time,
 simplifying the implementation.
@@
 multiple ptracer processes cannot use \texttt{ptrace} on a single target process simultaneously,
-which means only one of these systems can be used on a process at a time.
+which means only one of these systems can be used at a time.
 
@@
 Other aspects of our implementation would also be improved by new features in the Linux kernel;
-one examples is discussed in section \ref{execve}.
+one example is discussed in section \ref{execve}.
 \subsection{Portability to other Unix systems}
@@
 We've found that direct-style process creation is more expressive than fork-style and spawn-style,
-and even with an unoptimized implementation, has comparable performance.
+and even with an unoptimized implementation, has acceptable performance.
 We've used direct-style process creation successfully in production at \twosigma{}.

### 2020-05-28 research/paper/paper.tex
paper......

@@
 and arranges to call that syscall in the specified process,
-returning the response as a single register-sized integer.
+returning the result as a single register-sized integer.
 
@@
 to a userspace ``syscall server'' running in the target process,
-which performs the syscall and sends the response back.
+which performs the syscall and sends the result back.
 Communication to the syscall server happens over file descriptors,
@@
 and is the only userspace code running in an rsyscall-controlled child process;
-when not executing a syscall or writing out the response,
+when not executing a syscall or writing out the result,
 an rsyscall-controlled child process spends all its time blocked in \texttt{read},
@@
 
-\texttt{ptrace} already provides an in-kernel way to perform arbitrary actions on a target process,
+\texttt{ptrace}\cite{ptrace} already provides an in-kernel way to perform arbitrary actions on a target process,
 including system calls,
-but is not suitable for us\cite{ptrace}.
+but is not suitable for us.
 As we discuss in section \ref{related}, a number of systems, such as \texttt{gdb} and \texttt{strace},

### 2021-01-16 python/rsyscall/handle/__init__.py
handle: don't define constructor for FileDescriptor
__slots__ doesn't allow us to default valid to True, so previously we
set it in the constructor. But we only construct FileDescriptors in
one place anyway, so we can just pass True there instead.

We necessarily have to get rid of the constructor definitions in the
derived classes as well. Those constructors also narrow the type of
the `task' field through the type annotation on their argument. We can
achieve the same thing by just adding the field type hint to the
class body; looks like that doesn't duplicate the field, I guess
dataclasses d

@@ -125,9 +125,7 @@ class FileDescriptor(
 
     """
     __slots__ = ()
-    def __init__(self, task: Task, near: rsyscall.near.FileDescriptor) -> None:
-        super().__init__(task, near)
-        self.task: Task = task
+    task: Task
 
     def as_proc_path(self) -> Path:
         """Return the /proc/{pid}/fd/{num} path pointing to this FD.
@@ -215,7 +213,7 @@ class Task(
 
     def _file_descriptor_constructor(self, fd: rsyscall.near.FileDescriptor) -> FileDescriptor:
         # for extensibility
-        return FileDescriptor(self, fd)
+        return FileDescriptor(self, fd, True)
 
     def _make_fresh_address_space(self) -> None:
         self.address_space = rsyscall.far.AddressSpace(self.process.near.id)

### 2021-01-16 python/rsyscall/handle/fd.py

@@ -39,14 +39,6 @@ class BaseFileDescriptor:
     near: rsyscall.near.FileDescriptor
     valid: bool
 
-    def __init__(self,
-                 task: FileDescriptorTask,
-                 near: rsyscall.near.FileDescriptor,
-    ) -> None:
-        self.task = task
-        self.near = near
-        self.valid = True
-
     def _validate(self) -> None:
         if not self.valid:
             raise Exception("handle is no longer valid")

### 2021-01-16 python/rsyscall/sys/mman.py

@@ -64,9 +64,7 @@ class MemoryMappingTask(FileDescriptorTask):
         return MemoryMapping(self, ret, rsyscall.far.File())
 
 class MappableFileDescriptor(BaseFileDescriptor):
-    def __init__(self, task: MemoryMappingTask, near: near.FileDescriptor) -> None:
-        super().__init__(task, near)
-        self.task: MemoryMappingTask = task
+    task: MemoryMappingTask
 
     async def mmap(self, length: int, prot: PROT, flags: MAP,
                    offset: int=0,

### 2021-01-16 python/rsyscall/sys/mman.py
sys.mman: add __slots__ for MemoryMapping

@@ -34,6 +34,7 @@ import typing as t
 
 @dataclass(frozen=True)
 class MemoryMapping:
+    __slots__ = ('task', 'near', 'file')
     task: MemoryMappingTask
     near: rsyscall.near.MemoryMapping
     file: rsyscall.far.File

### 2021-01-16 python/rsyscall/sys/wait.py
sys.wait: minorly refactor ChildState.died
This, to my annoyance, is a significant-enough optimization to have
shown up in my profiling a while ago.

But I can actually persuade myself that this reads better than the old
style, so this is fine.

@@ -81,7 +81,7 @@ class ChildState:
         }[self.code])
 
     def died(self) -> bool:
-        return self.state(W.EXITED)
+        return self.code in {CLD.EXITED, CLD.KILLED, CLD.DUMPED}
 
     def clean(self) -> bool:
         return self.code == CLD.EXITED and self.exit_status == 0

### 2021-01-16 python/benchmark.py
paper stuff

@@ -1,71 +1,0 @@
-import trio
-from rsyscall import Path, Command
-import rsyscall.tasks.local as local
-import rsyscall.nix as nix
-from rsyscall.tasks.stdin_bootstrap import stdin_bootstrap, stdin_bootstrap_path_from_store
-from rsyscall.unistd import SEEK
-import csv
-import argparse
-from dataclasses import dataclass
-
-@dataclass
-class Run:
-    pid: int
-    # all in microseconds
-    pre_fork: int
-    in_child: int
-    in_parent: int
-    after_exit: int
-
-usec_in_sec = 1000*1000
-
-def analyze_data(data: str) -> None:
-    runs = []
-    for line in csv.DictReader(data.split()):
-        runs.append(Run(int(line['pid']),
-            pre_fork=int(line['pre_fork_sec'])*usec_in_sec + int(line['pre_fork_usec']),
-            in_child=int(line['in_child_sec'])*usec_in_sec + int(line['in_child_usec']),
-            in_parent=int(line['in_parent_sec'])*usec_in_sec + int(line['in_parent_usec']),
-            after_exit=int(line['after_exit_sec'])*usec_in_sec + int(line['after_exit_usec']),
-        ))
-
-    print(sum(run.after_exit - run.pre_fork for run in runs)/len(runs))
-
-async def main() -> None:
-    parser = argparse.ArgumentParser(description='Do benchmarking')
-    parser.add_argument('executable', type=Path)
-    parser.add_argument('num_runs', type=int)
-    
-    args = parser.parse_args()
-
-    cmd = Command(args.executable, [args.executable, "100"], {})
-
-    # ugh I need nicing
-    # let's do that privileged exec thing?
-    # then we'll need to support nicing things, too. but fine, this is good.
-    # we can hopefully get this done tonight.
-    # ugh uUGHHG ugh ugh.
-    # ok so sudo doesn't let you inherit higher file descriptors, because it sucks and is bad. hm.
-    # but, we could just do this all on the sudo side. nice, good hack!
-    stdin_bootstrap_path = await stdin_bootstrap_path_from_store(nix.local_store)
-    proc, privthread = await stdin_bootstrap(
-        local.thread, (await local.thread.environ.which("sudo")).args(stdin_bootstrap_path))
-    fd = await privthread.task.memfd_create(await privthread.ptr(Path("data")))
-    print("hi 0")
[…]

### 2021-01-16 research/paper/benchmark.py

@@ -1,0 +1,71 @@
+import trio
+from rsyscall import Path, Command
+import rsyscall.tasks.local as local
+import rsyscall.nix as nix
+from rsyscall.tasks.stdin_bootstrap import stdin_bootstrap, stdin_bootstrap_path_from_store
+from rsyscall.unistd import SEEK
+import csv
+import argparse
+from dataclasses import dataclass
+
+@dataclass
+class Run:
+    pid: int
+    # all in microseconds
+    pre_fork: int
+    in_child: int
+    in_parent: int
+    after_exit: int
+
+usec_in_sec = 1000*1000
+
+def analyze_data(data: str) -> None:
+    runs = []
+    for line in csv.DictReader(data.split()):
+        runs.append(Run(int(line['pid']),
+            pre_fork=int(line['pre_fork_sec'])*usec_in_sec + int(line['pre_fork_usec']),
+            in_child=int(line['in_child_sec'])*usec_in_sec + int(line['in_child_usec']),
+            in_parent=int(line['in_parent_sec'])*usec_in_sec + int(line['in_parent_usec']),
+            after_exit=int(line['after_exit_sec'])*usec_in_sec + int(line['after_exit_usec']),
+        ))
+
+    print(sum(run.after_exit - run.pre_fork for run in runs)/len(runs))
+
+async def main() -> None:
+    parser = argparse.ArgumentParser(description='Do benchmarking')
+    parser.add_argument('executable', type=Path)
+    parser.add_argument('num_runs', type=int)
+    
+    args = parser.parse_args()
+
+    cmd = Command(args.executable, [args.executable, "100"], {})
+
+    # ugh I need nicing
+    # let's do that privileged exec thing?
+    # then we'll need to support nicing things, too. but fine, this is good.
+    # we can hopefully get this done tonight.
+    # ugh uUGHHG ugh ugh.
+    # ok so sudo doesn't let you inherit higher file descriptors, because it sucks and is bad. hm.
+    # but, we could just do this all on the sudo side. nice, good hack!
+    stdin_bootstrap_path = await stdin_bootstrap_path_from_store(nix.local_store)
+    proc, privthread = await stdin_bootstrap(
+        local.thread, (await local.thread.environ.which("sudo")).args(stdin_bootstrap_path))
+    fd = await privthread.task.memfd_create(await privthread.ptr(Path("data")))
+    print("hi 0")
[…]

### 2021-01-16 research/paper/getpid_bench.py

@@ -1,0 +1,111 @@
+from dataclasses import dataclass
+from rsyscall import Command, Path, local_thread
+from rsyscall.sys.mman import PROT, MAP
+from rsyscall.sys.resource import PRIO
+from rsyscall.sys.wait import W
+from rsyscall.tasks.stdin_bootstrap import stdin_bootstrap, stdin_bootstrap_path_from_store
+from rsyscall.unistd import SEEK
+import statistics
+import argparse
+import os
+import rsyscall.nix as nix
+import subprocess
+import sys
+import time
+import trio
+import typing as t
+import json
+
+async def run_benchmark(mode: str, mmaps: int) -> t.Tuple[int, int]:
+    for _ in range(mmaps):
+        # set POPULATE so it actually gets faulted in; if it isn't
+        # already in the page tables, there's no slowdown
+        await local_thread.task.mmap(4096, PROT.READ|PROT.WRITE, MAP.PRIVATE|MAP.ANONYMOUS|MAP.POPULATE)
+    cmd = await local_thread.environ.which('cat')
+    await local_thread.environ.as_arglist(local_thread.ram)
+    cwd = await local_thread.ptr(Path("/dev"))
+    child = await local_thread.clone()
+    nest1_child = await child.clone()
+    nest2_child = await nest1_child.clone()
+    async def subp_run() -> None:
+        await local_thread.task.getpid()
+    async def rsys_run() -> None:
+        await child.task.getpid()
+    async def nest_run() -> None:
+        await child.task.getpid()
+    if mode == 'subprocess':
+        run = subp_run
+    elif mode == 'rsyscall':
+        run = rsys_run
+    elif mode == 'nest':
+        run = nest_run
+    prep_count = 20
+    count = 100
+    before_prep = time.time()
+    for _ in range(prep_count):
+        await run()
+    before = time.time()
+    for _ in range(count):
+        await run()
+    after = time.time()
+    prep_time = (before - before_prep)/prep_count
+    real_time = (after - before)/count
+    return real_time
+
+async def main() -> None:
+    parser = argparse.ArgumentParser(description='Do benchmarking of rsyscall vs subprocess.run')
+    parser.add_argument('--run-benchmark', choices=['subprocess', 'rsyscall', 'nest'])
+    parser.add_argument('--mmaps', type=int, default=0)
+    parser.add_argument('--no-use-setpriority', help="don't setpriority before benchmarking; doing that requires privileges,"
[…]

### 2021-01-16 research/paper/getpid_bench_plot.py

@@ -1,0 +1,22 @@
+import matplotlib.pyplot as plt
+import json
+import sys
+
+def main():
+    data = json.load(sys.stdin)
+    fig, ax = plt.subplots()
+    plt.xscale('log')
+    plt.yscale('log')
+    subprocess = data['subprocess']
+    rsyscall = data['rsyscall']
+    ax.plot([10**x for x, y in subprocess], [y for x, y in subprocess], 'o-', label="Python subprocess")
+    ax.plot([10**x for x, y in rsyscall], [y for x, y in rsyscall], '^-', label="rsyscall")
+    ax.legend()
+    
+    ax.set(xlabel='Pages mapped in memory', ylabel='time (us)')
+    ax.grid()
+    
+    fig.savefig(sys.stdout.buffer)
+
+if __name__ == "__main__":
+    main()

### 2021-01-16 python/rsyscall/handle/__init__.py
handle.Task: close SyscallInterface after calls to exec
We do this for exit, so we should do it for exec. Now that we don't
have rsyscall_exec dragging us down, there's no reason not to...

@@ -259,6 +259,7 @@ class Task(
             self._make_fresh_address_space()
             if isinstance(self.process, ChildProcess):
                 self.process.did_exec(command)
+        await self.sysif.close_interface()
 
     async def execve(self, filename: WrittenPointer[t.Union[str, os.PathLike]],
                      argv: WrittenPointer[ArgList],
@@ -281,6 +282,7 @@ class Task(
         self._make_fresh_address_space()
         if isinstance(self.process, ChildProcess):
             self.process.did_exec(command)
+        await self.sysif.close_interface()
 
     async def exit(self, status: int) -> None:
         self.manipulating_fd_table = True

### 2021-01-16 research/paper/rsys_vs_subp_bench.py
add rsys_vs_subp_bench

@@ -1,0 +1,41 @@
+from rsyscall import local_thread
+from rsyscall.sys.wait import W
+import subprocess
+import time
+import typing as t
+
+async def main():
+    count = 100
+    cmd = (await local_thread.environ.which("echo")).args("-n", "hello")
+    prep = 5
+    counts = ([], [], [], [])
+    async def rsys_run() -> None:
+        counts[0].append(time.time())
+        child = await local_thread.clone()
+        counts[1].append(time.time())
+        proc = await child.exec(cmd)
+        counts[2].append(time.time())
+        await proc.waitpid(W.EXITED)
+        counts[3].append(time.time())
+    async def subp_run() -> None:
+        subprocess.run(["echo", "-n", "hello"])
+    run = rsys_run
+    # run = subp_run
+    before_prep = time.time()
+    for _ in range(prep):
+        await run()
+    before = time.time()
+    for _ in range(count):
+        await run()
+    after = time.time()
+    print("prep time", (before - before_prep)/prep)
+    print("real time", (after - before)/count)
+
+    def process(starts: t.List[float], ends: t.List[float]) -> float:
+        return sum(end - start for start, end in zip(starts, ends))/count
+    print("clone", process(counts[0], counts[1]))
+    print("exec", process(counts[1], counts[2]))
+    print("waitpid", process(counts[2], counts[3]))
+
+if __name__ == "__main__":
+    trio.run(main)

### 2021-01-16 research/paper/getpid_bench_plot.py
paper: add getpid_bench

@@ -1,0 +1,22 @@
+import matplotlib.pyplot as plt
+import json
+import sys
+
+def main():
+    data = json.load(sys.stdin)
+    fig, ax = plt.subplots()
+    plt.xscale('log')
+    plt.yscale('log')
+    subprocess = data['subprocess']
+    rsyscall = data['rsyscall']
+    ax.plot([10**x for x, y in subprocess], [y for x, y in subprocess], 'o-', label="Python subprocess")
+    ax.plot([10**x for x, y in rsyscall], [y for x, y in rsyscall], '^-', label="rsyscall")
+    ax.legend()
+
+    ax.set(xlabel='Pages mapped in memory', ylabel='time (us)')
+    ax.grid()
+
+    fig.savefig(sys.stdout.buffer)
+
+if __name__ == "__main__":
+    main()

### 2021-01-16 python/rsysapps/miredo.py
rsysapps.miredo: rearrange for more clarity
This change is from many months ago, when I was working on the paper,
and I don't know whether it actually works, and I'm too lazy to test
it...

@@ -19,7 +19,7 @@ from rsyscall.memory.ram import RAMThread
 
 import rsyscall.tasks.local as local
 from rsyscall.sys.capability import CAP, CapHeader, CapData
-from rsyscall.sys.socket import AF, SOCK, SOL
+from rsyscall.sys.socket import AF, SOCK, SOL, Socketpair
 from rsyscall.sys.prctl import PR, PR_CAP_AMBIENT
 from rsyscall.fcntl import O
 from rsyscall.sched import CLONE
@@ -27,8 +27,7 @@ from rsyscall.netinet.in_ import SockaddrIn
 from rsyscall.netinet.ip import IP, IPPROTO
 from rsyscall.linux.netlink import SockaddrNl, NETLINK
 from rsyscall.linux.rtnetlink import RTMGRP
-from rsyscall.handle import Socketpair
-import rsyscall.net.if_ as netif
+from rsyscall.net.if_ import Ifreq, IFF_TUN, TUNSETIFF, SIOCGIFINDEX
 
 import rsyscall.nix as nix
 
@@ -58,37 +57,11 @@ class Miredo:
     # So, I'll just use a thread, which I do understand.
     # Hopefully we can get a more lightweight setns-based approach later?
     ns_thread: ChildThread
+    privproc_child: AsyncChildProcess
+    client_child: AsyncChildProcess
 
-async def exec_miredo_privproc(
-        miredo_exec: MiredoExecutables,
-        thread: ChildThread,
-        privproc_side: FileDescriptor, tun_index: int) -> AsyncChildProcess:
-    privproc_side = privproc_side.move(thread.task)
-    await thread.unshare_files(going_to_exec=True)
-    await thread.stdin.copy_from(privproc_side)
-    await thread.stdout.replace_with(privproc_side)
-    child = await thread.exec(miredo_exec.privproc.args(str(tun_index)))
-    return child
-
-async def exec_miredo_run_client(
-        miredo_exec: MiredoExecutables,
-        thread: ChildThread,
-        inet_sock: FileDescriptor,
-        tun_fd: FileDescriptor,
-        reqsock: FileDescriptor,
-        icmp6_fd: FileDescriptor,
-        client_side: FileDescriptor,
-        server_name: str) -> AsyncChildProcess:
-    fd_args = [fd.move(thread.task)
-               for fd in [inet_sock, tun_fd, reqsock, icmp6_fd, client_side]]
-    await thread.unshare_files()
-    child = await thread.exec(miredo_exec.run_client.args(
[…]

### 2021-04-07 research/paper/paper.tex
paper

--- 2020-05-27 research/paper/paper.tex
@@
 
-\usepackage{microtype}
-\usepackage{calc}
 \usepackage{tikz}
-\usepackage{dblfloatfix}
 \usepackage{amsmath}
-\usepackage{amssymb}
-\usepackage{pifont}
-\usepackage{comment}
 
 \usepackage{graphicx}
+\usepackage{hyperref}
 \usepackage{float}
@@
 
-\newcommand{\tbad}{\makebox[\widthof{\ding{51}}]{\ding{55}}:}
-\newcommand{\tgood}{\ding{51}:}
-
-% \titleformat*{\subsection}{\normalsize\bfseries}
-% \titleformat*{\subsubsection}{\normalsize\bfseries}
-% \titlespacing{\subsection}{0pt}{1ex}{0.5ex}
-% \titlespacing{\subsubsection}{0pt}{0.5ex}{0ex}
-\lstset{
-language=Python,
-basicstyle=\small\ttfamily,
-keywordstyle=,
-stringstyle=\color{red},
-frame=tb,
-columns=fullflexible,
-showstringspaces=false,
-belowskip=0pt,
-}
+\titleformat*{\subsection}{\normalsize\bfseries}
+\titleformat*{\subsubsection}{\normalsize\bfseries}
+\titlespacing{\subsection}{0pt}{1ex}{0.5ex}
+\titlespacing{\subsubsection}{0pt}{0.5ex}{0ex}
 
 \begin{document}
-\newcommand{\twosigma}{{[tech company]}}
-\newcommand{\githuburl}{[Github url removed]}
+\newcommand{\twosigma}{[tech company]} 
 \date{}
@@
 Traditional process creation interfaces,
-such as \texttt{fork} and \texttt{spawn},
+such as fork and spawn,
 are complex to use, limited in power, and difficult to abstract over.
 We develop a new process creation interface for Linux
-which allows a program to create a child process in a non-running state
+which allows a program to create a child process in an non-running state
 and initialize the new process by operating on it from the outside.
@@
 has better error handling,
-is comparably efficient,
+is more efficient,
 and is more amenable to abstraction.
@@
 \section{Introduction}\label{introduction}
-\begin{table*}
-\begin{tabular}{l|l|l|l}
-% BEGIN RECEIVE ORGTBL compare
- & fork-style & spawn-style & direct-style\\
-\hline
-Parent process requirements & \tbad Single thread, small memory & \tgood None & \tgood None\\
-Programming model & \tbad Complex (returns twice) & \tgood Simple (single call) & \tgood Simple (imperative)\\
-Maximally powerful & \tgood Yes, can call any syscall & \tbad No, limited interface & \tgood Yes, can call any syscall\\
-Reporting of results & \tbad Requires IPC & \tbad Not fine-grained & \tgood From individual child syscalls\\
-Non-inherited attributes & Mutated by code in child & Set by arguments & Mutated by code in parent\\
-% END RECEIVE ORGTBL compare
-\end{tabular}
-\caption{Features of fork-style vs spawn-style vs direct-style}
-\label{tab:styles}
-\end{table*}
-\begin{comment}
-#+ORGTBL: SEND compare orgtbl-to-latex :splice t
-|                             | fork-style                        | spawn-style                 | direct-style                          |
-|-----------------------------+-----------------------------------+-----------------------------+---------------------------------------|
-| Parent process requirements | \tbad Single thread, small memory | \tgood None                 | \tgood None                           |
-| Programming model           | \tbad Complex (returns twice)     | \tgood Simple (single call) | \tgood Simple (imperative)            |
-| Maximally powerful          | \tgood Yes, can call any syscall  | \tbad No, limited interface | \tgood Yes, can call any syscall      |
-| Reporting of results        | \tbad Requires IPC                | \tbad Not fine-grained      | \tgood From individual child syscalls |
-| Non-inherited attributes    | Mutated by code in child          | Set by arguments            | Mutated by code in parent             |
-% $
-\end{comment}
-The most well-known process creation interface on Linux, and Unix in general, is \texttt{fork}.
-When a parent process creates a child process through \texttt{fork} or other fork-style interfaces,
-both processes execute concurrently,
-and the child process typically inherits all but a few attributes from the parent process\cite{manfork}.
-The child process can call arbitrary system calls
-to mutate the new process until it reaches the desired state,
-typically ultimately calling \texttt{exec}.
-
-A user of \texttt{fork} must think carefully about the characteristics of the process
-from which \texttt{fork} is being called.
-\texttt{fork} can have poor performance when called from processes with many memory mappings\cite{forkroad}.
-This can be mitigated to some degree by using \texttt{vfork} or \texttt{CLONE\_VM},
-but those inflict further complexity on the caller\cite{vfork_dangerous}.
-Processes which are multi-threaded can encounter deadlocks and other issues if they call \texttt{fork}
-\cite{forkroad}\cite{tlpi}\cite{posix_spawn_error_pipe}.
-
-Some thread libraries provide partial mitigations for multi-threading issues,
-but it's up to user code to make use of those mitigations\cite{pthread_atfork}.
-
-Since the child process runs concurrently with the parent process,
-communicating events in the child back to the parent requires some form of IPC,
-adding complexity\cite{posix_spawn_error_pipe}.
-Even the basic interface of \texttt{fork} --- one function call which returns twice --- is unusual;
-such functions are generally considered to be complex, advanced topics in other settings
-\cite{callcc_fork}\cite{continuations}.
-
-Closely related to \texttt{fork}
-are process creation interfaces where the new process is launched running a function of the caller's choice,
-which can then call arbitrary system calls to mutate the new process
-\cite{clone}\cite{sfork}.
-Such interfaces have similar issues as \texttt{fork},
-and we class them together as ``fork-style'' interfaces.
-
-On the other hand, in a spawn-style interface,
+Existing process creation interfaces can be divided into three categories:
+Fork-style, spawn-style, and direct-style.
+
+Fork-style (such as \texttt{fork}) and spawn-style (such as \texttt{posix\_spawn}) are the most popular,
+but have a number of disadvantages.
+Correct, performant use of fork-style places significant constraints on the calling process,
+and also has a complex programming model.
+Spawn-style lifts many of these constraints, and is simpler to program against,
+but is fundamentally limited in the kinds of processes it can create.
+With both fork-style and spawn-style,
+it is difficult to
+branch based on results or errors encountered while initializing the new process:
+fork-style requires IPC if it wishes to communicate errors back to the rest of the program,
+and robust indication of errors in spawn-style requires substantial growth in the spawn interface.
+
+Direct-style is more rarely seen, but has a number of advantages.
+Direct-style involves creating a new child process in a non-running state,
+and then calling individual system calls from the parent to initialize the child.
+Direct-style does not constrain the calling process,
+easily branches at any step of the initialization,
+and can create any kind of process supported by the underlying system.
+
+Our contribution is an implementation of direct-style process creation for Linux,
+making use of our work on the \texttt{rsyscall} project.
+The \texttt{rsyscall} project develops language-specific ``process-independent'' system call libraries for Linux,
+which replace libc system call wrappers, and allow a program to perform cross-process system calls.
+rsyscall currently has best support for Python,
+but can be ported to other languages;
+the examples we show in this paper will be in Python,
+but generalize easily to other languages.
+
+Our direct-style process creation implementation uses \texttt{clone} to create a child process in an embryonic state
+which can then be manipulated from user code using syscalls with \texttt{rsyscall}.
+A direct-style \texttt{clone} has all the advantages of direct-style process creation
+over Linux's fork-style \texttt{fork} and spawn-style \texttt{posix\_spawn}:
+\begin{compactitem}
+\item
+  Processes can be created correctly and performantly from any parent process,
+  with a straightforward, familiar API.
+\item
+  Processes can be created with the full set of Linux features.
+\item
+  Syscall results are reported individually to the parent process,
+  rather than batched together as with spawn-style or manually reported via IPC as with fork-style.
+\end{compactitem}
+
+We've implemented a variety of useful applications using direct-style process creation on Linux,
+which we'll demonstrate in this paper.
+Direct-style \texttt{clone} has minimal performance cost in normal situations,
+and can outperform \texttt{fork} in some cases.
+We have deployed direct-style process creation in production internally at \twosigma.
+
+Our implementation is distributed as a feature of \texttt{rsyscall},
+and is immediately deployable on recent Linux kernels.
+\texttt{rsyscall} is open source and available from
+[Github url].
+% \url{https://github.com/catern/rsyscall}.
+
+This paper is organized in several sections.
+\begin{compactitem}
+\item In section \ref{background}, we provide background on fork-style, spawn-style, and direct-style process creation,
+and consider their advantages and disadvantages in more depth.
+\item In section \ref{overview}, we discuss the design of the API,
+including our usage of \texttt{rsyscall}.
+\item In section \ref{applications}, we show examples of usage of direct-style process creation in Python on Linux,
+including use of process features which are difficult to use on Linux with \texttt{fork} or \texttt{posix\_spawn}.
+\item In section \ref{implementation}, we explore our implementation in depth,
+and discuss several difficult process-creation related details of \texttt{rsyscall}'s implementation.
+\item In section \ref{evaluation}, we evaluate, somehow (TODO).
+\item In section \ref{related_work},
+we examine other work which also touches on the topics covered by this paper.
+\item In section \ref{future_work},
+we list several directions for future work related to direct-style process creation and to \texttt{rsyscall}.
+\item In section \ref{conclusions}, we briefly outline our conclusions about direct-style and the future of process creation.
+\end{compactitem}
+\section{Background}\label{background}
+As mentioned in the introduction,
+on most operating systems today,
+the available interfaces for process creation
+are either "fork-style" or "spawn-style".
+A few operating systems have a third style, which we refer to as "direct-style".
+\subsection{Fork-style}
+By far the most well-known fork-style interface is Unix's fork.\cite{forkhist}
+With a fork-style interface,
+a process, called the parent process, calls \texttt{fork},
+and a copy of that process is made,
+called the child process,
+differing only in a few attributes such as process id.
+The same program continues executing in both processes,
+with a different return value from fork to allow it to distinguish whether it is running in the parent or the child.
+Typically, the program will case on the return value
+and, if running in the child,
+call various system calls to modify the state of the new process.
+
+Fork has a number of flaws,
+and much ink has been spilled detailing them.
+We'll cover the most prominent briefly here,
+but see Baumann 2019 \cite{forkroad} for a more detailed evaluation.
+
+Fork, when called successfully, returns twice,
+creating two different processes which execute concurrently.
+This immediately makes fork difficult to integrate;
+for example, errors can't be easily communicated between the two processes.
+If there is an error in the initialization of the child process,
+the part of the program running in the child process
+needs to use some form of inter-process communication (IPC) to communicate it back to the parent process.
+The child process can run arbitrary code to make decisions about its initialization,
+but that code needs to use IPC if it wishes to communicate with the rest of the program,
+which is still in the parent process.
+
+Fork can have poor performance when called from processes with many memory mappings.\cite{forkroad}
+Fork copies many attributes about the parent process when creating the child process,
+including setting up copy-on-write memory mappings in the child process.
+This becomes slower as the parent process has more memory mappings,
+eventually taking a significant amount of time.
+This copying is required to robustly implement fork's model,
+where the same program continues executing without changes in both the parent process and child process;
+the copy of the program running in the child process might access any memory at any time.
+
+Multi-threaded programs generally cannot use fork safely.
+Typical Unix fork implementations duplicate only the thread calling fork from the parent process to the child process.
+In a multi-threaded program, this can cause various issues;
+for example, another thread might be holding a memory allocator lock at the time of the fork,
+which in the child process will never be unlocked,
+causing the child process to deadlock if it tries to allocate memory.
+Some thread libraries provide partial mitigations for this issue,
+but it's up to user code to make use of those mitigations.\cite{pthread_atfork}
+In combination with fork's poor performance in large-memory programs,
+a user of fork must think carefully
+about the characteristics of the process from which fork is being called.
+\subsection{Spawn-style}
+In a spawn-style interface,
 details about the new process are provided up-front as arguments to some function
 which creates the new process all at once.
-\texttt{posix\_spawn} is a typical example of this style
-\cite{posix_spawn}\cite{create_process}\cite{chainloading}.
-Other details, such as security context, are inherited from the parent process\cite{tlpi}\cite{create_process}.
-Spawn-style APIs
-can be called from any parent process without concerns about memory usage or multi-threading,
-when used correctly
-\cite{forkroad}\cite{posix_spawn_error_pipe}.
-
-Unlike \texttt{fork}, a spawn-style interface provides an API for process creation
-which is independent from normal system calls\cite{spawn_h}\cite{subprocess_run}.
-Spawn-style interfaces therefore often lack support for some features,
-and they also require learning a new interface distinct from the existing system call API
-\cite{tlpi}\cite{posix_spawn_lwn}.
-Also, spawn-style interfaces, by their nature as a single call,
-don't allow programs to branch on the results of individual modifications to the new process,
-and have worse error reporting for such modifications
-\cite{posix_spawn}\cite{posix_spawn_error_pipe}.
-
-Table \ref{tab:styles} summarizes the differences between fork-style and spawn-style.
-Each has its own advantages and disadvantages.
-
-A few academic operating systems
---- KeyKOS\cite{keykos} and seL4\cite{sel4}, for example ---
-use another style of process creation\cite{exokernel}\cite{fuchsia}\cite{singularity}.
-We use the new term ``direct-style'' to refer to this,
-since we know of no pre-existing generic term for this style of process creation.
+The most well known spawn-style interfaces are Unix's \texttt{posix\_spawn} \cite{posix_spawn}
+and Windows \texttt{CreateProcess} \cite{create_process};
+the Unix shell also serves as a spawn-style process creation interface,
+especially when used with Bernstein chaining\cite{chainloading}.
+
+Spawn-style interfaces typically still transparently copy some details from the parent process;
+for example, various security contexts,
+and any other process attribute which is not explicitly specified in the arguments to the spawn interface.
+The key difference between fork-style and spawn-style is not how much they copy,
+it is how the attributes which are not copied are specified:
+by mutation from code running in the new process in the case of fork-style,
+or by explicit argument passing in the case of spawn-style.
+
+Spawn-style interfaces lift the constraints on the calling process that fork-style interfaces impose.
+Spawn-style interfaces don't run user code in the new process during initialization.
+This means there is no need to copy memory,
+so performance is good when a spawn-style interface is used from a process with many memory mappings.
+Likewise, there are no inherent issues when a spawn-style interface is used from a multi-threaded program,
+though operating systems, such as Linux,
+which implement spawn-style interfaces on top of \texttt{fork} can still have threading bugs.
+
+However, the arguments that can be provided to a spawn-style process creation syscall
+do not cover all the possible attributes that one might want to set for the new process.
+Most systems have a large number of syscalls which can mutate the state of a process during its lifetime;
+for a spawn-style interface to work in all scenarios,
+all those possible mutations must be reproduced in the interface.
+For example, the \texttt{posix\_spawn} function provided by glibc does not support creating processes in new namespaces,
+as is required for container functionality on Linux.
+
+Spawn-style process creation
+essentially creates a new domain-specific language for specifying the attributes of the process.
+Fork-style allows existing system calls,
+with which programmers are already familiar,
+to be used to set up the new process.
+Spawn-style requires some form of declarative description of the new process,
+which must be learned in additional to the normal system call interface.
+
+Spawn-style process creation also does not allow for conditional logic during the setup.
+If the setup of the new process encounters an error at some point,
+the only option is to return from the entire spawn call with an error.
+Such errors returned from spawn-style calls
+are typically much less informative
+than the errors returned by the syscalls which directly mutate the process attributes.
+Other forms of conditional logic are also impossible in spawn-style;
+one modification to the process cannot depend on the result of some other modification.
+\subsection{Direct-style}
+A few academic operating systems, such as KeyKOS \cite{keykos}, seL4 \cite{sel4}
+and some others \cite{exokernel} \cite{fuschia} \cite{singularity},
+use another style of process creation, which we refer to as "direct-style".
 
@@
 the syscalls that can mutate a process
-take explicit arguments to indicate the process on which they should operate\cite{keykos}\cite{sel4}.
+take explicit arguments to indicate the process on which they should operate.
 In this way, the same syscalls that can mutate a process while it is running
@@
 
+We refer to this as "direct-style" process creation,
+because the parent creating the process operates on it directly and imperatively
+rather than dispatching a distinct unit of code to perform setup from inside the context of the new process,
+as in fork-style,
+or building up a declarative specification of what the new process should look like ahead of time,
+as in spawn-style.
+
+Since everything happens directly from the parent process,
+direct-style process creation is compatible with a variety of techniques
+which are difficult with fork-style or spawn-style.
+For example, a REPL could be used to interactively create the new process,
+or a program might initialize several processes at once and share data between them.
+
+Significantly,
+since all operations are performed through syscalls called directly from the main program,
+errors are indicated in the same way as any other syscall error.
+This is unlike fork-style, which needs IPC to indicate errors,
+and unlike spawn-style, which typically indicates errors at a very coarse-grained level.
+
+Like fork-style interfaces,
+direct-style interfaces can set up arbitrary attributes in the new process.
+Any attribute that can be changed by a syscall
+can be manipulated by a direct-style interface,
+just as with fork-style interfaces.
+This is unlike spawn-style interaces,
+which only can change attributes that are supported by the interface.
+
+Like spawn-style interfaces,
+direct-style interfaces have no constraints on the calling process.
+Also like spawn-style interface,
+no user code runs in the child process, so
+direct-style interfaces can be used from multi-threaded or large-memory processes without issues.
+
+Direct-style can be more complex to use;
+it is most typically used in capability-oriented operating systems,
+where a great deal of resources and information must be provided explicitly to initialize the new process.
+In a truly capability-oriented operating system,
+nothing is copied implicitly to the new process;
+everything must be explicitly specified.
+This can appear more complex
+when compared to process creation on Unix systems or Windows.
+
+However, we believe that a natural port of direct-style process creation to Linux
+provides an interface that is just as simple to use as fork-style or spawn-style on Linux,
+without their disadvantages.
+Such an interface, like fork-style and spawn-style,
+would implicitly copy some attributes to the new process,
+rather than being capability-oriented and requiring all attributes to be specified explicitly.
+This is the kind of direct-style interface that we contribute in this paper.
+\section{Overview}\label{overview}
+In most operating system APIs, including the ones available on Linux,
+when a program makes a syscall, it implicitly operates on the current process.
 To perform direct-style process creation on Linux,
-we need an API where we explicitly specify in which process we want to make a syscall.
-This differs from the normal mode of operation on Linux,
-where programs making syscalls implicitly operate on the current process.
-
-Our main contribution in this paper is such an API, in the form of the \texttt{rsyscall} library.
-\texttt{rsyscall} is a
-language-specific, object-capability-model, low-abstraction library for the Linux system call interface,
-bypassing libc\cite{capmyths}\cite{robust_composition}.
-The \texttt{rsyscall} project is currently focused on supporting Python,
-but the library can be ported to other languages.
-The examples we show in this paper will be in Python,
-but generalize easily.
-\texttt{rsyscall} is open source and available from
-\githuburl{}.
-
-Unlike typical C libraries such as glibc or musl,
-the \texttt{rsyscall} library is organized based on the object-capability model\cite{capmyths}\cite{robust_composition}.
-The capability to make syscalls in any specific process is reified as a language object.
+we need an API where we instead explicitly specify in which process we want to make a syscall.
+
+\texttt{rsyscall} provides this API.
+The \texttt{rsyscall} project develops
+language-specific, object-capability-model, low-abstraction libraries for the Linux syscall interface,
+bypassing libc.
+
+Unlike POSIX C libraries such as glibc or musl,
+an \texttt{rsyscall} library is organized based on the object-capability model.
+The capabilities to make syscalls in any given process are reified as language objects.
 If the capability to make syscalls in a specific process is not passed (in some way) to a function,
@@
 by manipulating capabilities for multiple processes.
-The relevant two types of capabilities for this paper are the initial capability for the ``local'' process
-and capabilities for child processes.
+The relevant two capabilities for this paper are the initial capability to the ``local'' process,
+and capabilities to child processes.
 The ``local'' process is the one which hosts the runtime for the running program,
 and in which a legacy libc would implicitly make syscalls.
-Every \texttt{rsyscall} program starts with the capability for the ``local'' process
-and uses it to bootstrap capabilities for other processes.
-
-\lstinputlisting[float,language=Python,label={lst:basic},caption={Create new process, change CWD, and exec}]
+Every \texttt{rsyscall} program starts with the ``local'' process
+and uses it to bootstrap the initial capabilities for other processes.
+
+To perform direct-style process creation,
+we use \texttt{rsyscall} to call \texttt{clone} on the ``local'' process to create a new child process.
+We call various syscalls in the child process to mutate it until it reaches the desired state,
+at which point we can call \texttt{execve} in the child process to start it running on its own.
+After that, the process is no longer under our control, and it functions like any other child process.
+The child process can be monitored using normal Linux child monitoring syscalls,
+such as \texttt{waitid}.
+
+\lstinputlisting[float=*ht,language=Python,label={basic},caption={Creating a new process, changing CWD, and execing}]
 {examples/basic/direct.py}
-Listing \ref{lst:basic} shows a Python program using \texttt{rsyscall}.
-We create a new child process using direct-style \texttt{clone} in the local process \verb|local|.
-Here, \texttt{local.clone()} does not create another task running the same user program,
-as with the usual \texttt{clone} system call,
-but instead returns a capability through which we can control the new child process.
-We can then call various syscalls in the child process to mutate it until it reachs the desired state.
-In this example,
-we call \texttt{chdir} in the child process to change its working directory,
-then call \texttt{execv} in the child to execute a new executable.
-Calling \texttt{execv} consumes the capability,
-releasing the process from our control;
-later use of this child process capability will fail with an exception.
-The child process can now be monitored using normal Linux child monitoring syscalls,
-such as \texttt{waitid}.
-
-These child process capabilities are created and managed in userspace,
-by launching new processes running a syscall server,
-which receives syscall requests, performs the syscall, and returns the response.
-No kernel modifications are required, and \texttt{rsyscall} is immediately deployable on recent Linux kernels.
-We discuss the implementation in depth in section \ref{implementation},
-and examine several difficult process-creation details.
-
-System calls called in a child process through \texttt{rsyscall}
-behave identically to system calls called implicitly in the current process through libc.
-A system call returning an error is reported in the standard way for system calls in Python:
-An exception is thrown at the point of the call.
-User code can branch as normal on the results of system calls,
-and implement fallbacks or other error handling logic.
-
-Since we can call any syscall,
-we can access any feature available in Linux; we are therefore at least as powerful as \texttt{fork}.
-As we'll show in section \ref{examples},
-we can in fact use Linux features in ways that are impractical with the usual fork-style or spawn-style interfaces.
-
-Direct-style \texttt{clone} has acceptable performance cost,
-and can outperform \texttt{fork} in some cases.
-Like a spawn-style interface,
-we can call \texttt{clone}
-without worse performance in the presence of large memory usage,
-and without the possibility of bugs in the presence of multi-threading.
-The performance overhead of creating processes with direct-style clone
-is comparable to the performance cost of creating processes while inside a Linux namespace.
-At \twosigma{}, we've used direct-style process creation
-to implement a production library for distributed system deployment.
-We evaluate in further detail in section \ref{evaluation}.
-
-In summary, this paper makes the following contributions:
-\begin{compactitem}
-\item We coin the term ``direct-style process creation''
-  to refer to a previously-unnamed style of process creation which is common on academic operating systems.
-\item We built \texttt{rsyscall}, a library for the Linux system call interface following the object-capability model.
-\item Using \texttt{rsyscall}, we built the first implementation of direct-style process creation on a Unix-like kernel.
-\end{compactitem}
-\texttt{rsyscall} is open source and available from \githuburl{}.
-\section{Examples}\label{examples}
+Listing \ref{basic} shows this in action in Python.
+We create a new child process using \texttt{clone} on the local process \verb|local|,
+and receive back a capability to control it.
+We call \texttt{chdir} in the child process to change its working directory to \texttt{/dev}.
+Then we call \texttt{execve},
+passing the path of the \texttt{cat} executable,
+the arguments to the new executable,
+and the unchanged set of environment variables.
+The name of the executable is passed as the first argument,
+as is typical in Unix.
+Calling \texttt{execve} consumes the capability,
+and later use of this child process capability will fail with an exception.
+
+The behavior of system calls made on a remote process through \texttt{rsyscall}
+is identical to the behavior of system calls made implicitly on the current process through libc.
+This is an API that systems programmers are already familiar with,
+and which is already thoroughly documented through Linux system call manpages.
+The novelty is constrained to the newly explicit process capabilities.
+
+The lack of novelty in the API is important not only for the learning curve,
+but also for power.
+Since we are using conventional Linux system calls,
+just in a cross-process way,
+direct-style \texttt{clone} will always be at least as powerful as \texttt{fork}.
+\section{Applications}\label{applications}
 In this section,
-we'll demonstrate direct-style process creation
-by using several sophisticated features of Linux processes.
-For concision, we assume that we are running with sufficient privileges,
-but these examples can all be run without privileges with appropriate use of \verb|CLONE_NEWUSER|
-\cite{user_namespaces}\cite{unpriv_fuse}.
-
-These examples are not novel, as such, as they use only conventional Linux system calls;
-but implementing them with \texttt{fork} or \verb|posix_spawn| requires substantially more code,
-or even a complete change in implementation approach as in section \ref{shared_fd_table}.
-We demonstrate this by comparison to fork-style implementations of these examples in section \ref{ease}.
-
-Several of these examples can be implemented by configuring and invoking existing software;
-such software effectively provides a specialized spawn-style interface.
-But such tools are often inflexible.
+we'll discuss several more advanced uses of processes
+using more sophisticated Linux features,
+and accompany them with Python examples using direct-style process creation.
+All of these examples use only conventional Linux system calls,
+and can therefore translated straightforwardly into any language with an \texttt{rsyscall} API.
+For concision, we assume that we are running with sufficient permissions in these examples,
+but these examples can all be run without privileges with appropriate use of \texttt{CLONE.NEWUSER}.
+
+Implementing these examples with \texttt{fork} or \verb|posix_spawn| requires substantially more boilerplate,
+or even a complete change in implementation approach,
+such as in section \ref{shared_fd_table}.
+We'll compare our direct-style implementations in this section
+to fork-style implementations
+in section \ref{evaluation}.
+
+Several of these examples demonstrate functionality that can be done today by specialized software and services.
+Such software often allows use of one or two of these techniques, but prevents the use of others.
 For example, a shell allows the creation of pipelines and container systems like Docker allow sandboxing,
-but the two are difficult to combine\cite{docker_pipe}.
-
-With improved process creation techniques,
-these features can be used directly by programmers
-instead of by configuring stand-alone software.
-In this way,
-individual real-world applications can pick and choose the features that are useful for them,
-as we describe in section \ref{realworld}.
-
-In these examples and in this paper,
-we will have multiple processes,
-all of which are under the control of a single program with a single flow of control.
-Because of the absence of multiple concurrent control flows,
-these examples could be executed interactively by a human operator,
-typing them in line by line in a REPL.
-We'll always be working from the perspective of this single program;
-never from the perspective of a child process, because (until we call \texttt{exec} and give up control)
-there is no meaningful program running in the child processes.
-We believe this absence of concurrency simplifies reasoning.
-\subsection{Abstraction through FD passing}\label{fd_abstraction}
+but the two are difficult to combine.
+
+One of the key advantages of direct-style process creation is that the techniques we demonstrate here
+can be easily used in combination,
+to meet the specific needs of individual real-world applications.
+\subsection{Abstraction through file descriptor passing}\label{fd_abstraction}
 In Linux (and Unix in general),
-a program can pass file descriptors (FDs) from the parent process to a child process
-by using file descriptor inheritance\cite{tlpi}.
-When a process is created, all open file descriptors are copied from the parent process to the child process.
-The child process can then make independent use of the file descriptors.
-We'll discuss this important feature further in section \ref{inheritance}.
+a program can pass file descriptors from the parent process to a child process
+by using file descriptor inheritance.
+All open file descriptors are copied at process creation time from the parent process to the child process;
+the child process can then make independent use of the file descriptors.
+We'll discuss this important feature further in section \ref{fdtables}.
 
 Most resources in Linux are managed through file descriptors,
-so this allows the parent process to pass a variety of resources to the child process,
-such as files, network connections, pipes, or other resources\cite{ucspi}.
-Since the resource is created outside the child process and passed down as an opqaue file descriptor,
-the precise details and type of the resource are not available to the child.
-This is a form of abstraction,
-and so we know ``for free''\cite{theoremsforfree}
-that the program running in the child process
-will not rely on the details of how the file descriptor was created,
-such as the filename that we opened or the hostname to which we connected.
-
-\lstinputlisting[float,language=Python,label={lst:fds},caption={Passing down FDs}]
+so this feature allows the parent process to pass a wide variety of resources to the child process.
+An executable program can be provided a file descriptor,
+which can correspond to any file in any filesystem,
+or to a network connection\cite{ucspi},
+or a pipe connected to another process,
+or a variety of other resources,
+without the process being aware of what specific resource it is accessing.
+In this way, we immediately know "for free"\cite{theoremsforfree}
+that the behavior of the program does not depend on the name of the file,
+or the hostname to it which the socket is connected,
+or the executable running in the process on the other end of the pipe,
+or other abstracted-away details.
+
+\lstinputlisting[float=*ht,language=Python,label={fds},caption={Passing down FDs}]
 {examples/fds/direct.py}
-In Listing \ref{lst:fds},
-we first open a file with read-write permission in the local process.
+In Listing \ref{fds},
+we first open a file with read-write permission.
 Then we create a child process,
@@
 We indicate that we want to operate on the child's inherited copy of the file descriptor with \verb|child.inherit_fd|.
-\texttt{inherit\_fd} performs no system calls,
-it just updates bookkeeping to return a new handle for the inherited file descriptor;
-we'll discuss it in more depth in section \ref{inheritance}.
 Then we disable \texttt{CLOEXEC} on the child's file descriptor,
-which on Linux we can do by clearing the FD flags with \verb|fcntl(fd, F_SETFD, 0)|;
-this ensures that the file descriptor will be usable by the new program after we call \texttt{exec}.
-Finally, we execute a new program in the child process,
-passing the file descriptor number as an argument.
-The new program will be able to use the file descriptor we opened through that file descriptor number.
+which on Linux we can do by clearing the FD flags with \verb|fcntl(fd, F_SETFD, 0)|.
+Finally, we execute some executable,
+passing the file descriptor number as an argument;
+since we disabled \texttt{CLOEXEC} on the file descriptor,
+the executable will be able to use that file descriptor through that file descriptor number.
+Unlike in listing \ref{basic}, we call the \texttt{execv} wrapper rather than \texttt{execve},
+so that we don't need to pass the (unmodified) environment to every \texttt{exec} call in these examples.
 \subsection{Non-shared-memory concurrency}
 Processes run concurrently,
-which enables modularity
-and allows exploiting the parallelism of the underlying hardware.
+which allows exploiting the parallelism of the underlying hardware.
 Since processes don't share memory,
 they can provide a less complex parallel programming environment
-than shared-memory thread-based approaches\cite{threads}.
-
-\lstinputlisting[float,language=Python,label={lst:pipe},caption={Creating a concurrent processing pipeline}]
+than shared-memory thread-based approaches.
+
+One of the most popular parallel programming environments is the Unix shell,
+which obtains its parallelism by running multiple processes connected via pipes.
+The Unix shell has a relatively constrained form of parallel processing,
+but it's also possible to create more complex webs of parallel processes,
+where, for example, one process might take multiple inputs over multiple pipes,
+or produce multiple outputs.
+
+\lstinputlisting[float=*ht,language=Python,label={pipe},caption={Creating a parallel processing pipeline}]
 {examples/pipe/direct.py}
-In listing \ref{lst:pipe},
-we execute a few programs concurrently,
+In listing \ref{pipe},
+we execute a few programs in parallel,
 connected by pipes.
-The \texttt{source} program generates two outputs, which go to \texttt{video\_sink} and \texttt{audio\_sink}.
-We first create two pipes in the local process,
-then inherit them down to several child processes using a helper function, \verb|argfd|,
+We first create two pipes,
+then inherit them down to several processes using a helper function, \verb|as_argument|,
 which uses \verb|inherit_fd| and \texttt{fcntl} as described in section \ref{fd_abstraction}.
-We call \texttt{execv} in each child process to run the desired programs,
-passing some file descriptors as arguments to each.
-The whole system of connected processes then runs concurrently.
+We exec each program passing some file descriptors as arguments.
+The whole system then runs in parallel,
+as normal child processes.
 \subsection{Customization without explicit support}
-While ideally all resources would be passed by file descriptor,
+While, ideally, all resources would be passed by file descriptor\cite{capsicum},
 Unix has a number of resources that are global,
-such as the process identifier space or the mount table,
-which cannot be changed on a per-process basis\cite{capsicum}.
-
-In Linux, many global resources have been made process-local
-through the namespaces system\cite{lwn_namespaces}.
+such as the pid space or the mount table,
+and which cannot be changed on a per-process basis.
+
+In Linux, most global resources have been made process-local
+through the namespaces system.
 Like other process-local resources such as the current working directory,
@@
 
-One key use of namespaces is to implement large-scale container systems such as Docker\cite{lwn_namespaces}.
-However, this is far from the only use.
+The most popular usage of namespaces is for large-scale container systems such as Docker;
+however, this is far from the only use.
 With an adaquate process creation primitive,
@@
 can be used to customize a program's behavior
-in ways that were not anticipated at development time
-by customizing the environment the process runs in\cite{plan9ns}.
-Other system calls such as \texttt{chdir} and \texttt{chroot} can also be used for this,
-but namespaces allow new ways of customization\cite{mount_namespaces}.
-
-\lstinputlisting[float,language=Python,label={lst:mount},caption={Overriding absolute path using a mount namespace}]
+in ways that were not anticipated by the author
+by customizing the environment the process runs in.
+Other system calls such as \texttt{chroot} can also be used for this.
+
+\lstinputlisting[float=*ht,language=Python,label={mount},caption={Overriding absolute path using a mount namespace}]
 {examples/mount/direct.py}
-One classic form of customization
-is overriding files at paths that were hardcoded into a program,
+One classic form of customization enabled by namespaces
+is overriding files at paths that were hardcoded into a program we can't change,
 without changing those files for the rest of the system.
-In listing \ref{lst:mount},
+In listing \ref{mount},
 we create a new child process as normal,
@@
 
-\verb|CLONE_NEWNS| causes \texttt{clone} to create the process in a new mount namespace,
-which allows us to manipulate the filesystem tree in a way that only this process will see,
-using the \texttt{mount} system call\cite{mount_namespaces}\cite{clone}.
-We call \texttt{mount}, passing \texttt{MS\_BIND}, to make the file \verb|/home/foo/custom_foo.conf|
-appear at the path \verb|/etc/foo.conf|;
-this is known as a bind mount\cite{mount}.
-Then we execute some program,
+\verb|CLONE_NEWNS| causes clone to create the process in a new mount namespace,
+which allows us to manipulate the filesystem tree in a way that only this process will see
+using the \texttt{mount} system call.
+We call \texttt{mount}, passing \texttt{MS.BIND}, to make the file \verb|/home/foo/custom_foo.conf|
+appear at the path \verb|/etc/foo.conf|.
+This is known as a bind mount.
+Then we execute some executable,
 which, when it opens \verb|/etc/foo.conf|, will see the contents of our \verb|custom_foo.conf|.
@@
 we can also deny the process access to resources that it otherwise receives by default.
-This is a key part of creating a secure sandbox for potentially malicious code\cite{seccomp}\cite{firejail}\cite{gvisor}.
-
-\lstinputlisting[float,language=Python,label={lst:unmount},caption={Unmount all and run executable via fexec}]
+This is a key part of creating a secure sandbox for potentially malicious code.
+
+\lstinputlisting[float=*ht,language=Python,label={unmount},caption={Unmount all and run executable via fexec}]
 {examples/unmount/direct.py}
-In listing \ref{lst:unmount}
+In listing \ref{unmount}
 we create a new child process,
-again in a new mount namespace using \verb|CLONE_NEWNS|.
-Since we won't be able to open files or use \texttt{execv} in the child process after the next step,
-we open several files in the child up-front.
+again in a new mount namespace using \texttt{CLONE.NEWNS}.
+We open the executable that we'll later exec in this child process,
+as well as another file as in listing \ref{fds},
+since we won't be able to do exec or open files normally after the next step.
 Then we use \texttt{umount},
@@
 removing it all from the view of this process.
-
-Then, as in listing \ref{lst:fds}, we prepare to pass \texttt{db\_fd} to the new program by unsetting \texttt{CLOEXC}.
-In this case we don't need to call \verb|inherit_fd|,
-since the file descriptor was opened directly from the child.
-We then run the executable we opened earlier using \texttt{fexec},
-which allows executing a file descriptor,
-and pass the database file descriptor number as an argument\cite{execveat}.
+As in listing \ref{fds}, we unset the \texttt{CLOEXEC} flag from the database file descriptor,
+so that it can be inherited across \texttt{exec};
+we don't need to call \verb|inherit_fd| since, in this case, the database fd was opened directly from the child.
+
+We then exec the file we opened earlier, using \texttt{fexec},
+which allows execing a file descriptor,
+and also pass the database file descriptor as an argument,
+as in listing \ref{fds}.
+(The underlying system call for \texttt{fexec} is \texttt{execveat};
+this \texttt{fexec} is a helper function.)
 Note that this executable must be statically linked,
-or it wouldn't work in an empty filesystem namespace
+or it wouldn't work in an empty filesystem namespace,
 with no libraries to dynamically link against.
@@
 that it won't be able to tamper with the rest of the system.
-A sandbox which is truly robust against malicious or compromised programs requires additional steps,
-but this is a substantial start\cite{firejail}\cite{gvisor}.
+A sandbox which is truly robust against malicious or compromised programs requires some additional steps,
+but this is a substantial start.
 Such a technique can also be used when a full sandbox is not relevant,
@@
 This allows us to approximate capability-based security\cite{capsicum}.
-\subsection{Nested clone and pid namespaces}\label{pidns}
+\subsection{Nested clone and pid namespaces}
 Process-local resources can also be used to control the lifetime of resources used by that process.
-Some Linux resources are not automatically cleaned up on process exit;
+Many Linux resources are not automatically cleaned up on process exit;
 a poorly coded program may allocate global resources
@@
 those subprocesses may unintentionally leak,
-and be left running on the system even after the program itself has stopped\cite{caternfork}.
+and be left running on the system even after the program itself has stopped.
 
@@
 When the init process dies,
-all other processes in the pid namespace are destroyed\cite{pid_namespaces}.
-
-\lstinputlisting[float,language=Python,label={lst:pidns},caption={Nested clone and pid namespace}]
+all other processes in the pid namespace are destroyed.
+
+\lstinputlisting[float=*ht,language=Python,label={pidns},caption={Nested clone and pid namespace}]
 {examples/pidns/direct.py}
-In listing \ref{lst:pidns},
-we create a child process which we'll name \texttt{init},
-passing \verb|CLONE_NEWPID| to create it in a new pid namespace\cite{clone}.
+In listing \ref{pidns},
+we create a child process which we call \texttt{init},
+passing \verb|CLONE_NEWPID| to create it in a new pid namespace.
 
@@
 which can be used exactly like a direct child process.
-We \texttt{exec} in the grandchild,
+We exec in the grandchild,
 and we can monitor the grandchild process from \texttt{init},
@@
 We can handle these responsibilities ourselves directly from our program,
-or we can continue on by \texttt{exec}ing an init daemon in \texttt{init} to handle it for us
-\cite{pid_namespaces}\cite{tini}\cite{dumb_init}.
+or we can continue on by execing an init daemon in \texttt{init} to handle it for us.
 
@@
 \subsection{Shared file descriptor tables}\label{shared_fd_table}
-We can create a child process which shares its file descriptor table with the parent process.
-This can be useful for a variety of purposes,
-such as accessing resources in other namespaces, or in multiple namespaces at once.
-We'll give one example using the Filesystem in Userspace (FUSE) Linux subsystem.
-
-FUSE\cite{fuse} allows a filesystem to be implemented with a userspace program.
-Many interesting filesystems\cite{sshfs}\cite{encfs}\cite{mp3fs} have been implemented using FUSE,
-representing a variety of resources as files.
-
-For improved modularity, we might want to mount and use a FUSE filesystem in our program,
-in a way that no other process can see it,
-without entering a new namespace ourselves.
-We can do this by creating a child process that shares its file descriptor table with the parent process.
-
-\lstinputlisting[float,language=Python,label={lst:fuse},caption={Shared file descriptor tables}]
+A library can create a child process to perform a useful service.
+This can sometimes be done more easily when the child process shares file descriptors with the parent.
+
+\lstinputlisting[float=*ht,language=Python,label={fuse},caption={Shared file descriptor tables}]
 {examples/fuse/direct.py}
-In listing \ref{lst:fuse},
+In listing \ref{fuse},
 we first create a child process \verb|ns_child| in a new mount namespace with \verb|CLONE_NEWNS|,
-and this time also pass \verb|CLONE_FILES|,
-which causes the file descriptor table to be shared between the parent process and the child process\cite{clone}.
-We create another child from \verb|ns_child|
-and use it to \texttt{exec} a FUSE filesystem,
+and, for the first time, also pass \verb|CLONE_FILES|,
+which causes the file descriptor table to be shared between the parent process and the child process.
+We create a grandchild off of \verb|ns_child|
+and use it to exec a FUSE filesystem,
 which will appear only in the mount namespace of \verb|ns_child| and its descendents.
 Then we can open FUSE files in \verb|ns_child|
-and use those files in \texttt{local},
+and use those files in the parent process,
 through the shared file descriptor table.
-We call \texttt{use\_fd} which, like \texttt{inherit\_fd},
-updates bookkeeping to create a new handle for the file descriptor existing in the shared fd table;
-we'll discuss \texttt{use\_fd} in section \ref{inheritance}.
 
@@
 can allow similar behavior without sharing the file descriptor table,
-but is substantially more complex\cite{scm_rights}.
+but is substantially more complex for the user.
 Nevertheless, for fork-style and spawn-style process creation,
 file descriptor passing over a Unix domain socket
-is our only option.
+is our only option to do something like this.
 \section{Implementation}\label{implementation}
+\subsection{Basics about \texttt{rsyscall}}
 In this section, we'll give a brief overview of the implementation of \texttt{rsyscall},
@@
 \texttt{rsyscall} can be conceptually divided in two parts:
-the basic syscall primitive,
+the basic cross-process syscall primitive,
 and a language-specific library built on top.
@@
 Such libraries only need to be able to call syscalls and explicitly specify a process in some way;
-they are, for the most part, agnostic to how the syscall primitive is implemented.
-The syscall primitive takes a syscall number, and some number of register-sized integer arguments,
-and arranges to call that syscall in the specified process,
-returning the response as a single register-sized integer.
-
-When making a syscall in the local process, the syscall is performed normally,
-directly on the local thread.
-
-When making a syscall in another process,
-\texttt{rsyscall}'s default userspace cross-process syscall primitive sends the syscall request
-to a userspace ``syscall server'' running in the target process,
-which performs the syscall and sends the response back.
-Communication to the syscall server happens over file descriptors,
-typically a pair of pipes.
-
-We use file descriptors for transport of data rather than shared memory
-to support straightforward integration with existing event loops;
-this is a key design constraint.
-While we only show synchronous programs in this paper,
-calling a syscall in another process may block at any time,
-and a complex program will likely have other requests to service concurrently.
-Even a relatively simple program may want to monitor multiple child processes at a time.
-To support this,
-our implementation fully supports asynchronous usage with a user-provided event loop,
-including the Python \texttt{async/await} coroutine support\cite{python_coroutines}.
-
-We have chosen to implement in userspace first,
-rather than immediately implementing this in the kernel,
-to aid deployability
-and to allow fast iteration while developing userspace code using these features.
-In the past, attempts to upstream kernel support for novel features without extensive userspace usage
-has had a poor reception in the Linux kernel community\cite{lwn_checkpoint}\cite{first_class_address_space}.
-We believe proving viability in userspace first will be better in the long term.
-
-\texttt{ptrace} provides a way to perform arbitrary actions on a target process,
-including system calls,
-but is not suitable for us\cite{ptrace}.
-As we discuss in section \ref{related}, a number of systems, such as \texttt{gdb} and \texttt{strace},
-use \texttt{ptrace} to implement debuggers.
-But multiple processes cannot use \texttt{ptrace} on a single target process at the same time;
-thus, if we used \texttt{ptrace} to implement \texttt{rsyscall},
-such debugging tools would not work on \texttt{rsyscall}-controlled processes,
+they are, for the most part, agnostic to how the cross-process syscall is implemented.
+
+On Linux \verb|x86_64|, a syscall is specified by a syscall number plus six register-sized arguments;
+a syscall returns one register-sized value.
+\texttt{rsyscall}'s default implementation of cross-process syscalls sends those seven integers over a pipe,
+and waits for a response on another pipe.
+Processes are created running a loop which, at each iteration,
+reads a syscall request off the pipe,
+performs that syscall,
+and writes the return value back over the return pipe.
+In this way, a cross-process syscall works much like a very primitive remote procedure call.
+
+We use pipes (or other file descriptors) for transport of data rather than a shared memory segment
+to support straightforward integration with event loops.
+
+ptrace provides an alternative means to perform arbitrary actions on other processes,
+including system calls.
+However, among other issues, it has the substantial disadvantage of not permitting multiple ptracers.
+A ptrace-based implementation would prevent using \texttt{strace} or \texttt{gdb} on \texttt{rsyscall}-controlled processes,
 which is an unacceptable limitation for a general-purpose utility.
@@
 The most simple way to access memory is for the local process to be in the same address space as the target process.
-This is the case most of the time; we pass \verb|CLONE_VM| to \texttt{clone} by default
-in the \texttt{rsyscall} wrapper for \texttt{clone}.
-
-Sometimes, the target process may be in another address space;
-for example, if the target process is at a different privilege level,
-we will want it to be in a different address space for security reasons\cite{vfork_dangerous}.
-There are a number of available techniques in that scenario;
-we choose to copy memory over a pair of pipes,
-again using file descriptors for the sake of easy event-loop integration.
+This is the case most of the time; we pass \texttt{CLONE.VM} to \texttt{clone} by default in the Python wrapper.
+However, the target process may be in another address.
+For example, if the target process is at a different privilege level,
+it must be in a different address space,
+as sharing an address space with a process at a different privilege level is unsafe.
+
+We implement access to the memory of processes in different address spaces through another set of pipes,
+by copying memory into and out of those pipes using \texttt{read} and \texttt{write} system calls in the target process.
+A \texttt{read} syscall corresponds to a write to memory,
+and a \texttt{write} syscall corresponds to a read from memory.
+
+The \verb|process_vm_readv| and \verb|process_vm_writev| system calls
+allow the caller to read and write memory from the virtual address space of other processes.
+However, they require that the caller have specific credentials relative to the process being accessed,
+which may not always be the case.
+Additionally, these system calls are disabled if ptrace is disabled system-wide,
+which is a niche but possible system configuration.
+To ensure that \texttt{rsyscall} can be used for arbitrary purposes and on arbitrary systems, we avoided these calls.
 \subsection{\texttt{clone}}\label{clone}
@@
 
-Besides \texttt{clone}, \texttt{vfork} and \texttt{fork} also create processes,
-but they are not suitable for flexible direct-style process creation.
+There are three Linux system calls which create processes:
+\texttt{fork}, \texttt{vfork} and \texttt{clone}.
 \texttt{vfork} suspends execution of the parent process
 while waiting for the child process to exit or call \texttt{execve},
-which is immediately unsuitable\cite{vfork}.
-\texttt{fork} lacks many features which are restricted to \texttt{clone}.
-For example, with \texttt{fork},
-we could not create child processes which share the parent's address space,
-which would complicate memory access.
-
-The raw \texttt{clone} system call creates a new process
+which is immediately unsuitable for direct-style process creation.
+\texttt{clone} provides a superset of the functionality of \texttt{fork},
+so we focused our attention on \texttt{clone}.
+
+\texttt{clone}
+(along with \texttt{fork}) creates a new process
 which immediately starts executing at the next instruction after the syscall instruction,
@@
 
-\texttt{clone} lets us change the stack for the new process.
-We can use this to make the new process call an arbitrary function,
-by storing its address on the new stack.
+Common Linux libcs provide a wrapper for \texttt{clone}
+which allow executing an arbitrary function with an arbitrary argument,
+but since we want to support remote execution,
+we can't use the libc wrapper.
+
+Instead, since \texttt{clone} lets us change the stack for the new process,
+we use this to return, after the system call to an arbitrary function address stored on the stack.
 Further arguments can also be passed on the stack,
@@
 
-We use this to create processes running the syscall server.
+We use this to create processes running the syscall request-response loop described in the previous section.
 After this,
-the parent process can be begin to call system calls in the child process.
-Most system calls work as normal;
-the new child process can be modified freely through \texttt{chdir}, \texttt{dup2}, and other system calls.
-From the system calls related to process creation,
-only \texttt{execve} needs substantial further attention.
+most system calls can be called as normal.
+The new process can be modified freely through \texttt{chdir}, \texttt{dup2}, and other system calls.
+Out of system calls related to process creation,
+only \texttt{execve} and \texttt{unshare} need substantial further attention.
 \subsection{\texttt{exec}}\label{execve}
@@
 because when it is successful, it does not return.
-We need a way to determine if \texttt{execve} is successful;
-naively waiting for a response to the syscall request may leave us waiting forever.
-
-There is a traditional technique used with \texttt{fork}
-to detect a successful \texttt{execve} using a pipe,
-which unfortunately won't work for us.
-With this technique, the parent process creates a pipe before forking,
-ensures both ends are marked \verb|O_CLOEXEC|,
-performs the \texttt{fork},
-closes the write end of the pipe,
-and \texttt{read}s the read end of the pipe.
-The child process either successfully calls \texttt{execve} or exits;
-either way, the last copy of the write end of the pipe will be closed,
-which causes the \texttt{read} in the parent process to return EOF.
-The parent can then check that the child process hasn't exited;
-if the child hasn't exited, then it must have successfully called exec.
-
-This trick works with \texttt{fork},
+Therefore we need a way to determine if \texttt{execve} is successful;
+naively waiting for a response to the syscall request will leave us waiting forever.
+
+There is a traditional technique to detect a successful \texttt{execve} using a pipe.
+We create a pipe before forking,
+ensure both ends are marked \verb|O_CLOEXEC|,
+perform the fork,
+call \texttt{execve} in the child,
+close the write end of the pipe,
+and wait for EOF on the read end.
+If the child process has neither successfully called \texttt{execve}, nor exited for some other reason,
+then the write end of the pipe will still be open in the child process's fd table,
+and the read end of the pipe will not return EOF.
+But once the child process calls \texttt{execve} successfully,
+\verb|O_CLOEXEC| will cause the write end of the pipe to be closed,
+and the read end of the pipe will return EOF.
+
+This trick works well with \texttt{fork};
 but it's not general enough to work with \texttt{clone}.
@@
 and the read end of the pipe will immediately return EOF,
-before the child has called \texttt{execve} or exited.
+regardless of whether the child has called \texttt{execve} or exited.
 
 Fortunately, there is an alternative solution, which does work with \verb|CLONE_FILES|.
-\texttt{clone} has an argument, \texttt{ctid}, which specifies a memory address.
-If the \verb|CLONE_CHILD_CLEARTID| \texttt{clone} flag is set,
-then when the child exits or execs,
-the kernel will set \texttt{ctid} to zero and then,
-crucially, perform a futex wakeup on it\cite{clone}.
+The \texttt{ctid} argument to \texttt{clone} specifies a memory address which,
+when the \verb|CLONE_CHILD_CLEARTID| flag is set,
+the kernel sets to zero when the child exits or execs,
+and then, crucially, performs a futex wakeup on.
 
@@
 but the relevant detail for us here is that we can wait on an address
-until a futex wakeup is performed on that address\cite{futex}.
+until a futex wakeup is performed on that address.
 This means we can wait on \texttt{ctid} until the futex wakeup is performed,
 and in this way get notified of the child process calling \texttt{execve} or exiting.
-A process can only wait on one futex at a time;
-to monitor multiple futexes from a single event loop,
-we need to create a dedicated child process for each futex we want to wait on,
+
+Unfortunately, futexes in current Linux integrate poorly:
+There is no way for a single process to wait for more than one futex at a time,
+and no way to monitor a futex with file-descriptor-monitoring syscalls such as \texttt{poll}.
+The best we can do is create a dedicated child process for each futex we want to wait on,
 and have this child process exit when the futex has a wakeup.
-We can then monitor these child processes from an event loop using standard techniques\cite{signalfd}\cite{pidfd}.
-
-So, we pass \texttt{ctid} whenever we call \texttt{clone},
+Monitoring child processes can be straightforwardly integrated into an event loop.
+
+While slightly complex to implement, this solution works well.
+We provide \texttt{ctid} whenever we call \texttt{clone},
 and set up a process to wait on that futex.
@@
 without the child process itself exiting,
-then we can catch that in \texttt{execve}
-and know that the child has successfully completed an \texttt{execve} call.
+then we can catch that in \texttt{execve} and know that the child has successfully called \texttt{execve}.
 
@@
 this ambiguity is unfortunate, but it is also present in the pipe-based approach.
-
-Some other Unix-like systems natively provide the ability to wait for a child's \texttt{execve};
-our implementation would be simplified, and the ambiguity eliminated,
-if we had this ability on Linux\cite{freebsd_kqueue}.
-One approach would be to add a new \texttt{clone} flag to
+This is, we believe, the best solution currently available.
+
+We would prefer for Linux to natively provide functionality to wait for a child's \texttt{execve}.
+Some other Unix-like systems provide this;
+kqueue, on FreeBSD, allows waiting for exec in arbitrary processes through kqueue's \verb|EVFILT_PROC|.
+
+One approach for Linux would be to add a new \texttt{clone} flag to
 opt-in to receiving \texttt{WEXECED} events through \texttt{waitid}.
@@
 when the \texttt{WEXECED} event happens.
-\subsection{Handling file descriptors}\label{fd_handles}
-In Unix, there are many types of resources maintained inside the kernel,
-such as file descriptors or memory mappings,
-which are referred to by identifiers which are valid only within a certain context\cite{tlpi}.
-For example, file descriptors are referred to by file descriptor numbers
-which are valid only in a certain file descriptor table;
-memory and memory mappings are referred to by memory addresses
-which are valid only in a certain address space;
-the same pattern is followed for a number of other resources.
-Each process has one file descriptor table, one address space, and one of each of the other contexts.
-Two or more processes can share individual contexts;
-for example, two processes might share an address space, but not a file descriptor table.
-
-The core issues here are the same for all such resources,
-so we'll focus on file descriptors.
-For concision and clarity, we'll abbreviate ``file descriptor'' as ``FD'' in this section.
-
-A typical POSIX program operates within only one process,
-and so only needs to concern itself with one of each kind of context.
-To identify and operate on an FD, it can use just an FD number,
-which is valid within the global implicit process's FD table.
-That's sufficient to make syscalls involving that FD.
-
-A program using \texttt{rsyscall}, however,
-operates in multiple processes,
-and has to deal with resources across multiple contexts.
-To identify an FD,
-it needs not just a FD number,
-but also some kind of identifier for an FD table.
-To then operate on that FD,
-it needs a capability for a process with that FD table,
-so that it can make syscalls in that process.
-
-To manage this, we add a new concept alongside FD, FD number, and FD table:
-The FD handle.
-An FD handle is a pair of an FD number and a process capability.
-This both precisely identifies an FD, and allows operating on it.
-We say an FD handle is associated with a process
-if that FD handle contains that process capability.
-
-All system calls which would otherwise return FD numbers or take FD numbers as an argument,
-now instead return FD handles or take FD handles as an argument.
-For convenience, as shown in section \ref{examples} for \texttt{fcntl},
-helper methods on the FD handle object are provided for many system calls dealing with a single FD;
-these methods simply call the corresponding underlying system call
-using the process capability contained in the FD handle.
-
-In this way, we can work with FDs without the ambiguity of dealing with raw FD numbers in a multi-process context.
-\subsubsection{File descriptor inheritance and table sharing}\label{inheritance}
-The picture is complicated by the need to support some FD behaviors which implicitly cross between processes.
-FDs are copied between processes at process creation time through FD inheritance,
-and two processes can share the same FDs through a shared FD table.
-
-FD inheritance is a behavior exhibited by \texttt{clone} (without \verb|CLONE_FILES|) and by \texttt{fork}.
-When these system calls are called,
-they create a new child process with a new FD table.
-The new FD table contains copies of all the FDs existing in the parent process's FD table at the time of the system call,
-at the same FD numbers.
-
-FD table sharing occurs when \texttt{clone} \textit{with} \texttt{CLONE\_FILES} is called;
-this creates a new child process that shares its FD table with its parent.
-As a result, either process can operate on any FD in the table,
-through the same FD numbers.
-
-These behaviors are unusual in that they operate on all open FDs at once,
-rather than individual FDs explicitly denoted by system call arguments.
-There are other ways to pass FDs between processes,
-such as \texttt{SCM\_RIGHTS}\cite{scm_rights},
-but those operate on individual FDs through explicit arguments to system calls.
-One or more FD handles are explicitly passed to a system call in one process,
-and one or more FD handles are explicitly returned from another system call in another process.
-The behaviors of FD inheritance and FD table sharing happen for all FDs, implicitly,
-and so they must be dealt with differently to be fully supported.
-
-\texttt{rsyscall} supports both behaviors in the same way,
-by allowing us to create new FD handles on-demand
-referring to already-open FDs created by FD inheritance or FD table sharing;
-by creating new FD handles in new processes, we can call syscalls in those processes on those FDs.
-\texttt{rsyscall} provides two functions
-\texttt{inherit\_fd} (used in listings \ref{lst:fds} and \ref{lst:pipe})
-and \texttt{use\_fd} (used in listing \ref{lst:fuse}), which support FD inheritance and FD table sharing, respectively.
-Both functions take as arguments a process capability $A$ and an FD handle $H$ associated with some other process $B$,
-and return a new FD handle associated with process $A$.
-For \texttt{inherit\_fd},
-this new handle refers to an \textit{inherited copy} of the FD which $H$ refers to;
-\texttt{inherit\_fd} will fail if $B$ is not $A$'s parent process,
-or if $H$ was not open at the time $A$ was created.
-For \texttt{use\_fd},
-this new handle refers to the \textit{same} FD that $H$ refers to, through the shared FD table;
-\texttt{use\_fd} will fail if $A$ and $B$ don't share the same FD table.
-\section{Evaluation}\label{evaluation}
-Our goal is a powerful approach to process creation that is easy to use correctly.
-Performance of our implementation is secondary,
-but the overhead of our implementation relative to alternative approaches
-must be low.
-
-We evaluate the degree to which we meet this goal by answering the following questions:
+\subsection{Managing file descriptor tables}\label{fdtables}
+As demonstrated in section \ref{shared_fd_table},
+the \verb|CLONE_FILES| flag can be passed to \texttt{clone},
+causing the file descriptor table to be shared between the parent process and child process.
+The same file descriptors are open in both processes at the same numbers,
+and if new file descriptors are opened in either process,
+they are also visible in the other process.
+The file descriptor table is unshared once the child calls \texttt{execve}.
+This is simple to model and convenient for many purposes.
+
+If \verb|CLONE_FILES| is not passed to \texttt{clone},
+then \texttt{clone} has the same behavior as \texttt{fork}:
+The new process has a new file descriptor table,
+containing copies of all the file descriptors existing in the parent at the time of the system call.
+This is known as file descriptor inheritance.
+
+This same behavior can also be triggered after process creation by calling \verb|unshare(CLONE_FILES)| or \texttt{execve}.
+If \verb|unshare(CLONE_FILES)|
+is called in a process currently sharing its file descriptor table with another process,
+then after the call that process will have a new, private file descriptor table,
+again with a copy of all the file descriptors existing at the time of the system call.
+
+If \texttt{execve} is called,
+the treatment of the file descriptor table is the same as \verb|unshare(CLONE_FILES)|,
+except that file descriptors with the \texttt{CLOEXEC} flag set on them
+are not copied into the newly created file descriptor table.
+We'll discuss the semantics of \texttt{CLOEXEC} more in section \ref{cloexec}.
+\subsubsection{Tracking file descriptors through handles}
+After a system call has copied all the file descriptors in the old table into the new table,
+we need to decide which file descriptors we want to keep open in the new table,
+and which file descriptors should be closed.
+
+This is a question of resource management.
+In a language with manual resource management,
+such as C,
+this would be delegated to the user,
+and the responsibilities of the \texttt{rsyscall} library would end there.
+Manual resource management in C has led to many bugs,
+many of which with significant security implications.
+
+Python, instead, has automatic resource management,
+and therefore we determine which file descriptors should stay open automatically:
+if a certain file descriptor is in use by the program,
+through a certain process,
+that file descriptor should stay open in that process's file descriptor table.
+
+To achieve this, file descriptors are used through garbage-collected handles;
+each handle is associated with a process and contains a specific file descriptor number.
+For a handle to remain valid,
+that file descriptor number must continue to refer to an open file descriptor in that process.
+We determine which handles are live (in use by the program)
+by recourse to the native Python garbage collection system.
+The file descriptors we keep open are then those which are necessary to keep the live handles valid.
+
+When a process changes file descriptor table through \verb|unshare(CLONE_FILES)|,
+we need to make sure all the live file descriptor handles associated with that process continue to be valid.
+This is straightforward:
+we enumerate all the handles associated with the process,
+and don't close the file descriptors that those handles refer to.
+
+Likewise, if a process changes file descriptor table through \texttt{execve}
+in a way that allows the program to keep control of the process,
+we make sure to keep all the file descriptors with live handles open.
+This requires careful management of \texttt{CLOEXEC}.
+
+When a process is newly created,
+no file descriptor handles yet exist.
+New handles can be created by opening files in the process,
+or by using the \verb|inherit_fd| function to create handles for inherited file descriptors.
+\verb|inherit_fd| takes a file descriptor handle associated with the parent process,
+and, if that file descriptor was open at the time the child process was created,
+creates a new handle associated with the child process
+for the child process's copy of that file descriptor.
+Once the inherited file descriptors are closed as described in section \ref{cloexec},
+\verb|inherit_fd| can't be used again until the process next changes file descriptor table.
+\subsubsection{Closing file descriptors after inheritance}\label{cloexec}
+After creating a new file descriptor table,
+and establishing which file descriptors should be kept open,
+we must promptly close other file descriptors.
+Leaving unwanted file descriptors open in the new table wastes resources,
+and can also cause bugs.
+For example, it's common practice to close the write end of a pipe
+and expect an EOF on the read end;
+if the write end is copied into a new file descriptor table between creating and closing,
+and the write end is never closed in the new table,
+the read end will never get an EOF.
+
+However, we can't simply close all other file descriptors,
+because we must preserve the semantics of implicit file descriptor inheritance.
+The possibility of implicit inheritance of file descriptors is an old Unix feature,
+which is useful in a wide variety of situations,
+in much the same way as environment variables, which are also implicitly inherited.
+As with environment variables,
+implicitly inherited file descriptors allow a resource to be passed down a process hierarchy
+without intervening programs being aware;
+this is a form of dynamic scope, which, though notorious, is often useful.
+
+Fortunately, file descriptors which should be implicitly inherited are marked:
+they have \texttt{CLOEXEC} turned off.
+For implicit inheritance to function, \texttt{CLOEXEC} must not be set;
+otherwise, the file descriptor will be closed, rather than inherited,
+the next time the process calls \texttt{execve}.
+And since \texttt{execve} can be called at any time in a general purpose program,
+any library that does not want its file descriptors to be inherited across \texttt{execve}
+must keep \texttt{CLOEXEC} turned on at all times.
+So, at any time, we know that the file descriptors with \texttt{CLOEXEC} off
+are exactly the file descriptors which should be implicitly inherited.
+
+Therefore, we simply need to close all file descriptors that have \texttt{CLOEXEC} set,
+and which are not referenced by a live file descriptor handle;
+we refer to this operation as \verb|close_cloexec|.
+
+For \verb|unshare(CLONE_FILES)|, we perform \verb|close_cloexec| automatically in userspace.
+For \texttt{clone}, we perform \verb|close_cloexec| only on user request,
+so that the user can control when \verb|inherit_fd| stops working.
+Even if the user never requests that unused file descriptors be closed,
+as long as they call \texttt{execve}, the table will be cleaned at that point.
+In both of these cases,
+we would benefit from a new system call which takes a list of file descriptors
+and closes all \texttt{CLOEXEC} file descriptors except for those on the list;
+this would replace our userspace implementation of \verb|close_cloexec|.
+
+For \texttt{execve}, \verb|close_cloexec| is, in some sense, done automatically for us by the kernel;
+except we don't get the chance to preserve live file descriptors handles,
+without first unsetting \texttt{CLOEXEC} on them.
+Again an extension to the kernel would be useful:
+A new argument to \texttt{execve} which allows specifying a list of file descriptors to inherit
+despite the \texttt{CLOEXEC} flag.
+This would allow programs to avoid the pattern
+of unsetting \texttt{CLOEXEC}, calling \texttt{execve}, and then immediately re-setting \texttt{CLOEXEC},
+just to inherit a file descriptor across a single exec.
+This might seem minor,
+but unsetting \texttt{CLOEXEC} before \texttt{execve} is not possible
+when the process calling \texttt{execve} has a shared file descriptor table,
+because some other process might call \texttt{execve} as well,
+and erroneously inherit the file descriptor.
+\section{Evaluation \textbf{TODO}}\label{evaluation}
+
+We quantitatively answer the following questions:
 \begin{compactitem}
@@
   Does direct-style process creation on Linux allow for simpler programs than the alternatives?
-  We compare implementations of similar functionality using both direct-style and fork-style in \ref{ease},
-  and find substantial benefits for direct-style.
+  We implement and compare similar functionality using both direct-style and fork-style in \ref{ease}.
 \item
-  Does our \texttt{rsyscall}-based direct-style process creation interface
-  have acceptable performance overhead relative to the alternatives?
-  We evaluate a number of microbenchmarks in \ref{subprocess_bench},
-  and find an acceptable level of overhead.
+  Does a clone-based implementation avoid the well-known performance overheads of fork?
+  We evaluate clone against fork on several microbenchmarks in \ref{microbench}.
 \item
-  Does direct-style process creation perform well in the ``real world''?
-  We discuss our positive experience with using direct-style process creation at \twosigma{} in \ref{realworld}.
+  Does a high-level Python direct-style process creation interface
+  have acceptable performance relative to the Python standard library's spawn-style interface?
+  We benchmark direct-style process creation against spawn-style in \ref{subprocess_bench}.
+% TODO have an item about use at TS? hm.
+% could be qualitative
 \end{compactitem}
@@
 \end{table}
-Does direct-style process creation on Linux allow for simpler programs than the alternatives?
-To answer this, we compare the direct-style examples shown in section \ref{examples}
-to programs with the same behavior implemented in fork-style.
-
-Most of the examples cannot be implemented with typical Linux spawn-style interfaces such as \verb|posix_spawn|,
-so we compare only to fork-style.
-
-The line counts of the direct-style and fork-style implementations are listed in table \ref{tab:ease}.
-Direct-style consistently takes under half the lines to express the same functionality.
-
-\lstinputlisting[float,language=Python,label={basic_fork},caption={Fork-style: Creating a new process, changing CWD, and execing}]
+We implement a variety of simple but powerful techniques with processes
+using both direct-style and fork.
+Our results are summarized in table \ref{tab:ease}.
+Direct-style consistently takes under half the lines the express the same thing.
+\lstinputlisting[float=*ht,language=Python,label={basic_fork},caption={Fork-style: Creating a new process, changing CWD, and execing}]
 {examples/basic/fork.py}
-Listing \ref{basic_fork} shows one of our fork-style implementations,
-corresponding to the direct-style implementation in listing \ref{lst:basic}.
+An example of one of the fork-style implementations is in listing \ref{basic_fork}.
+It corresponds to the direct-style example in listing \ref{basic}.
 
 In our fork-style implementations,
-we assume substantial infrastructure is available to make fork-style simpler;
-despite this, fork-style implementations are still significantly more verbose than direct-style.
+we assume a great deal of infrastructure is available to make fork-style simpler.
 We assume the existence of a robust IPC system,
-with communication channels set up automatically between the parent process and the child process,
 with an already-defined protocol which can cover all errors from all system calls.
-
-The main source of additional code in fork-style implementations
-is reporting errors back to the main program using IPC.
-This pattern is common in complex usage of \texttt{fork}\cite{posix_spawn_error_pipe}\cite{python_subprocess_errpipe}.
-Many users of \texttt{fork} avoid complex IPC
-by encoding a subset of the error information into the exit code of the child process
-when an error is encountered during child setup.
-Such an approach removes the need for IPC,
-but still requires similar code for catching errors, encoding them into the exit code,
-and detecting the error in the parent process by decoding the exit code.
-
-Direct-style does not require any extra work to report errors from the child process;
-errors are reported just like any other system call.
-This is the main immediately visible simplification of using direct-style process creation.
-
-It is possible to build an abstracted wrapper for fork-style process creation
-which abstracts away this error reporting code.
-We have built several such wrappers,
-but have ultimately discarded them in favor of direct-style process creation\cite{sfork}.
-
-We found that such wrappers, besides their complexity,
-don't provide any fundamental improvement
-for the issue of communication between the concurrently executing child process and parent process.
-As a result, while such wrappers can remove some boilerplate in easy cases,
-they remain complex in difficult cases like in sections \ref{pidns} and \ref{shared_fd_table},
-which rely on a single program being able to coordinate multiple processes.
-We believe only direct-style process creation is able to easily express such functionality.
-
-Direct-style also removes the need to be concerned about the state of the calling process,
-though this simplification is not immediately visible in a small example.
-The fork-style implementations may break if there are multiple threads in the calling process,
-and may be slow if the calling process has large amounts of memory mapped, as we'll see in section \ref{subprocess_bench}.
-The direct-style implementations do not have these limitations.
-\subsection{Microbenchmark results}\label{subprocess_bench}
-\subsubsection{Basic process creation}
+Despite this, fork-style implementations are still significantly more verbose than direct-style.
+\subsection{Clone versus fork}\label{microbench}
+\begin{figure}[h!]
+\centering
+ \includegraphics[width=0.5\textwidth]{microbench}
+ \caption{fork and clone performance under varying memory usage}
+ \label{fig:microbench}
+\end{figure}
+
+We run a benchmark which times how long process creation takes
+with \texttt{fork}, \texttt{clone}, and \verb|clone(CLONE_VM)|
+with varying amounts of anonymous memory mapped in the parent process
+before running the benchmark.
+\texttt{clone} without \verb|CLONE_VM| functions identically to fork.
+The results are in figure \ref{fig:microbench}.
+Without \verb|CLONE_VM|,
+process creation time is linear in the amount of memory mapped in the parent process.
+At the upper end of our benchmark, with 4 GB of memory mapped ($10^6$ pages),
+\texttt{fork} takes an average of 42.6 milliseconds to create a process,
+while \verb|clone(CLONE_VM)| stays flat at 21.1 microseconds.
+
+This is because \texttt{fork}, and \texttt{clone} without \verb|CLONE_VM|,
+must set up memory mappings in the child process for all the memory mapped in the parent process.
+\verb|clone(CLONE_VM)| does not have to do this bookkeeping,
+so its process creation time is constant in the amount of memory used in the parent process.
+
+\subsection{High-level direct-style versus spawn-style}\label{subprocess_bench}
 \begin{figure}[h!]
@@
  \includegraphics[width=0.5\textwidth]{subprocess_bench}
- \caption{Python spawn-style vs direct-style performance under varying memory usage}
+ \caption{Python subprocess and direct-style clone performance under varying memory usage}
  \label{fig:subprocess_bench}
 \end{figure}
-We evaluate a simple process creation workload:
-create a child process, exec the \texttt{true} binary, and wait for it to exit.
-We implement this using Python \texttt{rsyscall} direct-style \texttt{clone},
-the Python standard library's \texttt{subprocess.run} (which is implemented primarily with \texttt{fork}),
-\texttt{fork} from C,
-\texttt{vfork} from C,
-and \texttt{posix\_spawn} from C.
-We run on CPython 3.7.7, glibc 2.30, Linux 4.19.93,
-pinned to an isolated single core on an Intel i9-9900K CPU at standard clock speed,
-with 60GB of RAM.
-We vary the amount of anonymous memory mapped in the parent process
-to demonstrate how each implementation scales with memory usage.
-The results are summarized in figure \ref{fig:subprocess_bench}.
-
-Our true baseline for performance is the Python standard library's \texttt{subprocess.run};
-this takes an average of 1.4 milliseconds at low memory usage,
-while \texttt{rsyscall}'s \texttt{clone} takes an average of 2.2 milliseconds at any memory usage.
-As expected, the \texttt{fork}-based implementations scale linearly with memory usage,
-and the C implementations vastly outperform the Python implementations.
-\texttt{posix\_spawn} takes an average of 440 microseconds to start a process,
-and \texttt{vfork} takes 400 microseconds,
-regardless of the memory usage of the calling process,
-
-We perform another microbenchmark to evaluate the overhead of performing additional modifications of the child process.
-We call \texttt{getpid} from Python
-in both the child process and the parent process on the same benchmark setup.
-The average time per \texttt{getpid} call is
-3 microseconds when called in the parent process without going through \texttt{rsyscall},
-and 561 miroseconds when called through \texttt{rsyscall} in the child process.
-The difference, 558 microseconds,
-is the amount of overhead incurred
-by performing our child process setup system calls through \texttt{rsyscall}.
-
-These slowdowns in process creation and modification are substantial,
-but we found that this overhead is acceptable in practice.
-Process creation in Python is already slow, taking milliseconds of time,
-so it is not expected to be on the fast path.
-In that context, \texttt{rsyscall} has a reasonable cost compared to \texttt{subprocess.run},
-and avoids the bad scaling of \texttt{fork} which might be unexpected by the naive programmer.
-
-Furthermore, as we'll discuss in section \ref{realworld},
-we've found that the greater expressivity of direct-style
-provides for programs that are more efficient on a large scale,
-which makes up for the performance cost in micro-benchmarks.
-We've also found that, for many interesting applications,
-the performance overhead of direct-style process creation (or Python, for that matter)
-is dwarfed by the execution time of the native-code programs we ultimately run.
-
-As a result, though implementations in native code and in the kernel would likely remove most of this overhead,
-we have chosen to not invest effort into optimizing process creation at this micro-level,
-to preserve implementation simplicity;
-such optimization is reserved for future work.
-\subsubsection{Nested processes, \texttt{CLONE} flags}\label{clone_bench}
-\begin{figure}
-\centering
- \includegraphics[width=0.5\textwidth]{clone_bench}
- \caption{Time to create different processes from \texttt{rsyscall}, by variety of parent process and child process}
- \label{fig:clone_bench}
-\end{figure}
-As demonstrated in sections \ref{pidns} and \ref{shared_fd_table},
-\texttt{rsyscall} supports nested operation.
-That is, we can spawn a child from one of our children;
-this can impact performance.
-\texttt{clone} also supports specifying a variety of flags to create new namespaces when creating the child,
-which can also impact performance.
-
-We've benchmarked process creation using the same benchmark as in section \ref{subprocess_bench}
-from a variety of parent processes: the local process,
-a child (created with \texttt{clone()}),
-a grandchild (created with \texttt{clone().clone()}),
-or a child in a different namespace (created with \verb=clone(CLONE_NEWNS|CLONE_NEWPID)=).
-We've also specified different combinations of flags for the child process being created in the benchmark.
-We don't include the Python standard library in these benchmarks
-as there's no way to create nested child processes or specify \texttt{CLONE} flags in the Python standard library.
-The results are in figure \ref{fig:clone_bench}.
-
-Process creation in a child has a significant performance impact,
-primarily because of the overhead of performing syscalls in a child.
-Performance of process creation in a grandchild is the same as in a child,
-because there is no additional overhead for syscalls in a grandchild relative to a child.
-Benchmarking \texttt{getpid} in a grandchild
-confirms that the average time per \texttt{getpid} call is still 561 microseconds on our benchmark setup.
-
-The use of namespaces also has a significant impact,
-both when the parent process is in a namespace, and when the child process is created in a namespace.
-We perform the benchmark with no clone flags, and with the common clone flags
-\verb|CLONE_NEWNS| and \verb|CLONE_NEWPID|, first on their own and then together.
-This benchmark essentially measures the performance of the Linux namespaces system, not the work in this paper,
-but it is illustrative to compare it to the performance of \texttt{rsyscall}.
-
-Note that the cost of creating a new child process from a parent process that is already inside a namespace
-(the difference between the \texttt{clone()} and \texttt{clone(NS|PID)} parent processes)
-is around 2 milliseconds, comparable to the cost of running with \texttt{rsyscall}.
-Note also that the overhead of creating a new process inside a new namespace
-(the difference between the \texttt{clone()} and \texttt{clone(NS|PID)} child processes)
-is around 15 milliseconds; subtantially greater than the overhead of \texttt{rsyscall}.
-As namespaces are widely used,
-this suggests our overhead is acceptable.
-\subsection{Usage in the real world}\label{realworld}
-At \twosigma{} we have used \texttt{rsyscall} and direct-style process creation
-to implement an internal distributed system deployment library, written in Python,
-which we'll refer to here as Toplevel.
-Toplevel is in use in production,
-and is extensively used as part of testing and development.
-Toplevel consists of a collection of functions and modules
-for each component in our system,
-which start up components using direct-style process creation,
-all from a single Python parent process.
-
-One of the major benefits of direct-style process creation has been easy use of file descriptor inheritance.
-Our experience has been that with fork-style or spawn-style interfaces,
-file descriptor logic must ultimately be centralized,
-whereas file descriptors in a new process can be built up over time using direct-style,
-simplifying the implementation.
-We've found that even programmers without substantial prior experience with Unix programming
-are able to use file descriptor passing as a normal feature of their development process with relative ease,
-writing code to support new programs and passing down file descriptors to those programs without issue.
-
-Easy file descriptor passing has, in turn, allowed us to heavily use socket activation techniques
-to start up services in parallel,
-substantially speeding up system startup\cite{socketactivation}.
-We are able to bootstrap connections over our internal shared memory transports
-over pre-created, passed-down file descriptors,
-and our management interfaces listen on passed-down sockets;
-this allows starting up services in parallel, rather than in dependency order.
-Benchmarking a representative example built with an old library
-and its equivalent built with Toplevel,
-the old example took an average of 221 seconds to run,
-mostly spent in system startup,
-while the new example takes an average of 16 seconds.
-
-Easy file descriptor inheritance has also allowed us to treat file descriptors as a uniform interface in Toplevel.
-We pass around file descriptors freely inside Toplevel.
-This uniform interface allows us to flexibly swap implementations between in-process and out-of-process,
-allowing us to decide on a case-by-case basis
-whether we prefer the flexibility of an in-process Python implementation,
-or the performance of a native code implementation running in a child process.
-
-We've also made use of namespaces,
-chiefly user namespaces and pid namespaces.
-For legacy applications which directly start subprocesses,
-we prevent process leaking with a pid namespace.
-This is part of a strategy of noninterference and cleanup
-that ensures that multiple users of Toplevel
-can coexist simultaneously on the same host,
-even if running the same system.
-
-Toplevel also makes uses of other features of \texttt{rsyscall} which are not covered in this paper,
-and will be covered in future work.
-These include the ability to operate on remote hosts through capabilities for remote processes,
-and the ability to perform system calls in parallel across a pool of processes.
-These have significant synergy with direct-style process creation
-in forming an overall powerful system.
-
-All this has allowed us to run our systems in a much more dynamic way than before.
-We can freely start up arbitrary subsets of our systems on-demand,
-spread across one or more hosts,
-for development, testing, or production,
-regardless of the configuration of that host,
-without worrying about interference between instances,
-and without a dependency on any external privileged services,
-such as container runtimes.
-This has become a key part of our development workflow.
+
+We benchmark Python's \texttt{subprocess.run} against direct-style \texttt{clone} with \texttt{rsyscall}
+with varying memory usage, as in section \ref{microbench}.
+The results are in Figure \ref{fig:subprocess_bench}.
+TODO generate again to get the higher end where it crosses over.
 \section{Related work}\label{related_work}
+% TODO all this is just slapped onto the page without much thought, needs revising
 \subsection{Direct-style process creation}
-Baumann (2019) briefly suggests cross-process operations as a possible replacement for \texttt{fork} on Unix,
-positing many of the same advantages we find in practice\cite{forkroad}.
-Our work, though developed independently,
-is in many ways an elaboration on and implementation of that idea.
-
-As we discuss in section \ref{introduction},
+As we discuss in section \ref{background},
 many academic operating systems have direct-style process creation.
-We are not aware of any other instances of direct-style process creation in a Unix-like environment.
+We are not aware of any previous instances of direct-style process creation in a Unix-like environment.
 
 The closest related work known to us
-is efforts to build Unix compatibility environments
-on academic operating systems which natively use direct-style process creation;
-examples include Exokernel and Fuchsia\cite{exokernel}\cite{fuchsia_libc}.
-Such compatibility environments could be bypassed to perform direct-style process creation
+is efforts to build Unix compatibility environments, including fork-style process creation,
+on academic operating systems which natively use direct-style process creation.\cite{exokernel}
+Such compatibility environments could theoretically be bypassed to perform direct-style process creation
 using the underlying primitives,
@@
 in such an environment.
-\subsection{Remote system calls}\label{related}
-Many Unix-based systems have features described as ``remote system calls''.
-Most such systems do not allow a single program to manipulate multiple processes;
-rather, a program implicitly makes system calls in a single remote process
-which is distinct from the process the program's code runs in.
+\subsection{Remote system calls}
+There are many Unix-based systems which do something called "remote system calls".
+Most such systems do not allow a single program to manipulate multiple processes.
 Those systems that do allow manipulating multiple processes
@@
 
-% probably we should rework this whole section to say "X and Y, among other systems, do Z."
-HTCondor\cite{condor} and Popcorn\cite{popcorn}, among others,
-use system call forwarding to implement migration between hosts in a computing cluster.
-In these systems, processes can be live-migrated between hosts;
-when this occurs, the system will transparently forward IO-related system calls
+Many systems use system call forwarding to implement migration in a computing cluster.
+HTCondor\cite{condor} and Popcorn\cite{popcorn} are two examples.
+In both systems, processes can be live-migrated between hosts;
+when this occurs, the system will automatically forward IO-related system calls
 back to the original host.
 
-gvisor\cite{gvisor} and ViewOS\cite{viewos}, among others,
-use system call interception as a means of virtualization.
-In these systems,
-system calls made in one process are intercepted and forwarded to another process,
-which performs the specified system call and returns the results back to the original process.
-
-Many other systems have unusual system call invocation patterns,
-such as enclave systems like SCONE\cite{scone}\cite{eleos}
-or systems for exception-less system calls like FlexSC\cite{flexsc}.
-In these systems, for a variety of reasons,
-a system call in one process is not evaluated by directly entering the kernel,
-but instead is sent to some other process or thread to be evaluated,
-and the results received from that process or thread.
-
-A number of debugging or instrospection systems have the ability to perform or monitor system calls in other processes,
-typically using ptrace.
-Systems like gdb and CRIU
-retrieve information about the target process by forcing it to call various system calls to dump information.
-These systems typicaly use \texttt{ptrace},
-which can be used by a single ptracer process to operate on multiple target processes at once.
-Unfortunately, as discussed in section \ref{implementation},
-multiple ptracer processes cannot use \texttt{ptrace} on a single target process simultaneously,
-which means only one of these systems can be used on a process at a time.
-
-% maybe mention the notion of a formal interface definition for the linux syscall interface
-% that might go in future work even.
-% yeah it's not yet something developed.
-% a formal interface definition so that remote system calls can be employed without building a new system every time.
+Some systems intercept system calls and forward them to another process to be implemented
+as part of a virtualization system.
+gvisor\cite{gvisor} and ViewOS\cite{viewos} are two examples.
+In both systems,
+system calls made in one process are intercepted,
+and serviced by another process.
+
+Some systems forward system calls elsewhere for asynchronous execution.
+FlexSC\cite{flexsc} is one example.
+System calls in FlexSC are forwarded to a pool of dedicated system call execution threads.
+The recently added Linux feature \verb|io_uring| is, in some sense, another example;
+it supports sending a subset of system calls to the kernel,
+which executes them asynchronously,
+rather than blocking the thread.
+
+Some systems call system calls in other processes for debugging or introspection purposes.
+strace, gdb, and CRIU\cite{criu} are examples of this.
+These systems typically uses ptrace,
+which is capable of operating on multiple processes at once.
+As discussed in section \ref{implementation},
+a ptrace-based has approach has limitations which make it unsuitable for \texttt{rsyscall}.
 \subsection{Process capabilities}
-Many capability-oriented operating systems, such as KeyKOS\cite{keykos}, seL4\cite{sel4}, and others\cite{fuchsia},
-have process capabilities which allow one process to operate on another process.
-These systems are our main inspiration for this work.
-
-On several Unix systems,
-Capsicum\cite{capsicum} provides something called process descriptors,
+The Capsicum system provides something called process descriptors\cite{capsicum},
 which is a file descriptor handle for a process.
@@
 \subsection{Capability-secure libc replacements}
-Capsicum\cite{capsicum} provides a set of system calls
-which can be used to provide a capability-secure sandbox.
+The Capsicum system provides a set of system calls
+which can be used to provide a capability-secure sandbox.\cite{capsicum}
 Latter efforts\cite{oblivious}, including CloudABI\cite{cloudabi} and WASI\cite{wasi},
@@
 
-PLASH\cite{plash} provides a spawn-style API, in the form of a shell,
-to launch processes in a capability-secure environment.
+PLASH provides a spawn-style API, in the form of a shell,
+to launch processes in a capability-secure environment.\cite{plash}
 PLASH, like all spawn-style APIs, abstracts over the native Linux environment,
@@
 \subsection{Other applications of \texttt{rsyscall}}
+Most avenues of future work focus on \texttt{rsyscall}.
 \texttt{rsyscall} was not developed solely for the purpose of this paper,
 and it has other uses unrelated to direct-style process creation,
-such as asynchronous system calls, exceptionless system calls\cite{flexsc}, and cross-host operations, among others.
+such as asynchronous system calls, exceptionless system calls\cite{flexsc}, cross-host operations, among others.
 We are actively exploring such applications,
@@
 \texttt{rsyscall}'s cross-process syscalls can be performed entirely in userspace,
-which has substantial benefits for deployability and flexibility.
+which has substantial benefits for deployability.
 Nevertheless, direct support in the Linux kernel
 for creating a stub process and performing syscalls in the context of that process
-may provide efficiency benefits, as well as reducing overall complexity.
-
-Other aspects of our implementation would also be improved by new features in the Linux kernel;
-one examples is discussed in section \ref{execve}.
+may provide efficiency benefits, as well as reducing userspace-visible complexity.
+
+Several other aspects of our implementation would also be improved by new features in the Linux kernel,
+as discussed in sections \ref{execve} and \ref{cloexec}.
 \subsection{Portability to other Unix systems}
@@
 but others may wish to explore porting these techniques to other operating systems.
-% TODO maybe we could get rid of this paragraph?
 \subsection{Large scale open source usage}
-We would like to open source libraries built on top of direct-style process creation.
-From our experience using such libraries, discussed in section \ref{realworld},
-we believe this would make it easier to develop complex systems involving processes.
+We have made use of the techniques described in this paper
+in proprietary software at \twosigma.
+While this gives us personally greater confidence in these techniques,
+it would be beneficial to use them in a publicly available, open source system.
+Likely the best 
+Either porting an existing system to use these techniques,
+or using these techniques to create a substantial new system from scratch,
+would provide a substantial demonstration of the viability of these techniques.
 \section{Conclusions}\label{conclusions}
-We have introduced direct-style process creation on Linux for the first time,
-through \texttt{rsyscall}.
-This style of process creation is common on academic operating systems,
-but was previously not usable on Linux.
-We've provided a number of examples which demonstrate the usefulness
-of direct-style process creation in section \ref{examples}.
-Our current implementation of \texttt{rsyscall} has excellent support for Python,
-works entirely in userspace,
-and is immediately deployable on today's Linux systems.
-\texttt{rsyscall} is open source and available from \githuburl{}.
-
-We've found that direct-style process creation is more expressive than fork-style and spawn-style,
-and even with an unoptimized implementation, has comparable performance.
-We've used direct-style process creation successfully in production at \twosigma{}.
-Direct-style process creation is our technique of choice for new applications involving process management,
-including in container runtimes and distributed systems.
-Future work can make direct-style process creation more efficient
-and decrease the complexity of its implementation.
+Direct-style process creation is much less known and much less used than fork-style and spawn-style.
+We have implemented direct-style process creation for Linux.
+Our implementation is immediately deployable on today's Linux systems.
+We have discussed various applications of processes,
+and demonstrated the use of Linux direct-style process creation
+to implement them.
+
+We believe that direct-style process creation is superior to fork-style and spawn-style,
+including on Linux.
+Our technique is immediately deployable and can be used for a wide variety of complex applications.
 
 We hope that a better process creation mechanism
-will help encourage more creative use of processes and the features available in Linux.
-Though processes are a long-standing, widespread feature,
-we believe there is still much to be learned about how to use processes to their full potential.
+will help encourage more use of the process abstraction,
+which, though widespread,
+is still not used to its full potential.
+
 \bibliographystyle{plain}

### 2021-04-07 research/paper/paper.tex
paper

@@
 \usepackage{tikz}
+\usepackage{dblfloatfix}
 \usepackage{amsmath}
+\usepackage{amssymb}
+\usepackage{pifont}
+\usepackage{comment}
+
+\newcommand{\cmark}{\ding{51}}
+\newcommand{\xmark}{\ding{55}}
 
@@
 Fork-style, spawn-style, and direct-style.
+\begin{table*}
+\begin{tabular}{l|l|l|l}
+% BEGIN RECEIVE ORGTBL compare
+ & fork & spawn & direct-style\\
+\hline
+Requirements on parent process & \tbad Single-threaded, small memory & \tgood None & \tgood None\\
+Programming model & \tbad Complex (returns twice) & \tgood Simple (single call) & \tgood Simple (imperative)\\
+Maximally powerful & \tgood Yes, can call any syscall & \tbad No, limited interface & \tgood Yes, can call any syscall\\
+Reporting of results & \tbad Requires IPC & \tbad Not fine-grained & \tgood From individual child syscalls\\
+Non-inherited attributes & Mutated by code in child & Set by arguments & Mutated by code in parent\\
+% END RECEIVE ORGTBL compare
+\end{tabular}
+\caption{Features of fork-style vs spawn-style vs direct-style}
+\label{tab:styles}
+\end{table*}
+\begin{comment}
+#+ORGTBL: SEND compare orgtbl-to-latex :splice t
+|                                | fork                                | spawn                       | direct-style                          |
+|--------------------------------+-------------------------------------+-----------------------------+---------------------------------------|
+| Requirements on parent process | \tbad Single-threaded, small memory | \tgood None                 | \tgood None                           |
+| Programming model              | \tbad Complex (returns twice)       | \tgood Simple (single call) | \tgood Simple (imperative)            |
+| Maximally powerful             | \tgood Yes, can call any syscall    | \tbad No, limited interface | \tgood Yes, can call any syscall      |
+| Reporting of results           | \tbad Requires IPC                  | \tbad Not fine-grained      | \tgood From individual child syscalls |
+| Non-inherited attributes       | Mutated by code in child            | Set by arguments            | Mutated by code in parent             |
+% $
+\end{comment}
+Table \ref{tab:styles} summarizes the differences between each style.
 
@@
 
-Our contribution is an implementation of direct-style process creation for Linux,
-making use of our work on the \texttt{rsyscall} project.
-The \texttt{rsyscall} project develops language-specific ``process-independent'' system call libraries for Linux,
-which replace libc system call wrappers, and allow a program to perform cross-process system calls.
-rsyscall currently has best support for Python,
+We'll discuss each style in more detail in section \ref{background}.
+
+To perform direct-style process creation on Linux,
+we need an API where we explicitly specify in which process we want to make a syscall.
+This differs from the normal mode of operation on Linux,
+where programs making syscalls implicitly operate on the current process.
+
+Our contribution in this paper is this API, through the \texttt{rsyscall} library.
+The \texttt{rsyscall} project develops
+language-specific, object-capability-model, low-abstraction libraries for the Linux syscall interface,
+bypassing libc.
+\texttt{rsyscall} is currently focused on supporting Python,
 but can be ported to other languages;
@@
 but generalize easily to other languages.
-
-Our direct-style process creation implementation uses \texttt{clone} to create a child process in an embryonic state
-which can then be manipulated from user code using syscalls with \texttt{rsyscall}.
-A direct-style \texttt{clone} has all the advantages of direct-style process creation
-over Linux's fork-style \texttt{fork} and spawn-style \texttt{posix\_spawn}:
-\begin{compactitem}
-\item
-  Processes can be created correctly and performantly from any parent process,
-  with a straightforward, familiar API.
-\item
-  Processes can be created with the full set of Linux features.
-\item
-  Syscall results are reported individually to the parent process,
-  rather than batched together as with spawn-style or manually reported via IPC as with fork-style.
-\end{compactitem}
-
-We've implemented a variety of useful applications using direct-style process creation on Linux,
-which we'll demonstrate in this paper.
-Direct-style \texttt{clone} has minimal performance cost in normal situations,
-and can outperform \texttt{fork} in some cases.
-We have deployed direct-style process creation in production internally at \twosigma.
-
-Our implementation is distributed as a feature of \texttt{rsyscall},
-and is immediately deployable on recent Linux kernels.
 \texttt{rsyscall} is open source and available from
 [Github url].
-% \url{https://github.com/catern/rsyscall}.
-
-This paper is organized in several sections.
-\begin{compactitem}
-\item In section \ref{background}, we provide background on fork-style, spawn-style, and direct-style process creation,
-and consider their advantages and disadvantages in more depth.
-\item In section \ref{overview}, we discuss the design of the API,
-including our usage of \texttt{rsyscall}.
-\item In section \ref{applications}, we show examples of usage of direct-style process creation in Python on Linux,
+
+Unlike POSIX C libraries such as glibc or musl,
+an \texttt{rsyscall} library is organized based on the object-capability model.
+The capabilities to make syscalls in any given process are reified as language objects.
+If the capability to make syscalls in a specific process is not passed (in some way) to a function,
+then that function cannot make syscalls in that process.
+
+Due to this design, an \texttt{rsyscall} program can make use of multiple processes at once,
+by manipulating capabilities for multiple processes.
+The relevant two capabilities for this paper are the initial capability to the ``local'' process,
+and capabilities to child processes.
+The ``local'' process is the one which hosts the runtime for the running program,
+and in which a legacy libc would implicitly make syscalls.
+Every \texttt{rsyscall} program starts with the ``local'' process
+and uses it to bootstrap the initial capabilities for other processes.
+
+To perform direct-style process creation,
+we use \texttt{rsyscall} to call \texttt{clone} on the ``local'' process to create a new child process.
+We call various syscalls in the child process to mutate it until it reaches the desired state,
+at which point we can call \texttt{exec} in the child process to start it running on its own.
+After that, the process is no longer under our control, and it functions like any other child process.
+The child process can be monitored using normal Linux child monitoring syscalls,
+such as \texttt{waitid}.
+
+These child process capabilities are created and managed in userspace,
+by launching new processes running a syscall server,
+which receives syscall requests, performs the syscall, and returns the response.
+No kernel modifications are required, and \texttt{rsyscall} is immediately deployable on recent Linux kernel.
+We discuss the implementation in depth in section \ref{implementation},
+and examine several difficult process-creation related details.
+
+\lstinputlisting[float=h,language=Python,label={basic},caption={Create new process, change CWD, and exec}]
+{examples/basic/direct.py}
+Listing \ref{basic} shows a program using \texttt{rsyscall}.
+We create a new child process using \texttt{clone} on the local process \verb|local|,
+and receive back a capability to control it.
+We call \texttt{chdir} in the child process to change its working directory to \texttt{/dev}.
+Then we call \texttt{execv} in the child,
+passing the path of the \texttt{cat} executable
+and the arguments to the new executable.
+The name of the executable is passed as the first argument,
+as is typical in Unix.
+Calling \texttt{execv} consumes the capability,
+and later use of this child process capability will fail with an exception.
+
+The behavior of system calls made on a remote process through \texttt{rsyscall}
+is identical to the behavior of system calls made implicitly on the current process through libc.
+This is an API that systems programmers are already familiar with,
+and which is already thoroughly documented through Linux system call manpages.
+The novelty is constrained to the newly explicit process capabilities.
+Besides flattening the learning curve,
+this also ensures that direct-style \texttt{clone} will always be at least as powerful as \texttt{fork}.
+
+We show more examples of direct-style process creation in section \ref{examples},
 including use of process features which are difficult to use on Linux with \texttt{fork} or \texttt{posix\_spawn}.
-\item In section \ref{implementation}, we explore our implementation in depth,
-and discuss several difficult process-creation related details of \texttt{rsyscall}'s implementation.
-\item In section \ref{evaluation}, we evaluate, somehow (TODO).
-\item In section \ref{related_work},
-we examine other work which also touches on the topics covered by this paper.
-\item In section \ref{future_work},
-we list several directions for future work related to direct-style process creation and to \texttt{rsyscall}.
-\item In section \ref{conclusions}, we briefly outline our conclusions about direct-style and the future of process creation.
-\end{compactitem}
-\section{Background}\label{background}
+
+Direct-style \texttt{clone} has acceptable performance cost,
+and can outperform \texttt{fork} in some cases.
+The performance overhead of creating processes with direct-style clone
+is comparable to the performance cost of creating processes while inside a Linux namespace.
+At \twosigma, we have deployed direct-style process creation in production
+inside a library for distributed system deployment, Toplevel.
+We evaluate in further detail in section \ref{evaluation}.
+\section{Background on process-creation}\label{background}
 As mentioned in the introduction,
@@
 A few operating systems have a third style, which we refer to as "direct-style".
+
 \subsection{Fork-style}
@@
 This is the kind of direct-style interface that we contribute in this paper.
-\section{Overview}\label{overview}
-In most operating system APIs, including the ones available on Linux,
-when a program makes a syscall, it implicitly operates on the current process.
-To perform direct-style process creation on Linux,
-we need an API where we instead explicitly specify in which process we want to make a syscall.
-
-\texttt{rsyscall} provides this API.
-The \texttt{rsyscall} project develops
-language-specific, object-capability-model, low-abstraction libraries for the Linux syscall interface,
-bypassing libc.
-
-Unlike POSIX C libraries such as glibc or musl,
-an \texttt{rsyscall} library is organized based on the object-capability model.
-The capabilities to make syscalls in any given process are reified as language objects.
-If the capability to make syscalls in a specific process is not passed (in some way) to a function,
-then that function cannot make syscalls in that process.
-
-Due to this design, an \texttt{rsyscall} program can make use of multiple processes at once,
-by manipulating capabilities for multiple processes.
-The relevant two capabilities for this paper are the initial capability to the ``local'' process,
-and capabilities to child processes.
-The ``local'' process is the one which hosts the runtime for the running program,
-and in which a legacy libc would implicitly make syscalls.
-Every \texttt{rsyscall} program starts with the ``local'' process
-and uses it to bootstrap the initial capabilities for other processes.
-
-To perform direct-style process creation,
-we use \texttt{rsyscall} to call \texttt{clone} on the ``local'' process to create a new child process.
-We call various syscalls in the child process to mutate it until it reaches the desired state,
-at which point we can call \texttt{execve} in the child process to start it running on its own.
-After that, the process is no longer under our control, and it functions like any other child process.
-The child process can be monitored using normal Linux child monitoring syscalls,
-such as \texttt{waitid}.
-
-\lstinputlisting[float=*ht,language=Python,label={basic},caption={Creating a new process, changing CWD, and execing}]
-{examples/basic/direct.py}
-Listing \ref{basic} shows this in action in Python.
-We create a new child process using \texttt{clone} on the local process \verb|local|,
-and receive back a capability to control it.
-We call \texttt{chdir} in the child process to change its working directory to \texttt{/dev}.
-Then we call \texttt{execve},
-passing the path of the \texttt{cat} executable,
-the arguments to the new executable,
-and the unchanged set of environment variables.
-The name of the executable is passed as the first argument,
-as is typical in Unix.
-Calling \texttt{execve} consumes the capability,
-and later use of this child process capability will fail with an exception.
-
-The behavior of system calls made on a remote process through \texttt{rsyscall}
-is identical to the behavior of system calls made implicitly on the current process through libc.
-This is an API that systems programmers are already familiar with,
-and which is already thoroughly documented through Linux system call manpages.
-The novelty is constrained to the newly explicit process capabilities.
-
-The lack of novelty in the API is important not only for the learning curve,
-but also for power.
-Since we are using conventional Linux system calls,
-just in a cross-process way,
-direct-style \texttt{clone} will always be at least as powerful as \texttt{fork}.
-\section{Applications}\label{applications}
+\section{Examples}\label{examples}
 In this section,
-we'll discuss several more advanced uses of processes
-using more sophisticated Linux features,
-and accompany them with Python examples using direct-style process creation.
-All of these examples use only conventional Linux system calls,
-and can therefore translated straightforwardly into any language with an \texttt{rsyscall} API.
-For concision, we assume that we are running with sufficient permissions in these examples,
-but these examples can all be run without privileges with appropriate use of \texttt{CLONE.NEWUSER}.
-
-Implementing these examples with \texttt{fork} or \verb|posix_spawn| requires substantially more boilerplate,
+we'll demonstrate direct-style process creation
+by demonstrating several advanced uses of processes
+using more sophisticated Linux features.
+For concision, we assume in these examples that we are running with sufficient permissions in these examples,
+but these examples can all be run without privileges with appropriate use of \verb|CLONE_NEWUSER|.
+
+These examples are not novel, as such, as they use only conventional Linux system calls;
+but implementing them with \texttt{fork} or \verb|posix_spawn| requires substantially more code,
 or even a complete change in implementation approach,
-such as in section \ref{shared_fd_table}.
+as described in section \ref{shared_fd_table}.
 We'll compare our direct-style implementations in this section
@@
 
+% TODO need to clarify this more, this is v2, Larry didn't like v1 beyond first sentence
 Most resources in Linux are managed through file descriptors,
-so this feature allows the parent process to pass a wide variety of resources to the child process.
-An executable program can be provided a file descriptor,
-which can correspond to any file in any filesystem,
-or to a network connection\cite{ucspi},
-or a pipe connected to another process,
-or a variety of other resources,
-without the process being aware of what specific resource it is accessing.
-In this way, we immediately know "for free"\cite{theoremsforfree}
-that the behavior of the program does not depend on the name of the file,
-or the hostname to it which the socket is connected,
-or the executable running in the process on the other end of the pipe,
-or other abstracted-away details.
+so this feature allows the parent process to pass a variety of resources to the child process,
+such as files, network connections, pipes, or other resources.\cite{ucspi}
+Since the resource is created outside the child process and passed down as an opqaue file descriptor,
+the precise details and type of the resource are not available to the child.
+This is a form of abstraction,
+and we can know ``for free''\cite{theoremsforfree}
+that the program running in the child process
+will not rely on the details of how the file descriptor was created,
+such as the filename that was opened or the hostname that was connected to.
 
@@
 file descriptor passing over a Unix domain socket
-is our only option to do something like this.
+is our only option.
 \section{Implementation}\label{implementation}
-\subsection{Basics about \texttt{rsyscall}}
 In this section, we'll give a brief overview of the implementation of \texttt{rsyscall},
@@
 The most simple way to access memory is for the local process to be in the same address space as the target process.
-This is the case most of the time; we pass \texttt{CLONE.VM} to \texttt{clone} by default in the Python wrapper.
-However, the target process may be in another address.
-For example, if the target process is at a different privilege level,
-it must be in a different address space,
-as sharing an address space with a process at a different privilege level is unsafe.
-
-We implement access to the memory of processes in different address spaces through another set of pipes,
-by copying memory into and out of those pipes using \texttt{read} and \texttt{write} system calls in the target process.
-A \texttt{read} syscall corresponds to a write to memory,
-and a \texttt{write} syscall corresponds to a read from memory.
-
-The \verb|process_vm_readv| and \verb|process_vm_writev| system calls
-allow the caller to read and write memory from the virtual address space of other processes.
-However, they require that the caller have specific credentials relative to the process being accessed,
-which may not always be the case.
-Additionally, these system calls are disabled if ptrace is disabled system-wide,
-which is a niche but possible system configuration.
-To ensure that \texttt{rsyscall} can be used for arbitrary purposes and on arbitrary systems, we avoided these calls.
+This is the case most of the time; we pass \verb|CLONE_VM| to \texttt{clone} by default in the Python wrapper.
+
+Sometimes, the target process may be in another address space;
+for example, if the target process is at a different privilege level,
+it must be in a different address space.
+There are a number of available techniques in that scenario;
+we choose to copy memory over a pair of pipes,
+again using file descriptors for the sake of easy event-loop integration.
 \subsection{\texttt{clone}}\label{clone}
@@
 
-There are three Linux system calls which create processes:
-\texttt{fork}, \texttt{vfork} and \texttt{clone}.
+Besides \texttt{clone}, \texttt{vfork} and \texttt{fork} also create processes,
+but they are not suitable for flexible direct-style process creation.
 \texttt{vfork} suspends execution of the parent process
 while waiting for the child process to exit or call \texttt{execve},
-which is immediately unsuitable for direct-style process creation.
-\texttt{clone} provides a superset of the functionality of \texttt{fork},
-so we focused our attention on \texttt{clone}.
-
-\texttt{clone}
-(along with \texttt{fork}) creates a new process
+which is immediately unsuitable.
+\texttt{fork} lacks many features which are restricted to \texttt{clone}.
+For example, with \texttt{fork},
+we could not create child processes which share the parent's address space,
+which would complicate memory access.
+
+The raw \texttt{clone} system call creates a new process
 which immediately starts executing at the next instruction after the syscall instruction,
@@
 
-Common Linux libcs provide a wrapper for \texttt{clone}
-which allow executing an arbitrary function with an arbitrary argument,
-but since we want to support remote execution,
-we can't use the libc wrapper.
-
-Instead, since \texttt{clone} lets us change the stack for the new process,
-we use this to return, after the system call to an arbitrary function address stored on the stack.
+\texttt{clone} lets us change the stack for the new process.
+We can use this to make the new process call an arbitrary function,
+by storing its address on the new stack.
 Further arguments can also be passed on the stack,
@@
 
-We use this to create processes running the syscall request-response loop described in the previous section.
+We use this to create processes running the syscall server.
 After this,
-most system calls can be called as normal.
-The new process can be modified freely through \texttt{chdir}, \texttt{dup2}, and other system calls.
-Out of system calls related to process creation,
+the parent process can be begin to call system calls in the child process.
+Most system calls work as normal;
+the new child process can be modified freely through \texttt{chdir}, \texttt{dup2}, and other system calls.
+From the system calls related to process creation,
 only \texttt{execve} and \texttt{unshare} need substantial further attention.
@@
 because when it is successful, it does not return.
-Therefore we need a way to determine if \texttt{execve} is successful;
-naively waiting for a response to the syscall request will leave us waiting forever.
-
-There is a traditional technique to detect a successful \texttt{execve} using a pipe.
-We create a pipe before forking,
-ensure both ends are marked \verb|O_CLOEXEC|,
-perform the fork,
-call \texttt{execve} in the child,
-close the write end of the pipe,
-and wait for EOF on the read end.
-If the child process has neither successfully called \texttt{execve}, nor exited for some other reason,
-then the write end of the pipe will still be open in the child process's fd table,
-and the read end of the pipe will not return EOF.
-But once the child process calls \texttt{execve} successfully,
-\verb|O_CLOEXEC| will cause the write end of the pipe to be closed,
-and the read end of the pipe will return EOF.
-
-This trick works well with \texttt{fork};
+We need a way to determine if \texttt{execve} is successful;
+naively waiting for a response to the syscall request may leave us waiting forever.
+
+There is a traditional technique used with \texttt{fork}
+to detect a successful \texttt{execve} using a pipe,
+which unfortunately won't work for us.
+In this technique, the parent process creates a pipe before forking,
+ensures both ends are marked \verb|O_CLOEXEC|,
+performs the fork,
+closes the write end of the pipe,
+and \texttt{read}s the read end of the pipe.
+The child process either successfully calls \texttt{execve} or exits;
+either way, the last copy of the write end of the pipe will be closed,
+which causes the \texttt{read} in the parent process to return EOF.
+The parent can then check that the child process hasn't exited;
+if the child hasn't exited, then it must have successfully called exec.
+
+This trick works with \texttt{fork},
 but it's not general enough to work with \texttt{clone}.
@@
 and in this way get notified of the child process calling \texttt{execve} or exiting.
-
-Unfortunately, futexes in current Linux integrate poorly:
-There is no way for a single process to wait for more than one futex at a time,
-and no way to monitor a futex with file-descriptor-monitoring syscalls such as \texttt{poll}.
-The best we can do is create a dedicated child process for each futex we want to wait on,
+A process can only wait on one futex at a time;
+to monitor multiple futexes from a single event loop,
+we need to create a dedicated child process for each futex we want to wait on,
 and have this child process exit when the futex has a wakeup.
-Monitoring child processes can be straightforwardly integrated into an event loop.
-
-While slightly complex to implement, this solution works well.
-We provide \texttt{ctid} whenever we call \texttt{clone},
+We can then monitor these child processes from an event loop using standard techniques.\cite{signalfd}\cite{pidfd}
+
+So, we pass \texttt{ctid} whenever we call \texttt{clone},
 and set up a process to wait on that futex.
@@
 and erroneously inherit the file descriptor.
-\section{Evaluation \textbf{TODO}}\label{evaluation}
-
-We quantitatively answer the following questions:
+\section{Evaluation}\label{evaluation}
+Our goal is a powerful approach to process creation that is easy to use correctly.
+Performance of our implementation is secondary,
+but the overhead of our implementation relative to alternative approaches
+must be low.
+
+We evaluate the degree to which we meet this goal by answering the following questions:
 \begin{compactitem}
@@
   Does direct-style process creation on Linux allow for simpler programs than the alternatives?
-  We implement and compare similar functionality using both direct-style and fork-style in \ref{ease}.
+  We compare implementations of similar functionality using both direct-style and fork-style in \ref{ease},
+  and find substantial benefits for direct-style.
 \item
-  Does a clone-based implementation avoid the well-known performance overheads of fork?
-  We evaluate clone against fork on several microbenchmarks in \ref{microbench}.
+  In Python, does our \texttt{rsyscall}-based direct-style process creation interface
+  have acceptable performance relative to the standard library's spawn-style \texttt{subprocess.run}?
+  We benchmark direct-style process creation against spawn-style in \ref{subprocess_bench},
+  and find an acceptable level of overhead.
 \item
-  Does a high-level Python direct-style process creation interface
-  have acceptable performance relative to the Python standard library's spawn-style interface?
-  We benchmark direct-style process creation against spawn-style in \ref{subprocess_bench}.
-% TODO have an item about use at TS? hm.
-% could be qualitative
+  Does direct-style process creation perform well in the ``real world''?
+  We discuss our positive experience with using direct-style process creation at \twosigma in \ref{realworld}.
 \end{compactitem}
@@
 \end{table}
-We implement a variety of simple but powerful techniques with processes
-using both direct-style and fork.
-Our results are summarized in table \ref{tab:ease}.
-Direct-style consistently takes under half the lines the express the same thing.
-\lstinputlisting[float=*ht,language=Python,label={basic_fork},caption={Fork-style: Creating a new process, changing CWD, and execing}]
+Does direct-style process creation on Linux allow for simpler programs than the alternatives?
+To answer this, we compare the direct-style examples shown in section \ref{examples}
+to programs with the same behavior implemented in fork-style.
+
+Most of the examples cannot be implemented with typical Linux spawn-style interfaces such as \verb|posix_spawn|,
+so we compare only to fork-style.
+
+The line counts of the direct-style and fork-style implementations are listed in table \ref{tab:ease}.
+Direct-style consistently takes under half the lines to express the same functionality.
+
+\lstinputlisting[float=ht,language=Python,label={basic_fork},caption={Fork-style: Creating a new process, changing CWD, and execing}]
 {examples/basic/fork.py}
-An example of one of the fork-style implementations is in listing \ref{basic_fork}.
-It corresponds to the direct-style example in listing \ref{basic}.
+Listing \ref{basic_fork} gives an example of one of our fork-style implementations,
+corresponding to the direct-style example in listing \ref{basic}.
 
@@
 Despite this, fork-style implementations are still significantly more verbose than direct-style.
-\subsection{Clone versus fork}\label{microbench}
-\begin{figure}[h!]
-\centering
- \includegraphics[width=0.5\textwidth]{microbench}
- \caption{fork and clone performance under varying memory usage}
- \label{fig:microbench}
-\end{figure}
-
-We run a benchmark which times how long process creation takes
-with \texttt{fork}, \texttt{clone}, and \verb|clone(CLONE_VM)|
-with varying amounts of anonymous memory mapped in the parent process
-before running the benchmark.
-\texttt{clone} without \verb|CLONE_VM| functions identically to fork.
-The results are in figure \ref{fig:microbench}.
-Without \verb|CLONE_VM|,
-process creation time is linear in the amount of memory mapped in the parent process.
-At the upper end of our benchmark, with 4 GB of memory mapped ($10^6$ pages),
-\texttt{fork} takes an average of 42.6 milliseconds to create a process,
-while \verb|clone(CLONE_VM)| stays flat at 21.1 microseconds.
-
-This is because \texttt{fork}, and \texttt{clone} without \verb|CLONE_VM|,
-must set up memory mappings in the child process for all the memory mapped in the parent process.
-\verb|clone(CLONE_VM)| does not have to do this bookkeeping,
-so its process creation time is constant in the amount of memory used in the parent process.
-
-\subsection{High-level direct-style versus spawn-style}\label{subprocess_bench}
+
+The main source of additional code in the fork-style implementation
+is reporting errors back to the main program using IPC.
+If a program that is already reporting errors to some other destination,
+such code is not necessary;
+for example, a process supervisor or a shell might write process startup errors
+as textual logs which are read by a human.
+If a program does not already have such a destination for errors,
+then robustly dealing with errors in \texttt{fork} requires this additional IPC code.
+
+Direct-style does not require any extra work to report errors;
+errors are reported just like any other system call.
+This is the main immediately-visible simplification of using direct-style process creation.
+
+Direct-style also removes the need to be concerned about the state of the calling process,
+though this simplification is not immediately visible in a small example.
+The fork-style examples may break if there are multiple threads in the calling process,
+and may be slow if the calling process has large amounts of memory mapped, as we'll see in section \ref{subprocess_bench}.
+The direct-style examples do not have these limitations.
+\subsection{Microbenchmark results}\label{subprocess_bench}
 \begin{figure}[h!]
@@
  \includegraphics[width=0.5\textwidth]{subprocess_bench}
- \caption{Python subprocess and direct-style clone performance under varying memory usage}
+ \caption{Python spawn-style vs direct-style performance under varying memory usage}
  \label{fig:subprocess_bench}
 \end{figure}
-
-We benchmark Python's \texttt{subprocess.run} against direct-style \texttt{clone} with \texttt{rsyscall}
-with varying memory usage, as in section \ref{microbench}.
+In Python, does our \texttt{rsyscall}-based direct-style process creation interface
+have acceptable performance relative to the standard library's spawn-style \texttt{subprocess.run}?
+
+We first evaluate a simple process creation workload:
+create a child process, exec the \texttt{true} binary, and wait for it to exit.
+We implement this using both direct-style \texttt{clone}
+and the standard library's \texttt{subprocess.run}.
+We benchmark both implementations under CPython 3.7.7, on Linux 4.19.93,
+pinned to an isolated single core on an Intel i9-9900K CPU with 60GB of RAM.
+We vary the amount of anonymous memory mapped in the parent process
+to demonstrate how each implementation scales with memory usage.
 The results are in Figure \ref{fig:subprocess_bench}.
-TODO generate again to get the higher end where it crosses over.
+
+The direct-style implementation stays around 2.2 milliseconds regardless of memory usage.
+\texttt{subprocess.run}, which is implemented with \texttt{fork},
+starts at 1.4 milliseconds,
+and then scales linearly with the amount of memory mapped, as expected,
+up to 48.5 milliseconds at 4 GBs of mapped memory.
+The cross-over point for this micro-benchmark is about 40 megabytes of memory mapped.
+
+We perform another microbenchmark to evaluate the overhead of modifying the child process.
+We simply call \texttt{getpid} from Python
+in both the child process and the parent process on the same benchmark setup.
+The average time per \texttt{getpid} call is
+3 microseconds when called in the parent process without going through \texttt{rsyscall},
+and 561 miroseconds when called through \texttt{rsyscall} in the child process.
+The difference, 558 microseconds,
+is the amount of overhead incurred
+by performing our child process setup system calls through \texttt{rsyscall}.
+Through profiling, we've found that most of this per-syscall overhead is spent in Python,
+rather than waiting for syscall responses from the child task.
+
+These slowdowns in process creation and modification are substantial,
+but we found that this overhead is acceptable in practice.
+Process creation in Python is already slow, taking milliseconds of time,
+so it is not expected to be on the fast path.
+In that context, \texttt{rsyscall} has a reasonable cost compared to \texttt{subprocess.run},
+and avoids the bad scaling of \texttt{fork} which might be unexpected by the naive programmer.
+
+Furthermore, as we'll discuss in section \ref{realworld},
+we've found that the greater expressivity of direct-style
+provides for programs that are more efficient on a large scale,
+which makes up for the performance cost in micro-benchmarks.
+We've also found that, for many interesting applications,
+the performance overhead of direct-style process creation (or Python, for that matter)
+is dwarfed by the execution time of the native-code programs we ultimately run.
+
+As a result, though implementations in native code and in the kernel would likely remove most of this overhead,
+we have chosen to not invest effort into optimizing process creation at this micro-level,
+to preserve implementation simplicity;
+such optimization is reserved for future work.
+
+\subsection{Microbenchmarks for nested processes}\label{clone_bench}
+\begin{figure}[h!]
+\centering
+ \includegraphics[width=0.5\textwidth]{clone_bench}
+ \caption{Time to create different processes, by variety of parent process}
+ \label{fig:clone_bench}
+\end{figure}
+
+As demonstrated in sections \ref{pidns} and \ref{fuse},
+\texttt{rsyscall} supports nested operation.
+That is, we can spawn a child from one of our children;
+this can impact performance.
+\texttt{clone} also supports specifying a variety of flags to create new namespaces when creating the child,
+which can also impact performance.
+
+We've benchmarked process creation using the same benchmark as in section \ref{subprocess_bench}
+from a variety of parent processes: the local process,
+a child (created with \texttt{clone()}),
+a grandchild (created with \texttt{clone().clone()}),
+or a child in a different namespace (created with \verb=clone(CLONE_NEWNS|CLONE_NEWPID)=).
+We've also specified different combinations of flags for the process being created in the benchmark.
+The results are in figure \ref{fig:clone_bench}.
+We don't include the Python standard library in these benchmarks
+as there's no way to specify \texttt{CLONE} flags in the Python stdlib.
+
+Process creation in a child has a significant performance impact,
+primarily because of the overhead of performing syscalls in a child.
+Performance of process creation in a grandchild is the same as in a child,
+because there is no additional overhead for syscalls in a grandchild relative to a child.
+Benchmarking \texttt{getpid} in a grandchild
+confirms that the average time per \texttt{getpid} call is still 561 microseconds on our benchmark setup.
+
+The use of namespaces also has a significant impact,
+both when the parent process is in a namespace, and when the child process is created in a namespace.
+We perform the benchmark with no clone flags, and with the common clone flags
+\verb|CLONE_NEWNS| and \verb|CLONE_NEWPID|, first on their own and then together.
+This benchmark essentially measures the performance of the Linux namespaces system, not the work in this paper,
+but it is illustrative to compare it to the performance of \texttt{rsyscall}.
+Note that the cost of creating a new child process from a parent process that is already inside a namespace
+(the difference between the \texttt{clone()} and \texttt{clone(NS|PID)} parent processes)
+is around 2 milliseconds, comparable to the cost of running with \texttt{rsyscall}.
+Note also that the overhead of creating a new process inside a new namespace
+(the difference between the \texttt{clone()} and \texttt{clone(NS|PID)} child processes)
+is around 15 milliseconds; subtantially greater than the overhead of \texttt{rsyscall}.
+\subsection{Usage in the real world}\label{realworld}
+At \twosigma we have used \texttt{rsyscall} and direct-style process creation
+to implement an internal distributed system deployment library, written in Python,
+which we'll refer to here as Toplevel.
+Toplevel is in use in production,
+and is extensively used as part of testing and development.
+Toplevel consists of a collection of functions and modules
+for each component in our system,
+which start up components using direct-style process creation,
+all from a single Python parent process.
+
+One of the major benefits of direct-style process creation has been easy use of file descriptor inheritance.
+Our experience has been that with fork-style or spawn-style interfaces,
+all file descriptor logic must be centralized,
+whereas file descriptors in a new process can be built up over time using direct-style,
+simplifying the implementation.
+We've found that even programmers without substantial prior experience with Unix programming
+are able to use file descriptor passing as a normal feature of their development process with relative ease,
+writing code to support new programs and passing down file descriptors to those programs without issue.
+
+Easy file descriptor passing has, in turn, allowed us to heavily use socket activation techniques
+to start up services in parallel,
+substantially speeding up system startup.\cite{socketactivation}
+We are able to bootstrap connections over our internal shared memory transports
+over pre-created, passed-down file descriptors,
+and our management interfaces listen on passed-down sockets;
+this allows starting up services in parallel, rather than in dependency order.
+Benchmarking one test built with an old library and its equivalent built with Toplevel,
+the old test took an average of 221 seconds to run,
+mostly spent in system startup,
+while the new test takes an average of 16 seconds.
+
+Easy file descriptor inheritance has also allowed us to treat file descriptors as a uniform interface in Toplevel.
+We pass around file descriptors freely inside Toplevel.
+A given component can implemented with a separate process which has the file descriptor passed down,
+running a program written in native code for performance,
+or implemented in-process with a Python implementation for flexibility;
+either is possible behind a uniform file descriptor interface.
+
+We've also been able to create complicated pipelines in Toplevel,
+as in section \ref{pipe}.
+
+We've also made use of namespaces,
+chiefly user namespaces and pid namespaces.
+For legacy applications which directly start subprocesses,
+we prevent process leaking with a pid namespace.
+This is part of strategy of noninterference and cleanup
+that ensures that multiple users of Toplevel
+can coexist simultaneously on the same host,
+even if running the same system.
+
+Toplevel also makes uses of other features of \texttt{rsyscall} which are not covered in this paper.
+These include the ability to operate on remote hosts through capabilities for remote processes,
+and the ability to perform system calls in parallel across a pool of processes.
+These have significant synergy with direct-style process creation
+in forming an overall powerful system.
+
+All this has allowed us to run our systems in a much more dynamic way than before.
+We can freely start up arbitrary subsets of our systems on-demand,
+spread across one or more hosts,
+for development, testing, or production,
+regardless of the configuration of that host,
+without worrying about interference between instances,
+and without a dependency on any external privileged services,
+such as container runtimes.
+This has become a key part of our development workflow.
 \section{Related work}\label{related_work}
-% TODO all this is just slapped onto the page without much thought, needs revising
 \subsection{Direct-style process creation}
+Baumann 2019\cite{forkroad} briefly suggests cross-process operations as a possible replacement for fork on Unix,
+positing many of the same advantages we find in practice.
+Our work, though developed independently,
+is in many ways an elaboration on and implementation of that idea,
+to which we owe a great debt.
+
 As we discuss in section \ref{background},
@@
 but others may wish to explore porting these techniques to other operating systems.
+% TODO maybe we could get rid of this paragraph?
 \subsection{Large scale open source usage}
-We have made use of the techniques described in this paper
-in proprietary software at \twosigma.
-While this gives us personally greater confidence in these techniques,
-it would be beneficial to use them in a publicly available, open source system.
-Likely the best 
-Either porting an existing system to use these techniques,
-or using these techniques to create a substantial new system from scratch,
-would provide a substantial demonstration of the viability of these techniques.
+We would like to open source libraries built on top of direct-style process creation.
+From our experience using such libraries at \twosigma,
+we believe this would greatly ease the development of complex systems involving processes,
+and would provide a substantial demonstration of the power of direct-style process creation.
 \section{Conclusions}\label{conclusions}

### 2021-04-07 python/subprocess_bench.py
subprocess_bench: thing

@@ -144,28 +144,16 @@ async def main() -> None:
         return [(i, await run_many(mode, 10**i)) for i in range(3)]
     subprocess = [(0, 1662.7047061920166), (1, 1666.396141052246), (2, 1668.2496070861816), (3, 1670.4914569854736), (4, 2006.0789585113525), (5, 10209.86270904541), (6, 48588.73701095581)]
     rsyscall = [(0, 2229.5007705688477), (1, 2258.697509765625), (2, 2234.3757152557373), (3, 2238.978862762451), (4, 2196.8472003936768), (5, 2204.909563064575), (6, 2199.6195316314697)]
-    subprocess = await get_data("subprocess")
+    # subprocess = await get_data("subprocess")
+    # rsyscall = await get_data("rsyscall")
     print("subprocess =", subprocess)
-    rsyscall = await get_data("rsyscall")
     print("rsyscall =", rsyscall)
-    nest = await get_data("nest")
-    print("nest =", nest)
-    nestnest = await get_data("nestnest")
-    print("nestnest =", nestnest)
-    flags = await get_data("flags")
-    print("flags =", flags)
     fig, ax = plt.subplots(figsize=(6.4,3.2))
     plt.xscale('log')
     plt.yscale('log')
     ax.plot([10**x for x, y in subprocess], [y for x, y in subprocess], 'o-', label="Python subprocess",
             linewidth=4, markersize=12)
     ax.plot([10**x for x, y in rsyscall], [y for x, y in rsyscall], '^-', label="rsyscall clone",
-            linewidth=4, markersize=12)
-    ax.plot([10**x for x, y in nest], [y for x, y in nest], '^-', label="rsyscall nest",
-            linewidth=4, markersize=12)
-    ax.plot([10**x for x, y in nestnest], [y for x, y in nestnest], '^-', label="rsyscall nestnest",
-            linewidth=4, markersize=12)
-    ax.plot([10**x for x, y in flags], [y for x, y in flags], '^-', label="rsyscall clone(NEWPID|NEWNS)",
             linewidth=4, markersize=12)
     ax.legend()
     

### 2021-04-07 research/paper/paper.tex
paper

@@
 \titlespacing{\subsubsection}{0pt}{0.5ex}{0ex}
+\lstset{
+language=Python,
+basicstyle=\small\ttfamily,
+keywordstyle=,
+stringstyle=\color{red},
+frame=tb,
+columns=fullflexible,
+showstringspaces=false,
+belowskip=0pt,
+}
 
@@
 \section{Introduction}\label{introduction}
-Existing process creation interfaces can be divided into three categories:
-Fork-style, spawn-style, and direct-style.
 \begin{table*}
@@
 \end{comment}
-Table \ref{tab:styles} summarizes the differences between each style.
-
-Fork-style (such as \texttt{fork}) and spawn-style (such as \texttt{posix\_spawn}) are the most popular,
-but have a number of disadvantages.
-Correct, performant use of fork-style places significant constraints on the calling process,
-and also has a complex programming model.
-Spawn-style lifts many of these constraints, and is simpler to program against,
-but is fundamentally limited in the kinds of processes it can create.
-With both fork-style and spawn-style,
-it is difficult to
-branch based on results or errors encountered while initializing the new process:
-fork-style requires IPC if it wishes to communicate errors back to the rest of the program,
-and robust indication of errors in spawn-style requires substantial expansion of the spawn interface.
-
-Direct-style is more rarely seen, but gives a number of advantages.
-Direct-style involves creating a new child process in a non-running state,
-and then calling individual system calls from the parent to initialize the child.
-Direct-style does not constrain the calling process,
-easily branches at any step of the initialization,
-and can create any kind of process supported by the underlying system.
-
-We'll discuss each style in more detail in section \ref{background}.
+The most well-known process creation interface on Linux, and Unix in general, is \texttt{fork}.
+When a parent process creates a child process through \texttt{fork} or other fork-style interfaces,
+both processes execute concurrently,
+and the child process typically inherits all but a few attributes from the parent process.\cite{manfork}
+The child process can call arbitrary system calls
+to mutate the new process until it reaches the desired state,
+typically ultimately calling \texttt{exec}.
+
+A user of fork must think carefully about the characteristics of the process from which fork is being called.
+Fork can have poor performance when called from processes with many memory mappings.\cite{forkroad}
+This can be mitigated to some degree, but inflicts further complexity on the caller.\cite{vfork}\cite{vfork_dangerous}
+Processes which are multi-threaded can encounter deadlocks and other issues if they call fork.
+\cite{forkroad}\cite{tlpi}\cite{posix_spawn_error_pipe}
+Some thread libraries provide partial mitigations for multi-threading issues,
+but it's up to user code to make use of those mitigations.\cite{pthread_atfork}
+
+Since the child process runs concurrently with the parent process,
+communicating events in the child back to the parent requires some form of IPC,
+adding complexity.\cite{posix_spawn_error_pipe}
+Even the basic interface of fork - one function call which returns twice - is unusual;
+such functions are generally considered to be complex, advanced topics in other settings.
+\cite{callcc_fork}\cite{continuations}\cite{bashcc}
+
+In a spawn-style interface,
+details about the new process are provided up-front as arguments to some function
+which creates the new process all at once;
+\texttt{posix\_spawn} is a typical example of this style.\cite{posix_spawn}\cite{create_process}\cite{chainloading}
+Other details, such as security context, are inherited from the parent process.\cite{tlpi}\cite{create_process}
+Spawn-style APIs
+can be called from any parent process without concerns about memory usage or multi-threading,
+when used correctly.
+\cite{forkroad}\cite{posix_spawn_error_pipe}
+
+Unlike \texttt{fork}, a spawn-style interface provides an API for process creation
+which is independent from existing system calls.\cite{spawn_h}\cite{subprocess_run}
+Spawn-style interfaces therefore often lack support for some features,
+and they also require learning a new interface distinct from the normal system call API.\cite{tlpi}\cite{posix_spawn_lwn}
+Also, spawn-style interfaces, by their nature as a single call,
+don't allow programs to branch on the results of individual modifications to the new process,
+and have worse error reporting for such modifications.
+\cite{posix_spawn}\cite{posix_spawn_error_pipe}
+
+Table \ref{tab:styles} summarizes the differences between fork-style and spawn-style.
+Each has its own advantages and disadvantages.
+
+A few academic operating systems - KeyKOS, seL4, and some others, for example -
+use another style of process creation.\cite{keykos}\cite{sel4}\cite{exokernel}\cite{fuschia}\cite{singularity}
+We use the new term "direct-style" to refer to this,
+since we know of no pre-existing generic term for this style of process creation.
+
+In direct-style, a process is created by its parent in a non-running state,
+then supplied with various resources,
+and then started running once it is fully set up.
+In operating systems with true direct-style process creation,
+the syscalls that can mutate a process
+take explicit arguments to indicate the process on which they should operate.\cite{keykos}\cite{sel4}
+In this way, the same syscalls that can mutate a process while it is running
+are called by the parent process to mutate the process while it is being set up.
 
@@
 Every \texttt{rsyscall} program starts with the ``local'' process
-and uses it to bootstrap the initial capabilities for other processes.
-
-To perform direct-style process creation,
-we use \texttt{rsyscall} to call \texttt{clone} on the ``local'' process to create a new child process.
-We call various syscalls in the child process to mutate it until it reaches the desired state,
-at which point we can call \texttt{exec} in the child process to start it running on its own.
-After that, the process is no longer under our control, and it functions like any other child process.
-The child process can be monitored using normal Linux child monitoring syscalls,
+and uses it to bootstrap capabilities for other processes.
+
+\lstinputlisting[float,language=Python,label={basic},caption={Create new process, change CWD, and exec}]
+{examples/basic/direct.py}
+Listing \ref{basic} shows a Python program using \texttt{rsyscall}.
+We create a new child process using direct-style \texttt{clone} in the local process \verb|local|.
+Here, \texttt{local.clone()} does not create another task running the same user program,
+as with the usual \texttt{clone} system call,
+but instead returns a capability through which we can control the new child process.
+We can then call various syscalls in the child process to mutate it until it reachs the desired state.
+In this example,
+we call \texttt{chdir} in the child process to change its working directory,
+then call \texttt{execv} in the child to execute a new executable.
+Calling \texttt{execv} consumes the capability,
+releasing the process from our control;
+later use of this child process capability will fail with an exception.
+The child process can now be monitored using normal Linux child monitoring syscalls,
 such as \texttt{waitid}.
@@
 which receives syscall requests, performs the syscall, and returns the response.
-No kernel modifications are required, and \texttt{rsyscall} is immediately deployable on recent Linux kernel.
+No kernel modifications are required, and \texttt{rsyscall} is immediately deployable on recent Linux kernels.
 We discuss the implementation in depth in section \ref{implementation},
-and examine several difficult process-creation related details.
-
-\lstinputlisting[float=h,language=Python,label={basic},caption={Create new process, change CWD, and exec}]
-{examples/basic/direct.py}
-Listing \ref{basic} shows a program using \texttt{rsyscall}.
-We create a new child process using \texttt{clone} on the local process \verb|local|,
-and receive back a capability to control it.
-We call \texttt{chdir} in the child process to change its working directory to \texttt{/dev}.
-Then we call \texttt{execv} in the child,
-passing the path of the \texttt{cat} executable
-and the arguments to the new executable.
-The name of the executable is passed as the first argument,
-as is typical in Unix.
-Calling \texttt{execv} consumes the capability,
-and later use of this child process capability will fail with an exception.
-
-The behavior of system calls made on a remote process through \texttt{rsyscall}
-is identical to the behavior of system calls made implicitly on the current process through libc.
-This is an API that systems programmers are already familiar with,
-and which is already thoroughly documented through Linux system call manpages.
-The novelty is constrained to the newly explicit process capabilities.
-Besides flattening the learning curve,
-this also ensures that direct-style \texttt{clone} will always be at least as powerful as \texttt{fork}.
-
-We show more examples of direct-style process creation in section \ref{examples},
-including use of process features which are difficult to use on Linux with \texttt{fork} or \texttt{posix\_spawn}.
+and examine several difficult process-creation details.
+
+System calls called in a child process through \texttt{rsyscall}
+behave identically to system calls called implicitly in the current process through libc.
+A system call returning an error is reported in the standard way for system calls in Python:
+An exception is thrown at the point of the call.
+User code can branch as normal on the results of system calls,
+and implement fallbacks or other error handling logic.
+
+Since we can call any syscall,
+we can access any feature available in Linux; we are therefore at least as powerful as \texttt{fork}.
+As we'll show in section \ref{examples},
+we can in fact use Linux features in ways that are impractical with the usual fork-style or spawn-style interfaces.
 
@@
 and can outperform \texttt{fork} in some cases.
+Like a spawn-style interface,
+we can call \texttt{clone}
+without worse performance in the presence of large memory usage,
+and without the possibility of bugs in the presence of multi-threading.
 The performance overhead of creating processes with direct-style clone
 is comparable to the performance cost of creating processes while inside a Linux namespace.
-At \twosigma, we have deployed direct-style process creation in production
-inside a library for distributed system deployment, Toplevel.
+At \twosigma, we have deployed direct-style process creation
+as part of a library for distributed system deployment, used in production.
 We evaluate in further detail in section \ref{evaluation}.
-\section{Background on process-creation}\label{background}
-As mentioned in the introduction,
-on most operating systems today,
-the available interfaces for process creation
-are either "fork-style" or "spawn-style".
-A few operating systems have a third style, which we refer to as "direct-style".
-
-\subsection{Fork-style}
-By far the most well-known fork-style interface is Unix's fork.\cite{forkhist}
-With a fork-style interface,
-a process, called the parent process, calls \texttt{fork},
-and a copy of that process is made,
-called the child process,
-differing only in a few attributes such as process id.
-The same program continues executing in both processes,
-with a different return value from fork to allow it to distinguish whether it is running in the parent or the child.
-Typically, the program will branch based on the return value
-and, if running in the child,
-call various system calls to modify the state of the new process.
-
-Fork has a number of flaws,
-and much ink has been spilled detailing them.
-We'll cover the most egregious briefly here,
-but see Baumann 2019 \cite{forkroad} for a more detailed evaluation.
-
-Fork, when called successfully, returns twice,
-creating two different processes which execute concurrently.
-This immediately makes fork difficult to integrate;
-for example, errors can't be easily communicated between the two processes.
-If there is an error in the initialization of the child process,
-the part of the program running in the child process
-needs to use some form of inter-process communication (IPC) to communicate it back to the parent process.
-The child process can run arbitrary code to make decisions about its initialization,
-but that code needs to use IPC if it wishes to communicate with the rest of the program,
-which is still in the parent process.
-
-Fork can have poor performance when called from processes with many memory mappings.\cite{forkroad}
-Fork copies many attributes about the parent process when creating the child process,
-including setting up copy-on-write memory mappings in the child process.
-This becomes slower as the parent process has more memory mappings,
-eventually taking a significant amount of time.
-This copying is required to robustly implement fork's model,
-where the same program continues executing without changes in both the parent process and child process;
-the copy of the program running in the child process might access any memory at any time.
-
-Multi-threaded programs generally cannot use fork safely.
-Typical Unix fork implementations duplicate only the thread calling fork from the parent process to the child process.
-In a multi-threaded program, this can cause various issues;
-for example, another thread might be holding a memory allocator lock at the time of the fork,
-which in the child process will never be unlocked,
-causing the child process to deadlock if it tries to allocate memory.
-Some thread libraries provide partial mitigations for this issue,
-but it's up to user code to make use of those mitigations.\cite{pthread_atfork}
-In combination with fork's poor performance in large-memory programs,
-a user of fork must think carefully
-about the characteristics of the process from which fork is being called.
-\subsection{Spawn-style}
-In a spawn-style interface,
-details about the new process are provided up-front as arguments to some function
-which creates the new process all at once.
-The most well known spawn-style interfaces are Unix's \texttt{posix\_spawn} \cite{posix_spawn}
-and Windows \texttt{CreateProcess} \cite{create_process};
-the Unix shell also serves as a spawn-style process creation interface,
-especially when used with Bernstein chaining\cite{chainloading}.
-
-Spawn-style interfaces typically still transparently copy some details from the parent process;
-for example, various security contexts,
-and any other process attribute which is not explicitly specified in the arguments to the spawn interface.
-The key difference between fork-style and spawn-style is not how much they copy,
-it is how the attributes which are not copied are specified:
-by mutation from code running in the new process in the case of fork-style,
-or by explicit argument passing in the case of spawn-style.
-
-Spawn-style interfaces lift the constraints on the calling process that fork-style interfaces impose.
-Spawn-style interfaces don't run user code in the new process during initialization.
-This means there is no need to copy memory,
-so performance is good when a spawn-style interface is used from a process with many memory mappings.
-Likewise, there are no inherent issues when a spawn-style interface is used from a multi-threaded program,
-though operating systems, such as Linux,
-which implement spawn-style interfaces on top of \texttt{fork} can still have threading bugs.
-
-The arguments that can be provided to a spawn-style process creation syscall
-do not cover all the possible attributes that one might want to set for the new process.
-Most systems have a large number of syscalls which can mutate the state of a process during its lifetime;
-for a spawn-style interface to work in all scenarios,
-all those possible mutations must be reproduced in the interface.
-For example, the \texttt{posix\_spawn} function provided by glibc does not support creating processes in new namespaces,
-as is required for container functionality on Linux.
-
-Spawn-style process creation
-essentially creates a new domain-specific language for specifying the attributes of the process.
-Fork-style allows existing system calls,
-with which programmers are already familiar,
-to be used to set up the new process.
-Spawn-style requires some form of declarative description of the new process,
-which must be learned in addition to the normal system call interface.
-
-Spawn-style process creation also does not allow for conditional logic during the setup.
-If the setup of the new process encounters an error at some point,
-the only option is to return from the entire spawn call with an error.
-Such errors returned from spawn-style calls
-are typically much less informative
-than the errors returned by the syscalls which directly mutate the process attributes.
-Other forms of conditional logic are also impossible in spawn-style;
-one modification to the process cannot depend on the result of some other modification.
-\subsection{Direct-style}
-A few academic operating systems, such as KeyKOS \cite{keykos}, seL4 \cite{sel4}
-and some others \cite{exokernel} \cite{fuschia} \cite{singularity},
-use another style of process creation.
-We use the new term "direct-style" to refer to this,
-since we know of no pre-existing concise and generic term for this style of process creation.
-
-In direct-style, a process is created by its parent in a non-running state,
-then supplied with various resources,
-and then started running once it is fully set up.
-In operating systems with true direct-style process creation,
-the syscalls that can mutate a process
-take explicit arguments to indicate the process on which they should operate.
-In this way, the same syscalls that can mutate a process while it is running
-are called by the parent process to mutate the process while it is being set up.
-
-We refer to this as "direct-style" process creation,
-because the parent creating the process operates on it directly and imperatively
-rather than dispatching a distinct unit of code to perform setup from inside the context of the new process,
-as in fork-style,
-or building up a declarative specification of what the new process should look like ahead of time,
-as in spawn-style.
-
-Since everything happens directly from the parent process,
-direct-style process creation is compatible with a variety of techniques
-which are difficult with fork-style or spawn-style.
-For example, a REPL could be used to interactively create the new process,
-or a program might initialize several processes at once and share data between them.
-
-Significantly,
-since all operations are performed through syscalls called directly from the main program,
-errors are indicated in the same way as any other syscall error.
-This is unlike fork-style, which needs IPC to indicate errors,
-and unlike spawn-style, which typically indicates errors at a very coarse-grained level.
-
-Like fork-style interfaces,
-direct-style interfaces can set up arbitrary attributes in the new process.
-Any attribute that can be changed by a syscall
-can be manipulated by a direct-style interface,
-just as with fork-style interfaces.
-This is unlike spawn-style interaces,
-which only can change attributes that are supported by the interface.
-
-Like spawn-style interfaces,
-direct-style interfaces have no constraints on the calling process.
-Also like spawn-style interface,
-no user code runs in the child process, so
-direct-style interfaces can be used from multi-threaded or large-memory processes without issues.
-
-Direct-style can be more complex to use;
-it is most typically used in capability-oriented operating systems,
-where a great deal of resources and information must be provided explicitly to initialize the new process.
-In a truly capability-oriented operating system,
-nothing is copied implicitly to the new process;
-everything must be explicitly specified.
-This can appear more complex
-when compared to process creation on Unix systems or Windows.
-
-However, we believe that a natural port of direct-style process creation to Linux
-provides an interface that is just as simple to use as fork-style or spawn-style on Linux,
-without their disadvantages.
-Such an interface, like fork-style and spawn-style,
-would implicitly copy some attributes to the new process,
-rather than being capability-oriented and requiring all attributes to be specified explicitly.
-This is the kind of direct-style interface that we contribute in this paper.
 \section{Examples}\label{examples}
@@
 
-% TODO need to clarify this more, this is v2, Larry didn't like v1 beyond first sentence
 Most resources in Linux are managed through file descriptors,
-so this feature allows the parent process to pass a variety of resources to the child process,
+so this allows the parent process to pass a variety of resources to the child process,
 such as files, network connections, pipes, or other resources.\cite{ucspi}
@@
 in ways that were not anticipated at development time
-by customizing the environment the process runs in.
-Other system calls such as \texttt{chroot} can also be used for this.
-
-\lstinputlisting[float=*ht,language=Python,label={mount},caption={Overriding absolute path using a mount namespace}]
+by customizing the environment the process runs in.\cite{plan9ns}
+Other system calls such as \texttt{chdir} and \texttt{chroot} can also be used for this,
+but namespaces allow new ways of customization.\cite{mount_namespaces}
+
+\lstinputlisting[float,language=Python,label={mount},caption={Overriding absolute path using a mount namespace}]
 {examples/mount/direct.py}
-One classic form of customization enabled by namespaces
-is overriding files at paths that were hardcoded into a program we can't change,
+One classic form of customization
+is overriding files at paths that were hardcoded into a program,
 without changing those files for the rest of the system.
@@
 
-ptrace provides a way to perform arbitrary actions on other processes,
+We chose to implement in userspace first,
+rather than immediately implementing this in the kernel,
+to aid deployability
+and to allow fast iteration while developing userspace code using these features.
+In the past, attempts to upstream kernel support for novel features without extensive userspace usage
+has had a poor reception in the Linux kernel community.\cite{lwn_checkpoint}\cite{first_class_address_space}
+We believe proving viability in userspace first will be better in the long term.
+
+\texttt{ptrace} provides a way to perform arbitrary actions on other processes,
 including system calls,
-but is not suitable for us.
-Among other issues, it has the substantial disadvantage of not permitting multiple ptracers.
-A ptrace-based implementation would prevent using \texttt{strace} or \texttt{gdb} on \texttt{rsyscall}-controlled processes,
+but is not suitable for us.\cite{ptrace}
+Among other issues, it has the substantial disadvantage of not supporting multiple ptracers.
+A \texttt{ptrace}-based implementation would therefore prevent
+using \texttt{strace} or \texttt{gdb} on \texttt{rsyscall}-controlled processes,
 which is an unacceptable limitation for a general-purpose utility.
@@
 \item
-  In Python, does our \texttt{rsyscall}-based direct-style process creation interface
-  have acceptable performance relative to the standard library's spawn-style \texttt{subprocess.run}?
-  We benchmark direct-style process creation against spawn-style in \ref{subprocess_bench},
+  Does our \texttt{rsyscall}-based direct-style process creation interface
+  have acceptable performance overhead relative to the alternatives?
+  We evaluate a number of microbenchmarks in \ref{subprocess_bench},
   and find an acceptable level of overhead.
@@
 though this simplification is not immediately visible in a small example.
-The fork-style examples may break if there are multiple threads in the calling process,
+The fork-style implementations may break if there are multiple threads in the calling process,
 and may be slow if the calling process has large amounts of memory mapped, as we'll see in section \ref{subprocess_bench}.
-The direct-style examples do not have these limitations.
+The direct-style implementations do not have these limitations.
 \subsection{Microbenchmark results}\label{subprocess_bench}
+\subsubsection{Basic process creation}
 \begin{figure}[h!]
@@
 \end{figure}
-In Python, does our \texttt{rsyscall}-based direct-style process creation interface
-have acceptable performance relative to the standard library's spawn-style \texttt{subprocess.run}?
-
-We first evaluate a simple process creation workload:
+We evaluate a simple process creation workload:
 create a child process, exec the \texttt{true} binary, and wait for it to exit.
-We implement this using both direct-style \texttt{clone}
-and the standard library's \texttt{subprocess.run}.
-We benchmark both implementations under CPython 3.7.7, on Linux 4.19.93,
-pinned to an isolated single core on an Intel i9-9900K CPU with 60GB of RAM.
+We implement this using Python \texttt{rsyscall} direct-style \texttt{clone},
+the Python standard library's \texttt{subprocess.run} (which is implemented primarily with \texttt{fork}),
+\texttt{fork} from C,
+and \texttt{posix\_spawn} from C.
+We run on CPython 3.7.7, glibc 2.30, Linux 4.19.93,
+pinned to an isolated single core on an Intel i9-9900K CPU at standard clock speed,
+with 60GB of RAM.
 We vary the amount of anonymous memory mapped in the parent process
 to demonstrate how each implementation scales with memory usage.
-The results are in Figure \ref{fig:subprocess_bench}.
-
-The direct-style implementation stays around 2.2 milliseconds regardless of memory usage.
-\texttt{subprocess.run}, which is implemented with \texttt{fork},
-starts at 1.4 milliseconds,
-and then scales linearly with the amount of memory mapped, as expected,
-up to 48.5 milliseconds at 4 GBs of mapped memory.
-The cross-over point for this micro-benchmark is about 40 megabytes of memory mapped.
-
-We perform another microbenchmark to evaluate the overhead of modifying the child process.
-We simply call \texttt{getpid} from Python
+The results are summarized in figure \ref{fig:subprocess_bench}.
+
+Our true baseline for performance is the Python standard library's \texttt{subprocess.run};
+this takes an average of 1.4 milliseconds at low memory,
+while \texttt{rsyscall}'s \texttt{clone} takes an average of 2.2 milliseconds at any memory usage.
+As expected, the fork-based implementations scale poorly with memory usage,
+and the C implementations vastly outperform the Python implementations.
+\texttt{posix\_spawn} takes an average of 440 microseconds to start a process,
+regardless of the memory usage of the calling process.
+
+We perform another microbenchmark to evaluate the overhead of performing additional modifications of the child process.
+We call \texttt{getpid} from Python
 in both the child process and the parent process on the same benchmark setup.
@@
 or a child in a different namespace (created with \verb=clone(CLONE_NEWNS|CLONE_NEWPID)=).
-We've also specified different combinations of flags for the process being created in the benchmark.
+We've also specified different combinations of flags for the child process being created in the benchmark.
+We don't include the Python standard library in these benchmarks
+as there's no way to create nested child processes or specify \texttt{CLONE} flags in the Python standard library.
 The results are in figure \ref{fig:clone_bench}.
-We don't include the Python standard library in these benchmarks
-as there's no way to specify \texttt{CLONE} flags in the Python stdlib.
 
@@
 is around 15 milliseconds; subtantially greater than the overhead of \texttt{rsyscall}.
+As namespaces are widely used,
+this suggests our overhead is acceptable.
 \subsection{Usage in the real world}\label{realworld}
@@
 We pass around file descriptors freely inside Toplevel.
-A given component can implemented with a separate process which has the file descriptor passed down,
-running a program written in native code for performance,
-or implemented in-process with a Python implementation for flexibility;
-either is possible behind a uniform file descriptor interface.
-
-We've also been able to create complicated pipelines in Toplevel,
-as in section \ref{pipe}.
+This uniform interface allows us to flexibly swap implementations between in-process and out-of-process,
+allowing us to decide on a case-by-case basis
+whether we prefer the flexibility of an in-process Python implementation,
+or the performance of a native code implementation running in a child process.
 

### 2021-04-08 research/paper/analyze_clone_with_mmap.py
add clone_with_mmap benchmark

@@ -1,0 +1,32 @@
+from dataclasses import dataclass
+import csv
+import sys
+
+@dataclass
+class Run:
+    pid: int
+    # all in microseconds
+    pre_fork: int
+    in_child: int
+    in_parent: int
+    after_exit: int
+
+usec_in_sec = 1000*1000
+
+def analyze_data(data: str) -> None:
+    runs = []
+    for line in csv.DictReader(data.split()):
+        runs.append(Run(int(line['pid']),
+            pre_fork=int(line['pre_fork_sec'])*usec_in_sec + int(line['pre_fork_usec']),
+            in_child=int(line['in_child_sec'])*usec_in_sec + int(line['in_child_usec']),
+            in_parent=int(line['in_parent_sec'])*usec_in_sec + int(line['in_parent_usec']),
+            after_exit=int(line['after_exit_sec'])*usec_in_sec + int(line['after_exit_usec']),
+        ))
+
+    print(sum(run.after_exit - run.pre_fork for run in runs)/len(runs))
+
+def main() -> None:
+    analyze_data(sys.stdin.read())
+
+if __name__ == "__main__":
+    main()

### 2021-04-08 python/clone_bench.py
paper: some more benccchhmarrkss

@@ -1,0 +1,233 @@
+import time
+import sys
+import os
+try:
+    import trio
+except:
+    # ohhhh it's getting cleared by sudooooo
+    # got it.
+    print(os.environ, file=sys.stderr)
+    exit(1)
+import subprocess
+import rsyscall.tasks.local as local
+import sys
+from rsyscall.sys.wait import W
+from rsyscall import Command, Path
+from rsyscall.sched import CLONE
+
+import rsyscall.nix as nix
+from rsyscall.tasks.stdin_bootstrap import stdin_bootstrap, stdin_bootstrap_path_from_store
+from rsyscall.unistd import SEEK
+from rsyscall.sys.mman import PROT, MAP
+from rsyscall.sys.resource import PRIO
+from statistics import mean
+import csv
+import argparse
+from dataclasses import dataclass
+
+import matplotlib
+import matplotlib.pyplot as plt
+import numpy as np
+import typing as t
+
+import cProfile
+
+async def run_benchmark(run_mode: str, thing_mode: str) -> int:
+    cmd = await local.thread.environ.which('true')
+    await local.thread.environ.as_arglist(local.thread.ram)
+    if run_mode == 'stdlib':
+        if thing_mode == 'spawn':
+            async def run() -> None:
+                popen = subprocess.Popen([cmd.executable_path], preexec_fn=lambda: None)
+                popen.wait()
+        elif thing_mode in ['spawn_newns', 'spawn_newpid', 'spawn_newnspid']:
+            async def run() -> None:
+                pass
+        elif thing_mode == 'getpid':
+            async def run() -> None:
+                # hmm we want to bypass the pid cache.
+                os.getpgid(0)
+    else:
+        if run_mode == 'rsyscall':
+            thread = local.thread
+        elif run_mode == 'nest':
+            thread = await local.thread.clone()
+        elif run_mode == 'nestnest':
+            first_nesting_child = await local.thread.clone()
+            thread = await first_nesting_child.clone()
+        elif run_mode == 'flags':
+            thread = await local.thread.clone(CLONE.NEWNS|CLONE.NEWPID)
+        if thing_mode == 'getpid':
+            async def run() -> None:
+                await thread.task.getpgid()
+        else:
+            if thing_mode == 'spawn':
+                flags = CLONE.NONE
+            elif thing_mode == 'spawn_newns':
+                flags = CLONE.NEWNS
[…]

### 2021-04-08 python/subprocess_bench.py

@@ -13,6 +13,7 @@ import rsyscall.tasks.local as local
 import sys
 from rsyscall.sys.wait import W
 from rsyscall import Command, Path
+from rsyscall.sched import CLONE
 
 import rsyscall.nix as nix
 from rsyscall.tasks.stdin_bootstrap import stdin_bootstrap, stdin_bootstrap_path_from_store
@@ -36,20 +37,46 @@ async def run_benchmark(mode: str, mmaps: int) -> t.Tuple[int, int]:
         # set POPULATE so it actually gets faulted in; if it isn't
         # already in the page tables, there's no slowdown
         await local.thread.task.mmap(4096, PROT.READ|PROT.WRITE, MAP.PRIVATE|MAP.ANONYMOUS|MAP.POPULATE)
-    cmd = await local.thread.environ.which('true')
+    cmd = await local.thread.environ.which('cat')
     await local.thread.environ.as_arglist(local.thread.ram)
+    cwd = await local.thread.ptr(Path("/dev"))
     async def subp_run() -> None:
-        popen = subprocess.Popen([cmd.executable_path], preexec_fn=lambda: None)
+        popen = subprocess.Popen([cmd.executable_path, './null'], cwd=cwd.value)
         popen.wait()
     async def rsys_run() -> None:
-        thread = await local.thread.clone()
-        child = await thread.execv(cmd.executable_path, [cmd.executable_path])
-        await child.waitpid(W.EXITED)
-        await thread.close()
+        child = await local.thread.clone()
+        await child.task.chdir(cwd)
+        child_proc = await child.execv(cmd.executable_path, [cmd.executable_path, "./null"])
+        await child_proc.waitpid(W.EXITED)
+        await child.close()
     if mode == 'subprocess':
         run = subp_run
     elif mode == 'rsyscall':
         run = rsys_run
+    elif mode == 'nest':
+        nesting_child = await local.thread.clone()
+        async def run() -> None:
+            child = await nesting_child.clone()
+            await child.task.chdir(cwd)
+            child_proc = await child.execv(cmd.executable_path, [cmd.executable_path, "./null"])
+            await child_proc.waitpid(W.EXITED)
+            await child.close()
+    elif mode == 'nestnest':
+        first_nesting_child = await local.thread.clone()
+        second_nesting_child = await first_nesting_child.clone()
+        async def run() -> None:
[…]

### 2021-04-08 research/paper/paper.tex
paper

--- 2021-04-07 research/paper/paper.tex
@@
 \usepackage{tikz}
-\usepackage{dblfloatfix}
 \usepackage{amsmath}
@@
 \usepackage{titlesec}
-
-% Ugh, these aren't the same size right now.
-\newcommand{\tbad}{\xmark:}
-\newcommand{\tgood}{\cmark:}
-% \newcommand{\tbad}{\text{\sffamily X} :}
-% \newcommand{\tgood}{\checkmark:}
 
@@
 Fork-style, spawn-style, and direct-style.
-\begin{table*}
-\begin{tabular}{l|l|l|l}
-% BEGIN RECEIVE ORGTBL compare
- & fork & spawn & direct-style\\
-\hline
-Requirements on parent process & \tbad Single-threaded, small memory & \tgood None & \tgood None\\
-Programming model & \tbad Complex (returns twice) & \tgood Simple (single call) & \tgood Simple (imperative)\\
-Maximally powerful & \tgood Yes, can call any syscall & \tbad No, limited interface & \tgood Yes, can call any syscall\\
-Reporting of results & \tbad Requires IPC & \tbad Not fine-grained & \tgood From individual child syscalls\\
-Non-inherited attributes & Mutated by code in child & Set by arguments & Mutated by code in parent\\
-% END RECEIVE ORGTBL compare
-\end{tabular}
-\caption{Features of fork-style vs spawn-style vs direct-style}
-\label{tab:styles}
-\end{table*}
-\begin{comment}
-#+ORGTBL: SEND compare orgtbl-to-latex :splice t
-|                                | fork                                | spawn                       | direct-style                          |
-|--------------------------------+-------------------------------------+-----------------------------+---------------------------------------|
-| Requirements on parent process | \tbad Single-threaded, small memory | \tgood None                 | \tgood None                           |
-| Programming model              | \tbad Complex (returns twice)       | \tgood Simple (single call) | \tgood Simple (imperative)            |
-| Maximally powerful             | \tgood Yes, can call any syscall    | \tbad No, limited interface | \tgood Yes, can call any syscall      |
-| Reporting of results           | \tbad Requires IPC                  | \tbad Not fine-grained      | \tgood From individual child syscalls |
-| Non-inherited attributes       | Mutated by code in child            | Set by arguments            | Mutated by code in parent             |
-% $
-\end{comment}
-Table \ref{tab:styles} summarizes the differences between each style.
 
@@
 
-We'll discuss each style in more detail in section \ref{background}.
-
-To perform direct-style process creation on Linux,
-we need an API where we explicitly specify in which process we want to make a syscall.
-This differs from the normal mode of operation on Linux,
-where programs making syscalls implicitly operate on the current process.
-
-Our contribution in this paper is this API, through the \texttt{rsyscall} library.
-The \texttt{rsyscall} project develops
-language-specific, object-capability-model, low-abstraction libraries for the Linux syscall interface,
-bypassing libc.
-\texttt{rsyscall} is currently focused on supporting Python,
+Our contribution is an implementation of direct-style process creation for Linux,
+making use of our work on the \texttt{rsyscall} project.
+The \texttt{rsyscall} project develops language-specific ``process-independent'' system call libraries for Linux,
+which replace libc system call wrappers,
+and allow a program to perform cross-process system calls.
+\texttt{rsyscall} works entirely in userspace
+and can be deployed on existing Linux systems without kernel modifications.
+\texttt{rsyscall} is currently focused on supporting, and is implemented in, Python,
 but can be ported to other languages;
@@
 but generalize easily to other languages.
+
+Our direct-style process creation implementation uses \texttt{clone} to create a child process in an embryonic state
+which can then be manipulated from user code in the parent process using syscalls with \texttt{rsyscall}.
+A direct-style \texttt{clone} has all the advantages of direct-style process creation
+over Linux's fork-style \texttt{fork} and spawn-style \texttt{posix\_spawn}:
+\begin{compactitem}
+\item
+  Processes can be created correctly and performantly from any parent process,
+  with a straightforward, familiar API.
+\item
+  Processes can be created with the full set of Linux features.
+\item
+  Syscall results are reported individually to the parent process.
+  % rather than batched together as with spawn-style or manually reported via IPC as with fork-style.
+\end{compactitem}
+
+We've implemented a variety of useful applications using direct-style process creation on Linux,
+which we'll demonstrate in this paper.
+Direct-style \texttt{clone} has minimal performance cost in normal situations,
+and can outperform \texttt{fork} in some cases.
+We have deployed direct-style process creation in production internally at \twosigma.
+
+Our implementation is distributed as a feature of \texttt{rsyscall},
+and is immediately deployable, entirely in userspace, on recent Linux kernels.
 \texttt{rsyscall} is open source and available from
 [Github url].
-
-Unlike POSIX C libraries such as glibc or musl,
-an \texttt{rsyscall} library is organized based on the object-capability model.
-The capabilities to make syscalls in any given process are reified as language objects.
-If the capability to make syscalls in a specific process is not passed (in some way) to a function,
-then that function cannot make syscalls in that process.
-
-Due to this design, an \texttt{rsyscall} program can make use of multiple processes at once,
-by manipulating capabilities for multiple processes.
-The relevant two capabilities for this paper are the initial capability to the ``local'' process,
-and capabilities to child processes.
-The ``local'' process is the one which hosts the runtime for the running program,
-and in which a legacy libc would implicitly make syscalls.
-Every \texttt{rsyscall} program starts with the ``local'' process
-and uses it to bootstrap the initial capabilities for other processes.
-
-To perform direct-style process creation,
-we use \texttt{rsyscall} to call \texttt{clone} on the ``local'' process to create a new child process.
-We call various syscalls in the child process to mutate it until it reaches the desired state,
-at which point we can call \texttt{exec} in the child process to start it running on its own.
-After that, the process is no longer under our control, and it functions like any other child process.
-The child process can be monitored using normal Linux child monitoring syscalls,
-such as \texttt{waitid}.
-
-These child process capabilities are created and managed in userspace,
-by launching new processes running a syscall server,
-which receives syscall requests, performs the syscall, and returns the response.
-No kernel modifications are required, and \texttt{rsyscall} is immediately deployable on recent Linux kernel.
-We discuss the implementation in depth in section \ref{implementation},
-and examine several difficult process-creation related details.
-
-\lstinputlisting[float=h,language=Python,label={basic},caption={Create new process, change CWD, and exec}]
-{examples/basic/direct.py}
-Listing \ref{basic} shows a program using \texttt{rsyscall}.
-We create a new child process using \texttt{clone} on the local process \verb|local|,
-and receive back a capability to control it.
-We call \texttt{chdir} in the child process to change its working directory to \texttt{/dev}.
-Then we call \texttt{execv} in the child,
-passing the path of the \texttt{cat} executable
-and the arguments to the new executable.
-The name of the executable is passed as the first argument,
-as is typical in Unix.
-Calling \texttt{execv} consumes the capability,
-and later use of this child process capability will fail with an exception.
-
-The behavior of system calls made on a remote process through \texttt{rsyscall}
-is identical to the behavior of system calls made implicitly on the current process through libc.
-This is an API that systems programmers are already familiar with,
-and which is already thoroughly documented through Linux system call manpages.
-The novelty is constrained to the newly explicit process capabilities.
-Besides flattening the learning curve,
-this also ensures that direct-style \texttt{clone} will always be at least as powerful as \texttt{fork}.
-
-We show more examples of direct-style process creation in section \ref{examples},
+% \url{https://github.com/catern/rsyscall}.
+
+This paper is organized in several sections.
+\begin{compactitem}
+\item Section \ref{background} provides background on fork-style, spawn-style, and direct-style process creation,
+and considers their advantages and disadvantages in more depth.
+\item Section \ref{overview} discusses the design of the API,
+including our usage of \texttt{rsyscall}.
+\item Section \ref{examples} shows examples of direct-style process creation in Python on Linux,
 including use of process features which are difficult to use on Linux with \texttt{fork} or \texttt{posix\_spawn}.
-
-Direct-style \texttt{clone} has acceptable performance cost,
-and can outperform \texttt{fork} in some cases.
-The performance overhead of creating processes with direct-style clone
-is comparable to the performance cost of creating processes while inside a Linux namespace.
-At \twosigma, we have deployed direct-style process creation in production
-inside a library for distributed system deployment, Toplevel.
-We evaluate in further detail in section \ref{evaluation}.
+\item Section \ref{implementation} explores our implementation in depth,
+and discusses several difficult process-creation related details of \texttt{rsyscall}.'s implementation.
+\item Section \ref{evaluation} evaluates our results using both quantitative and qualitative approaches.
+\item Section \ref{related_work} examines other work which also touches on the topics covered by this paper.
+\item Section \ref{future_work} lists several directions for future work
+related to direct-style process creation and to \texttt{rsyscall}.
+\item Section \ref{conclusions} briefly outlines our conclusions about direct-style and the future of process creation.
+\end{compactitem}
 \section{Background on process-creation}\label{background}
@@
 
+% Ugh, these aren't the same size right now.
+\newcommand{\tbad}{\xmark:}
+\newcommand{\tgood}{\cmark:}
+% \newcommand{\tbad}{\text{\sffamily X} :}
+% \newcommand{\tgood}{\checkmark:}
+
+\begin{table*}
+\begin{tabular}{l|l|l|l}
+% BEGIN RECEIVE ORGTBL compare
+ & fork-style & spawn-style & direct-style\\
+\hline
+Works with any caller & \tbad No threads or large memory & \tgood Yes & \tgood Yes\\
+Programming model & \tbad Complex (returns twice) & \tgood Simple (single call) & \tgood Simple (imperative)\\
+Maximally powerful & \tgood Yes, can call any syscall & \tbad No, limited interface & \tgood Can call any syscall\\
+Error handling & \tbad Poor, requires IPC & \tbad Poor, not fine-grained & \tgood Reports individual errors\\
+Non-copied attributes & Mutated by code in child & Set by arguments & Mutated by code in parent\\
+% END RECEIVE ORGTBL compare
+\end{tabular}
+\caption{Features of fork-style vs spawn-style vs direct-style}
+\label{tab:ease}
+\end{table*}
+\begin{comment}
+#+ORGTBL: SEND compare orgtbl-to-latex :splice t
+|                       | fork-style                       | spawn-style                  | direct-style                     |
+|-----------------------+----------------------------------+------------------------------+----------------------------------|
+| Works with any caller | \tbad No threads or large memory | \tgood Yes                   | \tgood Yes                       |
+| Programming model     | \tbad Complex (returns twice)    | \tgood Simple (single call)  | \tgood Simple (imperative)       |
+| Maximally powerful    | \tgood Yes, can call any syscall | \tbad No, limited interface  | \tgood Can call any syscall      |
+| Error handling        | \tbad Poor, requires IPC         | \tbad Poor, not fine-grained | \tgood Reports individual errors |
+| Non-copied attributes | Mutated by code in child         | Set by arguments             | Mutated by code in parent        |
+% $
+\end{comment}
 \subsection{Fork-style}
@@
 This is the kind of direct-style interface that we contribute in this paper.
+\section{Overview of API}\label{overview}
+In most operating system APIs, including the ones available on Linux,
+when a program makes a syscall, it implicitly operates on the current process.
+To perform direct-style process creation on Linux,
+we need an API where we instead explicitly specify in which process we want to make a syscall.
+
+\texttt{rsyscall} provides this API.
+The \texttt{rsyscall} project develops
+language-specific, object-capability-model, low-abstraction libraries for the Linux syscall interface,
+bypassing libc.
+
+Unlike POSIX C libraries such as glibc or musl,
+an \texttt{rsyscall} library is organized based on the object-capability model.
+The capabilities to make syscalls in any given process are reified as language objects.
+If the capability to make syscalls in a specific process is not passed (in some way) to a function,
+then that function cannot make syscalls in that process.
+
+Due to this design, an \texttt{rsyscall} program can make use of multiple processes at once,
+by manipulating capabilities for multiple processes.
+The relevant two capabilities for this paper are the initial capability to the ``local'' process,
+and capabilities to child processes.
+The ``local'' process is the one which hosts the runtime for the running program,
+and in which a legacy libc would implicitly make syscalls.
+Every \texttt{rsyscall} program starts with the ``local'' process
+and uses it to bootstrap the initial capabilities for other processes.
+
+To perform direct-style process creation,
+we use \texttt{rsyscall} to call \texttt{clone} on the ``local'' process to create a new child process.
+We call various syscalls in the child process to mutate it until it reaches the desired state,
+at which point we can call \texttt{execve} in the child process to start it running on its own.
+After that, the process is no longer under our control, and it functions like any other child process.
+The child process can be monitored using normal Linux child monitoring syscalls,
+such as \texttt{waitid}.
+
+\lstinputlisting[float=ht,language=Python,label={basic},caption={Creating a new process, changing CWD, and execing}]
+{examples/basic/direct.py}
+Listing \ref{basic} shows this in action in Python.
+We create a new child process using \texttt{clone} on the local process \verb|local|,
+and receive back a capability to control it.
+We call \texttt{chdir} in the child process to change its working directory to \texttt{/dev}.
+Then we call \texttt{execve} in the child,
+passing the path of the \texttt{cat} executable,
+the arguments to the new executable,
+and the unchanged set of environment variables.
+The name of the executable is passed as the first argument,
+as is typical in Unix.
+Calling \texttt{execve} consumes the capability,
+and later use of this child process capability will fail with an exception.
+
+The behavior of system calls made on a remote process through \texttt{rsyscall}
+is identical to the behavior of system calls made implicitly on the current process through libc.
+This is an API that systems programmers are already familiar with,
+and which is already thoroughly documented through Linux system call manpages.
+The novelty is constrained to the newly explicit process capabilities.
+
+The lack of novelty in the API is important not only for the learning curve,
+but also for power.
+Since we are using conventional Linux system calls,
+just in a cross-process way,
+direct-style \texttt{clone} will always be at least as powerful as \texttt{fork}.
 \section{Examples}\label{examples}
@@
 the executable will be able to use that file descriptor through that file descriptor number.
+Unlike in listing \ref{basic}, we call the \texttt{execv} wrapper rather than \texttt{execve},
+so that we don't need to pass the (unmodified) environment to every \texttt{exec} call in these examples.
 \subsection{Non-shared-memory concurrency}
@@
 At \twosigma we have used \texttt{rsyscall} and direct-style process creation
-to implement an internal distributed system deployment library, written in Python,
-which we'll refer to here as Toplevel.
-Toplevel is in use in production,
+as a component of an internal distributed system deployment library, written in Python.
+This library is in use in production,
 and is extensively used as part of testing and development.
-Toplevel consists of a collection of functions and modules
-for each component in our system,
-which start up components using direct-style process creation,
-all from a single Python parent process.
-
-One of the major benefits of direct-style process creation has been easy use of file descriptor inheritance.
-Our experience has been that with fork-style or spawn-style interfaces,
-all file descriptor logic must be centralized,
-whereas file descriptors in a new process can be built up over time using direct-style,
-simplifying the implementation.
-We've found that even programmers without substantial prior experience with Unix programming
-are able to use file descriptor passing as a normal feature of their development process with relative ease,
-writing code to support new programs and passing down file descriptors to those programs without issue.
-
-Easy file descriptor passing has, in turn, allowed us to heavily use socket activation techniques
-to start up services in parallel,
-substantially speeding up system startup.\cite{socketactivation}
-We are able to bootstrap connections over our internal shared memory transports
-over pre-created, passed-down file descriptors,
-and our management interfaces listen on passed-down sockets;
-this allows starting up services in parallel, rather than in dependency order.
-Benchmarking one test built with an old library and its equivalent built with Toplevel,
-the old test took an average of 221 seconds to run,
-mostly spent in system startup,
-while the new test takes an average of 16 seconds.
-
-Easy file descriptor inheritance has also allowed us to treat file descriptors as a uniform interface in Toplevel.
-We pass around file descriptors freely inside Toplevel.
-A given component can implemented with a separate process which has the file descriptor passed down,
-running a program written in native code for performance,
-or implemented in-process with a Python implementation for flexibility;
-either is possible behind a uniform file descriptor interface.
-
-We've also been able to create complicated pipelines in Toplevel,
-as in section \ref{pipe}.
-
-We've also made use of namespaces,
-chiefly user namespaces and pid namespaces.
-For legacy applications which directly start subprocesses,
+
+Each service deployed by this library is started with custom code,
+written using direct-style process creation.
+This allows us to pass down file descriptors and initialize process attributes
+according to each program's needs;
+which in turn has let us make optimizations which previously required too much program-specific customization.
+
+For example,
+we are able to bootstrap connections over our internal shared memory transports over passed-down file descriptors,
+and our management interfaces listen on passed-down sockets.
+This allows substantial parallelism in the startup of services.\cite{socketactivation}
+Taking one representative example,
+one test built with an old library took an average of 221 seconds to run,
+mostly spent in system startup time as processes started up serially.
+With parallelized system startup, the same test takes an average of 16 seconds.
+
+We've found that programmers
+are able to use file descriptor passing as a normal feature of their development process with relative ease.
+Programmers without substantial prior experience with Unix programming
+are able to write code to support new services
+nd pass down file descriptors to the program without encountering significant issues.
+
+We also flexibly use namespaces where appropriate.
+For example, for legacy applications which directly start subprocesses,
 we prevent process leaking with a pid namespace.
-This is part of strategy of noninterference and cleanup
-that ensures that multiple users of Toplevel
-can coexist simultaneously on the same host,
-even if running the same system.
-
-Toplevel also makes uses of other features of \texttt{rsyscall} which are not covered in this paper.
-These include the ability to operate on remote hosts through capabilities for remote processes,
-and the ability to perform system calls in parallel across a pool of processes.
-These have significant synergy with direct-style process creation
-in forming an overall powerful system.
 

### 2021-04-08 research/paper/ease.py
paper

@@ -9,7 +9,7 @@ def extract_label(line: str) -> str:
     label, _ = rest.split("},caption")
     return label
 
-listing_labels = [extract_label(line) for line in open(dirname/"paper.tex") if line.startswith("\\lstinputlisting")]
+listing_labels = [extract_label(line) for line in open(dirname/"paper.tex") if line.startswith("\\lstinputlisting") and "fork" not in line]
 
 headings = (
     "Name",

### 2021-04-08 python/clone_bench.py
paper: benchmark: improvements

@@ -89,9 +89,9 @@ async def run_benchmark(run_mode: str, thing_mode: str) -> int:
 
 async def main() -> None:
     parser = argparse.ArgumentParser(description='Do benchmarking of rsyscall vs subprocess.run')
-    run_modes = ['stdlib', 'rsyscall', 'nest', 'nestnest', 'flags']
+    run_modes = ['rsyscall', 'nest', 'nestnest', 'flags']
     parser.add_argument('--run-mode', choices=run_modes)
-    thing_modes = ['spawn', 'spawn_newns', 'spawn_newpid', 'spawn_newnspid', 'getpid']
+    thing_modes = ['spawn', 'spawn_newns', 'spawn_newpid', 'spawn_newnspid']
     parser.add_argument('--thing-mode', choices=thing_modes)
     parser.add_argument('--no-use-setpriority', help="don't setpriority before benchmarking; doing that requires privileges,"
                         " which are attained by running the benchmark with sudo (handled internally)",
@@ -138,13 +138,17 @@ async def main() -> None:
             time = float(data)*1000*1000
             times.append(time)
         return(mean(times))
-    data = {}
-    # for run_mode in run_modes:
-    #     for thing_mode in thing_modes:
-    #         result = await run_many(run_mode, thing_mode)
-    #         print(run_mode, thing_mode, result)
-    #         data.setdefault(run_mode, dict())[thing_mode] = result
[…]

### 2021-04-08 python/subprocess_bench.py

@@ -68,7 +68,7 @@ async def run_benchmark(mode: str, mmaps: int) -> t.Tuple[int, int]:
 
 async def main() -> None:
     parser = argparse.ArgumentParser(description='Do benchmarking of rsyscall vs subprocess.run')
-    parser.add_argument('--run-benchmark', choices=['subprocess', 'rsyscall', 'fork', 'posix_spawn'])
+    parser.add_argument('--run-benchmark', choices=['subprocess', 'rsyscall', 'fork', 'vfork', 'posix_spawn'])
     parser.add_argument('--mmaps', type=int, default=0)
     parser.add_argument('--no-use-setpriority', help="don't setpriority before benchmarking; doing that requires privileges,"
                         " which are attained by running the benchmark with sudo (handled internally)",
@@ -102,7 +102,7 @@ async def main() -> None:
             # strace - the betrayer! - claims that we're passing -20. lies!
             await child.task.setpriority(PRIO.PROCESS, 40)
         await child.inherit_fd(fd).dup2(child.stdout)
-        if mode in ['fork', 'posix_spawn']:
+        if mode in ['fork', 'vfork', 'posix_spawn']:
             proc = await child.exec(native_cmd.args(mode, str(mmaps)))
         else:
             proc = await child.exec(cmd.args(
@@ -131,10 +131,12 @@ async def main() -> None:
     # subprocess = await get_data("subprocess")
     # rsyscall = await get_data("rsyscall")
     # fork = await get_data("fork")
+    vfork = await get_data("vfork")
     # posix_spawn = await get_data("posix_spawn")
     print("subprocess =", subprocess)
     print("rsyscall =", rsyscall)
     print("fork =", fork)
+    print("vfork =", vfork)
     print("posix_spawn =", posix_spawn)
     fig, ax = plt.subplots(figsize=(6.4,3.2))
     plt.xscale('log')
@@ -144,6 +146,8 @@ async def main() -> None:
     ax.plot([10**x for x, y in subprocess], [y for x, y in subprocess], 'o-', label="Python subprocess",
             linewidth=4, markersize=12)
     ax.plot([10**x for x, y in fork], [y for x, y in fork], 'P-', label="C fork",
+            linewidth=4, markersize=12)
+    ax.plot([10**x for x, y in vfork], [y for x, y in vfork], 'X-', label="C vfork",
             linewidth=4, markersize=12)
[…]

### 2021-04-08 research/paper/paper.tex
paper

--- 2021-04-07 research/paper/paper.tex
@@
 \subsection{Usage in the real world}\label{realworld}
-At \twosigma we have used \texttt{rsyscall} and direct-style process creation
+At \twosigma{} we have used \texttt{rsyscall} and direct-style process creation
 to implement an internal distributed system deployment library, written in Python,
@@
 We would like to open source libraries built on top of direct-style process creation.
-From our experience using such libraries at \twosigma,
+From our experience using such libraries at \twosigma{},
 we believe this would greatly ease the development of complex systems involving processes,

### 2021-04-08 python/subprocess_bench.py
benchmark: tweak

@@ -17,7 +17,7 @@ from rsyscall.sched import CLONE
 
 import rsyscall.nix as nix
 from rsyscall.tasks.stdin_bootstrap import stdin_bootstrap, stdin_bootstrap_path_from_store
-from rsyscall.unistd import SEEK
+from rsyscall.unistd import SEEK, OK
 from rsyscall.sys.mman import PROT, MAP
 from rsyscall.sys.resource import PRIO
 from statistics import mean
@@ -37,11 +37,11 @@ async def run_benchmark(mode: str, mmaps: int) -> t.Tuple[int, int]:
         # set POPULATE so it actually gets faulted in; if it isn't
         # already in the page tables, there's no slowdown
         await local.thread.task.mmap(4096, PROT.READ|PROT.WRITE, MAP.PRIVATE|MAP.ANONYMOUS|MAP.POPULATE)
-    cmd = await local.thread.environ.which('cat')
+    cmd = await local.thread.environ.which('true')
     await local.thread.environ.as_arglist(local.thread.ram)
     cwd = await local.thread.ptr(Path("/dev"))
     async def subp_run() -> None:
-        popen = subprocess.Popen([cmd.executable_path, './null'], cwd=cwd.value)
+        popen = subprocess.Popen([cmd.executable_path])
         popen.wait()
     async def rsys_run() -> None:
         child = await local.thread.clone()
@@ -53,30 +53,6 @@ async def run_benchmark(mode: str, mmaps: int) -> t.Tuple[int, int]:
         run = subp_run
     elif mode == 'rsyscall':
         run = rsys_run
-    elif mode == 'nest':
-        nesting_child = await local.thread.clone()
-        async def run() -> None:
-            child = await nesting_child.clone()
-            await child.task.chdir(cwd)
-            child_proc = await child.execv(cmd.executable_path, [cmd.executable_path, "./null"])
-            await child_proc.waitpid(W.EXITED)
-            await child.close()
-    elif mode == 'nestnest':
-        first_nesting_child = await local.thread.clone()
-        second_nesting_child = await first_nesting_child.clone()
-        async def run() -> None:
-            child = await second_nesting_child.clone()
-            await child.task.chdir(cwd)
-            child_proc = await child.execv(cmd.executable_path, [cmd.executable_path, "./null"])
-            await child_proc.waitpid(W.EXITED)
-            await child.close()
[…]

### 2021-04-08 research/paper/native.c
add miscellaneous figures I forgot how to reproduce lol

@@ -1,0 +1,100 @@
+#define _GNU_SOURCE
+#include <unistd.h>
+#include <stdio.h>
+#include <stdlib.h>
+#include <sys/wait.h>
+#include <sys/time.h>
+#include <sys/mman.h>
+#include <sys/syscall.h>
+#include <sched.h>
+#include <string.h>
+#include <signal.h>
+#include <err.h>
+#include <sched.h>
+#include <spawn.h>
+
+enum proc_create {
+        FORK,
+        VFORK,
+        POSIX_SPAWN,
+};
+
+void run(enum proc_create mode, char **envp, int count) {
+        for (int i = 0; i < count; i++) {
+                int pid;
+                switch (mode) {
+                case FORK: {
+                        char *argv[] = {"true", NULL};
+                        pid = fork();
+                        if (pid == 0) {
+                                execve(TRUE_PATH, argv, envp);
+                                err(1, "execve");
+                        }
+                } break;
+                case VFORK: {
+                        char *argv[] = {"true", NULL};
+                        pid = vfork();
+                        if (pid == 0) {
+                                execve(TRUE_PATH, argv, envp);
+                                err(1, "execve");
+                        }
+                } break;
+                case POSIX_SPAWN: {
+                        char *argv[] = {"true", NULL};
+                        int rc = posix_spawn(&pid, TRUE_PATH, NULL, NULL, argv, envp);
+                        if (rc < 0) {
+                                err(1, "posix_spawn");
+                        }
+                } break;
+                }
+                waitpid(pid, NULL, 0);
+        }
+}
+
+struct options {
+        enum proc_create mode;
+        int num_mappings;
+};
+
+static struct options parse_options(int argc, char **argv) {
+        void usage() {
+                printf("Usage: %s mode num_mappings\n", argc >= 1 ? argv[0] : "native_bench");
+                exit(1);
+        }
+        if (argc < 3) usage();
+        enum proc_create mode;
+        if (!strcmp(argv[1], "fork")) {
+                mode = FORK;
+        } else if (!strcmp(argv[1], "vfork")) {
+                mode = VFORK;
+        } else if (!strcmp(argv[1], "posix_spawn")) {
+                mode = POSIX_SPAWN;
+        } else {
+                errx(1, "invalid mode %s", argv[1]);
+        }
+        return (struct options) {
+                .mode = mode,
[…]
